### 2025-05-12 docs/paper.Rmd
add AEA LaTeX class, BibTeX style, and supporting files for paper compilation

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Beckerian Quantity-Quality Trade-off in the Philippine Context"
journal: "2025 Schupf Fellowship Project" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 1
issue: 1
jel:
  - A10
  - A11
keywords:
  - first keyword
  - second keyword
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides,
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  Acknowledgements
abstract: |
  Abstract goes here
output: rticles::aea_article
---

```{r}
#| label: template-pointers
#| include: false
#| eval: false

\begin{itemize}
\item Do not use an "Introduction" heading. Begin your introductory material
before the first section heading.

\item Avoid style markup (except sparingly for emphasis).

\item Avoid using explicit vertical or horizontal space.

\item Captions are short and go below figures but above tables.

\item The tablenotes or figurenotes environments may be used below tables
or figures, respectively, as demonstrated below.

\item If you have difficulties with the mathtime package, adjust the package
options appropriately for your platform. If you can't get it to work, just
remove the package or see our technical support document online (please
refer to the author instructions).

\item If you are using an appendix, it goes last, after the bibliography.
Use regular section headings to make the appendix headings.

\item If you are not using an appendix, you may delete the appendix command
and sample appendix section heading.

\item Either the natbib package or the harvard package may be used with bibtex.
To include one of these packages, uncomment the appropriate usepackage command
above. Note: you can't use both packages at once or compile-time errors will result.

\end{itemize}
```

\section{Introduction}

Sample figure:

\begin{figure}
Figure here.

\caption{Caption for figure below.}
\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\begin{figurenotes}[Source]
Figure notes with optional leadin (Source, in this case).
\end{figurenotes}
\end{figure}

Sample table:

\begin{table}
\caption{Caption for table above.}

\begin{tabular}{lll}
& Heading 1 & Heading 2 \\
Row 1 & 1 & 2 \\
Row 2 & 3 & 4%
\end{tabular}
\begin{tablenotes}
Table notes environment without optional leadin.
\end{tablenotes}
\begin{tablenotes}[Source]
Table notes environment with optional leadin (Source, in this case).
\end{tablenotes}
\end{table}

References here (manual or bibTeX). If you are using bibTeX, add your bib file name in place of BibFile in the bibliography command. % Remove or comment out the next two lines if you are not using bibtex.

\bibliographystyle{aea}
\bibliography{references}

\% The appendix command is issued once, prior to all appendices, if any. \appendix

\section{Mathematical Appendix}

### 2025-05-12 docs/paper.tex
add AEA LaTeX class, BibTeX style, and supporting files for paper compilation

% AER-Article.tex for AEA last revised 22 June 2011
\documentclass[AER]{AEA}

% The mathtime package uses a Times font instead of Computer Modern.
% Uncomment the line below if you wish to use the mathtime package:
%\usepackage[cmbold]{mathtime}
% Note that miktex, by default, configures the mathtime package to use commercial fonts
% which you may not have. If you would like to use mathtime but you are seeing error
% messages about missing fonts (mtex.pfb, mtsy.pfb, or rmtmi.pfb) then please see
% the technical support document at http://www.aeaweb.org/templates/technical_support.pdf
% for instructions on fixing this problem.

% Note: you may use either harvard or natbib (but not both) to provide a wider
% variety of citation commands than latex supports natively. See below.

% Uncomment the next line to use the natbib package with bibtex
\usepackage{natbib}

% Uncomment the next line to use the harvard package with bibtex
%\usepackage[abbr]{harvard}

% This command determines the leading (vertical space between lines) in draft mode
% with 1.5 corresponding to "double" spacing.
\draftSpacing{1.5}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Add imagehandling

\usepackage{hyperref}

\begin{document}

\title{Nutritional Deficits and the Quantity-Quality Trade-off: Evidence
from an Exogenous Fertility Shock in Low-Income Urban Settings in the
Philippines}
\shortTitle{A shorter title}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

\author{
  Erika Salvador\\
  Caroline Theoharides,\thanks{
  Salvador: Amherst
College, \href{mailto:esalvador28@amherst.edu}{esalvador28@amherst.edu}.
  Theoharides: Amherst
College, \href{mailto:ctheoharides@amherst.edu}{ctheoharides@amherst.edu}.
  Acknowledgements
}
}

\date{\today}
\pubMonth{05}
\pubYear{2025}
\pubVolume{1}
\pubIssue{1}
\JEL{A10, A11}
\Keywords{first keyword, second keyword}

\begin{abstract}
Abstract goes here
\end{abstract}

\maketitle

\section{First Section in Body}

Sample figure:

\begin{figure}
Figure here.

\caption{Caption for figure below.}
\begin{figurenotes}
Figure notes without optional leadin.
\end{figurenotes}
\begin{figurenotes}[Source]
Figure notes with optional leadin (Source, in this case).
\end{figurenotes}
\end{figure}

Sample table:

\begin{table}
\caption{Caption for table above.}

\begin{tabular}{lll}
& Heading 1 & Heading 2 \\
Row 1 & 1 & 2 \\
Row 2 & 3 & 4%
\end{tabular}
\begin{tablenotes}
Table notes environment without optional leadin.
\end{tablenotes}
\begin{tablenotes}[Source]
Table notes environment with optional leadin (Source, in this case).
\end{tablenotes}
\end{table}

References here (manual or bibTeX). If you are using bibTeX, add your
bib file name in place of BibFile in the bibliography command. \% Remove
or comment out the next two lines if you are not using bibtex.

\bibliographystyle{aea}
\bibliography{references}

\% The appendix command is issued once, prior to all appendices, if any.
\appendix

\section{Mathematical Appendix}

\end{document}

### 2025-05-12 docs/paper.Rmd
draft introduction and link references.bib for article output

@@
 output: rticles::aea_article
+bibliography: references.bib
 ---
@@
 \section{Introduction}
+
+The quantity–quality trade-off is one of the most influential ideas in the economics of the family.
+This framework, introduced by Becker \citep{becker1960economic, becker1991allocation}, asserts that parents who have more children must allocate fewer resources to each child.
+Constraints on income and attention lead to reductions in per-child investments, especially in settings where public provision of services is weak or inconsistent.
+The trade-off helps explain long-run trends in fertility decline, human capital accumulation, and intergenerational poverty in low- and middle-income countries (LMICs) \citep{rosenzweig1980testing, schwarze2003does}.
+
+Despite this theoretical foundation, most empirical tests of the trade-off have prioritized educational outcomes as the primary measure of child “quality.” Studies often rely on proxies such as school enrollment, standardized test scores, or years of completed schooling \citep{black2005more, angrist2010effects}.
+These indicators indeed encapsulate important aspects of human capital formation, yet they obscure other equally vital dimensions.
+Among these, nutrition stands out as both foundational and predictive.
+Adequate nutrition supports cognitive development and raises productivity later in life \citep{hoddinott2013adult}.
+Children who experience chronic undernutrition face biological constraints that limit their ability to benefit from schooling, regardless of enrollment status or household income \citep{grantham2007development}.
+Analyses that exclude nutritional outcomes therefore risk understating the full scope of the quantity–quality trade-off.
+
+This paper addresses this gap by studying how an exogenous increase in fertility affected child nutrition in the Philippines.
+In 1990, the mayor of Manila issued an executive order that prohibited modern contraceptives in the city’s public health system.
+Health centers could no longer distribute birth control pills, condoms, or IUDs.
+Local officials also removed family planning materials and instructed providers to offer only natural methods.
+Because the national government did not impose similar restrictions, this policy created a natural experiment.
+Poor families in Manila experienced a sudden reduction in access to contraceptive services, while other urban households in comparable regions retained access \citep{dumas2019fertility}.
+
+I use this policy discontinuity to examine the relationship between fertility and child nutrition.
+The analysis draws on multiple waves of the Philippine Demographic and Health Surveys (DHS), which offer nationally representative data on household structure, maternal characteristics, and child anthropometrics.
+I focus on stunting (low height-for-age) and wasting (low weight-for-height) as outcome variables.
+These indicators capture long-term and short-term nutritional stress, respectively, and predict later-life productivity, disease risk, and mortality \citep{victora2008maternal}.
+
+The empirical strategy proceeds in three steps.
+First, I replicate prior work to confirm that the contraceptive ban increased fertility among affected women \citep{dumas2019fertility}.
+Second, I estimate reduced-form effects of exposure to the policy on child nutrition.
+Third, I examine heterogeneity by maternal education, household wealth, and access to prenatal care.
+Families with fewer resources may face tighter constraints when household size increases.
+If the quantity–quality trade-off holds in the nutritional domain, then children in these households should face a higher risk of malnutrition.
+
+This study contributes to the literature in several ways.
+It provides rare causal evidence that links fertility shocks to nutrition rather than education.
+It expands the definition of child quality to reflect physiological outcomes.
+It also highlights the long-run consequences of local reproductive policy in a middle-income democracy.
+In the wider policy landscape, the Philippine case offers a warning.
+Governments that restrict reproductive autonomy may unintentionally weaken child health and human capital formation.
+As countries pursue goals related to nutrition, health equity, and gender rights, evidence from natural experiments such as this one can inform the design of more inclusive and sustainable population policies.
 

### 2025-05-12 docs/paper.tex
draft introduction and link references.bib for article output

@@
 % AER-Article.tex for AEA last revised 22 June 2011
-\documentclass[AER]{AEA}
+\documentclass[2025 Schupf Fellowship Project]{AEA}
 
@@
 Philippines}
-\shortTitle{A shorter title}
+\shortTitle{The Beckerian Quantity-Quality Trade-off in the Philippine
+Context}
 % \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
@@
 
-\section{First Section in Body}
+\section{Introduction}
+
+The quantity--quality trade-off is one of the most influential ideas in
+the economics of the family. This framework, introduced by Becker
+\citep{becker1960economic, becker1991allocation}, asserts that parents
+who have more children must allocate fewer resources to each child.
+Constraints on income and attention lead to reductions in per-child
+investments, especially in settings where public provision of services
+is weak or inconsistent. The trade-off helps explain long-run trends in
+fertility decline, human capital accumulation, and intergenerational
+poverty in low- and middle-income countries (LMICs)
+\citep{rosenzweig1980testing, schwarze2003does}.
+
+Despite this theoretical foundation, most empirical tests of the
+trade-off have prioritized educational outcomes as the primary measure
+of child ``quality.'' Studies often rely on proxies such as school
+enrollment, standardized test scores, or years of completed schooling
+\citep{black2005more, angrist2010effects}. These indicators indeed
+encapsulate important aspects of human capital formation, yet they
+obscure other equally vital dimensions. Among these, nutrition stands
+out as both foundational and predictive. Adequate nutrition supports
+cognitive development and raises productivity later in life
+\citep{hoddinott2013adult}. Children who experience chronic
+undernutrition face biological constraints that limit their ability to
+benefit from schooling, regardless of enrollment status or household
+income \citep{grantham2007development}. Analyses that exclude
+nutritional outcomes therefore risk understating the full scope of the
+quantity--quality trade-off.
+
+This paper addresses this gap by studying how an exogenous increase in
+fertility affected child nutrition in the Philippines. In 1990, the
+mayor of Manila issued an executive order that prohibited modern
+contraceptives in the city's public health system. Health centers could
+no longer distribute birth control pills, condoms, or IUDs. Local
+officials also removed family planning materials and instructed
+providers to offer only natural methods. Because the national government
+did not impose similar restrictions, this policy created a natural
+experiment. Poor families in Manila experienced a sudden reduction in
+access to contraceptive services, while other urban households in
+comparable regions retained access \citep{dumas2019fertility}.
+
+I use this policy discontinuity to examine the relationship between
+fertility and child nutrition. The analysis draws on multiple waves of
+the Philippine Demographic and Health Surveys (DHS), which offer
+nationally representative data on household structure, maternal
+characteristics, and child anthropometrics. I focus on stunting (low
+height-for-age) and wasting (low weight-for-height) as outcome
+variables. These indicators capture long-term and short-term nutritional
+stress, respectively, and predict later-life productivity, disease risk,
+and mortality \citep{victora2008maternal}.
+
+The empirical strategy proceeds in three steps. First, I replicate prior
+work to confirm that the contraceptive ban increased fertility among
+affected women \citep{dumas2019fertility}. Second, I estimate
+reduced-form effects of exposure to the policy on child nutrition.
+Third, I examine heterogeneity by maternal education, household wealth,
+and access to prenatal care. Families with fewer resources may face
+tighter constraints when household size increases. If the
+quantity--quality trade-off holds in the nutritional domain, then
+children in these households should face a higher risk of malnutrition.
+
+This study contributes to the literature in several ways. It provides
+rare causal evidence that links fertility shocks to nutrition rather
+than education. It expands the definition of child quality to reflect
+physiological outcomes. It also highlights the long-run consequences of
+local reproductive policy in a middle-income democracy. In the wider
+policy landscape, the Philippine case offers a warning. Governments that
+restrict reproductive autonomy may unintentionally weaken child health
+and human capital formation. As countries pursue goals related to
+nutrition, health equity, and gender rights, evidence from natural
+experiments such as this one can inform the design of more inclusive and
+sustainable population policies.
 

### 2025-06-10 docs/paper.Rmd
revise introduction and add working abstract and acknowledgements

--- 2025-05-12 docs/paper.Rmd
@@
 jel:
-  - A10
-  - A11
+  - J13  # Fertility; Childbearing; Family Planning; Population Policy
+  - I15  # Health and Economic Development
+  - O15  # Human Resources; Human Development; Income Distribution; Migration
 keywords:
-  - first keyword
-  - second keyword
+  - fertility shocks
+  - child nutrition
+  - quantity-quality trade-off
+  - urban poverty
+  - Philippines
 author:
@@
     affiliation: Amherst College
-  - name: Caroline Theoharides,
+  - name: Caroline Theoharides
     firstname: Caroline
@@
 acknowledgements: |
-  Acknowledgements
+  This research was made possible by the Schupf Fellowship at Amherst College.
+I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project.
+I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work.
+I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support.
+The views expressed and any errors contained herein are entirely my own responsibility.
 abstract: |
-  Abstract goes here
-output: rticles::aea_article
+  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households.
+I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities.
+Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height.
+[Will add more after data analysis]
+output:
+  rticles::aea_article:
+    keep_tex: true
+    citation_package: natbib
+bibliography: references.bib
 ---
@@
 
-Sample figure:
+The trade-off between child quantity and child quality is a foundational concept in the economics of the family.
+First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{beckerlewis1973, beckertomes1976}, this framework posits that parents allocate finite resources—both financial and non-financial—across children.
+An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital.
+This mechanism has served as an explanatory model for changes in fertility behavior and the evolution of population structures in low- and middle-income countries.
 
-\begin{figure}
-Figure here.
+Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes.
+Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010effects}.
+These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality.
+Other outcomes, such as nutritional status, are early-onset or biologically constrained.
+They also tend to be less responsive to remediation later in life.
+A child who suffers from chronic malnutrition may exhibit permanently reduced cognitive capacity and face limits in physical development that affect long-run productivity regardless of subsequent educational access \citep{hoddinott2013adult, grantham2007development}.
 
-\caption{Caption for figure below.}
-\begin{figurenotes}
-Figure notes without optional leadin.
-\end{figurenotes}
-\begin{figurenotes}[Source]
-Figure notes with optional leadin (Source, in this case).
-\end{figurenotes}
-\end{figure}
+The exclusion of nutritional outcomes from much of the empirical literature leaves an important dimension of the trade-off untested.
+Nutritional investments in early childhood are essential to early childhood development and long-term outcomes \citep{victora2008maternal, hoddinott2013adult}.
+They shape brain development, immune system functioning, and physical stature, and they have been shown to predict later-life earnings and health outcomes across a wide range of settings \citep{grantham2007development, alderman2006long}.
+The biological irreversibility of early-life nutritional deficits further distinguishes them from other forms of investment.
+Educational deficits may be partially remediable; nutritional failures often are not.
+A credible estimate of the trade-off between fertility and child quality must account for nutrition if it aims to assess the full set of consequences associated with fertility shocks.
 
-Sample table:
+This study addresses this gap by examining the nutritional effects of a localized, exogenous increase in fertility in the Philippines.
+In 1990, the mayor of Manila implemented an executive policy that prohibited the provision of modern contraceptives in all city-run health facilities.
+The order removed access to pills, condoms, intrauterine devices, and related public health materials and instructed healthcare providers to offer only natural family planning methods.
+This policy remained in place for nearly a decade and affected only the jurisdiction of the Manila city government.
+The national government did not implement a comparable restriction, and surrounding cities within Metro Manila continued to provide access to modern contraceptives.
+The policy thus created a spatial and temporal discontinuity in contraceptive access that was uncorrelated with underlying fertility preferences or concurrent shifts in household income or governance.
+As a result, the Manila ban serves as a quasi-experimental source of variation in fertility exposure among poor urban households.
 
-\begin{table}
-\caption{Caption for table above.}
+I use this natural experiment to estimate the causal effect of increased fertility on child nutrition.
+The analysis relies on nationally representative data from multiple waves of the Philippine Demographic and Health Survey (DHS), which provide data on household structure and maternal characteristics, as well as measurements of child anthropometry.
+The outcomes of interest are height-for-age and weight-for-height z-scores, which serve as standardized indicators of chronic and acute malnutrition, respectively.
+These outcomes are widely used in the global health and development literature and capture nutritional deprivation over both long and short time horizons \citep{victora2008maternal}.
+The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.
 
-\begin{tabular}{lll}
-& Heading 1 & Heading 2 \\
-Row 1 & 1 & 2 \\
-Row 2 & 3 & 4%
-\end{tabular}
-\begin{tablenotes}
-Table notes environment without optional leadin.
-\end{tablenotes}
-\begin{tablenotes}[Source]
-Table notes environment with optional leadin (Source, in this case).
-\end{tablenotes}
-\end{table}
+The identification strategy rests on two key assumptions.
+First, in the absence of the contraceptive ban, nutritional trends in Manila would have evolved in parallel with those in comparison cities.
+Second, any other policy or economic shocks affecting Manila during the study period must not coincide precisely with the timing and scope of the contraceptive policy.
+I test these assumptions using falsification checks, placebo comparisons, and robustness specifications that include city-specific time trends, maternal fixed effects, and controls for baseline demographic differences.
 
-References here (manual or bibTeX).
-If you are using bibTeX, add your bib file name in place of BibFile in the bibliography command.
-% Remove or comment out the next two lines if you are not using bibtex.
+The analysis proceeds in three stages.
+I first replicate existing work \citep{dumas2010fertility} to confirm that the contraceptive ban led to an increase in fertility among affected women.
+I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age.
+Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care.
+These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.
+
+This research contributes to the literature in several respects.
+It introduces new evidence on the nutritional consequences of fertility shocks in an urban developing-country context.
+It expands the measurement of child quality beyond educational outcomes to include biological indicators.
+It also adds to the limited body of work that uses plausibly exogenous variation in reproductive health policy to study downstream effects on human capital.
+More broadly, the findings underscore the potential for local family planning restrictions to generate unintended consequences for child well-being in settings characterized by poverty, food insecurity, and institutional fragility.
 

### 2025-06-10 docs/paper.tex
revise introduction and add working abstract and acknowledgements

--- 2025-05-12 docs/paper.tex
@@
   Erika Salvador\\
-  Caroline Theoharides,\thanks{
+  Caroline Theoharides\thanks{
   Salvador: Amherst
@@
 College, \href{mailto:ctheoharides@amherst.edu}{ctheoharides@amherst.edu}.
-  Acknowledgements
+  This research was made possible by the Schupf Fellowship at Amherst
+  College. I am indebted to my faculty adviser, Professor Caroline
+  Theoharides, for her mentorship and support throughout the development
+  of this project. I also wish to thank Faculty Director Professor
+  Amelie Hastie and the campus partners whose institutional support
+  enabled this work. I am further grateful to the Departments of
+  Economics and Mathematics \& Statistics for their academic support.
+  The views expressed and any errors contained herein are entirely my
+  own responsibility.
 }
@@
 \date{\today}
-\pubMonth{05}
+\pubMonth{06}
 \pubYear{2025}
@@
 \pubIssue{1}
-\JEL{A10, A11}
-\Keywords{first keyword, second keyword}
+\JEL{J13, I15, O15}
+\Keywords{fertility shocks, child nutrition, quantity-quality
+trade-off, urban poverty, Philippines}
 
 \begin{abstract}
-Abstract goes here
+This paper examines whether increased fertility affects early-life
+nutritional outcomes in low-income urban households. I exploit a natural
+experiment created by a 1990 policy in Manila, Philippines, which banned
+modern contraceptives from city-run health facilities. Using a
+difference-in-differences framework and nationally representative data
+from the Philippine Demographic and Health Surveys, I estimate the
+reduced-form impact of the policy on child height-for-age and
+weight-for-height. {[}Will add more after data analysis{]}
 \end{abstract}
@@
 
-\section{First Section in Body}
+\section{Introduction}
 
-Sample figure:
+The trade-off between child quantity and child quality is a foundational
+concept in the economics of the family. First articulated by
+\citet{becker1960economic} and extended in subsequent models of
+household behavior \citep{beckerlewis1973, beckertomes1976}, this
+framework posits that parents allocate finite resources---both financial
+and non-financial---across children. An increase in fertility reduces
+the resources available per child and, under binding constraints, may
+lead to lower investments in health, education, and other forms of human
+capital. This mechanism has served as an explanatory model for changes
+in fertility behavior and the evolution of population structures in low-
+and middle-income countries.
 
-\begin{figure}
-Figure here.
+Empirical investigations of the quantity--quality trade-off have focused
+primarily on educational outcomes. Studies in both high-income and
+low-income settings have examined the effects of fertility on school
+enrollment, grade progression, test scores, and completed years of
+schooling
+\citep{rosenzweig1980testing, black2005more, angrist2010effects}. These
+outcomes serve as accessible proxies for long-run human capital
+accumulation, but they represent only one dimension of child quality.
+Other outcomes, such as nutritional status, are early-onset or
+biologically constrained. They also tend to be less responsive to
+remediation later in life. A child who suffers from chronic malnutrition
+may exhibit permanently reduced cognitive capacity and face limits in
+physical development that affect long-run productivity regardless of
+subsequent educational access
+\citep{hoddinott2013adult, grantham2007development}.
 
-\caption{Caption for figure below.}
-\begin{figurenotes}
-Figure notes without optional leadin.
-\end{figurenotes}
-\begin{figurenotes}[Source]
-Figure notes with optional leadin (Source, in this case).
-\end{figurenotes}
-\end{figure}
+The exclusion of nutritional outcomes from much of the empirical
+literature leaves an important dimension of the trade-off untested.
+Nutritional investments in early childhood are essential to early
+childhood development and long-term outcomes
+\citep{victora2008maternal, hoddinott2013adult}. They shape brain
+development, immune system functioning, and physical stature, and they
+have been shown to predict later-life earnings and health outcomes
+across a wide range of settings
+\citep{grantham2007development, alderman2006long}. The biological
+irreversibility of early-life nutritional deficits further distinguishes
+them from other forms of investment. Educational deficits may be
+partially remediable; nutritional failures often are not. A credible
+estimate of the trade-off between fertility and child quality must
+account for nutrition if it aims to assess the full set of consequences
+associated with fertility shocks.
 
-Sample table:
+This study addresses this gap by examining the nutritional effects of a
+localized, exogenous increase in fertility in the Philippines. In 1990,
+the mayor of Manila implemented an executive policy that prohibited the
+provision of modern contraceptives in all city-run health facilities.
+The order removed access to pills, condoms, intrauterine devices, and
+related public health materials and instructed healthcare providers to
+offer only natural family planning methods. This policy remained in
+place for nearly a decade and affected only the jurisdiction of the
+Manila city government. The national government did not implement a
+comparable restriction, and surrounding cities within Metro Manila
+continued to provide access to modern contraceptives. The policy thus
+created a spatial and temporal discontinuity in contraceptive access
+that was uncorrelated with underlying fertility preferences or
+concurrent shifts in household income or governance. As a result, the
+Manila ban serves as a quasi-experimental source of variation in
+fertility exposure among poor urban households.
 
-\begin{table}
-\caption{Caption for table above.}
+I use this natural experiment to estimate the causal effect of increased
+fertility on child nutrition. The analysis relies on nationally
+representative data from multiple waves of the Philippine Demographic
+and Health Survey (DHS), which provide data on household structure and
+maternal characteristics, as well as measurements of child
+anthropometry. The outcomes of interest are height-for-age and
+weight-for-height z-scores, which serve as standardized indicators of
+chronic and acute malnutrition, respectively. These outcomes are widely
+used in the global health and development literature and capture
+nutritional deprivation over both long and short time horizons
+\citep{victora2008maternal}. The empirical strategy follows a
+difference-in-differences design that compares child outcomes in Manila
+and comparable urban areas before and after the onset of the policy.
 
-\begin{tabular}{lll}
-& Heading 1 & Heading 2 \\
-Row 1 & 1 & 2 \\
-Row 2 & 3 & 4%
-\end{tabular}
-\begin{tablenotes}
-Table notes environment without optional leadin.
-\end{tablenotes}
-\begin{tablenotes}[Source]
-Table notes environment with optional leadin (Source, in this case).
-\end{tablenotes}
-\end{table}
+The identification strategy rests on two key assumptions. First, in the
+absence of the contraceptive ban, nutritional trends in Manila would
+have evolved in parallel with those in comparison cities. Second, any
+other policy or economic shocks affecting Manila during the study period
+must not coincide precisely with the timing and scope of the
+contraceptive policy. I test these assumptions using falsification
+checks, placebo comparisons, and robustness specifications that include
+city-specific time trends, maternal fixed effects, and controls for
+baseline demographic differences.
 
-References here (manual or bibTeX). If you are using bibTeX, add your
-bib file name in place of BibFile in the bibliography command. \% Remove
-or comment out the next two lines if you are not using bibtex.
+The analysis proceeds in three stages. I first replicate existing work
+\citep{dumas2010fertility} to confirm that the contraceptive ban led to
+an increase in fertility among affected women. I then estimate
+reduced-form effects of policy exposure on nutritional outcomes for
+children under five years of age. Finally, I examine heterogeneity in
+effects across subsamples defined by maternal education, household
+wealth, and access to prenatal care. These dimensions serve as proxies
+for household resource availability and capacity to buffer the
+nutritional consequences of fertility increases.
+
+This research contributes to the literature in several respects. It
+introduces new evidence on the nutritional consequences of fertility
+shocks in an urban developing-country context. It expands the
+measurement of child quality beyond educational outcomes to include
+biological indicators. It also adds to the limited body of work that
+uses plausibly exogenous variation in reproductive health policy to
+study downstream effects on human capital. More broadly, the findings
+underscore the potential for local family planning restrictions to
+generate unintended consequences for child well-being in settings
+characterized by poverty, food insecurity, and institutional fragility.
 

### 2025-06-10 docs/paper.Rmd
override AER default to display SCHUPF as issue name

@@
 short: "The Beckerian Quantity-Quality Trade-off in the Philippine Context"
-journal: "2025 Schupf Fellowship Project" # AER, AEJ, PP, JEL
+journal: "" # AER, AEJ, PP, JEL
 month: "`r format(Sys.Date(), '%m')`"
@@
 [Will add more after data analysis]
-output:
-  rticles::aea_article:
-    keep_tex: true
-    citation_package: natbib
+output: rticles::aea_article
 bibliography: references.bib

### 2025-06-10 docs/paper.Rmd
start RRL section with theoretical background

@@
 title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
-short: "The Beckerian Quantity-Quality Trade-off in the Philippine Context"
+short: "The Q-Q Trade-off in the Philippine Context"
 journal: "" # AER, AEJ, PP, JEL
@@
 The analysis proceeds in three stages.
-I first replicate existing work \citep{dumas2010fertility} to confirm that the contraceptive ban led to an increase in fertility among affected women.
+I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women.
 I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age.
@@
 
+\section{Review of Related Literature}
+
+\subsection{Theoretical Background}
+
+The quantity–quality (Q–Q) trade-off is a core idea in the economics of fertility.
+It formalizes the constraint that when parents face a fixed income, they must allocate limited resources between the number of children they have and the amount invested in each child's development.
+In this framework, “quality” refers to any investment that raises a child’s human capital—such as nutrition, education, healthcare—while “quantity” refers simply to the number of children.
+Because these investments are resource-intensive, increasing one often comes at the expense of the other.
+
+Becker’s foundational model \citep{becker1960economic} formalizes this as a constrained optimization problem.
+A household maximizes utility over the number of children \(n\), child quality \(q\), and consumption \(c\):
+
+\[\max_{n,q,c} \quad U(n, q, c)\]
+
+subject to the budget constraint:
+
+\[I = c + n(p_n + p_q q),\]
+
+where \(I\) is household income, \(p_n\) is the fixed cost per child (e.g., food, shelter), and \(p_q\) is the marginal cost of investing in child quality.
+The budget constraint makes the trade-off explicit: increasing \(n\) reduces how much can be spent on \(q\), and vice versa.
+
 \bibliographystyle{aea}

### 2025-06-10 docs/paper.tex
override AER default to display SCHUPF as issue name
start RRL section with theoretical background

@@
 % AER-Article.tex for AEA last revised 22 June 2011
-\documentclass[2025 Schupf Fellowship Project]{AEA}
+\documentclass[]{AEA}
 
@@
 Philippines}
-\shortTitle{The Beckerian Quantity-Quality Trade-off in the Philippine
-Context}
+\shortTitle{The Q-Q Trade-off in the Philippine Context}
 % \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
@@
 The analysis proceeds in three stages. I first replicate existing work
-\citep{dumas2010fertility} to confirm that the contraceptive ban led to
-an increase in fertility among affected women. I then estimate
-reduced-form effects of policy exposure on nutritional outcomes for
-children under five years of age. Finally, I examine heterogeneity in
-effects across subsamples defined by maternal education, household
-wealth, and access to prenatal care. These dimensions serve as proxies
-for household resource availability and capacity to buffer the
-nutritional consequences of fertility increases.
+\citep{dumas2019sex} to confirm that the contraceptive ban led to an
+increase in fertility among affected women. I then estimate reduced-form
+effects of policy exposure on nutritional outcomes for children under
+five years of age. Finally, I examine heterogeneity in effects across
+subsamples defined by maternal education, household wealth, and access
+to prenatal care. These dimensions serve as proxies for household
+resource availability and capacity to buffer the nutritional
+consequences of fertility increases.
 
@@
 
+\section{Review of Related Literature}
+
+\subsection{Theoretical Background}
+
+The quantity--quality (Q--Q) trade-off is a core idea in the economics
+of fertility. It formalizes the constraint that when parents face a
+fixed income, they must allocate limited resources between the number of
+children they have and the amount invested in each child's development.
+In this framework, ``quality'' refers to any investment that raises a
+child's human capital---such as nutrition, education, healthcare---while
+``quantity'' refers simply to the number of children. Because these
+investments are resource-intensive, increasing one often comes at the
+expense of the other.
+
+Becker's foundational model \citep{becker1960economic} formalizes this
+as a constrained optimization problem. A household maximizes utility
+over the number of children \(n\), child quality \(q\), and consumption
+\(c\):
+
+\[\max_{n,q,c} \quad U(n, q, c)\]
+
+subject to the budget constraint:
+
+\[I = c + n(p_n + p_q q),\]
+
+where \(I\) is household income, \(p_n\) is the fixed cost per child
+(e.g., food, shelter), and \(p_q\) is the marginal cost of investing in
+child quality. The budget constraint makes the trade-off explicit:
+increasing \(n\) reduces how much can be spent on \(q\), and vice versa.
+
 \bibliographystyle{aea}

### 2025-06-13 docs/paper.Rmd
expand theoretical background w/ full income constraint and dynamic model (partial draft)

--- 2025-06-10 docs/paper.Rmd
@@
 year: "`r format(Sys.Date(), '%Y')`"
-vol: 1
-issue: 1
 jel:
@@
 [Will add more after data analysis]
-output: rticles::aea_article
+output: 
+  rticles::aea_article:
+    keep_tex: true
+    latex_engine: pdflatex
 bibliography: references.bib
@@
 
-This research contributes to the literature in several respects.
-It introduces new evidence on the nutritional consequences of fertility shocks in an urban developing-country context.
-It expands the measurement of child quality beyond educational outcomes to include biological indicators.
-It also adds to the limited body of work that uses plausibly exogenous variation in reproductive health policy to study downstream effects on human capital.
-More broadly, the findings underscore the potential for local family planning restrictions to generate unintended consequences for child well-being in settings characterized by poverty, food insecurity, and institutional fragility.
+This study contributes to the literature in several important ways.
+It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities.
+Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting.
+It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being.
+More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.
+
+\section{Review of Related Literature}
+
+\subsection{Theoretical Background}
+
+Gary Becker’s early work reframed fertility as an economic decision and drew on microeconomic theory to how families choose whether and how many children to have.
+In his 1960 paper, Becker challenged the then-dominant view that fertility declines with income were simply the result of better access to contraception or changing social values.
+Instead, he proposed that childbearing decisions could be modeled using the tools of consumer choice theory.
+Within this framework, children are treated as goods that provide utility, with households allocating resources to balance their desire for more children (quantity) against the desire to invest in each child (quality).
+
+Becker argued that children serve both consumption and production purposes.
+They bring direct satisfaction to parents and may also contribute economically, especially in agrarian or informal settings.
+In the Philippines, many children aged 5–17 work informally—through street vending or helping in family businesses—to support household income, with an estimated 1.09 million children working in 2023 \citep{manilabulletin2024}.
+This economic role aligns with Becker’s framework and may partly explain why some low-income households continue to prefer larger families.
+At the same time, fertility tends to decline as income rises and per-child investments increase.
+According to the 2017 National Demographic and Health Survey, women in the poorest wealth quintile had an average of 4.5 children, compared to just 2.0 in the richest \citep{psa2018}.
+This inverse relationship between income and fertility, observed globally, also appears in the Philippine context: poorer households often maintain larger families, while wealthier ones prioritize child quality.
+
+\citet{beckerlewis1973} build on these conceptual foundations to introduce the formal concept of a quantity–quality (Q–Q) trade-off.
+They develop a utility-maximizing framework in which parents allocate resources between the number of children and the quality of investments in each child.
+The model formalizes how parents maximize utility from consumption $c$, number of children $n$, and child quality $q$:
+
+$$
+U = U(c, n, q)
+$$
+
+subject to the full income constraint: $$ I = c+n(p_n + q \times p_q) $$ where $p_n$ denotes the direct cost associated with bearing each additional child (e.g., delivery, basic needs), and $p_q$ represents the marginal cost of investing in quality per child, such as health care or education.
+The trade-off is reflected in the structure of the cost function, which reveals interaction effects: the marginal cost of improving child quality rises as the number of children increases, and similarly, the cost of an additional child rises with the level of quality investment per child.
+This complementarity is captured through the positive cross-derivatives:
+
+$$
+\frac{\partial MC_q}{\partial n} > 0 \quad \text{and} \quad \frac{\partial MC_n}{\partial q} > 0
+$$
+
+In short, the more children a family has, the more difficult it becomes to allocate sufficient resources—such as time, nutrition, and education—to each child.
+
+While the static Q–Q model explains the immediate trade-offs families face between the number and quality of children, it remains limited in scope by assuming a single-period decision framework, which fails to account for intertemporal linkages—for instance, how early investments in child health or education may constrain or enhance later household choices—and overlooks how forward-looking behavior, such as anticipated returns to child quality in the form of future earnings or social mobility, may influence present-day household choices.
+To address this, \citet{becker1976child} developed a dynamic model that explicitly incorporates intergenerational mobility.
+In this model, parents account for how their fertility and investment choices shape the human capital outcomes of their children—for example, how a decision to space births more closely can constrain parental resources and reduce per-child investments in health and education.
+The resulting dynastic utility function captures both present and future-oriented preferences:
+
+$$
+U = u(c) + \beta \cdot n \cdot V(q)
+$$
+
+where $\beta$ denotes child altruism, or the extent to which parents value their children’s future well-being relative to their own consumption.
+A higher $\beta$ indicates stronger intergenerational concern, meaning parents are more willing to sacrifice current resources to improve their children’s prospects.
+For example, a low-income family in the Philippines might choose to send a child to private school despite financial constraints to increase the child’s future earnings or social mobility.
+Conversely, a lower $\beta$ implies a stronger emphasis on present needs and diminished willingness to invest in long-term child outcomes.
+This may occur in households facing severe income insecurity, where sending children to work may be prioritized over education.
+On the other hand, $V(q)$ represents the expected return on child quality.
+A high $V(q)$ implies that parents expect substantial future benefits from quality investments, such as access to better-paying jobs or improved life expectanc, making such investments more worthwhile than having additional children.
+A low $V(q)$, in contrast, suggests that even with significant investments, the perceived or actual gains are limited, often due to barriers like poor school quality, weak labor demand, or geographic constraints.
+
+To formalize the trade-off similarly as in the static model, we again impose the full income constraint. In optimizing this problem, the first-order condition for child quality becomes:
+
+$$
+\beta \cdot V'(q) = p_q + n \cdot \frac{\partial p_q}{\partial q}
+$$
+
+This condition implies that the marginal benefit of child quality, scaled by altruism $\beta$, must equal the marginal cost of improving that quality.
+As family size $n$ increases, the cost of maintaining or increasing quality also rises.
+This reinforces the quantity–quality trade-off: families with more children may face sharper constraints in deepening investments per child.
+In contexts like the Philippines, where income constraints are binding and child quality has high private returns, this model predicts lower fertility and increased education spending among upwardly mobile households.
 

### 2025-06-13 docs/paper.tex
expand theoretical background w/ full income constraint and dynamic model (partial draft)

--- 2025-06-10 docs/paper.tex
@@
 \pubYear{2025}
-\pubVolume{1}
-\pubIssue{1}
+\pubVolume{}
+\pubIssue{}
 \JEL{J13, I15, O15}
@@
 The analysis proceeds in three stages. I first replicate existing work
-\citep{dumas2010fertility} to confirm that the contraceptive ban led to
-an increase in fertility among affected women. I then estimate
-reduced-form effects of policy exposure on nutritional outcomes for
-children under five years of age. Finally, I examine heterogeneity in
-effects across subsamples defined by maternal education, household
-wealth, and access to prenatal care. These dimensions serve as proxies
-for household resource availability and capacity to buffer the
-nutritional consequences of fertility increases.
+\citep{dumas2019sex} to confirm that the contraceptive ban led to an
+increase in fertility among affected women. I then estimate reduced-form
+effects of policy exposure on nutritional outcomes for children under
+five years of age. Finally, I examine heterogeneity in effects across
+subsamples defined by maternal education, household wealth, and access
+to prenatal care. These dimensions serve as proxies for household
+resource availability and capacity to buffer the nutritional
+consequences of fertility increases.
 
-This research contributes to the literature in several respects. It
-introduces new evidence on the nutritional consequences of fertility
-shocks in an urban developing-country context. It expands the
-measurement of child quality beyond educational outcomes to include
-biological indicators. It also adds to the limited body of work that
-uses plausibly exogenous variation in reproductive health policy to
-study downstream effects on human capital. More broadly, the findings
-underscore the potential for local family planning restrictions to
-generate unintended consequences for child well-being in settings
-characterized by poverty, food insecurity, and institutional fragility.
+This study contributes to the literature in several important ways. It
+provides new evidence on how increases in fertility---caused by policy
+restrictions on family planning---can affect child nutrition in poor,
+urban communities. Most past research has focused on education, but this
+study expands the idea of child quality to include biological measures
+such as stunting and wasting. It also adds to the small number of
+studies that use unexpected changes in reproductive health policy to
+examine long-term effects on children's well-being. More broadly, the
+results show that local restrictions on family planning can
+unintentionally harm children's health, especially in settings where
+families already face poverty, food insecurity, and limited public
+services.
+
+\section{Review of Related Literature}
+
+\subsection{Theoretical Background}
+
+Gary Becker's early work reframed fertility as an economic decision and
+drew on microeconomic theory to how families choose whether and how many
+children to have. In his 1960 paper, Becker challenged the then-dominant
+view that fertility declines with income were simply the result of
+better access to contraception or changing social values. Instead, he
+proposed that childbearing decisions could be modeled using the tools of
+consumer choice theory. Within this framework, children are treated as
+goods that provide utility, with households allocating resources to
+balance their desire for more children (quantity) against the desire to
+invest in each child (quality).
+
+Becker argued that children serve both consumption and production
+purposes. They bring direct satisfaction to parents and may also
+contribute economically, especially in agrarian or informal settings. In
+the Philippines, many children aged 5--17 work informally---through
+street vending or helping in family businesses---to support household
+income, with an estimated 1.09 million children working in 2023
+\citep{manilabulletin2024}. This economic role aligns with Becker's
+framework and may partly explain why some low-income households continue
+to prefer larger families. At the same time, fertility tends to decline
+as income rises and per-child investments increase. According to the
+2017 National Demographic and Health Survey, women in the poorest wealth
+quintile had an average of 4.5 children, compared to just 2.0 in the
+richest \citep{psa2018}. This inverse relationship between income and
+fertility, observed globally, also appears in the Philippine context:
+poorer households often maintain larger families, while wealthier ones
+prioritize child quality.
+
+\citet{beckerlewis1973} build on these conceptual foundations to
+introduce the formal concept of a quantity--quality (Q--Q) trade-off.
+They develop a utility-maximizing framework in which parents allocate
+resources between the number of children and the quality of investments
+in each child. The model formalizes how parents maximize utility from
+consumption \(c\), number of children \(n\), and child quality \(q\):
+
+\[
+U = U(c, n, q)
+\]
+
+subject to the full income constraint: \[ I = c+n(p_n + q \times p_q) \]
+where \(p_n\) denotes the direct cost associated with bearing each
+additional child (e.g., delivery, basic needs), and \(p_q\) represents
+the marginal cost of investing in quality per child, such as health care
+or education. The trade-off is reflected in the structure of the cost
+function, which reveals interaction effects: the marginal cost of
+improving child quality rises as the number of children increases, and
+similarly, the cost of an additional child rises with the level of
+quality investment per child. This complementarity is captured through
+the positive cross-derivatives:
+
+\[
+\frac{\partial MC_q}{\partial n} > 0 \quad \text{and} \quad \frac{\partial MC_n}{\partial q} > 0
+\]
+
+In short, the more children a family has, the more difficult it becomes
+to allocate sufficient resources---such as time, nutrition, and
+education---to each child.
+
+While the static Q--Q model explains the immediate trade-offs families
+face between the number and quality of children, it remains limited in
+scope by assuming a single-period decision framework, which fails to
+account for intertemporal linkages---for instance, how early investments
+in child health or education may constrain or enhance later household
+choices---and overlooks how forward-looking behavior, such as
+anticipated returns to child quality in the form of future earnings or
+social mobility, may influence present-day household choices. To address
+this, \citet{becker1976child} developed a dynamic model that explicitly
+incorporates intergenerational mobility. In this model, parents account
+for how their fertility and investment choices shape the human capital
+outcomes of their children---for example, how a decision to space births
+more closely can constrain parental resources and reduce per-child
+investments in health and education. The resulting dynastic utility
+function captures both present and future-oriented preferences:
+
+\[
+U = u(c) + \beta \cdot n \cdot V(q)
+\]
+
+where \(\beta\) denotes child altruism, or the extent to which parents
+value their children's future well-being relative to their own
+consumption. A higher \(\beta\) indicates stronger intergenerational
+concern, meaning parents are more willing to sacrifice current resources
+to improve their children's prospects. For example, a low-income family
+in the Philippines might choose to send a child to private school
+despite financial constraints to increase the child's future earnings or
+social mobility. Conversely, a lower \(\beta\) implies a stronger
+emphasis on present needs and diminished willingness to invest in
+long-term child outcomes. This may occur in households facing severe
+income insecurity, where sending children to work may be prioritized
+over education. On the other hand, \(V(q)\) represents the expected
+return on child quality. A high \(V(q)\) implies that parents expect
+substantial future benefits from quality investments, such as access to
+better-paying jobs or improved life expectanc, making such investments
+more worthwhile than having additional children. A low \(V(q)\), in
+contrast, suggests that even with significant investments, the perceived
+or actual gains are limited, often due to barriers like poor school
+quality, weak labor demand, or geographic constraints.
+
+To formalize the trade-off similarly as in the static model, we again
+impose the full income constraint. In optimizing this problem, the
+first-order condition for child quality becomes:
+
+\[
+\beta \cdot V'(q) = p_q + n \cdot \frac{\partial p_q}{\partial q}
+\]
+
+This condition implies that the marginal benefit of child quality,
+scaled by altruism \(\beta\), must equal the marginal cost of improving
+that quality. As family size \(n\) increases, the cost of maintaining or
+increasing quality also rises. This reinforces the quantity--quality
+trade-off: families with more children may face sharper constraints in
+deepening investments per child. In contexts like the Philippines, where
+income constraints are binding and child quality has high private
+returns, this model predicts lower fertility and increased education
+spending among upwardly mobile households.
 

### 2025-06-15 docs/paper.Rmd
replace volume/issue template with publication month and year only

--- 2025-06-10 docs/paper.Rmd
@@
 The trade-off between child quantity and child quality is a foundational concept in the economics of the family.
-First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{beckerlewis1973, beckertomes1976}, this framework posits that parents allocate finite resources—both financial and non-financial—across children.
+First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{becker1973interaction, becker1976child}, this framework posits that parents allocate finite resources—both financial and non-financial—across children.
 An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital.
@@
 
-This research contributes to the literature in several respects.
-It introduces new evidence on the nutritional consequences of fertility shocks in an urban developing-country context.
-It expands the measurement of child quality beyond educational outcomes to include biological indicators.
-It also adds to the limited body of work that uses plausibly exogenous variation in reproductive health policy to study downstream effects on human capital.
-More broadly, the findings underscore the potential for local family planning restrictions to generate unintended consequences for child well-being in settings characterized by poverty, food insecurity, and institutional fragility.
+This study contributes to the literature in several important ways.
+It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities.
+Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting.
+It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being.
+More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.
+
+\section{Review of Related Literature}
+
+The quantity–quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each.
+Quality in this context refers to the human capital of each child: attributes like education, health, and nutrition that enhance a child’s future productivity and well-being.
+The genesis of this idea traces back to Gary Becker’s seminal work around 1960, which for the first time treated children as economic goods subject to parental choice and budget constraints \citep{becker1960economic}.
+Becker argued that as families become wealthier, they may not simply want more children, but rather better-raised children, much as a household might prefer a higher-quality car or house over a greater quantity of them.
+This proposition led to a formal theory in which increases in income or changes in economic conditions cause parents to substitute child quality for quantity, consistent with historical patterns of lower fertility and higher educational attainment during economic development \citep{galor2000population}.
+
+In what follows, I review the theoretical foundations of the Q-Q model and its evolution in the literature.
+I begin with the static models of Becker \citep{becker1960economic} and Becker–Lewis \citep{becker1973interaction}, which first formalized the trade-off within a household utility maximization framework.
+We then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}.
+Next, I turn to macroeconomic and unified growth models, notably Galor and Weil \citep{galor2000population} and Galor and Moav \citep{galor2002natural}, which integrate the Q–Q mechanism into a general theory of demographic and economic transformation.
+
+Finally, I discuss more recent refinements that enrich the basic model by incorporating credit constraints \citep{doepke2004accounting}, intra-household bargaining \citep{doepke2019bargaining}, and multi-dimensional child quality \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis on health and nutrition.
+My focus is on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.
+
+\subsection{Theoretical Background}
 

### 2025-06-16 docs/paper.Rmd
write theoretical background (exposition and static q-q model)

@@
 
+Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}.
+In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$).
+A simple representation is a utility function:
+
+$$
+U = U(n, q, y),
+$$
+
+with $U$ increasing in each argument up to some satiation point.
+Here quality $q$ can be thought of as the expenditure or investment per child (e.g.
+education spending, health care, nutrition), assumed for now to be the same for each child.
+Parents face a budget constraint that links quantity and quality: raising more children dilutes the resources available per child.
+A prototypical budget constraint (in static form) can be written as:
+
+$$
+p_y y + p_n n + p_q n q = I,
+$$
+
+where $I$ is total family income (or full income), $p_n$ represents baseline, non-discretionary costs associated with each additional child (e.g.
+expenditures on food, shelter, or clothing that are incurred irrespective of quality-enhancing investments), and $p_q$ denotes the marginal cost of investing in one unit of quality per child.The term $p_q n q$ captures total expenditure on quality for all children and is linear in $n$.
+As the number of children rises, parents must extend any chosen level of $q$ across a broader base, which amplifies the total cost of quality.
+Conversely, the term $p_n n$ implies that the cost of an additional child rises with the quality level $q$ already chosen, since each child must meet a higher standard of care or investment.
+For instance, a household that chooses to provide more education or better health care per child incurs an additional burden when it expands family size, as each child must receive the same enhanced level of investment.
+Similarly, a larger family increases the cumulative cost of quality, even if $q$ remains fixed, due to the need to replicate expenditures across more children.
+In short, the shadow price of child quality increases with $n$, and the shadow price of child quantity increases with $q$.
+The cost structure induces a mutual dependence between quantity and quality, such that any adjustment along one dimension alters the effective cost of the other.
+
+Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity–quality cost structure.
+An increase in $n$ raises the total cost required to sustain a given level of $q$ for each child, while a higher level of $q$ raises the marginal cost associated with having an additional child.
+For example, allocating more resources to education or health per child increases the financial burden of expanding family size.
+This interdependence links the two decisions directly.
+The household cannot choose $n$ and $q$ in isolation; each choice alters the marginal cost of the other.
+
+Mathematically, the trade-off appears in the first-order conditions of the household’s optimization problem. Let $\lambda$ represent the Lagrange multiplier on the full-income constraint.
+
+$$
+\mathcal{L} = U(n, q, y) + \lambda \left( I - p_y y - p_n n - p_q n q \right).
+$$
+
+The first-order conditions are:
+
+$$
+\frac{\partial \mathcal{L}}{\partial n} = U_n - \lambda(p_n + p_q q) = 0, \quad
+\frac{\partial \mathcal{L}}{\partial q} = U_q - \lambda p_q n = 0, \quad
+\frac{\partial \mathcal{L}}{\partial y} = U_y - \lambda p_y = 0.
+$$
+
+Combining the first two yields:
+
+$$
+\frac{U_n}{U_q} = \frac{p_n + p_q q}{p_q n}.
+$$
+This condition equates the marginal rate of substitution between quantity and quality to the ratio of their full marginal costs.
+The numerator rises with $q$, and the denominator rises with $n$.
+As one choice increases, the relative cost of the other becomes higher.
+This relationship induces substitution toward the less costly dimension.
+The trade-off between quantity and quality arises from the structure of the budget itself.
+It does not rely on specific assumptions about utility curvature or intrinsic substitutability \citep{becker1973interaction}.
+
+This formulation implies two core predictions.
+Firstly, although both child quantity and child quality may rise with income, the household’s budget constraint can generate a negative relationship between income and fertility.
+As income increases, total spending on children tends to rise, but the allocation often favors quality over quantity.
+Becker illustrated this with the analogy of durable goods: wealthier households tend to upgrade the quality of a house or a car rather than acquire additional units.
+In a similar way, higher-income families often direct additional resources toward education, nutrition, or health per child.
+Within the model, an income increase ($\mathrm{d}I > 0$) produces a direct effect that makes children more affordable and an indirect effect that discourages fertility.
+As $q$ rises, the shadow price of an additional child also rises.
+If the marginal utility from higher quality exceeds that from larger family size, then the substitution effect outweighs the income effect, leading to a lower optimal $n$.
+This mechanism offers a structural explanation for the demographic transition: fertility tends to fall as households become richer, even when preferences remain unchanged.
+
+Furthermore, a similar logic applies to changes in the cost parameters $p_q$ and $p_n$.
+A decline in $p_q$, such as through a policy that lowers the price of education or health care, increases $q$ and raises the marginal cost of quantity.
+This effect reduces optimal fertility.
+A rise in $p_n$, which may reflect higher child-rearing costs or a greater opportunity cost of parental time, reduces the appeal of larger families and can shift resources toward child quality.
+These outcomes follow from the structure of the budget constraint, without requiring any explicit preference for quality over quantity.
+Becker and Lewis noted that these comparative static results align with observed patterns.
+For example, increases in women’s wages often reduce fertility more than they reduce educational spending per child.
+This asymmetry reflects the model’s central feature: quantity and quality are linked through their cost structure.
+An increase in one raises the marginal cost of the other.
+The model explains how households make trade-offs between the number of children and investments in each.
+
+While the early Q–Q models were static (one-period) representations, subsequent contributions extended the framework to consider fertility and child investment over multiple periods or even multiple generations.
+A central development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring.
+These extensions allowed child quality to evolve endogenously across time, rather than being determined solely within a single period.
+
 \bibliographystyle{aea}

### 2025-06-16 docs/paper.tex
write theoretical background (exposition and static q-q model)

--- 2025-06-10 docs/paper.tex
@@
 \citet{becker1960economic} and extended in subsequent models of
-household behavior \citep{beckerlewis1973, beckertomes1976}, this
+household behavior \citep{becker1973interaction, becker1976child}, this
 framework posits that parents allocate finite resources---both financial
@@
 The analysis proceeds in three stages. I first replicate existing work
-\citep{dumas2010fertility} to confirm that the contraceptive ban led to
-an increase in fertility among affected women. I then estimate
-reduced-form effects of policy exposure on nutritional outcomes for
-children under five years of age. Finally, I examine heterogeneity in
-effects across subsamples defined by maternal education, household
-wealth, and access to prenatal care. These dimensions serve as proxies
-for household resource availability and capacity to buffer the
-nutritional consequences of fertility increases.
+\citep{dumas2019sex} to confirm that the contraceptive ban led to an
+increase in fertility among affected women. I then estimate reduced-form
+effects of policy exposure on nutritional outcomes for children under
+five years of age. Finally, I examine heterogeneity in effects across
+subsamples defined by maternal education, household wealth, and access
+to prenatal care. These dimensions serve as proxies for household
+resource availability and capacity to buffer the nutritional
+consequences of fertility increases.
 
-This research contributes to the literature in several respects. It
-introduces new evidence on the nutritional consequences of fertility
-shocks in an urban developing-country context. It expands the
-measurement of child quality beyond educational outcomes to include
-biological indicators. It also adds to the limited body of work that
-uses plausibly exogenous variation in reproductive health policy to
-study downstream effects on human capital. More broadly, the findings
-underscore the potential for local family planning restrictions to
-generate unintended consequences for child well-being in settings
-characterized by poverty, food insecurity, and institutional fragility.
+This study contributes to the literature in several important ways. It
+provides new evidence on how increases in fertility---caused by policy
+restrictions on family planning---can affect child nutrition in poor,
+urban communities. Most past research has focused on education, but this
+study expands the idea of child quality to include biological measures
+such as stunting and wasting. It also adds to the small number of
+studies that use unexpected changes in reproductive health policy to
+examine long-term effects on children's well-being. More broadly, the
+results show that local restrictions on family planning can
+unintentionally harm children's health, especially in settings where
+families already face poverty, food insecurity, and limited public
+services.
+
+\section{Review of Related Literature}
+
+The quantity--quality (Q--Q) theory, a central idea in modern family
+economics, holds that parents face a trade-off between the number of
+children and the ``quality'' of investment---such as education or
+health---they can provide to each. Quality in this context refers to the
+human capital of each child: attributes like education, health, and
+nutrition that enhance a child's future productivity and well-being. The
+genesis of this idea traces back to Gary Becker's seminal work around
+1960, which for the first time treated children as economic goods
+subject to parental choice and budget constraints
+\citep{becker1960economic}. Becker argued that as families become
+wealthier, they may not simply want more children, but rather
+better-raised children, much as a household might prefer a
+higher-quality car or house over a greater quantity of them. This
+proposition led to a formal theory in which increases in income or
+changes in economic conditions cause parents to substitute child quality
+for quantity, consistent with historical patterns of lower fertility and
+higher educational attainment during economic development
+\citep{galor2000population}.
+
+In what follows, I review the theoretical foundations of the Q-Q model
+and its evolution in the literature. I begin with the static models of
+Becker \citep{becker1960economic} and Becker--Lewis
+\citep{becker1973interaction}, which first formalized the trade-off
+within a household utility maximization framework. We then examine
+extensions to dynamic, intergenerational settings, including the
+contributions of Becker and Tomes \citep{becker1976child} on child
+endowments and the altruistic dynastic model associated with Barro and
+Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and
+unified growth models, notably Galor and Weil
+\citep{galor2000population} and Galor and Moav \citep{galor2002natural},
+which integrate the Q--Q mechanism into a general theory of demographic
+and economic transformation.
+
+Finally, I discuss more recent refinements that enrich the basic model
+by incorporating credit constraints \citep{doepke2004accounting},
+intra-household bargaining \citep{doepke2019bargaining}, and
+multi-dimensional child quality
+\citep{hoddinott2013economic, kalemli2002does}, with a special emphasis
+on health and nutrition. My focus is on how the Q--Q framework has been
+applied to understand fertility and child investment patterns,
+especially in developing country contexts where resource constraints and
+health outcomes are paramount.
+
+\subsection{Theoretical Background}
+
+Becker's early work introduced an economic approach to fertility,
+treating children analogous to ``durable goods'' that yield utility to
+parents but come at a cost \citep{doepke2015gary}. In Becker's 1960
+model, a household derives satisfaction from the number of children
+(\(n\)) and from the quality of each child (\(q\)), alongside
+conventional consumption of other goods (\(y\)). A simple representation
+is a utility function:
+
+\[
+U = U(n, q, y),
+\]
+
+with \(U\) increasing in each argument up to some satiation point. Here
+quality \(q\) can be thought of as the expenditure or investment per
+child (e.g.~education spending, health care, nutrition), assumed for now
+to be the same for each child. Parents face a budget constraint that
+links quantity and quality: raising more children dilutes the resources
+available per child. A prototypical budget constraint (in static form)
+can be written as:
+
+\[
+p_y y + p_n n + p_q n q = I,
+\]
+
+where \(I\) is total family income (or full income), \(p_n\) represents
+baseline, non-discretionary costs associated with each additional child
+(e.g.~expenditures on food, shelter, or clothing that are incurred
+irrespective of quality-enhancing investments), and \(p_q\) denotes the
+marginal cost of investing in one unit of quality per child.The term
+\(p_q n q\) captures total expenditure on quality for all children and
+is linear in \(n\). As the number of children rises, parents must extend
+any chosen level of \(q\) across a broader base, which amplifies the
+total cost of quality. Conversely, the term \(p_n n\) implies that the
+cost of an additional child rises with the quality level \(q\) already
+chosen, since each child must meet a higher standard of care or
+investment. For instance, a household that chooses to provide more
+education or better health care per child incurs an additional burden
+when it expands family size, as each child must receive the same
+enhanced level of investment. Similarly, a larger family increases the
+cumulative cost of quality, even if \(q\) remains fixed, due to the need
+to replicate expenditures across more children. In short, the shadow
+price of child quality increases with \(n\), and the shadow price of
+child quantity increases with \(q\). The cost structure induces a mutual
+dependence between quantity and quality, such that any adjustment along
+one dimension alters the effective cost of the other.
+
+Becker and Lewis (1973) formalize the mutually reinforcing nature of the
+quantity--quality cost structure. An increase in \(n\) raises the total
+cost required to sustain a given level of \(q\) for each child, while a
+higher level of \(q\) raises the marginal cost associated with having an
+additional child. For example, allocating more resources to education or
+health per child increases the financial burden of expanding family
+size. This interdependence links the two decisions directly. The
+household cannot choose \(n\) and \(q\) in isolation; each choice alters
+the marginal cost of the other.
+
+Mathematically, the trade-off appears in the first-order conditions of
+the household's optimization problem. Let \(\lambda\) represent the
+Lagrange multiplier on the full-income constraint.
+
+\[
+\mathcal{L} = U(n, q, y) + \lambda \left( I - p_y y - p_n n - p_q n q \right).
+\]
+
+The first-order conditions are:
+
+\[
+\frac{\partial \mathcal{L}}{\partial n} = U_n - \lambda(p_n + p_q q) = 0, \quad
+\frac{\partial \mathcal{L}}{\partial q} = U_q - \lambda p_q n = 0, \quad
+\frac{\partial \mathcal{L}}{\partial y} = U_y - \lambda p_y = 0.
+\]
+
+Combining the first two yields:
+
+\[
+\frac{U_n}{U_q} = \frac{p_n + p_q q}{p_q n}.
+\] This condition equates the marginal rate of substitution between
+quantity and quality to the ratio of their full marginal costs. The
+numerator rises with \(q\), and the denominator rises with \(n\). As one
+choice increases, the relative cost of the other becomes higher. This
+relationship induces substitution toward the less costly dimension. The
+trade-off between quantity and quality arises from the structure of the
+budget itself. It does not rely on specific assumptions about utility
+curvature or intrinsic substitutability \citep{becker1973interaction}.
+
+This formulation implies two core predictions. Firstly, although both
+child quantity and child quality may rise with income, the household's
+budget constraint can generate a negative relationship between income
+and fertility. As income increases, total spending on children tends to
+rise, but the allocation often favors quality over quantity. Becker
+illustrated this with the analogy of durable goods: wealthier households
+tend to upgrade the quality of a house or a car rather than acquire
+additional units. In a similar way, higher-income families often direct
+additional resources toward education, nutrition, or health per child.
+Within the model, an income increase (\(\mathrm{d}I > 0\)) produces a
+direct effect that makes children more affordable and an indirect effect
+that discourages fertility. As \(q\) rises, the shadow price of an
+additional child also rises. If the marginal utility from higher quality
+exceeds that from larger family size, then the substitution effect
+outweighs the income effect, leading to a lower optimal \(n\). This
+mechanism offers a structural explanation for the demographic
+transition: fertility tends to fall as households become richer, even
+when preferences remain unchanged.
+
+Furthermore, a similar logic applies to changes in the cost parameters
+\(p_q\) and \(p_n\). A decline in \(p_q\), such as through a policy that
+lowers the price of education or health care, increases \(q\) and raises
+the marginal cost of quantity. This effect reduces optimal fertility. A
+rise in \(p_n\), which may reflect higher child-rearing costs or a
+greater opportunity cost of parental time, reduces the appeal of larger
+families and can shift resources toward child quality. These outcomes
+follow from the structure of the budget constraint, without requiring
+any explicit preference for quality over quantity. Becker and Lewis
+noted that these comparative static results align with observed
+patterns. For example, increases in women's wages often reduce fertility
+more than they reduce educational spending per child. This asymmetry
+reflects the model's central feature: quantity and quality are linked
+through their cost structure. An increase in one raises the marginal
+cost of the other. The model explains how households make trade-offs
+between the number of children and investments in each.
+
+While the early Q--Q models were static (one-period) representations,
+subsequent contributions extended the framework to consider fertility
+and child investment over multiple periods or even multiple generations.
+A central development in this literature was the incorporation of
+intergenerational human capital dynamics, where parents derive utility
+not only from the number and quality of children in the present, but
+also from the long-run outcomes of their offspring. These extensions
+allowed child quality to evolve endogenously across time, rather than
+being determined solely within a single period.
 

### 2025-06-17 docs/paper.Rmd
write theoretical background (intergenerational q-q model)

@@
 
+\subsubsection{Becker’s Static Model}
 Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}.
@@
 
+\subsubsection{Intergenerational Models: Altruism and Child Endowments}
+
 While the early Q–Q models were static (one-period) representations, subsequent contributions extended the framework to consider fertility and child investment over multiple periods or even multiple generations.
-A central development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring.
+The main development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring.
 These extensions allowed child quality to evolve endogenously across time, rather than being determined solely within a single period.
+
+One of the earliest and most influential models of this kind was proposed by \citet{becker1976child}.
+In their formulation, each child enters the world with an exogenous endowment $E$, which may reflect factors such as cognitive ability (e.g., measured IQ or language acquisition speed), early health status (e.g., birth weight or incidence of neonatal complications), genetic predispositions (e.g., risk for chronic illness, temperament, or neurodevelopmental traits), or family background characteristics (e.g., parental education, household stability, or neighborhood conditions).
+Furthermore, parents can augment this endowment by investing resources $q$ in the form of education, nutrition, and other quality-enhancing inputs.
+The effective adult human capital of the child might be expressed as $H = E + f(q)$ (in a simple additive form) or a multiplicative variant $H = E \cdot f(q)$, where $f(q)$ is an increasing concave function of parental investment.
+
+\citet{becker1976child} emphasized that variation in endowments $E$ can shape how parents allocate investments $q$ across children.
+When the productivity of investment increases with endowment, parents may concentrate resources on children with higher $E$, who are more likely to convert additional investment into future success.
+In other cases, parents may attempt to compensate for lower endowments by directing greater investment toward disadvantaged children.
+Put simply, child-specific variation in initial conditions affects not only outcomes but also strategic parental choices.
+Because of this, the relationship between income and the demand for child quality is not uniform.
+The income elasticity of demand for quality may differ across households, depending on the distribution of endowments within the family and across the broader population.
+
+Furthermore, \citet{becker1976child} showed that at low income levels, much of what constitutes child quality comes from exogenous endowments, i.e., factors like public education, neighborhood environment, or access to basic healthcare that are not privately purchased.
+In these settings, small increases in parental income may not lead to significant changes in fertility or investment behavior.
+Since most of the child’s future outcomes are determined by the fixed endowment component, marginal investment plays a smaller role.
+However, as income rises, private resources become a larger part of what determines quality, and the classic Q–Q trade-off begins to shape behavior.
+Parents begin to allocate more income toward fewer children in order to enhance quality through direct investment.
+Under certain theoretical conditions, such as equal utility elasticities for quantity and quality, this framework produces a non-monotonic relationship between income and fertility.
+Fertility may decline as income rises at first, which reflects the desire to invest more intensively per child.
+However, beyond some point, once the marginal return to investment begins to flatten or saturate relative to the fixed endowment, fertility may increase again.
+
+The U-shaped prediction emerges only under specific assumptions, and its validity depends on both the shape of the utility function and how endowments relate to parental background.
+More broadly, \citet{becker1976child} enriched the Q–Q framework by incorporating elements that reflect real-world variation.
+They argued that not all differences in child outcomes are the result of deliberate parental choice.
+Random factors, biological traits, and socioeconomic settings play a role.
+In this light, public policies, such as subsidized schooling, early childhood programs, or universal healthcare, can influence private fertility and investment decisions by shifting the effective value of $E$ across the population.
+If government programs raise the floor for child endowments, then even low-income parents can achieve better outcomes without large private sacrifices.
+These policy-induced shifts in $E$ alter the perceived return to having more children or investing more per child.
+
+\citet{becker1976child} also considered the possibility that endowment is not randomly assigned but may vary systematically with income.
+Higher-income households may produce children with higher $E$ due to better maternal nutrition, access to prenatal care, lower exposure to environmental risk, or assortative matching on traits associated with educational or occupational success.
+In these families, not only are the resources available for investment greater, but the potential gains from investment may also be higher, because children are better positioned to benefit from those inputs.
+This interaction deepens the divide between high- and low-income households, making it harder for disadvantaged families to catch up.
+As a result, inequality can persist or even widen across generations.
+
+Finally, the model provides a mechanism for understanding how imperfections in credit markets can lead to persistent disadvantages.
+If parents with low income and low-$E$ children cannot borrow to finance quality-enhancing investment, then the next generation begins life with the same disadvantage.
+Without external intervention or structural change, this loop continues, which results in a pattern where poor families remain poor and rich families accumulate further advantage.
+The Becker-Tomes framework thus connects household-level decisions to bigger questions about the intergenerational transmission of human capital.
+
+Parallel to Becker and Tomes’s static analysis of endowments, another strand of the literature developed a fully dynamic version of the Q-Q model by incorporating parental altruism toward children’s welfare.
+In this framework, introduced by \citet{barro1989fertility}, parents care not only about the number and quality of their children but also about the utility their descendants will enjoy in the future.
+Altruism in this context means that parents treat their children's utility as part of their own, thus extending the household’s objective across generations.
+For example, a parent may reduce personal consumption to pay for a child’s schooling, motivated not just by the child's immediate benefit but by the satisfaction the parent gains from the child’s long-term success.
+This leads to a formulation of dynastic utility, where the household's objective spans infinitely many periods and takes the form of a recursive altruistic structure.
+A representative formulation is
+
+$$
+U_{0} = \sum_{t=0}^{\infty} \beta^{t}\,u(c_{t}, n_{t}),
+$$
+
+where $c_t$ denotes the consumption of the $t$-th generation, $n_t$ the number of children, and $\beta \in (0,1)$ the intertemporal discount factor.
+Given this structure, having an extra child $n_t$ enters utility positively, but each child is assumed to receive the same utility as the parent if raised at a comparable standard of living.
+As a result, parents confront an intertemporal trade-off: having more children expands the number of future utility streams but also stretches current resources, since each child requires support.
+This trade-off gives rise to an Euler equation for optimal fertility choice, analogous to an optimal growth condition.
+
+An implication of the dynastic model is that fertility decisions are sensitive to macroeconomic conditions, such as the interest rate or the rate of return on capital.
+A rise in interest rates increases the opportunity cost of channeling resources into children rather than saving, which tends to reduce current fertility—a substitution effect across generations.
+At the same time, higher returns make future generations wealthier, and this anticipated prosperity enters the utility calculations of parents in more complex ways.
+\citet{barro1989fertility} demonstrated that the model can account for observed fertility responses to economic fluctuations and policy interventions It can also explain historical phenomena such as postwar baby booms and subsequent fertility declines through shifts in returns or labor‐market opportunities.
+
+In many dynastic models, child quality appears indirectly, often through the child’s future human capital or income.
+One variant assumes parents value the aggregate human‐capital stock of their children.
+This specification, combined with altruism, produces a similar trade‐off: concentrating resources in fewer children raises each child’s human capital, which raises the dynasty’s long‐run welfare.
+These intergenerational extensions link micro‐level fertility decisions to macroeconomic outcomes.
+By the late 1980s, work by Becker, Barro, and others had recast fertility as an endogenous choice that interacts with capital accumulation, income distribution, and policy.
+This laid the foundation for unified growth theories, which view the quantity–quality mechanism as central to demographic transition and long-run development.
 

### 2025-06-17 docs/paper.tex
write theoretical background (intergenerational q-q model)

@@
 and child investment over multiple periods or even multiple generations.
-A central development in this literature was the incorporation of
+The main development in this literature was the incorporation of
 intergenerational human capital dynamics, where parents derive utility
@@
 being determined solely within a single period.
+
+One of the earliest and most influential models of this kind was
+proposed by \citet{becker1976child}. In their formulation, each child
+enters the world with an exogenous endowment \(E\), which may reflect
+factors such as cognitive ability (e.g., measured IQ or language
+acquisition speed), early health status (e.g., birth weight or incidence
+of neonatal complications), genetic predispositions (e.g., risk for
+chronic illness, temperament, or neurodevelopmental traits), or family
+background characteristics (e.g., parental education, household
+stability, or neighborhood conditions). Furthermore, parents can augment
+this endowment by investing resources \(q\) in the form of education,
+nutrition, and other quality-enhancing inputs. The effective adult human
+capital of the child might be expressed as \(H = E + f(q)\) (in a simple
+additive form) or a multiplicative variant \(H = E \cdot f(q)\), where
+\(f(q)\) is an increasing concave function of parental investment.
+
+\citet{becker1976child} emphasized that variation in endowments \(E\)
+can shape how parents allocate investments \(q\) across children. When
+the productivity of investment increases with endowment, parents may
+concentrate resources on children with higher \(E\), who are more likely
+to convert additional investment into future success. In other cases,
+parents may attempt to compensate for lower endowments by directing
+greater investment toward disadvantaged children. Put simply,
+child-specific variation in initial conditions affects not only outcomes
+but also strategic parental choices. Because of this, the relationship
+between income and the demand for child quality is not uniform. The
+income elasticity of demand for quality may differ across households,
+depending on the distribution of endowments within the family and across
+the broader population.
+
+Furthermore, \citet{becker1976child} showed that at low income levels,
+much of what constitutes child quality comes from exogenous endowments,
+i.e., factors like public education, neighborhood environment, or access
+to basic healthcare that are not privately purchased. In these settings,
+small increases in parental income may not lead to significant changes
+in fertility or investment behavior. Since most of the child's future
+outcomes are determined by the fixed endowment component, marginal
+investment plays a smaller role. However, as income rises, private
+resources become a larger part of what determines quality, and the
+classic Q--Q trade-off begins to shape behavior. Parents begin to
+allocate more income toward fewer children in order to enhance quality
+through direct investment. Under certain theoretical conditions, such as
+equal utility elasticities for quantity and quality, this framework
+produces a non-monotonic relationship between income and fertility.
+Fertility may decline as income rises at first, which reflects the
+desire to invest more intensively per child. However, beyond some point,
+once the marginal return to investment begins to flatten or saturate
+relative to the fixed endowment, fertility may increase again.
+
+The U-shaped prediction emerges only under specific assumptions, and its
+validity depends on both the shape of the utility function and how
+endowments relate to parental background. More broadly,
+\citet{becker1976child} enriched the Q--Q framework by incorporating
+elements that reflect real-world variation. They argued that not all
+differences in child outcomes are the result of deliberate parental
+choice. Random factors, biological traits, and socioeconomic settings
+play a role. In this light, public policies, such as subsidized
+schooling, early childhood programs, or universal healthcare, can
+influence private fertility and investment decisions by shifting the
+effective value of \(E\) across the population. If government programs
+raise the floor for child endowments, then even low-income parents can
+achieve better outcomes without large private sacrifices. These
+policy-induced shifts in \(E\) alter the perceived return to having more
+children or investing more per child.
+
+\citet{becker1976child} also considered the possibility that endowment
+is not randomly assigned but may vary systematically with income.
+Higher-income households may produce children with higher \(E\) due to
+better maternal nutrition, access to prenatal care, lower exposure to
+environmental risk, or assortative matching on traits associated with
+educational or occupational success. In these families, not only are the
+resources available for investment greater, but the potential gains from
+investment may also be higher, because children are better positioned to
+benefit from those inputs. This interaction deepens the divide between
+high- and low-income households, making it harder for disadvantaged
+families to catch up. As a result, inequality can persist or even widen
+across generations.
+
+Finally, the model provides a mechanism for understanding how
+imperfections in credit markets can lead to persistent disadvantages. If
+parents with low income and low-\(E\) children cannot borrow to finance
+quality-enhancing investment, then the next generation begins life with
+the same disadvantage. Without external intervention or structural
+change, this loop continues, which results in a pattern where poor
+families remain poor and rich families accumulate further advantage. The
+Becker-Tomes framework thus connects household-level decisions to bigger
+questions about the intergenerational transmission of human capital.
+
+Parallel to Becker and Tomes's static analysis of endowments, another
+strand of the literature developed a fully dynamic version of the Q-Q
+model by incorporating parental altruism toward children's welfare. In
+this framework, introduced by \citet{barro1989fertility}, parents care
+not only about the number and quality of their children but also about
+the utility their descendants will enjoy in the future. Altruism in this
+context means that parents treat their children's utility as part of
+their own, thus extending the household's objective across generations.
+For example, a parent may reduce personal consumption to pay for a
+child's schooling, motivated not just by the child's immediate benefit
+but by the satisfaction the parent gains from the child's long-term
+success. This leads to a formulation of dynastic utility, where the
+household's objective spans infinitely many periods and takes the form
+of a recursive altruistic structure. A representative formulation is
+
+\[
+U_{0} = \sum_{t=0}^{\infty} \beta^{t}\,u(c_{t}, n_{t}),
+\]
+
+where \(c_t\) denotes the consumption of the \(t\)-th generation,
+\(n_t\) the number of children, and \(\beta \in (0,1)\) the
+intertemporal discount factor. Given this structure, having an extra
+child \(n_t\) enters utility positively, but each child is assumed to
+receive the same utility as the parent if raised at a comparable
+standard of living. As a result, parents confront an intertemporal
+trade-off: having more children expands the number of future utility
+streams but also stretches current resources, since each child requires
+support. This trade-off gives rise to an Euler equation for optimal
+fertility choice, analogous to an optimal growth condition.
+
+An implication of the dynastic model is that fertility decisions are
+sensitive to macroeconomic conditions, such as the interest rate or the
+rate of return on capital. A rise in interest rates increases the
+opportunity cost of channeling resources into children rather than
+saving, which tends to reduce current fertility---a substitution effect
+across generations. At the same time, higher returns make future
+generations wealthier, and this anticipated prosperity enters the
+utility calculations of parents in more complex ways.
+\citet{barro1989fertility} demonstrated that the model can account for
+observed fertility responses to economic fluctuations and policy
+interventions It can also explain historical phenomena such as postwar
+baby booms and subsequent fertility declines through shifts in returns
+or labor‐market opportunities.
+
+In many dynastic models, child quality appears indirectly, often through
+the child's future human capital or income. One variant assumes parents
+value the aggregate human‐capital stock of their children. This
+specification, combined with altruism, produces a similar trade‐off:
+concentrating resources in fewer children raises each child's human
+capital, which raises the dynasty's long‐run welfare. These
+intergenerational extensions link micro‐level fertility decisions to
+macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and
+others had recast fertility as an endogenous choice that interacts with
+capital accumulation, income distribution, and policy. This laid the
+foundation for unified growth theories, which view the quantity--quality
+mechanism as central to demographic transition and long-run development.
 

### 2025-06-21 docs/paper.Rmd
write theoretical background (unified growth models)

@@
 Finally, I discuss more recent refinements that enrich the basic model by incorporating credit constraints \citep{doepke2004accounting}, intra-household bargaining \citep{doepke2019bargaining}, and multi-dimensional child quality \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis on health and nutrition.
-My focus is on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.
+The literature review focuses on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.
 
@@
 \subsubsection{Becker’s Static Model}
-Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}.
+
+Becker’s early work introduced an economic model of fertility and treated children as durable goods that provide utility to parents but impose costs \citep{doepke2015gary}.
 In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$).
@@
 with $U$ increasing in each argument up to some satiation point.
-Here quality $q$ can be thought of as the expenditure or investment per child (e.g.
+Here, quality $q$ can be thought of as the expenditure or investment per child (e.g.
 education spending, health care, nutrition), assumed for now to be the same for each child.
@@
 
+\subsubsection{Unified Growth Models}
+
+The unified growth theory, developed in the late 1990s and 2000s (notably by Oded Galor and co-authors), seeks to explain in one framework the entire sweep of economic development – from Malthusian stagnation, through the demographic transition, to modern growth.
+A central puzzle it addresses is why fertility rates, which were historically high and invariant to income in the Malthusian era, began to decline sharply in tandem with industrialization and rising incomes, eventually stabilizing at much lower levels in developed economies.
+The Q–Q trade-off provides a key part of the answer in these models.
+\citet{galor2000population} and \citet{galor2002natural} explicitly incorporate parental choices about the quantity and quality of children and show how changes in the economic environment alter those choices and trigger demographic transitions.
+
+In \citet{galor2000population}’s model, for instance, technological progress gradually increases the return to human capital, especially in skilled occupations.
+In the early stages, when production relies on basic tools and techniques, unskilled labor holds more value.
+Under these conditions, parents have little reason to invest in formal schooling.
+Children are expected to contribute economically through agricultural work, domestic tasks, or low-skill jobs in workshops and factories.
+Fertility remains high because children impose a low financial burden and generate immediate returns.
+As technology becomes more advanced—such as during the Industrial Revolution—the earnings gap between skilled and unskilled labor widens.
+Education begins to offer significant advantages in the labor market.
+In response, parents adjust by having fewer children and placing greater emphasis on each child’s development, including school attendance and better health care.
+
+Evidently, industrialization raises the economic value of skilled labor, which alters household incentives.
+As returns to education increase, parents begin to favor investments in child quality over child quantity.
+This shift results in declining fertility because families choose to have fewer children and allocate more resources to each.
+The feedback effect is significant: higher educational investment raises productivity in the next generation, which in turn accelerates technological advancement and further increases the returns to human capital.
+Over time, the economy moves from a state of high fertility and low growth to one characterized by low fertility and sustained growth.
+
+Furthermore, \citet{galor2002natural} introduced an evolutionary refinement to the unified growth framework by accounting for heterogeneity in parental preferences.
+During the Malthusian period, some families placed greater emphasis on child quality, such as education, while others prioritized quantity.
+In a stagnant economy with limited returns to education, high-fertility lineages maintained a numerical advantage and suppressed average human capital.
+As technological progress increased the returns to education, families that valued quality gained an economic edge.
+Their children acquired more human capital and achieved higher income and survival rates.
+These advantages allowed such families to grow in relative size.
+Over time, this process resembled a form of evolutionary selection, gradually favoring quality-oriented parental types and shifting the population toward greater emphasis on child human capital.
+These dynamics strengthened the shift from high-fertility, low-education regimes to low-fertility, high-investment family structures.
+In formal overlapping-generations models, \citet{galor2002natural} demonstrate that this evolutionary adaptation accelerates the demographic transition.
+Their framework accounts for the rapid and widespread drop in fertility once it takes effect.
+Higher returns to human capital push parents to favor quality, while preferences for quality begin to dominate within the population.
+These forces support the emergence of a low-fertility, high-investment equilibrium and establish a unified explanation for both economic development and demographic change.
+
+Importantly, unified growth models identify several complementary mechanisms that reinforce the basic Q–Q trade-off during development.
+One is the decline of child labor.
+As the economy modernizes, the value of child labor falls, both because legal reforms often restrict child labor and because parents realize the earnings their children could make as unskilled laborers are paltry compared to the potential returns if those children instead spend time in school.
+\citet{hazan2002child} formally show that when child labor becomes less profitable relative to adult (skilled) labor, parents further reduce fertility and invest more in each child’s education.
+Historical evidence from England, for instance, indicates that during industrialization the wages of children (relative to adults) dropped significantly, especially in skilled families, and this was accompanied by parents pulling children out of work to send them to school.
+\citet{galor2006human} even argue that capitalist industrialists supported public education laws and child labor bans as a way to increase the human capital of the workforce, inadvertently hastening the fertility transition.
+
+Another mechanism is the rise in life expectancy and child survival.
+Improvements in sanitation, nutrition, and medical knowledge in developing societies led to more children surviving to adulthood.
+While the earliest unified growth models treated mortality as exogenous or ignored it, later research demonstrated that declining child mortality can trigger lower fertility as well– parents no longer need “extra” births for insurance once they are confident their existing children will survive.
+In other words, increased child survival and the quality–quantity trade-off are complementary explanations for fertility decline that operate in tandem.
+When fewer births are lost to disease, parents can achieve a desired number of surviving offspring with fewer total births, and they tend to reallocate effort into each child’s health and education.
+
+The overall effect is a reinforcing cycle: better health raises the returns to schooling (healthier children can learn more effectively and have longer working lives), which further encourages educational investments and reduces fertility.
+Indeed, Galor notes that human capital should be interpreted broadly to include health as well as schooling; in unified growth theory, improvements in nutrition and physical well-being were crucial to making labor more productive and thus were part and parcel of the rise in demand for human capital.
+
+The unified growth literature places the Q–Q model within a more general account of economic and demographic change.
+In this framework, higher income or stronger returns to child quality reduce fertility and help shift economies from stagnation toward sustained growth.
+Several mechanisms support this transition, such as a fall in child labor, a drop in child mortality, and a shift in parental priorities.
+These models explain not only the presence of a quantity–quality trade-off but also its rising influence at a specific point in history.
+The evidence supports these claims: countries that saw earlier increases in returns to education experienced earlier fertility decline, while delays in reforms, such as public education or health access, corresponded to prolonged high fertility.
+As a result, the Q–Q mechanism forms a key component of unified growth theory.
+
 \bibliographystyle{aea}

### 2025-06-21 docs/paper.tex
write theoretical background (unified growth models)

@@
 \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis
-on health and nutrition. My focus is on how the Q--Q framework has been
-applied to understand fertility and child investment patterns,
-especially in developing country contexts where resource constraints and
-health outcomes are paramount.
+on health and nutrition. The literature review focuses on how the Q--Q
+framework has been applied to understand fertility and child investment
+patterns, especially in developing country contexts where resource
+constraints and health outcomes are paramount.
 
@@
 
-Becker's early work introduced an economic approach to fertility,
-treating children analogous to ``durable goods'' that yield utility to
-parents but come at a cost \citep{doepke2015gary}. In Becker's 1960
-model, a household derives satisfaction from the number of children
-(\(n\)) and from the quality of each child (\(q\)), alongside
-conventional consumption of other goods (\(y\)). A simple representation
-is a utility function:
+\subsubsection{Becker’s Static Model}
+
+Becker's early work introduced an economic model of fertility and
+treated children as durable goods that provide utility to parents but
+impose costs \citep{doepke2015gary}. In Becker's 1960 model, a household
+derives satisfaction from the number of children (\(n\)) and from the
+quality of each child (\(q\)), alongside conventional consumption of
+other goods (\(y\)). A simple representation is a utility function:
 
@@
 
-with \(U\) increasing in each argument up to some satiation point. Here
+with \(U\) increasing in each argument up to some satiation point. Here,
 quality \(q\) can be thought of as the expenditure or investment per
@@
 between the number of children and investments in each.
+
+\subsubsection{Intergenerational Models: Altruism and Child Endowments}
 
@@
 
+\subsubsection{Unified Growth Models}
+
+The unified growth theory, developed in the late 1990s and 2000s
+(notably by Oded Galor and co-authors), seeks to explain in one
+framework the entire sweep of economic development -- from Malthusian
+stagnation, through the demographic transition, to modern growth. A
+central puzzle it addresses is why fertility rates, which were
+historically high and invariant to income in the Malthusian era, began
+to decline sharply in tandem with industrialization and rising incomes,
+eventually stabilizing at much lower levels in developed economies. The
+Q--Q trade-off provides a key part of the answer in these models.
+\citet{galor2000population} and \citet{galor2002natural} explicitly
+incorporate parental choices about the quantity and quality of children
+and show how changes in the economic environment alter those choices and
+trigger demographic transitions.
+
+In \citet{galor2000population}'s model, for instance, technological
+progress gradually increases the return to human capital, especially in
+skilled occupations. In the early stages, when production relies on
+basic tools and techniques, unskilled labor holds more value. Under
+these conditions, parents have little reason to invest in formal
+schooling. Children are expected to contribute economically through
+agricultural work, domestic tasks, or low-skill jobs in workshops and
+factories. Fertility remains high because children impose a low
+financial burden and generate immediate returns. As technology becomes
+more advanced---such as during the Industrial Revolution---the earnings
+gap between skilled and unskilled labor widens. Education begins to
+offer significant advantages in the labor market. In response, parents
+adjust by having fewer children and placing greater emphasis on each
+child's development, including school attendance and better health care.
+
+Evidently, industrialization raises the economic value of skilled labor,
+which alters household incentives. As returns to education increase,
+parents begin to favor investments in child quality over child quantity.
+This shift results in declining fertility because families choose to
+have fewer children and allocate more resources to each. The feedback
+effect is significant: higher educational investment raises productivity
+in the next generation, which in turn accelerates technological
+advancement and further increases the returns to human capital. Over
+time, the economy moves from a state of high fertility and low growth to
+one characterized by low fertility and sustained growth.
+
+Furthermore, \citet{galor2002natural} introduced an evolutionary
+refinement to the unified growth framework by accounting for
+heterogeneity in parental preferences. During the Malthusian period,
+some families placed greater emphasis on child quality, such as
+education, while others prioritized quantity. In a stagnant economy with
+limited returns to education, high-fertility lineages maintained a
+numerical advantage and suppressed average human capital. As
+technological progress increased the returns to education, families that
+valued quality gained an economic edge. Their children acquired more
+human capital and achieved higher income and survival rates. These
+advantages allowed such families to grow in relative size. Over time,
+this process resembled a form of evolutionary selection, gradually
+favoring quality-oriented parental types and shifting the population
+toward greater emphasis on child human capital. These dynamics
+strengthened the shift from high-fertility, low-education regimes to
+low-fertility, high-investment family structures. In formal
+overlapping-generations models, \citet{galor2002natural} demonstrate
+that this evolutionary adaptation accelerates the demographic
+transition. Their framework accounts for the rapid and widespread drop
+in fertility once it takes effect. Higher returns to human capital push
+parents to favor quality, while preferences for quality begin to
+dominate within the population. These forces support the emergence of a
+low-fertility, high-investment equilibrium and establish a unified
+explanation for both economic development and demographic change.
+
+Importantly, unified growth models identify several complementary
+mechanisms that reinforce the basic Q--Q trade-off during development.
+One is the decline of child labor. As the economy modernizes, the value
+of child labor falls, both because legal reforms often restrict child
+labor and because parents realize the earnings their children could make
+as unskilled laborers are paltry compared to the potential returns if
+those children instead spend time in school. \citet{hazan2002child}
+formally show that when child labor becomes less profitable relative to
+adult (skilled) labor, parents further reduce fertility and invest more
+in each child's education. Historical evidence from England, for
+instance, indicates that during industrialization the wages of children
+(relative to adults) dropped significantly, especially in skilled
+families, and this was accompanied by parents pulling children out of
+work to send them to school. \citet{galor2006human} even argue that
+capitalist industrialists supported public education laws and child
+labor bans as a way to increase the human capital of the workforce,
+inadvertently hastening the fertility transition.
+
+Another mechanism is the rise in life expectancy and child survival.
+Improvements in sanitation, nutrition, and medical knowledge in
+developing societies led to more children surviving to adulthood. While
+the earliest unified growth models treated mortality as exogenous or
+ignored it, later research demonstrated that declining child mortality
+can trigger lower fertility as well-- parents no longer need ``extra''
+births for insurance once they are confident their existing children
+will survive. In other words, increased child survival and the
+quality--quantity trade-off are complementary explanations for fertility
+decline that operate in tandem. When fewer births are lost to disease,
+parents can achieve a desired number of surviving offspring with fewer
+total births, and they tend to reallocate effort into each child's
+health and education.
+
+The overall effect is a reinforcing cycle: better health raises the
+returns to schooling (healthier children can learn more effectively and
+have longer working lives), which further encourages educational
+investments and reduces fertility. Indeed, Galor notes that human
+capital should be interpreted broadly to include health as well as
+schooling; in unified growth theory, improvements in nutrition and
+physical well-being were crucial to making labor more productive and
+thus were part and parcel of the rise in demand for human capital.
+
+The unified growth literature places the Q--Q model within a more
+general account of economic and demographic change. In this framework,
+higher income or stronger returns to child quality reduce fertility and
+help shift economies from stagnation toward sustained growth. Several
+mechanisms support this transition, such as a fall in child labor, a
+drop in child mortality, and a shift in parental priorities. These
+models explain not only the presence of a quantity--quality trade-off
+but also its rising influence at a specific point in history. The
+evidence supports these claims: countries that saw earlier increases in
+returns to education experienced earlier fertility decline, while delays
+in reforms, such as public education or health access, corresponded to
+prolonged high fertility. As a result, the Q--Q mechanism forms a key
+component of unified growth theory.
+
 \bibliographystyle{aea}

### 2025-07-01 docs/paper.Rmd
write theoretical background (recent refinements - credit constraints)

@@
 
+\subsubsection{Recent Refinements to the Q-Q Model}
+
+Contemporary research has further refined the quantity–quality model by relaxing some of its initial simplifying assumptions.
+Three important extensions involve (1) capital market imperfections that constrain parents’ ability to invest in child quality, (2) intra-household conflict and bargaining between mothers and fathers over fertility choices, and (3) recognition that child quality is multi-dimensional, which extends beyond schooling to include health, nutrition, and other facets of human capital.
+
+To begin, I examine how credit constraints can give rise to poverty traps.
+The canonical quantity–quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high.
+In practice, especially in low-income settings, credit markets function imperfectly: poor households typically cannot secure loans to cover children’s education or medical costs even when such investments would yield substantial future gains.
+This market failure magnifies the Q–Q trade-off.
+Becker, Lewis, and Willis–already noted by \citet{grawe2008quality} as emphasizing “resource limitations”–implicitly recognized this issue, but contemporary models make it explicit by imposing a borrowing constraint.
+Parents must fund childrearing from current income alone; they cannot collateralize a child’s future wages to pay today’s school fees.
+Consequently, when income is low, each additional birth directly reduces the attainable quality per child, potentially trapping families in a low-income–high-fertility equilibrium.
+
+Formal models show that “the quality–quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism.
+\citet{kremer2002income} and \citet{de2003inequality} document that countries facing tighter liquidity constraints tend to display higher fertility and lower educational attainment, consistent with liquidity-constrained parents favoring quantity over quality.
+Cross-country evidence likewise indicates that where financial frictions are more severe, the negative correlation between fertility and schooling is stronger.
+Theoretically, introducing a borrowing limit can generate multiple steady states: one with low fertility and high investment when incomes suffice to cover quality costs, and another with high fertility and minimal investment when they do not.
+Policy instruments such as education subsidies or conditional cash transfers effectively relax these constraints, which nudge households toward the low-fertility, high-investment equilibrium.
+In short, incorporating credit market imperfections deepens the explanatory power of the Q-Q model: economic growth alone may not reduce fertility if households remain too cash-poor to afford schooling, whereas targeted quality-enhancing transfers can catalyze both demographic and human-capital transitions.
+
 \bibliographystyle{aea}

### 2025-07-01 docs/paper.tex
write theoretical background (recent refinements - credit constraints)

@@
 
+\subsubsection{Recent Refinements to the Q-Q Model}
+
+Contemporary research has further refined the quantity--quality model by
+relaxing some of its initial simplifying assumptions. Three important
+extensions involve (1) capital market imperfections that constrain
+parents' ability to invest in child quality, (2) intra-household
+conflict and bargaining between mothers and fathers over fertility
+choices, and (3) recognition that child quality is multi-dimensional,
+which extends beyond schooling to include health, nutrition, and other
+facets of human capital.
+
+To begin, I examine how credit constraints can give rise to poverty
+traps. The canonical quantity--quality (Q--Q) framework assumes that
+parents can reallocate resources freely; borrowing against future
+earnings to finance schooling or health investments whenever the
+expected return is high. In practice, especially in low-income settings,
+credit markets function imperfectly: poor households typically cannot
+secure loans to cover children's education or medical costs even when
+such investments would yield substantial future gains. This market
+failure magnifies the Q--Q trade-off. Becker, Lewis, and Willis--already
+noted by \citet{grawe2008quality} as emphasizing ``resource
+limitations''--implicitly recognized this issue, but contemporary models
+make it explicit by imposing a borrowing constraint. Parents must fund
+childrearing from current income alone; they cannot collateralize a
+child's future wages to pay today's school fees. Consequently, when
+income is low, each additional birth directly reduces the attainable
+quality per child, potentially trapping families in a
+low-income--high-fertility equilibrium.
+
+Formal models show that ``the quality--quantity trade-off arises from a
+binding credit constraint that prevents parents from borrowing against
+future child income.'' Empirical work supports this mechanism.
+\citet{kremer2002income} and \citet{de2003inequality} document that
+countries facing tighter liquidity constraints tend to display higher
+fertility and lower educational attainment, consistent with
+liquidity-constrained parents favoring quantity over quality.
+Cross-country evidence likewise indicates that where financial frictions
+are more severe, the negative correlation between fertility and
+schooling is stronger. Theoretically, introducing a borrowing limit can
+generate multiple steady states: one with low fertility and high
+investment when incomes suffice to cover quality costs, and another with
+high fertility and minimal investment when they do not. Policy
+instruments such as education subsidies or conditional cash transfers
+effectively relax these constraints, which nudge households toward the
+low-fertility, high-investment equilibrium. In short, incorporating
+credit market imperfections deepens the explanatory power of the Q-Q
+model: economic growth alone may not reduce fertility if households
+remain too cash-poor to afford schooling, whereas targeted
+quality-enhancing transfers can catalyze both demographic and
+human-capital transitions.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.Rmd
write theoretical background (recent refinements - intrahousehold bargaining)s)

@@
 
+A further refinement of the quantity–quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions.
+The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \).
+In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines.
+According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health.
+These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.
+
+In such models, the mother is typically assumed to have a stronger preference for child quality—such as health and schooling—while the father may favor either more children or alternative uses of household resources.
+The resolution of these conflicting preferences depends on bargaining power, which can be shaped by income contributions, legal rights, cultural norms, or access to external resources.
+When the mother’s bargaining power increases, theoretical models predict a shift toward lower fertility and higher per-child investment, holding other factors constant.
+This prediction is consistent with empirical findings: \citet{iyigun2007endogenous, doepke2019bargaining} show that greater female empowerment—via education or labor force participation—correlates with reduced fertility and increased investment in child human capital.
+
+Mathematically, the household’s first-order condition for fertility in a bargaining model can be written as:
+
+\[
+\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
+\]
+
+with a corresponding condition for \( q \).
+Here, \( \alpha \in [0,1] \) represents the wife’s bargaining weight.
+When \( \alpha \) increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.
+
+Recent work also explores dynamic bargaining, in which spouses negotiate sequential decisions over time, potentially leading to strategic behavior (e.g., one partner may accelerate or delay subsequent births).
+Although these models introduce complexity, their core implication for the Q–Q framework is clear: household power dynamics fundamentally shape the trade-off between child quantity and quality.
+In societies where women have limited decision-making autonomy—due to lack of access to contraception, or social norms—fertility tends to remain high and per-child investment low, which stalls demographic transition.
+Conversely, when women gain bargaining power—through legal reforms, labor market participation, or targeted transfers—the household often reallocates resources toward fewer but higher-quality children.
+
+This theoretical insight is corroborated by policy experiments.
+For instance, cash transfer programs directed to mothers consistently lead to greater spending on children’s health, education, and nutrition compared to equivalent transfers given to fathers \citep{duflo2003grandmothers, thomas1990intra} .
+Such outcomes support the hypothesis that mothers place higher weight on child quality, and that who controls the purse strings matters deeply.
+In sum, introducing intra-household bargaining into the Q–Q model enriches its explanatory scope: it highlights how family structure and power asymmetries—not just income levels or prices—generate variation in fertility and human capital outcomes across and within societies.
+
+\subsection{Health and Nutrition in the Q–Q Model}
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.tex
write theoretical background (recent refinements - intrahousehold bargaining)s)

@@
 \date{\today}
-\pubMonth{06}
+\pubMonth{07}
 \pubYear{2025}
@@
 
+A further refinement of the quantity--quality (Q--Q) model considers the
+question of \emph{who} within the household makes fertility and child
+investment decisions. The original Beckerian framework adopts a unitary
+model of the family, and assumes a single utility function and complete
+agreement between spouses over optimal fertility \(n\) and child quality
+\(q\). In practice, however, empirical evidence reveals significant
+heterogeneity in preferences between household members along gender
+lines. According to \citet{oppenheim1987impact, thomas1990intra}, for
+instance, men often desire more children than women and may differ in
+their willingness to invest in each child's education or health. These
+discrepancies have motivated game-theoretic models of intra-household
+bargaining, in which fertility and investment outcomes reflect the
+relative influence of each parent's preferences.
+
+In such models, the mother is typically assumed to have a stronger
+preference for child quality---such as health and schooling---while the
+father may favor either more children or alternative uses of household
+resources. The resolution of these conflicting preferences depends on
+bargaining power, which can be shaped by income contributions, legal
+rights, cultural norms, or access to external resources. When the
+mother's bargaining power increases, theoretical models predict a shift
+toward lower fertility and higher per-child investment, holding other
+factors constant. This prediction is consistent with empirical findings:
+\citet{iyigun2007endogenous, doepke2019bargaining} show that greater
+female empowerment---via education or labor force
+participation---correlates with reduced fertility and increased
+investment in child human capital.
+
+Mathematically, the household's first-order condition for fertility in a
+bargaining model can be written as:
+
+\[
+\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
+\]
+
+with a corresponding condition for \(q\). Here, \(\alpha \in [0,1]\)
+represents the wife's bargaining weight. When \(\alpha\) increases, the
+composite marginal utility of additional children typically
+decreases---especially if the wife prefers fewer children---leading to
+lower equilibrium fertility and a shift along the Q--Q frontier toward
+higher quality.
+
+Recent work also explores dynamic bargaining, in which spouses negotiate
+sequential decisions over time, potentially leading to strategic
+behavior (e.g., one partner may accelerate or delay subsequent births).
+Although these models introduce complexity, their core implication for
+the Q--Q framework is clear: household power dynamics fundamentally
+shape the trade-off between child quantity and quality. In societies
+where women have limited decision-making autonomy---due to lack of
+access to contraception, or social norms---fertility tends to remain
+high and per-child investment low, which stalls demographic transition.
+Conversely, when women gain bargaining power---through legal reforms,
+labor market participation, or targeted transfers---the household often
+reallocates resources toward fewer but higher-quality children.
+
+This theoretical insight is corroborated by policy experiments. For
+instance, cash transfer programs directed to mothers consistently lead
+to greater spending on children's health, education, and nutrition
+compared to equivalent transfers given to fathers
+\citep{duflo2003grandmothers, thomas1990intra} . Such outcomes support
+the hypothesis that mothers place higher weight on child quality, and
+that who controls the purse strings matters deeply. In sum, introducing
+intra-household bargaining into the Q--Q model enriches its explanatory
+scope: it highlights how family structure and power asymmetries---not
+just income levels or prices---generate variation in fertility and human
+capital outcomes across and within societies.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.Rmd
write health and nutrition in the q-q model
Expanded the paper to include a new subsection discussing the role of health and nutrition in the quantity–quality (Q–Q) model of fertility and child investment. Added theoretical discussion, empirical evidence, and policy implications, along with new references and citations supporting the extended framework.

@@
 
+The original Q–Q models often used a single catch-all variable for child quality, typically thought of as education or “expenditure per child.” Recent work emphasizes that child quality is multi-faceted, and that parents make trade-offs along several dimensions of investment—cognitive development, health, nutrition, etc.
+This is particularly salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.
+
+The theoretical question is how these dimensions interact with the quantity decision.
+If parents allocate a budget across, say, schooling \( q_{\text{edu}} \) and nutrition/health \( q_{\text{health}} \) for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them).
+In some models, health and education are complementary: a healthier child benefits more from education, and an educated mother might raise a healthier child.
+This complementarity can amplify the Q–Q trade-off—investing in one dimension (health) increases the returns to investing in the other (education), so a high-quality strategy becomes more focused on fewer children.
+
+On the other hand, if one dimension has diminishing returns more quickly than another, parents might prioritize achieving a threshold level of health for all children before adding more education, which creates a nonlinear effect on fertility.
+One especially important aspect of health in the Q–Q framework is child survival.
+The probability that a child survives to adulthood effectively multiplies the utility of having that child.
+Historically, high child mortality led to a strategy of “quantity for insurance,” where parents had additional births to ensure survivorship.
+As mortality falls due to public health improvements, parents can shift toward quality without risking childlessness \citep{kalemli2002does, kalemli2000mortality}.
+
+\citet{kalemli2002does} developed a stochastic model in which fertility choices are made under uncertainty about child survival.
+Her results show that declining mortality causally reduces fertility and increases educational investment per child, in line with the Q–Q trade-off.
+In unified growth models, declining child mortality reinforces the demand-for-human-capital channel of the demographic transition \citep{galor2004physical}.
+
+Beyond survival, nutritional status is a critical indicator of quality in many low-income settings.
+Parents often face choices about how to allocate food or medical care, especially under resource scarcity.
+Empirical evidence supports the Q–Q model in this context: children with many siblings often exhibit poorer nutritional outcomes.
+A recent study using Vietnam Young Lives data finds that an additional sibling reduces height-for-age and weight-for-age z-scores by about 0.3 standard deviations on average \citep{chen2021population}  This effect is substantial, suggesting that children in larger families tend to be shorter and lighter than their peers from smaller families, likely due to the more limited distribution of healthcare and parental attention
+
+Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia , where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs.
+Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:
+
+\[
+H = g(F, M)
+\]
+
+and \( H \) enters either utility directly or the child’s future productivity.
+Parents then choose \( n \), \( F \), \( M \), and possibly schooling per child.
+Larger \( n \) reduces \( F \) and \( M \) per child, lowering \( H \); if parents place high value on \( H \), or if \( H \) enhances returns to education, they will opt for smaller families.
+
+Importantly, improvements in nutrition (through income growth or public programs) can first lead to both better health and higher fertility (since healthier women can bear more children).
+But over time, as standards of living rise, better nutrition increases the returns to investing in fewer, healthier children.
+Economic historian \citet{fogel1994economic} argues that Europe’s demographic transition was partly driven by better diets: improved nutrition raised the productivity of educated workers, which made education more valuable and shifted family preferences toward quality.
+
+These propositions are important for modern development policy.
+Family planning programs that encourage lower fertility without improving child health and education may have limited long-term effects.
+Conversely, health and nutrition interventions—such as vaccinations or food supplementation—can enhance the returns to education and induce parents to reduce fertility voluntarily.
+The extended Q–Q framework thus serves as a unifying tool to understand how diverse interventions—ranging from school fees to nutrition programs—shape long-run development trajectories.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.tex
write health and nutrition in the q-q model
Expanded the paper to include a new subsection discussing the role of health and nutrition in the quantity–quality (Q–Q) model of fertility and child investment. Added theoretical discussion, empirical evidence, and policy implications, along with new references and citations supporting the extended framework.

@@
 
+\subsection{Health and Nutrition in the Q–Q Model}
+
+The original Q--Q models often used a single catch-all variable for
+child quality, typically thought of as education or ``expenditure per
+child.'' Recent work emphasizes that child quality is multi-faceted, and
+that parents make trade-offs along several dimensions of
+investment---cognitive development, health, nutrition, etc. This is
+particularly salient in developing countries, where basic health and
+nutrition are pressing concerns alongside schooling.
+
+The theoretical question is how these dimensions interact with the
+quantity decision. If parents allocate a budget across, say, schooling
+\(q_{\text{edu}}\) and nutrition/health \(q_{\text{health}}\) for each
+child, then having more children forces cutbacks in both dimensions
+(unless parents reallocate across them). In some models, health and
+education are complementary: a healthier child benefits more from
+education, and an educated mother might raise a healthier child. This
+complementarity can amplify the Q--Q trade-off---investing in one
+dimension (health) increases the returns to investing in the other
+(education), so a high-quality strategy becomes more focused on fewer
+children.
+
+On the other hand, if one dimension has diminishing returns more quickly
+than another, parents might prioritize achieving a threshold level of
+health for all children before adding more education, which creates a
+nonlinear effect on fertility. One especially important aspect of health
+in the Q--Q framework is child survival. The probability that a child
+survives to adulthood effectively multiplies the utility of having that
+child. Historically, high child mortality led to a strategy of
+``quantity for insurance,'' where parents had additional births to
+ensure survivorship. As mortality falls due to public health
+improvements, parents can shift toward quality without risking
+childlessness \citep{kalemli2002does, kalemli2000mortality}.
+
+\citet{kalemli2002does} developed a stochastic model in which fertility
+choices are made under uncertainty about child survival. Her results
+show that declining mortality causally reduces fertility and increases
+educational investment per child, in line with the Q--Q trade-off. In
+unified growth models, declining child mortality reinforces the
+demand-for-human-capital channel of the demographic transition
+\citep{galor2004physical}.
+
+Beyond survival, nutritional status is a critical indicator of quality
+in many low-income settings. Parents often face choices about how to
+allocate food or medical care, especially under resource scarcity.
+Empirical evidence supports the Q--Q model in this context: children
+with many siblings often exhibit poorer nutritional outcomes. A recent
+study using Vietnam Young Lives data finds that an additional sibling
+reduces height-for-age and weight-for-age z-scores by about 0.3 standard
+deviations on average \citep{chen2021population} This effect is
+substantial, suggesting that children in larger families tend to be
+shorter and lighter than their peers from smaller families, likely due
+to the more limited distribution of healthcare and parental attention
+
+Similar patterns emerge in sub-Saharan Africa
+\citep{bishwakarma2019first} and South Asia , where first-born children
+experience slower growth once younger siblings arrive---again pointing
+to intrahousehold trade-offs. Theoretically, one can extend the Becker
+model with a nutrition production function, where a child's health \(H\)
+depends on food \(F\) and medical care \(M\), such that:
+
+\[
+H = g(F, M)
+\]
+
+and \(H\) enters either utility directly or the child's future
+productivity. Parents then choose \(n\), \(F\), \(M\), and possibly
+schooling per child. Larger \(n\) reduces \(F\) and \(M\) per child,
+lowering \(H\); if parents place high value on \(H\), or if \(H\)
+enhances returns to education, they will opt for smaller families.
+
+Importantly, improvements in nutrition (through income growth or public
+programs) can first lead to both better health and higher fertility
+(since healthier women can bear more children). But over time, as
+standards of living rise, better nutrition increases the returns to
+investing in fewer, healthier children. Economic historian
+\citet{fogel1994economic} argues that Europe's demographic transition
+was partly driven by better diets: improved nutrition raised the
+productivity of educated workers, which made education more valuable and
+shifted family preferences toward quality.
+
+These propositions are important for modern development policy. Family
+planning programs that encourage lower fertility without improving child
+health and education may have limited long-term effects. Conversely,
+health and nutrition interventions---such as vaccinations or food
+supplementation---can enhance the returns to education and induce
+parents to reduce fertility voluntarily. The extended Q--Q framework
+thus serves as a unifying tool to understand how diverse
+interventions---ranging from school fees to nutrition programs---shape
+long-run development trajectories.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.Rmd
write q-q evidence in the philippines and southeast asia

@@
 Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes.
-Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010effects}.
+Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}.
 These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality.
@@
 
-\subsubsection{Recent Refinements to the Q-Q Model}
+\subsection{Recent Refinements to the Q-Q Model}
 
@@
 
+\subsection{Philippine Evidence on the Q–Q Trade-off}
+
+Early empirical work in the Philippines provided suggestive support for this theory.
+For example, an influential study by \citet{horton1986child} used Philippine household data to jointly examine fertility and child nutrition and treated nutritional status as a measure of child quality.
+Horton found that better-educated mothers and fathers tended to choose smaller families and achieved better-nourished children–evidence of substitution away from “quantity” toward child “quality.” Notably, she observed strong birth-order effects on nutrition (first-born children faring better than later-born), which hint that parents may not evenly distribute quality investments among all children.
+This early work indicated a quantity–quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.
+
+Subsequent studies in the Philippines have tackled the critical issue of causality, using innovative research designs to isolate exogenous changes in family size.
+One seminal contribution is \citet{orbeta2010number}, who examined the impact of family size on children’s schooling using a nationally representative survey.
+Recognizing that family size is endogenous (parents’ fertility choices may reflect unobserved preferences or constraints), Orbeta employed an instrumental-variable (IV) approach grounded in Becker’s framework.
+Specifically, he used the sex composition of the first two children as an instrument for having additional children – an approach pioneered by \citet{angrist1996children} – leveraging the fact that Filipino parents often desire a mixed-gender sibset (and are more likely to have a third child if the first two are the same sex) \citep{vicerra2013fertility}.
+This strategy aims to generate random-like variation in family size uncorrelated with parental characteristics.
+
+The IV estimates confirmed a significant negative causal effect of higher fertility on educational outcomes.
+\citet{orbeta2010number} found that each additional child in the household reduced the proportion of school-age children (6–24 years) attending school by roughly 19% of the baseline attendance rate.
+The trade-off was especially pronounced at higher education levels: for example, the estimated drop in school attendance was about 26% at the secondary level and 57% at the tertiary level for each additional sibling.
+These are sizable effects, implying that children from large families are substantially less likely to remain in school, presumably due to tighter household budget constraints or diluted parental attention.
+Moreover, the burden of the trade-off appeared regressive: Orbeta’s results showed much larger schooling deficits from an extra child in poorer households than in richer ones.
+For instance, in the poorest quintile, an additional sibling reduced school attendance by an estimated 24% (for ages 6–24), compared to a 16% reduction in the richest quintile.
+This regressive pattern aligns with Becker’s theory that resource constraints bind more tightly for low-income families, which makes the Q–Q trade-off more acute.
+In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity–quality trade-off in this context.
+Large family size thus emerges as one mechanism contributing to poverty, by impeding children’s human capital accumulation in the Philippines.
+
+Further compelling evidence comes from a natural experiment studied by \citet{dumas2019sex}.
+They exploit a unique policy shock in metropolitan Manila to identify the trade-off.
+In 1998, the Mayor of Manila city imposed a sudden ban on modern contraceptives in public facilities, drastically curtailing access to family planning for residents of Manila city (but not in surrounding municipalities).
+Dumas and Lefranc use this policy as a quasi-experiment: comparing families in Manila city (treated by the ban) to similar families in other cities unaffected by the ban before and after 1998.
+This difference-in-differences design, coupled with the fact that older mothers were naturally less fecund during the ban, isolates an exogenous fertility increase.
+
+The results are striking.
+The contraceptives ban led to a significant rise in births and family size in Manila city relative to the control areas.
+Correspondingly, children born in Manila during the ban era experienced a sizable decline in educational attainment – clear evidence of a Q–Q trade-off precipitated by the shock.
+In the authors’ words, the policy-driven increase in family size “provide[s] evidence of a quality–quantity trade-off”: larger families, forced by the ban, resulted in lower schooling outcomes per child.
+By leveraging an actual anti-contraception policy, this study offers causal confirmation that constraints on family planning can worsen child quality.
+Together with Orbeta’s findings, it reinforces the relevance of Becker’s hypothesis in the Philippine setting: whether via unintended fertility (Manila’s case) or ordinary variation in sibling sex mix, more children mean fewer resources per child and thus worse education outcomes.
+Consistent with theory, Filipino parents facing exogenous increases in quantity were unable to maintain the same level of quality per child.
+
+Beyond education, researchers have also examined health and other child outcomes in relation to family size.
+Evidence generally suggests the Q–Q trade-off extends to child health and nutrition.
+For instance, \citet{horton1986child} already hinted that large families may compromise child nutrition for later-born siblings.
+More recent regional research resonates with this.
+\citet{hatton2018fertility} analyze the effect of fertility on child height (a long-run health indicator) using longitudinal data from Indonesia – a neighboring Southeast Asian country with comparable developmental challenges.
+They address endogeneity by exploiting Indonesia’s family planning program rollout and exposure to mass media as instruments for fertility.
+The authors find a significant negative impact of family size on child health: each additional sibling is associated with about a one-third standard deviation reduction in a child’s height-for-age Z-score, after controlling for other factors.
+This health penalty from having more children is strongest in low-education households and appears in both urban and rural areas.
+Such findings mirror the Philippine evidence that the harms of large family size are most pronounced among disadvantaged families.
+In economic terms, poorer parents with many children struggle to provide adequate nutrition and schooling to all, which highlight the equity dimension of the Q–Q trade-off.
+
+\subsection{Family Size and Child Outcomes in Southeast Asia}
+
+The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity–quality trade-off with diverse outcomes and methods.
+In Vietnam, for example, rapid fertility decline alongside rising education led to questions about a Q–Q mechanism.
+\citet{anh1998family} documented a negative correlation between family size and children’s school enrollment in Vietnam, though their analysis could not fully establish causality.
+
+More rigorously, \citet{dang2016decision} used distance to the nearest family planning center as an instrument to study Vietnamese households’ investments in education.
+They introduced a novel measure of child quality—spending on private tutoring, a prevalent form of educational investment in Vietnam—alongside traditional indicators like schooling expenses.
+The IV estimates confirmed that children with more siblings receive significantly lower educational investments from their families.
+In particular, Vietnamese families of larger size spent less on each school-age child’s schooling and tutoring, even after controlling for community factors.
+This effect was robust across different definitions of family size and model specifications, indicating that Vietnamese parents do indeed trade off quantity for quality when faced with resource constraints.
+Such evidence aligns squarely with Becker’s model: as Vietnamese family size increases, per-child education spending falls; parents are prioritizing “quality” less when they have more offspring.
+
+Indonesia and Thailand show similar patterns.
+Demographic research in Thailand during its fertility transition found that large families had markedly worse educational outcomes.
+\citet{knodel1990family}, studying Thai data in the 1990s, observed that once family size exceeded about 4–5 children, the likelihood of a child progressing to or staying in secondary school dropped precipitously compared to smaller families.
+Although these early Thai studies were based on correlations, they strongly suggested that limited family resources were being spread thin in big families and hurt children’s schooling attainment.
+
+In Indonesia, cohort analyses and natural experiments reinforce the trade-off.
+\citet{maralani2008changing} showed that the relationship between sibship size and schooling evolved from neutral or even positive for older cohorts (born when education opportunities were limited) to negative for more recent cohorts, consistent with a growing importance of education in a modernizing economy.
+More concretely, the aforementioned study by \citet{hatton2018fertility} in Indonesia provides causal evidence that mirrors the Philippine findings in health and nutrition.
+Likewise, an analysis of Indonesian census data \citep{feng2021effect} found that having additional siblings significantly lowers children’s educational attainment once birth order effects are accounted for, paralleling results from China and Vietnam.
+
+These regional studies share a commonality: in resource-constrained settings of Southeast Asia, increased child quantity tends to come at the expense of child quality, be it years of schooling, academic spending, or health status.
+The consistency of this pattern—across countries with different cultures and policies—highlights the fundamental economic logic identified by \citet{becker1973interaction}.
+Parents with finite resources face difficult choices, and many appear to balance quantity and quality in a way that confirms the trade-off hypothesis.
+
+Despite these similarities, there are some noteworthy nuances and gaps in the ASEAN literature.
+One is the role of public policy and development level.
+Evidence suggests that the Q–Q trade-off may be mitigated in contexts with strong public support for education and health.
+For instance, studies in developed countries (e.g.
+Israel, Norway) often find little or no trade-off once factors like birth order are accounted for \citep{black2005more, kristensen2010educational, angrist2010multiple}.
+In Southeast Asia, however, public education quality and social safety nets are still developing, and the cost of raising children (education fees, food, etc.) is largely borne by families themselves \citep{oecd2024sigi}.
+This may explain why the trade-off emerges so clearly in the Philippines, Vietnam, Indonesia, and Thailand.
+
+Another nuance is methodological: more recent studies employ credible identification strategies (IVs, twins, policy shocks) and consistently find a causal negative effect of family size on child outcomes, whereas older studies without such controls sometimes found weaker effects or none at all.
+This highlights the importance of accounting for endogeneity.
+
+Finally, there remain gaps for future research.
+Most ASEAN studies focus on education and early-life health indicators as measures of child quality; there is relatively little evidence on long-term outcomes such as children’s eventual earnings or income in adulthood.
+It is not yet fully clear whether the schooling and health disadvantages observed in larger families translate into significantly lower adult productivity or income—a link that \citet{becker1976child} theorized but which could be explored further in this region.
+
+Additionally, while the trade-off appears pervasive, its magnitude can vary: for example, the penalty of an extra child may be larger in poorer rural areas than in urban or wealthier settings, suggesting that local context (poverty, gender norms, access to services) can modulate the trade-off.
+Comparative studies across ASEAN are still somewhat limited, and a common challenge is disentangling related factors like birth order, sibling composition, and parental preferences.
+
+Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity–quality conjecture.
+As families have fewer children, they appear to invest more in each child’s education—investments crucial for human capital development and economic growth.
+Conversely, high-fertility households risk under-investing per child, which reinforces cycles of poverty and inequality.
+This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity–quality trade-off and help families achieve both manageable size and better outcomes for the next generation.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.tex
write q-q evidence in the philippines and southeast asia

@@
 schooling
-\citep{rosenzweig1980testing, black2005more, angrist2010effects}. These
+\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
 outcomes serve as accessible proxies for long-run human capital
@@
 
+\subsection{Philippine Evidence on the Q–Q Trade-off}
+
+Early empirical work in the Philippines provided suggestive support for
+this theory. For example, an influential study by
+\citet{horton1986child} used Philippine household data to jointly
+examine fertility and child nutrition and treated nutritional status as
+a measure of child quality. Horton found that better-educated mothers
+and fathers tended to choose smaller families and achieved
+better-nourished children--evidence of substitution away from
+``quantity'' toward child ``quality.'' Notably, she observed strong
+birth-order effects on nutrition (first-born children faring better than
+later-born), which hint that parents may not evenly distribute quality
+investments among all children. This early work indicated a
+quantity--quality (Q--Q) trade-off in Philippine families, though it
+largely documented correlations rather than definitive causation.
+
+Subsequent studies in the Philippines have tackled the critical issue of
+causality, using innovative research designs to isolate exogenous
+changes in family size. One seminal contribution is
+\citet{orbeta2010number}, who examined the impact of family size on
+children's schooling using a nationally representative survey.
+Recognizing that family size is endogenous (parents' fertility choices
+may reflect unobserved preferences or constraints), Orbeta employed an
+instrumental-variable (IV) approach grounded in Becker's framework.
+Specifically, he used the sex composition of the first two children as
+an instrument for having additional children -- an approach pioneered by
+\citet{angrist1996children} -- leveraging the fact that Filipino parents
+often desire a mixed-gender sibset (and are more likely to have a third
+child if the first two are the same sex) \citep{vicerra2013fertility}.
+This strategy aims to generate random-like variation in family size
+uncorrelated with parental characteristics.
+
+The IV estimates confirmed a significant negative causal effect of
+higher fertility on educational outcomes. \citet{orbeta2010number} found
+that each additional child in the household reduced the proportion of
+school-age children (6--24 years) attending school by roughly 19\% of
+the baseline attendance rate. The trade-off was especially pronounced at
+higher education levels: for example, the estimated drop in school
+attendance was about 26\% at the secondary level and 57\% at the
+tertiary level for each additional sibling. These are sizable effects,
+implying that children from large families are substantially less likely
+to remain in school, presumably due to tighter household budget
+constraints or diluted parental attention. Moreover, the burden of the
+trade-off appeared regressive: Orbeta's results showed much larger
+schooling deficits from an extra child in poorer households than in
+richer ones. For instance, in the poorest quintile, an additional
+sibling reduced school attendance by an estimated 24\% (for ages 6--24),
+compared to a 16\% reduction in the richest quintile. This regressive
+pattern aligns with Becker's theory that resource constraints bind more
+tightly for low-income families, which makes the Q--Q trade-off more
+acute. In summary, Orbeta's study -- the first in the Philippines to
+account for fertility endogeneity -- provides robust evidence that
+increases in family size cause significant declines in child educational
+attainment and validates the quantity--quality trade-off in this
+context. Large family size thus emerges as one mechanism contributing to
+poverty, by impeding children's human capital accumulation in the
+Philippines.
+
+Further compelling evidence comes from a natural experiment studied by
+\citet{dumas2019sex}. They exploit a unique policy shock in metropolitan
+Manila to identify the trade-off. In 1998, the Mayor of Manila city
+imposed a sudden ban on modern contraceptives in public facilities,
+drastically curtailing access to family planning for residents of Manila
+city (but not in surrounding municipalities). Dumas and Lefranc use this
+policy as a quasi-experiment: comparing families in Manila city (treated
+by the ban) to similar families in other cities unaffected by the ban
+before and after 1998. This difference-in-differences design, coupled
+with the fact that older mothers were naturally less fecund during the
+ban, isolates an exogenous fertility increase.
+
+The results are striking. The contraceptives ban led to a significant
+rise in births and family size in Manila city relative to the control
+areas. Correspondingly, children born in Manila during the ban era
+experienced a sizable decline in educational attainment -- clear
+evidence of a Q--Q trade-off precipitated by the shock. In the authors'
+words, the policy-driven increase in family size ``provide{[}s{]}
+evidence of a quality--quantity trade-off'': larger families, forced by
+the ban, resulted in lower schooling outcomes per child. By leveraging
+an actual anti-contraception policy, this study offers causal
+confirmation that constraints on family planning can worsen child
+quality. Together with Orbeta's findings, it reinforces the relevance of
+Becker's hypothesis in the Philippine setting: whether via unintended
+fertility (Manila's case) or ordinary variation in sibling sex mix, more
+children mean fewer resources per child and thus worse education
+outcomes. Consistent with theory, Filipino parents facing exogenous
+increases in quantity were unable to maintain the same level of quality
+per child.
+
+Beyond education, researchers have also examined health and other child
+outcomes in relation to family size. Evidence generally suggests the
+Q--Q trade-off extends to child health and nutrition. For instance,
+\citet{horton1986child} already hinted that large families may
+compromise child nutrition for later-born siblings. More recent regional
+research resonates with this. \citet{hatton2018fertility} analyze the
+effect of fertility on child height (a long-run health indicator) using
+longitudinal data from Indonesia -- a neighboring Southeast Asian
+country with comparable developmental challenges. They address
+endogeneity by exploiting Indonesia's family planning program rollout
+and exposure to mass media as instruments for fertility. The authors
+find a significant negative impact of family size on child health: each
+additional sibling is associated with about a one-third standard
+deviation reduction in a child's height-for-age Z-score, after
+controlling for other factors. This health penalty from having more
+children is strongest in low-education households and appears in both
+urban and rural areas. Such findings mirror the Philippine evidence that
+the harms of large family size are most pronounced among disadvantaged
+families. In economic terms, poorer parents with many children struggle
+to provide adequate nutrition and schooling to all, which highlight the
+equity dimension of the Q--Q trade-off.
+
+\subsection{Family Size and Child Outcomes in Southeast Asia}
+
+The Philippine experience is echoed in other ASEAN countries, where
+researchers have probed the quantity--quality trade-off with diverse
+outcomes and methods. In Vietnam, for example, rapid fertility decline
+alongside rising education led to questions about a Q--Q mechanism.
+\citet{anh1998family} documented a negative correlation between family
+size and children's school enrollment in Vietnam, though their analysis
+could not fully establish causality.
+
+More rigorously, \citet{dang2016decision} used distance to the nearest
+family planning center as an instrument to study Vietnamese households'
+investments in education. They introduced a novel measure of child
+quality---spending on private tutoring, a prevalent form of educational
+investment in Vietnam---alongside traditional indicators like schooling
+expenses. The IV estimates confirmed that children with more siblings
+receive significantly lower educational investments from their families.
+In particular, Vietnamese families of larger size spent less on each
+school-age child's schooling and tutoring, even after controlling for
+community factors. This effect was robust across different definitions
+of family size and model specifications, indicating that Vietnamese
+parents do indeed trade off quantity for quality when faced with
+resource constraints. Such evidence aligns squarely with Becker's model:
+as Vietnamese family size increases, per-child education spending falls;
+parents are prioritizing ``quality'' less when they have more offspring.
+
+Indonesia and Thailand show similar patterns. Demographic research in
+Thailand during its fertility transition found that large families had
+markedly worse educational outcomes. \citet{knodel1990family}, studying
+Thai data in the 1990s, observed that once family size exceeded about
+4--5 children, the likelihood of a child progressing to or staying in
+secondary school dropped precipitously compared to smaller families.
+Although these early Thai studies were based on correlations, they
+strongly suggested that limited family resources were being spread thin
+in big families and hurt children's schooling attainment.
+
+In Indonesia, cohort analyses and natural experiments reinforce the
+trade-off. \citet{maralani2008changing} showed that the relationship
+between sibship size and schooling evolved from neutral or even positive
+for older cohorts (born when education opportunities were limited) to
+negative for more recent cohorts, consistent with a growing importance
+of education in a modernizing economy. More concretely, the
+aforementioned study by \citet{hatton2018fertility} in Indonesia
+provides causal evidence that mirrors the Philippine findings in health
+and nutrition. Likewise, an analysis of Indonesian census data
+\citep{feng2021effect} found that having additional siblings
+significantly lowers children's educational attainment once birth order
+effects are accounted for, paralleling results from China and Vietnam.
+
+These regional studies share a commonality: in resource-constrained
+settings of Southeast Asia, increased child quantity tends to come at
+the expense of child quality, be it years of schooling, academic
+spending, or health status. The consistency of this pattern---across
+countries with different cultures and policies---highlights the
+fundamental economic logic identified by \citet{becker1973interaction}.
+Parents with finite resources face difficult choices, and many appear to
+balance quantity and quality in a way that confirms the trade-off
+hypothesis.
+
+Despite these similarities, there are some noteworthy nuances and gaps
+in the ASEAN literature. One is the role of public policy and
+development level. Evidence suggests that the Q--Q trade-off may be
+mitigated in contexts with strong public support for education and
+health. For instance, studies in developed countries (e.g.~Israel,
+Norway) often find little or no trade-off once factors like birth order
+are accounted for
+\citep{black2005more, kristensen2010educational, angrist2010multiple}.
+In Southeast Asia, however, public education quality and social safety
+nets are still developing, and the cost of raising children (education
+fees, food, etc.) is largely borne by families themselves
+\citep{oecd2024sigi}. This may explain why the trade-off emerges so
+clearly in the Philippines, Vietnam, Indonesia, and Thailand.
+
+Another nuance is methodological: more recent studies employ credible
+identification strategies (IVs, twins, policy shocks) and consistently
+find a causal negative effect of family size on child outcomes, whereas
+older studies without such controls sometimes found weaker effects or
+none at all. This highlights the importance of accounting for
+endogeneity.
+
+Finally, there remain gaps for future research. Most ASEAN studies focus
+on education and early-life health indicators as measures of child
+quality; there is relatively little evidence on long-term outcomes such
+as children's eventual earnings or income in adulthood. It is not yet
+fully clear whether the schooling and health disadvantages observed in
+larger families translate into significantly lower adult productivity or
+income---a link that \citet{becker1976child} theorized but which could
+be explored further in this region.
+
+Additionally, while the trade-off appears pervasive, its magnitude can
+vary: for example, the penalty of an extra child may be larger in poorer
+rural areas than in urban or wealthier settings, suggesting that local
+context (poverty, gender norms, access to services) can modulate the
+trade-off. Comparative studies across ASEAN are still somewhat limited,
+and a common challenge is disentangling related factors like birth
+order, sibling composition, and parental preferences.
+
+Nonetheless, the prevailing evidence from the Philippines and its
+regional neighbors strongly supports Becker's quantity--quality
+conjecture. As families have fewer children, they appear to invest more
+in each child's education---investments crucial for human capital
+development and economic growth. Conversely, high-fertility households
+risk under-investing per child, which reinforces cycles of poverty and
+inequality. This literature thus provides an important empirical
+foundation for policies in the Philippines and ASEAN---from family
+planning programs to education subsidies---that aim to ease the
+quantity--quality trade-off and help families achieve both manageable
+size and better outcomes for the next generation.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.Rmd
write conceptual framework & hypothesis

@@
 
+\section{Conceptual Framework and Hypotheses}
+
+This study builds upon the foundational framework of the quantity–quality (Q–Q) trade-off, as articulated in \citet{becker1960economic} seminal work and later refined by \citet{becker1973interaction}.
+Within this theoretical model, fertility and child quality are jointly determined under a full-income constraint, such that any increase in the number of children—absent a commensurate expansion in resources—necessitates a reduction in the average investment per child.
+While prior empirical applications of this framework have largely centered on educational attainment, child quality may also be understood along biological dimensions, especially along nutritional status, which emerges earlier in the life course and may be less subject to later remediation.
+
+The current analysis focuses exclusively on early-life nutrition as a proxy for child quality.
+Nutrition, unlike cognitive or educational outcomes, is characterized by low substitutability and a narrow temporal window during which investments are most productive.
+The trade-off is thus expected to manifest strongly in settings where fertility rises exogenously while household resources remain constrained.
+In such environments, families facing an unexpected increase in sibship size may be forced to reallocate scarce food, time, and health inputs across a larger number of dependents, potentially resulting in lower nutritional attainment for each child.
+
+To formalize this relationship, consider a household with total income \( I \), choosing nutritional investments \(\kappa_j\) for each of \( n \) children, along with consumption \( Z \) unrelated to child quality:
+
+\[
+I = \sum_{j=1}^{n} \kappa_j + Z.
+\]
+
+In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers.
+If nutritional investments per child decline, and if such inputs are critical for growth, observable deficits in height-for-age (HAZ) or weight-for-height (WHZ) may emerge.
+The first hypothesis emerging from this framework is that an exogenous increase in
+
+fertility leads to a reduction in per-child nutritional investment, particularly in low-income, urban households where slack in the family budget is limited.
+A second, related hypothesis posits that this reduction in inputs translates into measurable nutritional deficits among young children, specifically in the form of lower anthropometric outcomes such as height-for-age (HAZ) and weight-for-height (WHZ) z-scores.
+These two hypotheses—concerning, respectively, the behavioral reallocation of resources and its biological consequence—jointly seek to identify a causal link between fertility shocks and early-life deprivation.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.tex
write conceptual framework & hypothesis

@@
 
+\section{Conceptual Framework and Hypotheses}
+
+This study builds upon the foundational framework of the
+quantity--quality (Q--Q) trade-off, as articulated in
+\citet{becker1960economic} seminal work and later refined by
+\citet{becker1973interaction}. Within this theoretical model, fertility
+and child quality are jointly determined under a full-income constraint,
+such that any increase in the number of children---absent a commensurate
+expansion in resources---necessitates a reduction in the average
+investment per child. While prior empirical applications of this
+framework have largely centered on educational attainment, child quality
+may also be understood along biological dimensions, especially along
+nutritional status, which emerges earlier in the life course and may be
+less subject to later remediation.
+
+The current analysis focuses exclusively on early-life nutrition as a
+proxy for child quality. Nutrition, unlike cognitive or educational
+outcomes, is characterized by low substitutability and a narrow temporal
+window during which investments are most productive. The trade-off is
+thus expected to manifest strongly in settings where fertility rises
+exogenously while household resources remain constrained. In such
+environments, families facing an unexpected increase in sibship size may
+be forced to reallocate scarce food, time, and health inputs across a
+larger number of dependents, potentially resulting in lower nutritional
+attainment for each child.
+
+To formalize this relationship, consider a household with total income
+\(I\), choosing nutritional investments \(\kappa_j\) for each of \(n\)
+children, along with consumption \(Z\) unrelated to child quality:
+
+\[
+I = \sum_{j=1}^{n} \kappa_j + Z.
+\]
+
+In the presence of a fertility shock, holding \(I\) constant, the
+average \(\kappa_j\) must fall unless offset by changes in parental
+behavior or external transfers. If nutritional investments per child
+decline, and if such inputs are critical for growth, observable deficits
+in height-for-age (HAZ) or weight-for-height (WHZ) may emerge. The first
+hypothesis emerging from this framework is that an exogenous increase in
+
+fertility leads to a reduction in per-child nutritional investment,
+particularly in low-income, urban households where slack in the family
+budget is limited. A second, related hypothesis posits that this
+reduction in inputs translates into measurable nutritional deficits
+among young children, specifically in the form of lower anthropometric
+outcomes such as height-for-age (HAZ) and weight-for-height (WHZ)
+z-scores. These two hypotheses---concerning, respectively, the
+behavioral reallocation of resources and its biological
+consequence---jointly seek to identify a causal link between fertility
+shocks and early-life deprivation.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.Rmd
write data & empirical strategy

@@
 
+\section{Data and Empirical Strategy}
+
+The analysis draws on multiple nationally representative data sources, with the Philippine Censuses serving as the primary dataset.
+Microdata from the 1990, 1995, 2000, 2007, and 2010 waves are used to construct baseline comparisons and perform robustness checks.
+The 1995 wave is treated as the pre-policy reference period, while 2007 serves as the post-policy period.
+Due to the possibility that reductions in contraceptive access began prior to the formal issuance of Executive Order No.
+003 in 2000, the 2000 census wave is excluded from the baseline specification, though it is incorporated in secondary analyses.
+Placebo estimations using the 1990 census are also implemented to validate the identifying assumptions.
+
+While the census provides rich information on household composition and allows for city-level fixed effects with considerable statistical power, it does not include direct measures of child health or anthropometry.
+To address this limitation, two additional datasets are employed.
+The Philippine Demographic and Health Surveys (DHS), pooled across the 1993, 1998, 2003, and 2008 rounds, offer standardized anthropometric data for children under five, alongside detailed maternal characteristics and fertility histories.
+The DHS enables construction of child-level outcomes—height-for-age z-scores (HAZ) and weight-for-height z-scores (WHZ)—and supports temporal comparisons across the pre- and post-ban periods.
+These outcomes serve as internationally recognized indicators of chronic and acute malnutrition, respectively.
+
+To complement these sources, the Annual Poverty Indicators Survey (APIS) is used to assess household-level economic conditions, particularly food expenditure, living standards, and maternal education.
+Although smaller in scale, the APIS provides contextual variables necessary for exploring heterogeneity in treatment effects across wealth strata or educational backgrounds.
+
+The empirical strategy adopts a difference-in-differences (DiD) framework.
+It utilizes the 2000 contraceptive policy in Manila as a natural experiment.
+The policy—enacted through Executive Order No.
+003—effectively banned modern contraceptive distribution in public health facilities across Manila City, but left surrounding municipalities in the National Capital Region (NCR) unaffected.
+This jurisdictional variation enables identification of the causal effect of fertility increases on child nutrition, under the assumption that, in the absence of the policy, nutritional trends in Manila would have followed a parallel trajectory to those in comparison cities.
+
+The main estimating equation is specified as follows:
+
+\[
+\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
+\]
+
+where \(\text{Nutrition}_{ict}\) denotes the anthropometric outcome (HAZ or WHZ) for child \( i \) in city \( c \) at time \( t \); \(\text{Post}_t\) is an indicator for the post-policy period; \(\text{Manila}_c\) is a treatment group dummy; \(X_{ict}\) is a vector of controls including child age, sex, birth order, maternal age and education, household wealth quintile, and urban barangay indicators; \(\lambda_c\) and \(\tau_t\) are city and year fixed effects, respectively.
+The coefficient \(\beta\) estimates the average treatment effect of the fertility shock on nutritional outcomes.
+
+To strengthen identification, additional specifications are estimated.
+City-specific linear trends are introduced to account for differential temporal patterns in nutrition outcomes.
+A triple-difference extension interacts the DiD term with maternal age cohort indicators, under the assumption that older women—being less fecund—were less likely to experience changes in fertility behavior.
+This approach allows for control of unobserved city-level shocks that may disproportionately affect certain demographic groups.
+Sensitivity analyses also test alternative control groups: first, using only Quezon City, which is geographically contiguous and socio-economically comparable to Manila; and second, using all other NCR cities excluding Quezon.
+
+Standard errors are clustered at the city-year level to allow for serial correlation and heteroskedasticity.
+Placebo tests using the 1990–1995 period confirm that pre-policy trends in nutritional outcomes were parallel across treatment and control groups.
+Additional robustness checks include falsification tests using children born outside of the city jurisdiction and exclusion of households with missing maternal linkage or implausible anthropometric values.
+
+Although the analysis is subject to the usual limitations of observational designs—including residual confounding and measurement error—the combined use of census, DHS, and APIS data enables a multi-faceted examination of the Q–Q trade-off in nutrition.
+The strategy mirrors that of \citet{dumas2019sex}, who investigated educational outcomes using the same policy shock, but extends their framework to explore a dimension of child quality that is both biologically foundational and policy-relevant in the context of urban poverty.
+
 \bibliographystyle{aea}

### 2025-07-03 docs/paper.tex
write data & empirical strategy

@@
 
+\section{Data and Empirical Strategy}
+
+The analysis draws on multiple nationally representative data sources,
+with the Philippine Censuses serving as the primary dataset. Microdata
+from the 1990, 1995, 2000, 2007, and 2010 waves are used to construct
+baseline comparisons and perform robustness checks. The 1995 wave is
+treated as the pre-policy reference period, while 2007 serves as the
+post-policy period. Due to the possibility that reductions in
+contraceptive access began prior to the formal issuance of Executive
+Order No.~003 in 2000, the 2000 census wave is excluded from the
+baseline specification, though it is incorporated in secondary analyses.
+Placebo estimations using the 1990 census are also implemented to
+validate the identifying assumptions.
+
+While the census provides rich information on household composition and
+allows for city-level fixed effects with considerable statistical power,
+it does not include direct measures of child health or anthropometry. To
+address this limitation, two additional datasets are employed. The
+Philippine Demographic and Health Surveys (DHS), pooled across the 1993,
+1998, 2003, and 2008 rounds, offer standardized anthropometric data for
+children under five, alongside detailed maternal characteristics and
+fertility histories. The DHS enables construction of child-level
+outcomes---height-for-age z-scores (HAZ) and weight-for-height z-scores
+(WHZ)---and supports temporal comparisons across the pre- and post-ban
+periods. These outcomes serve as internationally recognized indicators
+of chronic and acute malnutrition, respectively.
+
+To complement these sources, the Annual Poverty Indicators Survey (APIS)
+is used to assess household-level economic conditions, particularly food
+expenditure, living standards, and maternal education. Although smaller
+in scale, the APIS provides contextual variables necessary for exploring
+heterogeneity in treatment effects across wealth strata or educational
+backgrounds.
+
+The empirical strategy adopts a difference-in-differences (DiD)
+framework. It utilizes the 2000 contraceptive policy in Manila as a
+natural experiment. The policy---enacted through Executive Order
+No.~003---effectively banned modern contraceptive distribution in public
+health facilities across Manila City, but left surrounding
+municipalities in the National Capital Region (NCR) unaffected. This
+jurisdictional variation enables identification of the causal effect of
+fertility increases on child nutrition, under the assumption that, in
+the absence of the policy, nutritional trends in Manila would have
+followed a parallel trajectory to those in comparison cities.
+
+The main estimating equation is specified as follows:
+
+\[
+\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
+\]
+
+where \(\text{Nutrition}_{ict}\) denotes the anthropometric outcome (HAZ
+or WHZ) for child \(i\) in city \(c\) at time \(t\); \(\text{Post}_t\)
+is an indicator for the post-policy period; \(\text{Manila}_c\) is a
+treatment group dummy; \(X_{ict}\) is a vector of controls including
+child age, sex, birth order, maternal age and education, household
+wealth quintile, and urban barangay indicators; \(\lambda_c\) and
+\(\tau_t\) are city and year fixed effects, respectively. The
+coefficient \(\beta\) estimates the average treatment effect of the
+fertility shock on nutritional outcomes.
+
+To strengthen identification, additional specifications are estimated.
+City-specific linear trends are introduced to account for differential
+temporal patterns in nutrition outcomes. A triple-difference extension
+interacts the DiD term with maternal age cohort indicators, under the
+assumption that older women---being less fecund---were less likely to
+experience changes in fertility behavior. This approach allows for
+control of unobserved city-level shocks that may disproportionately
+affect certain demographic groups. Sensitivity analyses also test
+alternative control groups: first, using only Quezon City, which is
+geographically contiguous and socio-economically comparable to Manila;
+and second, using all other NCR cities excluding Quezon.
+
+Standard errors are clustered at the city-year level to allow for serial
+correlation and heteroskedasticity. Placebo tests using the 1990--1995
+period confirm that pre-policy trends in nutritional outcomes were
+parallel across treatment and control groups. Additional robustness
+checks include falsification tests using children born outside of the
+city jurisdiction and exclusion of households with missing maternal
+linkage or implausible anthropometric values.
+
+Although the analysis is subject to the usual limitations of
+observational designs---including residual confounding and measurement
+error---the combined use of census, DHS, and APIS data enables a
+multi-faceted examination of the Q--Q trade-off in nutrition. The
+strategy mirrors that of \citet{dumas2019sex}, who investigated
+educational outcomes using the same policy shock, but extends their
+framework to explore a dimension of child quality that is both
+biologically foundational and policy-relevant in the context of urban
+poverty.
+
 \bibliographystyle{aea}

### 2025-07-06 docs/paper.Rmd
revise data and empirical strategy section
Updated the Data and Empirical Strategy section to clarify the primary use of DHS data, detail the empirical design, and specify the role of census and APIS data. The new text provides a more precise description of the identification strategy, sample construction, and robustness checks, and removes a prior citation. PDF and TeX files regenerated accordingly.

--- 2025-07-03 docs/paper.Rmd
@@
 
+\section{Data and Empirical Strategy}
+
+The primary dataset used in this analysis is the Philippine Demographic and Health Surveys (DHS), specifically the 1993, 1998, 2003, 2008, and 2013 waves.
+These surveys are nationally representative and contain detailed, child-level anthropometric indicators, such as height-for-age z-scores (HAZ) and weight-for-height z-scores (WHZ).
+These metrics are internationally recognized as proxies for chronic and acute malnutrition, respectively.
+The DHS also provides rich information on maternal characteristics, household wealth, fertility histories, and geographic location, which enabled a quasi-experimental research design centered on the policy-induced fertility shock in Manila City.
+
+The empirical strategy exploits the 2000 issuance of Executive Order No.
+003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City.
+This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR).
+Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group.
+The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.
+
+The identification approach employs a difference-in-differences (DiD) framework. The main specification is:
+
+\[
+\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
+\]
+
+where \(\text{Nutrition}_{ict}\) denotes the HAZ or WHZ score for child \(i\) in city \(c\) born in year \(t\); \(\text{Post}_t\) is an indicator for children born after 2000; \(\text{Manila}_c\) is a treatment group dummy; and \(X_{ict}\) includes controls for child age in months, sex, birth order, maternal age and education, household wealth quintile, and urban barangay indicators.
+City and year fixed effects \((\lambda_c, \tau_t)\) are included to account for time-invariant differences and common macroeconomic shocks.
+Standard errors are clustered at the city-year level to address potential autocorrelation and heteroskedasticity.
+
+The 1998 DHS wave is treated as the pre-policy benchmark, while 2003 serves as the primary post-policy round for estimating short-run effects.
+The 2008 survey is incorporated into robustness checks to evaluate medium-run persistence of any observed nutritional gaps.
+The 1993 DHS is used in placebo tests to evaluate pre-policy parallel trends between Manila and control cities.
+Although included cautiously due to possible confounding from attrition and mobility, the 2013 DHS is optionally used to examine longer-term cohort outcomes and nutritional trajectories.
+
+To validate the existence and magnitude of the fertility shock itself, and to assess potential violations of the identifying assumptions, Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 are used.
+The 1995 wave is treated as the pre-policy reference period for fertility, household size, and demographic composition, while 2007 captures the post-policy landscape.
+The 2000 census wave is excluded from baseline comparisons due to uncertainty about when contraceptive access began declining, though it is retained for robustness checks involving changes in fertility behavior.
+The 2010 wave is used to examine longer-run demographic patterns.
+A placebo comparison between 1990 and 1995 confirms that fertility rates in Manila and comparison cities followed parallel trajectories prior to the policy change.
+
+To supplement both the DHS and Census, the Annual Poverty Indicators Survey (APIS) is used to provide household-level context on food expenditure, maternal education, and living standards.
+While more limited in sample size and coverage, the APIS enables exploratory analyses of heterogeneity in treatment effects by wealth and education, and supports the triangulation of maternal resource constraints during the post-policy period.
+
 \bibliographystyle{aea}

### 2025-07-06 docs/paper.tex
revise data and empirical strategy section
Updated the Data and Empirical Strategy section to clarify the primary use of DHS data, detail the empirical design, and specify the role of census and APIS data. The new text provides a more precise description of the identification strategy, sample construction, and robustness checks, and removes a prior citation. PDF and TeX files regenerated accordingly.

--- 2025-07-03 docs/paper.tex
@@
 
+\section{Data and Empirical Strategy}
+
+The primary dataset used in this analysis is the Philippine Demographic
+and Health Surveys (DHS), specifically the 1993, 1998, 2003, 2008, and
+2013 waves. These surveys are nationally representative and contain
+detailed, child-level anthropometric indicators, such as height-for-age
+z-scores (HAZ) and weight-for-height z-scores (WHZ). These metrics are
+internationally recognized as proxies for chronic and acute
+malnutrition, respectively. The DHS also provides rich information on
+maternal characteristics, household wealth, fertility histories, and
+geographic location, which enabled a quasi-experimental research design
+centered on the policy-induced fertility shock in Manila City.
+
+The empirical strategy exploits the 2000 issuance of Executive Order
+No.~003, which effectively banned the distribution of modern
+contraceptives in public health centers across Manila City. This abrupt
+and localized withdrawal of contraceptive access resulted in a fertility
+increase concentrated within the city, but did not affect surrounding
+municipalities within the National Capital Region (NCR). Children born
+in Manila following the policy are considered exposed to a fertility
+shock, while children in other NCR cities over the same period serve as
+the control group. The analytical sample is restricted to children under
+five years old at the time of each survey round to ensure accurate
+measurement of early-life nutritional outcomes.
+
+The identification approach employs a difference-in-differences (DiD)
+framework. The main specification is:
+
+\[
+\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
+\]
+
+where \(\text{Nutrition}_{ict}\) denotes the HAZ or WHZ score for child
+\(i\) in city \(c\) born in year \(t\); \(\text{Post}_t\) is an
+indicator for children born after 2000; \(\text{Manila}_c\) is a
+treatment group dummy; and \(X_{ict}\) includes controls for child age
+in months, sex, birth order, maternal age and education, household
+wealth quintile, and urban barangay indicators. City and year fixed
+effects \((\lambda_c, \tau_t)\) are included to account for
+time-invariant differences and common macroeconomic shocks. Standard
+errors are clustered at the city-year level to address potential
+autocorrelation and heteroskedasticity.
+
+The 1998 DHS wave is treated as the pre-policy benchmark, while 2003
+serves as the primary post-policy round for estimating short-run
+effects. The 2008 survey is incorporated into robustness checks to
+evaluate medium-run persistence of any observed nutritional gaps. The
+1993 DHS is used in placebo tests to evaluate pre-policy parallel trends
+between Manila and control cities. Although included cautiously due to
+possible confounding from attrition and mobility, the 2013 DHS is
+optionally used to examine longer-term cohort outcomes and nutritional
+trajectories.
+
+To validate the existence and magnitude of the fertility shock itself,
+and to assess potential violations of the identifying assumptions,
+Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 are
+used. The 1995 wave is treated as the pre-policy reference period for
+fertility, household size, and demographic composition, while 2007
+captures the post-policy landscape. The 2000 census wave is excluded
+from baseline comparisons due to uncertainty about when contraceptive
+access began declining, though it is retained for robustness checks
+involving changes in fertility behavior. The 2010 wave is used to
+examine longer-run demographic patterns. A placebo comparison between
+1990 and 1995 confirms that fertility rates in Manila and comparison
+cities followed parallel trajectories prior to the policy change.
+
+To supplement both the DHS and Census, the Annual Poverty Indicators
+Survey (APIS) is used to provide household-level context on food
+expenditure, maternal education, and living standards. While more
+limited in sample size and coverage, the APIS enables exploratory
+analyses of heterogeneity in treatment effects by wealth and education,
+and supports the triangulation of maternal resource constraints during
+the post-policy period.
+
 \bibliographystyle{aea}

### 2025-07-07 data-analysis/scripts/01_dhs_clean.R
add DHS data cleaning and merging script
Introduces 01_dhs_clean.R to load, harmonize, and merge KR, IR, and HR DHS files for the Philippines (1993–2013). The script standardizes variables, merges datasets, creates derived variables, and saves the cleaned data for further analysis.

@@ -1,0 +1,98 @@
+# ==============================================================================
+# 01_dhs_clean.R -
+# Purpose : Load every KR, IR, HR DHS file for PH (1993–2013), harmonise, merge,
+#           flag treatment/control, and save to clean-data/.
+# Author  : Erika Salvador '28 (esalvador28@amherst.edu) 
+# Project : Manila Contraceptive Ban & the Quantity-Quality Trade-off (nutrition arm)
+# ==============================================================================
+
+# Load required libraries
+library(haven)        # = for reading .dta files (Stata format)
+library(dplyr)        # = for data manipulation (filter, mutate, join, etc.)
+library(labelled)     # = for handling variable labels from Stata
+library(janitor)      # = for cleaning column names (snake_case)
+library(stringr)      # = for string operations
+library(purrr)        # = for functional programming (map functions)
+library(glue)         # = for formatted strings
+library(readr)        # = for writing output files
+library(crayon)       # = for colorful messages
+library(cli)          # = for pretty status symbols
+
+
+# Step 1: Set up paths and years =============================================
+
+rounds <- c(1993, 1998, 2003, 2008, 2013)                # = list of survey years
+file_dir <- "../raw-data/dhs_raw"                        # = folder with raw files
+
+suffixes <- c(
+  kr = "kr_children",                                    # = Kids Recode files
+  ir = "ir_individual",                                  # = Individual Recode files
+  hr = "hr_household"                                    # = Household Recode files
+)
+
+# Function to read and standardize DHS files
+read_dhs <- function(year, suf) {
+  file.path(file_dir, glue("dhs_{year}_{suf}.DTA")) |>  # = build file path
+    read_stata() |>                                     # = read Stata file
+    clean_names() |>                                    # = clean column names
+    mutate(survey_year = year)                          # = tag with year
+}
+
+# Step 2: Load and select variables from each file type ======================
+
+# Children (KR)
+kr_vars <- c("survey_year", "v001", "v002", "bidx", "hw70", "hw71",
+             "b4", "b5", "b9", "bord", "caseid")
+
+kr_all <- map_dfr(rounds, read_dhs, suf = suffixes["kr"]) |>
+  select(any_of(kr_vars))
+
+# Women (IR)
+ir_vars <- c("survey_year", "caseid", "v001", "v002", "v003",    # = IDs
[…]

### 2025-07-07 data-analysis/scripts/00_vars.R
add script to generate dhs variable catalogs
Introduces 00_vars.R to create variable catalogs for PH DHS rounds (1993–2013) across Kids-Recode, Individual-Recode, and Household-Recode files. The script reads Stata files, extracts variable metadata, and outputs CSV catalogs for each file type.

@@ -1,0 +1,66 @@
+# ==============================================================================
+# 00_vars.R - 
+# Purpose  = Create variable catalogs for PH DHS rounds
+#            (1993, 1998, 2003, 2008, 2013) – separately for
+#            Kids-Recode (KR), Individual-Recode (IR), and
+#            Household-Recode (HR) files.
+# Author   = Erika Salvador ’28  <esalvador28@amherst.edu>
+# Project  = Manila Contraceptive Ban & the Q–Q Trade-off
+# ==============================================================================
+
+# Load required libraries 
+library(haven)      # read Stata .dta
+library(dplyr)      # filter / mutate / summarise
+library(purrr)      # map functions
+library(stringr)    # string helpers
+library(janitor)    # clean_names
+library(tidyr)      # unnest / pivot_longer
+library(glue)       # formatted paths
+library(readr)      # write_csv
+
+# 1. Set up survey rounds & file types =========================================
+rounds   = c(1993, 1998, 2003, 2008, 2013)
+file_dir = "../raw-data/dhs_raw"         
+
+suffixes = c(
+  kr = "kr_children",     # Kids Recode (child level)
+  ir = "ir_individual",   # Individual Recode (woman level)
+  hr = "hr_household"     # Household Recode (household level)
+)
+
+# 2. Define helper function to read data and pull variable meta ================
+
+grab_vars = function(year, suf) {
+  
+  # Build full file path (`glue` fills in {year} & {suf})
+  f_path = glue("{file_dir}/dhs_{year}_{suf}.DTA")
+  
+  # Haven returns a data-frame-like object with labelled columns
+  dat    = read_dta(f_path)
+  
+  tibble(
+    survey_year = year,
+    file_type   = names(suffixes)[suffixes == suf],     # kr / ir / hr
+    variable    = names(dat),
+    type        = map_chr(dat, ~ class(.x)[1])          # extract first class
+  )
+}
+
+# 3. Build combined meta table =================================================
+catalog_all =
+  expand_grid(year = rounds, suf = suffixes) |>         # all combos
+  mutate(meta = map2(year, suf, grab_vars)) |>          # read each file
+  unnest(meta) |>                                       # drop the list column
+  select(survey_year, file_type, variable, type) |>     # tidy order
+  arrange(file_type, survey_year, variable)
+
+# 4. Split and save to CSV =====================================================
+output_dir = "../outputs"      
[…]

### 2025-07-07 docs/paper.Rmd
revise data and empirical strategy section
The revision details dataset construction, outcome definitions, covariates, treatment assignment, empirical approach, and robustness checks. Also updated the difference-in-differences specification and missing data handling. Removed outdated content and improved methodological transparency.

--- 2025-07-03 docs/paper.Rmd
@@
 
+\section{Data and Empirical Strategy}
+
+This study constructs a harmonized, child-level analytic dataset using five rounds of the Philippine Demographic and Health Survey (DHS): 1993, 1998, 2003, 2008, and 2013.
+The DHS is a nationally representative, cross-sectional survey series conducted by the Philippine Statistics Authority in collaboration with ICF International.
+Each round follows a stratified two-stage cluster design and collects standardized information on child health, fertility, maternal characteristics, and household infrastructure.
+For each wave, I merge the child-level (KR), mother-level (IR), and household-level (HR) files using cluster, household, and maternal identifiers, ensuring that each observation corresponds to a live birth occurring within five years of the survey.
+The analytic sample is restricted to children aged 0–59 months to standardize developmental exposure windows and conform to the World Health Organization (WHO) reference growth standards.
+
+The primary outcomes are height-for-age (HAZ) and weight-for-height (WHZ) z-scores, internationally recognized proxies for chronic and acute malnutrition, respectively.
+These anthropometric indicators are cleaned following DHS protocols: biologically implausible values are excluded, and remaining z-scores are rescaled from hundredths to continuous units.
+In addition to anthropometry, the child record includes a suite of health-related variables.
+These include indicators for whether the child has received basic vaccinations—such as BCG, DPT, polio, and measles—as well as whether the child experienced episodes of diarrhea, fever, or respiratory illness in the two weeks prior to the survey.
+These supplementary health measures serve as intermediate outcomes in secondary analyses to capture short-run morbidity and preventive health investment.
+
+Maternal-level information is drawn from the individual recode file.
+Core covariates include maternal age, completed years of education, and total number of children ever born (parity), each of which captures socioeconomic status and fertility history.
+I additionally include maternal body mass index (BMI) and anemia status, which serve as both controls and potential mediators of child health.
+These indicators reflect the mother’s nutritional status and physiological capacity to support healthy fetal and postnatal development.
+In robustness checks, I examine whether declines in maternal nutritional status following the fertility shock account for observed changes in child anthropometry.
+
+Household-level characteristics are used to control for environmental and infrastructural determinants of health.
+Specifically, I include household size, urban residence, wealth quintile, access to electricity, type of toilet facility, whether the toilet is shared, and source of drinking water.
+These variables proxy for household-level resources and exposure to disease environments.
+In extension analyses, I interact treatment status with these environmental indicators to assess whether the nutritional consequences of the fertility shock are magnified under infrastructural deprivation.
+
+The treatment is defined by exposure to Executive Order No.
+003, which was issued by the Manila City government in January 2000 and effectively suspended the public distribution of modern contraceptives.
+This policy change resulted in a localized increase in fertility that was not mirrored in surrounding municipalities within the National Capital Region (NCR).
+Children born in Manila City after the implementation of the policy are classified as treated, while children in other NCR cities serve as the control group.
+This spatially and temporally bounded intervention provides quasi-experimental variation that facilitates identification of causal effects.
+
+The empirical approach adopts a difference-in-differences framework that compares trends in child nutritional outcomes between treated and control cities before and after the policy change.
+The identifying assumption is that, absent the policy, nutritional trajectories in Manila and non-Manila NCR cities would have evolved in parallel.
+This assumption is supported by visual inspection and pre-trend placebo tests using the 1993 and 1998 waves.
+The 1998 DHS serves as the primary pre-policy baseline, while the 2003 survey captures short-run effects.
+The 2008 and 2013 waves are used in robustness checks to assess medium- and long-run persistence of nutritional deficits.
+All specifications include city and birth-year fixed effects, and standard errors are clustered at the city-year level to account for within-cluster correlation.
+
+The empirical strategy exploits the 2000 issuance of Executive Order No.
+003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City.
+This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR).
+Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group.
+The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.
+
+I estimate the following difference-in-differences specification:
+
+\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict}
+
+In this equation, $\text{Nutrition}_{ict}$ refers to either height-for-age (HAZ) or weight-for-height (WHZ) for child $i$ in city $c$ born in year $t$.
+The indicator $\text{Post}_t$ equals 1 if the child was born after the implementation of Executive Order No.
+003 in 2000, and $\text{Manila}c$ equals 1 if the child resides in Manila City.
+The interaction term captures the effect of being born in Manila after the contraceptive policy was enacted.
+The vector $X{ict}$ includes controls for child age in months, sex, birth order, maternal age, education, parity, body mass index (BMI), anemia status, household wealth quintile, household size, electricity access, water source, toilet type, shared sanitation status, and urban residence.
+City fixed effects $\lambda_c$ control for time-invariant characteristics at the locality level, while year fixed effects $\tau_t$ absorb aggregate time shocks.
+The coefficient of interest, $\beta$, represents the average treatment effect of the policy-induced fertility shock on child nutrition.
+Standard errors are clustered at the city-year level to account for potential autocorrelation within geographic units over time.
+
+To support the internal validity of the design, I use Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude and timing of the fertility shock.
+The 1995 and 2007 waves provide clean pre- and post-policy benchmarks, while the 2000 census—although temporally proximate to the policy—is retained for robustness checks that test the sensitivity of estimates to the exact onset of fertility changes.
+A placebo comparison between 1990 and 1995 shows no divergence in fertility trends between Manila and control cities prior to the policy, strengthening the credibility of the parallel trends assumption.
+
+To supplement the DHS and Census data, I incorporate the Annual Poverty Indicators Survey (APIS), a nationally representative household survey that captures detailed information on food expenditure, maternal education, and living standards.
+Although the APIS lacks anthropometric data and has more limited geographic coverage, it enables exploratory analyses of treatment effect heterogeneity by income and education strata and helps triangulate changes in maternal resource constraints following the policy shock.
+
+Finally, missing data are handled following standard practices in applied microeconometrics.
+Observations with missing or flagged anthropometric outcomes are excluded from the analysis.
+For other covariates with partial nonresponse—such as maternal BMI, anemia, or household infrastructure—I adopt the missing-indicator method: missing values are imputed with a constant, and a binary indicator is included in all regression models to control for the fact of missingness.
+This approach preserves sample size while allowing the data to inform whether nonresponse is systematically related to the outcomes of interest.
+As robustness checks, I estimate alternative specifications using multiple imputation and inverse-probability weighting to test the sensitivity of results to different missing data assumptions.
+All descriptive and inferential statistics are computed using DHS-provided sampling weights, and models account for the complex survey design by clustering standard errors at the primary sampling unit.
+
 \bibliographystyle{aea}

### 2025-07-07 docs/paper.tex
revise data and empirical strategy section
The revision details dataset construction, outcome definitions, covariates, treatment assignment, empirical approach, and robustness checks. Also updated the difference-in-differences specification and missing data handling. Removed outdated content and improved methodological transparency.

--- 2025-07-03 docs/paper.tex
@@
 
+\section{Data and Empirical Strategy}
+
+This study constructs a harmonized, child-level analytic dataset using
+five rounds of the Philippine Demographic and Health Survey (DHS): 1993,
+1998, 2003, 2008, and 2013. The DHS is a nationally representative,
+cross-sectional survey series conducted by the Philippine Statistics
+Authority in collaboration with ICF International. Each round follows a
+stratified two-stage cluster design and collects standardized
+information on child health, fertility, maternal characteristics, and
+household infrastructure. For each wave, I merge the child-level (KR),
+mother-level (IR), and household-level (HR) files using cluster,
+household, and maternal identifiers, ensuring that each observation
+corresponds to a live birth occurring within five years of the survey.
+The analytic sample is restricted to children aged 0--59 months to
+standardize developmental exposure windows and conform to the World
+Health Organization (WHO) reference growth standards.
+
+The primary outcomes are height-for-age (HAZ) and weight-for-height
+(WHZ) z-scores, internationally recognized proxies for chronic and acute
+malnutrition, respectively. These anthropometric indicators are cleaned
+following DHS protocols: biologically implausible values are excluded,
+and remaining z-scores are rescaled from hundredths to continuous units.
+In addition to anthropometry, the child record includes a suite of
+health-related variables. These include indicators for whether the child
+has received basic vaccinations---such as BCG, DPT, polio, and
+measles---as well as whether the child experienced episodes of diarrhea,
+fever, or respiratory illness in the two weeks prior to the survey.
+These supplementary health measures serve as intermediate outcomes in
+secondary analyses to capture short-run morbidity and preventive health
+investment.
+
+Maternal-level information is drawn from the individual recode file.
+Core covariates include maternal age, completed years of education, and
+total number of children ever born (parity), each of which captures
+socioeconomic status and fertility history. I additionally include
+maternal body mass index (BMI) and anemia status, which serve as both
+controls and potential mediators of child health. These indicators
+reflect the mother's nutritional status and physiological capacity to
+support healthy fetal and postnatal development. In robustness checks, I
+examine whether declines in maternal nutritional status following the
+fertility shock account for observed changes in child anthropometry.
+
+Household-level characteristics are used to control for environmental
+and infrastructural determinants of health. Specifically, I include
+household size, urban residence, wealth quintile, access to electricity,
+type of toilet facility, whether the toilet is shared, and source of
+drinking water. These variables proxy for household-level resources and
+exposure to disease environments. In extension analyses, I interact
+treatment status with these environmental indicators to assess whether
+the nutritional consequences of the fertility shock are magnified under
+infrastructural deprivation.
+
+The treatment is defined by exposure to Executive Order No.~003, which
+was issued by the Manila City government in January 2000 and effectively
+suspended the public distribution of modern contraceptives. This policy
+change resulted in a localized increase in fertility that was not
+mirrored in surrounding municipalities within the National Capital
+Region (NCR). Children born in Manila City after the implementation of
+the policy are classified as treated, while children in other NCR cities
+serve as the control group. This spatially and temporally bounded
+intervention provides quasi-experimental variation that facilitates
+identification of causal effects.
+
+The empirical approach adopts a difference-in-differences framework that
+compares trends in child nutritional outcomes between treated and
+control cities before and after the policy change. The identifying
+assumption is that, absent the policy, nutritional trajectories in
+Manila and non-Manila NCR cities would have evolved in parallel. This
+assumption is supported by visual inspection and pre-trend placebo tests
+using the 1993 and 1998 waves. The 1998 DHS serves as the primary
+pre-policy baseline, while the 2003 survey captures short-run effects.
+The 2008 and 2013 waves are used in robustness checks to assess medium-
+and long-run persistence of nutritional deficits. All specifications
+include city and birth-year fixed effects, and standard errors are
+clustered at the city-year level to account for within-cluster
+correlation.
+
+The empirical strategy exploits the 2000 issuance of Executive Order
+No.~003, which effectively banned the distribution of modern
+contraceptives in public health centers across Manila City. This abrupt
+and localized withdrawal of contraceptive access resulted in a fertility
+increase concentrated within the city, but did not affect surrounding
+municipalities within the National Capital Region (NCR). Children born
+in Manila following the policy are considered exposed to a fertility
+shock, while children in other NCR cities over the same period serve as
+the control group. The analytical sample is restricted to children under
+five years old at the time of each survey round to ensure accurate
+measurement of early-life nutritional outcomes.
+
+I estimate the following difference-in-differences specification:
+
+\text{Nutrition}\_\{ict\} = \alpha + \beta \cdot (\text{Post}\_t
+\times \text{Manila}\emph{c) + X}\{ict\}'\gamma + \lambda\_c +
+\tau\emph{t + \varepsilon}\{ict\}
+
+In this equation, \(\text{Nutrition}_{ict}\) refers to either
+height-for-age (HAZ) or weight-for-height (WHZ) for child \(i\) in city
+\(c\) born in year \(t\). The indicator \(\text{Post}_t\) equals 1 if
+the child was born after the implementation of Executive Order No.~003
+in 2000, and \(\text{Manila}c\) equals 1 if the child resides in Manila
+City. The interaction term captures the effect of being born in Manila
+after the contraceptive policy was enacted. The vector \(X{ict}\)
+includes controls for child age in months, sex, birth order, maternal
+age, education, parity, body mass index (BMI), anemia status, household
+wealth quintile, household size, electricity access, water source,
+toilet type, shared sanitation status, and urban residence. City fixed
+effects \(\lambda_c\) control for time-invariant characteristics at the
+locality level, while year fixed effects \(\tau_t\) absorb aggregate
+time shocks. The coefficient of interest, \(\beta\), represents the
+average treatment effect of the policy-induced fertility shock on child
+nutrition. Standard errors are clustered at the city-year level to
+account for potential autocorrelation within geographic units over time.
+
+To support the internal validity of the design, I use Philippine Census
+microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude
+and timing of the fertility shock. The 1995 and 2007 waves provide clean
+pre- and post-policy benchmarks, while the 2000 census---although
+temporally proximate to the policy---is retained for robustness checks
+that test the sensitivity of estimates to the exact onset of fertility
+changes. A placebo comparison between 1990 and 1995 shows no divergence
+in fertility trends between Manila and control cities prior to the
+policy, strengthening the credibility of the parallel trends assumption.
+
+To supplement the DHS and Census data, I incorporate the Annual Poverty
+Indicators Survey (APIS), a nationally representative household survey
+that captures detailed information on food expenditure, maternal
+education, and living standards. Although the APIS lacks anthropometric
+data and has more limited geographic coverage, it enables exploratory
+analyses of treatment effect heterogeneity by income and education
+strata and helps triangulate changes in maternal resource constraints
+following the policy shock.
+
+Finally, missing data are handled following standard practices in
+applied microeconometrics. Observations with missing or flagged
+anthropometric outcomes are excluded from the analysis. For other
+covariates with partial nonresponse---such as maternal BMI, anemia, or
+household infrastructure---I adopt the missing-indicator method: missing
+values are imputed with a constant, and a binary indicator is included
+in all regression models to control for the fact of missingness. This
+approach preserves sample size while allowing the data to inform whether
+nonresponse is systematically related to the outcomes of interest. As
+robustness checks, I estimate alternative specifications using multiple
+imputation and inverse-probability weighting to test the sensitivity of
+results to different missing data assumptions. All descriptive and
+inferential statistics are computed using DHS-provided sampling weights,
+and models account for the complex survey design by clustering standard
+errors at the primary sampling unit.
+
 \bibliographystyle{aea}

### 2025-07-07 data-analysis/scripts/01_dhs_clean.R
replace and refactor dhs cleaning script
Deleted 01_dhs_clean.R and added 01_dhs_merge.R with a refactored approach for building the under-five child-level analytic panel from Philippine DHS data. The new script improves variable management, adds column prefixes to avoid name collisions, and outputs both .rds and .csv files for downstream analysis.

@@ -1,98 +1,0 @@
-# ==============================================================================
-# 01_dhs_clean.R -
-# Purpose : Load every KR, IR, HR DHS file for PH (1993–2013), harmonise, merge,
-#           flag treatment/control, and save to clean-data/.
-# Author  : Erika Salvador '28 (esalvador28@amherst.edu) 
-# Project : Manila Contraceptive Ban & the Quantity-Quality Trade-off (nutrition arm)
-# ==============================================================================
-
-# Load required libraries
-library(haven)        # = for reading .dta files (Stata format)
-library(dplyr)        # = for data manipulation (filter, mutate, join, etc.)
-library(labelled)     # = for handling variable labels from Stata
-library(janitor)      # = for cleaning column names (snake_case)
-library(stringr)      # = for string operations
-library(purrr)        # = for functional programming (map functions)
-library(glue)         # = for formatted strings
-library(readr)        # = for writing output files
-library(crayon)       # = for colorful messages
-library(cli)          # = for pretty status symbols
-
-
-# Step 1: Set up paths and years =============================================
-
-rounds <- c(1993, 1998, 2003, 2008, 2013)                # = list of survey years
-file_dir <- "../raw-data/dhs_raw"                        # = folder with raw files
-
-suffixes <- c(
-  kr = "kr_children",                                    # = Kids Recode files
-  ir = "ir_individual",                                  # = Individual Recode files
-  hr = "hr_household"                                    # = Household Recode files
-)
-
-# Function to read and standardize DHS files
-read_dhs <- function(year, suf) {
-  file.path(file_dir, glue("dhs_{year}_{suf}.DTA")) |>  # = build file path
-    read_stata() |>                                     # = read Stata file
-    clean_names() |>                                    # = clean column names
-    mutate(survey_year = year)                          # = tag with year
-}
-
-# Step 2: Load and select variables from each file type ======================
-
-# Children (KR)
-kr_vars <- c("survey_year", "v001", "v002", "bidx", "hw70", "hw71",
-             "b4", "b5", "b9", "bord", "caseid")
-
-kr_all <- map_dfr(rounds, read_dhs, suf = suffixes["kr"]) |>
-  select(any_of(kr_vars))
-
-# Women (IR)
-ir_vars <- c("survey_year", "caseid", "v001", "v002", "v003",    # = IDs
[…]

### 2025-07-07 data-analysis/scripts/01_dhs_merge.R

@@ -1,0 +1,104 @@
+# ==============================================================================
+# 01_dhs_clean.R -
+# Purpose = Build an under-five child-level analytic panel from Philippine DHS
+#           rounds 1993-2013.  Steps: read KR / IR / HR, keep core variables,
+#           merge, derive nutrition flags, save cleaned datasets.
+# Author  : Erika Salvador '28 (esalvador28@amherst.edu) 
+# Project : Manila Contraceptive Ban & the Quantity-Quality Trade-off 
+#           (nutrition arm)
+#
+# In plain English this script:
+#   1) Reads the raw child (KR), mother (IR), and household (HR)
+#      files for each survey year.
+#   2) Keeps only the variables we actually need.
+#   3) Adds prefixes (“m_”, “h_”) so mother- and household-level
+#      columns cannot collide with child-level names when merged.
+#   4) Merges the three files so that every row is a child.
+#   5) Stacks (appends) all years together.
+#   6) Saves the result as both .rds (R-native) and .csv (universal).
+# ==============================================================================
+
+# Load libraries
+# install.packages(c("haven","dplyr","purrr","readr"), dependencies = TRUE)
+library(haven)   # read_dta()
+library(dplyr)   # mutate(), select(), left_join(), etc.
+library(purrr)   # map_dfr() for looping
+library(readr)   # write_rds(), write_csv()
+
+# 1. Folders ===================================================================
+raw_path  <- "../raw-data/dhs_raw"  # where PHKR1993FL.DTA etc. live
+save_path <- "../clean-data"        # where clean files will be written
+
+# 2. Survey waves to process ===================================================
+survey_years <- c(1993, 1998, 2003, 2008, 2013)
+
+# 3. Short-lists of columns to retain ==========================================
+# Why short-list? Shrinking the file up-front speeds everything up,
+# lowers memory use, and avoids accidentally analyzing the wrong variable.
+
+kr_keep <- c(                                   # CHILD file
+  "survey_year", "v001", "v002", "v003", "bidx",          # IDs
+  "b4", "b5", "bord", "b9",                               # sex, survival, birth order, age (mos.)
+  "hw70", "hw71", "hw72",                                 # HAZ, WAZ, WHZ ×100
[…]

### 2025-07-07 data-analysis/scripts/01_dhs_merge.R
update comments and variable descriptions
Replaces curly quotes and clarifies comments regarding variable prefixes and merge steps. No functional code changes were made; updates are limited to documentation and code comments for improved clarity.

@@ -1,5 +1,5 @@
 # ==============================================================================
-# 01_dhs_clean.R -
+# 01_dhs_merge.R -
 # Purpose = Build an under-five child-level analytic panel from Philippine DHS
 #           rounds 1993-2013.  Steps: read KR / IR / HR, keep core variables,
 #           merge, derive nutrition flags, save cleaned datasets.
@@ -11,7 +11,7 @@
 #   1) Reads the raw child (KR), mother (IR), and household (HR)
 #      files for each survey year.
 #   2) Keeps only the variables we actually need.
-#   3) Adds prefixes (“m_”, “h_”) so mother- and household-level
+#   3) Adds prefixes (???m_???, ???h_???) so mother- and household-level
 #      columns cannot collide with child-level names when merged.
 #   4) Merges the three files so that every row is a child.
 #   5) Stacks (appends) all years together.
@@ -39,7 +39,7 @@ survey_years <- c(1993, 1998, 2003, 2008, 2013)
 kr_keep <- c(                                   # CHILD file
   "survey_year", "v001", "v002", "v003", "bidx",          # IDs
   "b4", "b5", "bord", "b9",                               # sex, survival, birth order, age (mos.)
-  "hw70", "hw71", "hw72",                                 # HAZ, WAZ, WHZ ×100
+  "hw70", "hw71", "hw72",                                 # HAZ, WAZ, WHZ ??100
   "m14", "m15", "m3a", "m3b", "m3c", "m3h",               # ANC & delivery
   "h2","h3","h4","h5","h6","h7","h8","h9",                # vaccinations
   "h11", "h22", "h31",                                    # illness
@@ -76,7 +76,7 @@ load_and_clean_year <- function(year) {
     mutate(survey_year = year) |>
     select(any_of(ir_keep)) |>
     rename_with(~ paste0("m_", .x), -c(survey_year, v001, v002, v003))
-  # Prefix “m_” ensures we don't overwrite shared variable names like v024 (region)
+  # Prefix ???m_??? ensures we don't overwrite shared variable names like v024 (region)
   
   # HOUSEHOLD (HR)
   hr_file <- file.path(raw_dir, paste0("dhs_", year, "_hr_household.DTA"))
@@ -84,9 +84,9 @@ load_and_clean_year <- function(year) {
     mutate(survey_year = year) |>
[…]

### 2025-07-27 docs/paper.Rmd
fix math equation formatting in empirical strategy
Corrected the display of the difference-in-differences equation in both the R Markdown and LaTeX source to use proper math environments.
revise paper title and introduction for clarity
Updated the paper's title, subtitle, and abstract to focus on early-life health outcomes and the impact of Manila's 2000 contraceptive ban. The introduction was rewritten to emphasize health outcomes (birth weight, perceived birth size, vaccinations) rather than nutrition alone, clarify the empirical strategy, and provide more context on the policy and its quasi

@@
 ---
-title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
+title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment"
+subtitile: "Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
 short: "The Q-Q Trade-off in the Philippine Context"
@@
 abstract: |
-  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households.
-I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities.
-Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height.
-[Will add more after data analysis]
+  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households.
+I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No.
+003—which effectively banned the provision of modern contraceptives in all city-run health facilities.
+Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy.
+The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
 output: 
@@
 First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{becker1973interaction, becker1976child}, this framework posits that parents allocate finite resources—both financial and non-financial—across children.
-An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital.
-This mechanism has served as an explanatory model for changes in fertility behavior and the evolution of population structures in low- and middle-income countries.
+An increase in fertility dilutes the resources available per child and, under binding constraints, may lead to reduced investments in each child’s human capital.
 
-Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes.
-Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}.
+Empirical investigations of the quantity-quality trade-off have focused primarily on educational outcomes.
+Most studies examine how increases in maternal fertility—or equivalently, the number of siblings a child has—affect school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}.
 These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality.
-Other outcomes, such as nutritional status, are early-onset or biologically constrained.
-They also tend to be less responsive to remediation later in life.
-A child who suffers from chronic malnutrition may exhibit permanently reduced cognitive capacity and face limits in physical development that affect long-run productivity regardless of subsequent educational access \citep{hoddinott2013adult, grantham2007development}.
+Other dimensions of child quality–such as health at birth and early childhood health investments—are early-onset or biologically constrained, and they tend to be less amenable to remediation later in life.
+For instance, babies born with very low birth weight face higher risks of infant mortality and long-term developmental deficits that cannot be fully reversed \citep{cook2015understanding}.
+Similarly, a child who misses critical early vaccinations is more vulnerable to infectious diseases that can kill or disable them, which results to lifelong negative health impacts \citep{costa2024child}.
+Such early-life health setbacks may permanently impair cognitive development and physical growth.
+This limits long-run productivity regardless of subsequent improvements in schooling or incom
 
-The exclusion of nutritional outcomes from much of the empirical literature leaves an important dimension of the trade-off untested.
-Nutritional investments in early childhood are essential to early childhood development and long-term outcomes \citep{victora2008maternal, hoddinott2013adult}.
+The omission of these health outcomes from much of the empirical literature leaves important aspects of the quantity-quality trade-off unexamined.
+Investments in maternal and child health—such as adequate prenatal care leading to healthy birth weight and timely childhood immunizations—are important to early development and long-term outcomes \citep{victora2008maternal, hoddinott2013adult}.
 They shape brain development, immune system functioning, and physical stature, and they have been shown to predict later-life earnings and health outcomes across a wide range of settings \citep{grantham2007development, alderman2006long}.
-The biological irreversibility of early-life nutritional deficits further distinguishes them from other forms of investment.
-Educational deficits may be partially remediable; nutritional failures often are not.
-A credible estimate of the trade-off between fertility and child quality must account for nutrition if it aims to assess the full set of consequences associated with fertility shocks.
+Unlike educational gaps, which may be partially addressed later in life, early health deficits—such as poor nutrition or missed vaccinations—are often irreversible \citep{victora2008maternal, hoddinott2013adult}.
+As a result, any credible evaluation of the quantity-quality trade-off must include early-life health outcomes to fully capture the long-term consequences of different fertility shocks.
 
-This study addresses this gap by examining the nutritional effects of a localized, exogenous increase in fertility in the Philippines.
-In 1990, the mayor of Manila implemented an executive policy that prohibited the provision of modern contraceptives in all city-run health facilities.
-The order removed access to pills, condoms, intrauterine devices, and related public health materials and instructed healthcare providers to offer only natural family planning methods.
-This policy remained in place for nearly a decade and affected only the jurisdiction of the Manila city government.
-The national government did not implement a comparable restriction, and surrounding cities within Metro Manila continued to provide access to modern contraceptives.
-The policy thus created a spatial and temporal discontinuity in contraceptive access that was uncorrelated with underlying fertility preferences or concurrent shifts in household income or governance.
-As a result, the Manila ban serves as a quasi-experimental source of variation in fertility exposure among poor urban households.
+This study addresses that gap by examining how a localized, exogenous increase in fertility influenced early-life health outcomes in the Philippines.
+In 2000, then–Mayor José “Lito” Atienza of Manila City issued Executive Order No.
+003, which effectively barred the provision and promotion of modern contraceptives in all city-run health facilities.
+The directive eliminated access to pills, condoms, intrauterine devices, and other forms of modern family planning, and mandated that public healthcare providers offer only natural methods.
+The policy remained in effect for nearly a decade and applied solely within the jurisdiction of the Manila city government.
+No comparable restriction was implemented at the national level, and neighboring cities within Metro Manila continued to provide access to modern contraceptives.
+This policy created a clear spatial and temporal discontinuity in access to family planning services—one that was plausibly unrelated to changes in fertility preferences, household income, or broader governance trends.
+As such, the Manila contraceptive ban provides a quasi-experimental shock to fertility exposure among poor urban households.
 
-I use this natural experiment to estimate the causal effect of increased fertility on child nutrition.
-The analysis relies on nationally representative data from multiple waves of the Philippine Demographic and Health Survey (DHS), which provide data on household structure and maternal characteristics, as well as measurements of child anthropometry.
-The outcomes of interest are height-for-age and weight-for-height z-scores, which serve as standardized indicators of chronic and acute malnutrition, respectively.
-These outcomes are widely used in the global health and development literature and capture nutritional deprivation over both long and short time horizons \citep{victora2008maternal}.
+I use this natural experiment to estimate the causal impact of increased family size on early-childhood health outcomes.
+The analysis uses nationally representative data from multiple rounds of the Philippine Demographic and Health Survey (DHS), which include detailed information on household structure, maternal characteristics, and child health indicators.
+The primary outcomes are birth weight (measured in grams, when reported) and the mother’s subjective assessment of the child’s size at birth—both of which serve as indicators of neonatal health status.
+In addition, I assess whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment.
+These outcomes are widely recognized indicators of child well-being and reflect both biological and behavioral dimensions of parental investment in early development.
 The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.
@@
 The identification strategy rests on two key assumptions.
-First, in the absence of the contraceptive ban, nutritional trends in Manila would have evolved in parallel with those in comparison cities.
-Second, any other policy or economic shocks affecting Manila during the study period must not coincide precisely with the timing and scope of the contraceptive policy.
-I test these assumptions using falsification checks, placebo comparisons, and robustness specifications that include city-specific time trends, maternal fixed effects, and controls for baseline demographic differences.
+First, in the absence of the contraceptive ban, trends in child health outcomes in Manila would have evolved in parallel with those in the comparison cities.
+Second, no other policy changes or economic shocks in Manila during the study period coincided with the timing and scope of the contraceptive policy.
+I probe these assumptions using falsification tests, placebo comparisons, and robustness checks that include city-specific time trends, mother fixed effects, and controls for baseline demographic differences.
 
@@
 I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women.
-I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age.
-Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care.
-These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.
+Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, perceived birth size, and vaccination uptake).
+Finally, I examine heterogeneity in these effects across subgroups defined by maternal education, household wealth, and access to prenatal care.
+These characteristics proxy for a household’s resources and its capacity to buffer the impact of increased family size on child health.
 
 This study contributes to the literature in several important ways.
-It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities.
-Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting.
-It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being.
-More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.
+It provides new evidence on how an externally induced increase in family size, resulting from a local restriction on contraceptive access,  can affect child health in poor urban communities.
+Whereas most past research on the quantity-quality trade-off has focused on education, this study broadens the notion of child “quality” to include early-life health measures such as birth weight and immunization status.
+It also adds to the small number of studies that leverage unexpected changes in reproductive health policy to examine long-term effects on children’s well-being.
+More broadly, the findings indicate that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public health services.
 
@@
 
-\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict}
+$$ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} $$
 

### 2025-07-27 docs/paper.tex
fix math equation formatting in empirical strategy
Corrected the display of the difference-in-differences equation in both the R Markdown and LaTeX source to use proper math environments.
revise paper title and introduction for clarity
Updated the paper's title, subtitle, and abstract to focus on early-life health outcomes and the impact of Manila's 2000 contraceptive ban. The introduction was rewritten to emphasize health outcomes (birth weight, perceived birth size, vaccinations) rather than nutrition alone, clarify the empirical strategy, and provide more context on the policy and its quasi

@@
 
-\title{Nutritional Deficits and the Quantity-Quality Trade-off: Evidence
-from an Exogenous Fertility Shock in Low-Income Urban Settings in the
-Philippines}
+\title{Womb for One More? Fertility Expansion and the Limits of Early
+Health Investment}
 \shortTitle{The Q-Q Trade-off in the Philippine Context}
@@
 \begin{abstract}
-This paper examines whether increased fertility affects early-life
-nutritional outcomes in low-income urban households. I exploit a natural
-experiment created by a 1990 policy in Manila, Philippines, which banned
-modern contraceptives from city-run health facilities. Using a
-difference-in-differences framework and nationally representative data
-from the Philippine Demographic and Health Surveys, I estimate the
-reduced-form impact of the policy on child height-for-age and
-weight-for-height. {[}Will add more after data analysis{]}
+This paper examines whether increased fertility, driven by reduced
+access to modern contraception, affects early-life health outcomes in
+low-income urban households. I exploit a natural experiment created by a
+2000 policy in Manila, Philippines---Executive Order No.~003---which
+effectively banned the provision of modern contraceptives in all
+city-run health facilities. Using a difference-in-differences design and
+nationally representative data from the Philippine Demographic and
+Health Surveys, I estimate the policy's impact on child health by
+comparing outcomes in Manila before and after the ban to those in
+comparable urban areas unaffected by the policy. The outcomes of
+interest include birth weight (in grams), the mother's perceived size of
+the child at birth, and receipt of key childhood vaccinations.
 \end{abstract}
@@
 framework posits that parents allocate finite resources---both financial
-and non-financial---across children. An increase in fertility reduces
+and non-financial---across children. An increase in fertility dilutes
 the resources available per child and, under binding constraints, may
-lead to lower investments in health, education, and other forms of human
-capital. This mechanism has served as an explanatory model for changes
-in fertility behavior and the evolution of population structures in low-
-and middle-income countries.
-
-Empirical investigations of the quantity--quality trade-off have focused
-primarily on educational outcomes. Studies in both high-income and
-low-income settings have examined the effects of fertility on school
-enrollment, grade progression, test scores, and completed years of
-schooling
+lead to reduced investments in each child's human capital.
+
+Empirical investigations of the quantity-quality trade-off have focused
+primarily on educational outcomes. Most studies examine how increases in
+maternal fertility---or equivalently, the number of siblings a child
+has---affect school enrollment, grade progression, test scores, and
+completed years of schooling
 \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
@@
 accumulation, but they represent only one dimension of child quality.
-Other outcomes, such as nutritional status, are early-onset or
-biologically constrained. They also tend to be less responsive to
-remediation later in life. A child who suffers from chronic malnutrition
-may exhibit permanently reduced cognitive capacity and face limits in
-physical development that affect long-run productivity regardless of
-subsequent educational access
-\citep{hoddinott2013adult, grantham2007development}.
-
-The exclusion of nutritional outcomes from much of the empirical
-literature leaves an important dimension of the trade-off untested.
-Nutritional investments in early childhood are essential to early
-childhood development and long-term outcomes
-\citep{victora2008maternal, hoddinott2013adult}. They shape brain
-development, immune system functioning, and physical stature, and they
-have been shown to predict later-life earnings and health outcomes
+Other dimensions of child quality--such as health at birth and early
+childhood health investments---are early-onset or biologically
+constrained, and they tend to be less amenable to remediation later in
+life. For instance, babies born with very low birth weight face higher
+risks of infant mortality and long-term developmental deficits that
+cannot be fully reversed \citep{cook2015understanding}. Similarly, a
+child who misses critical early vaccinations is more vulnerable to
+infectious diseases that can kill or disable them, which results to
+lifelong negative health impacts \citep{costa2024child}. Such early-life
+health setbacks may permanently impair cognitive development and
+physical growth. This limits long-run productivity regardless of
+subsequent improvements in schooling or incom
+
+The omission of these health outcomes from much of the empirical
+literature leaves important aspects of the quantity-quality trade-off
+unexamined. Investments in maternal and child health---such as adequate
+prenatal care leading to healthy birth weight and timely childhood
+immunizations---are important to early development and long-term
+outcomes \citep{victora2008maternal, hoddinott2013adult}. They shape
+brain development, immune system functioning, and physical stature, and
+they have been shown to predict later-life earnings and health outcomes
 across a wide range of settings
-\citep{grantham2007development, alderman2006long}. The biological
-irreversibility of early-life nutritional deficits further distinguishes
-them from other forms of investment. Educational deficits may be
-partially remediable; nutritional failures often are not. A credible
-estimate of the trade-off between fertility and child quality must
-account for nutrition if it aims to assess the full set of consequences
-associated with fertility shocks.
-
-This study addresses this gap by examining the nutritional effects of a
-localized, exogenous increase in fertility in the Philippines. In 1990,
-the mayor of Manila implemented an executive policy that prohibited the
-provision of modern contraceptives in all city-run health facilities.
-The order removed access to pills, condoms, intrauterine devices, and
-related public health materials and instructed healthcare providers to
-offer only natural family planning methods. This policy remained in
-place for nearly a decade and affected only the jurisdiction of the
-Manila city government. The national government did not implement a
-comparable restriction, and surrounding cities within Metro Manila
-continued to provide access to modern contraceptives. The policy thus
-created a spatial and temporal discontinuity in contraceptive access
-that was uncorrelated with underlying fertility preferences or
-concurrent shifts in household income or governance. As a result, the
-Manila ban serves as a quasi-experimental source of variation in
-fertility exposure among poor urban households.
-
-I use this natural experiment to estimate the causal effect of increased
-fertility on child nutrition. The analysis relies on nationally
-representative data from multiple waves of the Philippine Demographic
-and Health Survey (DHS), which provide data on household structure and
-maternal characteristics, as well as measurements of child
-anthropometry. The outcomes of interest are height-for-age and
-weight-for-height z-scores, which serve as standardized indicators of
-chronic and acute malnutrition, respectively. These outcomes are widely
-used in the global health and development literature and capture
-nutritional deprivation over both long and short time horizons
-\citep{victora2008maternal}. The empirical strategy follows a
-difference-in-differences design that compares child outcomes in Manila
-and comparable urban areas before and after the onset of the policy.
+\citep{grantham2007development, alderman2006long}. Unlike educational
+gaps, which may be partially addressed later in life, early health
+deficits---such as poor nutrition or missed vaccinations---are often
+irreversible \citep{victora2008maternal, hoddinott2013adult}. As a
+result, any credible evaluation of the quantity-quality trade-off must
+include early-life health outcomes to fully capture the long-term
+consequences of different fertility shocks.
+
+This study addresses that gap by examining how a localized, exogenous
+increase in fertility influenced early-life health outcomes in the
+Philippines. In 2000, then--Mayor José ``Lito'' Atienza of Manila City
+issued Executive Order No.~003, which effectively barred the provision
+and promotion of modern contraceptives in all city-run health
+facilities. The directive eliminated access to pills, condoms,
+intrauterine devices, and other forms of modern family planning, and
+mandated that public healthcare providers offer only natural methods.
+The policy remained in effect for nearly a decade and applied solely
+within the jurisdiction of the Manila city government. No comparable
+restriction was implemented at the national level, and neighboring
+cities within Metro Manila continued to provide access to modern
+contraceptives. This policy created a clear spatial and temporal
+discontinuity in access to family planning services---one that was
+plausibly unrelated to changes in fertility preferences, household
+income, or broader governance trends. As such, the Manila contraceptive
+ban provides a quasi-experimental shock to fertility exposure among poor
+urban households.
+
+I use this natural experiment to estimate the causal impact of increased
+family size on early-childhood health outcomes. The analysis uses
+nationally representative data from multiple rounds of the Philippine
+Demographic and Health Survey (DHS), which include detailed information
+on household structure, maternal characteristics, and child health
+indicators. The primary outcomes are birth weight (measured in grams,
+when reported) and the mother's subjective assessment of the child's
+size at birth---both of which serve as indicators of neonatal health
+status. In addition, I assess whether the child received key early-life
+vaccinations, including those against polio,
+diphtheria--pertussis--tetanus (DPT), and measles, as a measure of
+essential postnatal health investment. These outcomes are widely
+recognized indicators of child well-being and reflect both biological
+and behavioral dimensions of parental investment in early development.
+The empirical strategy follows a difference-in-differences design that
+compares child outcomes in Manila and comparable urban areas before and
+after the onset of the policy.
 
 The identification strategy rests on two key assumptions. First, in the
-absence of the contraceptive ban, nutritional trends in Manila would
-have evolved in parallel with those in comparison cities. Second, any
-other policy or economic shocks affecting Manila during the study period
-must not coincide precisely with the timing and scope of the
-contraceptive policy. I test these assumptions using falsification
-checks, placebo comparisons, and robustness specifications that include
-city-specific time trends, maternal fixed effects, and controls for
+absence of the contraceptive ban, trends in child health outcomes in
+Manila would have evolved in parallel with those in the comparison
+cities. Second, no other policy changes or economic shocks in Manila
+during the study period coincided with the timing and scope of the
+contraceptive policy. I probe these assumptions using falsification
+tests, placebo comparisons, and robustness checks that include
+city-specific time trends, mother fixed effects, and controls for
 baseline demographic differences.
@@
 \citep{dumas2019sex} to confirm that the contraceptive ban led to an
-increase in fertility among affected women. I then estimate reduced-form
-effects of policy exposure on nutritional outcomes for children under
-five years of age. Finally, I examine heterogeneity in effects across
-subsamples defined by maternal education, household wealth, and access
-to prenatal care. These dimensions serve as proxies for household
-resource availability and capacity to buffer the nutritional
-consequences of fertility increases.
+increase in fertility among affected women. Next, I estimate the
+reduced-form effects of exposure to the policy on the selected child
+health outcomes (birth weight, perceived birth size, and vaccination
+uptake). Finally, I examine heterogeneity in these effects across
+subgroups defined by maternal education, household wealth, and access to
+prenatal care. These characteristics proxy for a household's resources
+and its capacity to buffer the impact of increased family size on child
+health.
 
 This study contributes to the literature in several important ways. It
-provides new evidence on how increases in fertility---caused by policy
-restrictions on family planning---can affect child nutrition in poor,
-urban communities. Most past research has focused on education, but this
-study expands the idea of child quality to include biological measures
-such as stunting and wasting. It also adds to the small number of
-studies that use unexpected changes in reproductive health policy to
-examine long-term effects on children's well-being. More broadly, the
-results show that local restrictions on family planning can
-unintentionally harm children's health, especially in settings where
-families already face poverty, food insecurity, and limited public
-services.
+provides new evidence on how an externally induced increase in family
+size, resulting from a local restriction on contraceptive access, can
+affect child health in poor urban communities. Whereas most past
+research on the quantity-quality trade-off has focused on education,
+this study broadens the notion of child ``quality'' to include
+early-life health measures such as birth weight and immunization status.
+It also adds to the small number of studies that leverage unexpected
+changes in reproductive health policy to examine long-term effects on
+children's well-being. More broadly, the findings indicate that local
+restrictions on family planning can unintentionally harm children's
+health, especially in settings where families already face poverty, food
+insecurity, and limited public health services.
 
@@
 
-\text{Nutrition}\_\{ict\} = \alpha + \beta \cdot (\text{Post}\_t
-\times \text{Manila}\emph{c) + X}\{ict\}'\gamma + \lambda\_c +
-\tau\emph{t + \varepsilon}\{ict\}
+\[ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} \]
 

### 2025-07-27 docs/paper.Rmd
revise theory and empirical sections in paper draft
Expanded and clarified the literature review, conceptual framework, and empirical strategy in the paper. Standardized terminology from 'quantity–quality' to 'quantity-quality', added new references and recent evidence, and restructured sections for improved flow. Updated title and subtitle for consistency across Rmd, tex, and PDF outputs.

@@
 ---
-title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment"
-subtitile: "Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
+title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
 short: "The Q-Q Trade-off in the Philippine Context"
@@
 
-The quantity–quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each.
+The quantity-quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each.
 Quality in this context refers to the human capital of each child: attributes like education, health, and nutrition that enhance a child’s future productivity and well-being.
@@
 
-Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity–quality cost structure.
+Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity-quality cost structure.
 An increase in $n$ raises the total cost required to sustain a given level of $q$ for each child, while a higher level of $q$ raises the marginal cost associated with having an additional child.
@@
 By the late 1980s, work by Becker, Barro, and others had recast fertility as an endogenous choice that interacts with capital accumulation, income distribution, and policy.
-This laid the foundation for unified growth theories, which view the quantity–quality mechanism as central to demographic transition and long-run development.
+This laid the foundation for unified growth theories, which view the quantity-quality mechanism as central to demographic transition and long-run development.
 
@@
 Several mechanisms support this transition, such as a fall in child labor, a drop in child mortality, and a shift in parental priorities.
-These models explain not only the presence of a quantity–quality trade-off but also its rising influence at a specific point in history.
+These models explain not only the presence of a quantity-quality trade-off but also its rising influence at a specific point in history.
 The evidence supports these claims: countries that saw earlier increases in returns to education experienced earlier fertility decline, while delays in reforms, such as public education or health access, corresponded to prolonged high fertility.
@@
 
-Contemporary research has further refined the quantity–quality model by relaxing some of its initial simplifying assumptions.
+Contemporary research has further refined the quantity-quality model by relaxing some of its initial simplifying assumptions.
 Three important extensions involve (1) capital market imperfections that constrain parents’ ability to invest in child quality, (2) intra-household conflict and bargaining between mothers and fathers over fertility choices, and (3) recognition that child quality is multi-dimensional, which extends beyond schooling to include health, nutrition, and other facets of human capital.
@@
 To begin, I examine how credit constraints can give rise to poverty traps.
-The canonical quantity–quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high.
+The canonical quantity-quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high.
 In practice, especially in low-income settings, credit markets function imperfectly: poor households typically cannot secure loans to cover children’s education or medical costs even when such investments would yield substantial future gains.
@@
 
-Formal models show that “the quality–quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism.
+Formal models show that “the quality-quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism.
 \citet{kremer2002income} and \citet{de2003inequality} document that countries facing tighter liquidity constraints tend to display higher fertility and lower educational attainment, consistent with liquidity-constrained parents favoring quantity over quality.
@@
 
-A further refinement of the quantity–quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions.
+A further refinement of the quantity-quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions.
 The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \).
@@
 The original Q–Q models often used a single catch-all variable for child quality, typically thought of as education or “expenditure per child.” Recent work emphasizes that child quality is multi-faceted, and that parents make trade-offs along several dimensions of investment—cognitive development, health, nutrition, etc.
-This is particularly salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.
+This is salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.
 
@@
 
-Beyond survival, nutritional status is a critical indicator of quality in many low-income settings.
-Parents often face choices about how to allocate food or medical care, especially under resource scarcity.
-Empirical evidence supports the Q–Q model in this context: children with many siblings often exhibit poorer nutritional outcomes.
-A recent study using Vietnam Young Lives data finds that an additional sibling reduces height-for-age and weight-for-age z-scores by about 0.3 standard deviations on average \citep{chen2021population}  This effect is substantial, suggesting that children in larger families tend to be shorter and lighter than their peers from smaller families, likely due to the more limited distribution of healthcare and parental attention
+Beyond survival, early-life health indicators such as birth weight, perceived size at birth, and vaccination status are important dimensions of child quality, especially in low-income settings.
+When resources are limited, parents may struggle to meet the basic needs of each child.
+Empirical evidence aligns with the quantity–quality framework: children from larger families often have worse health outcomes at birth and are less likely to receive core immunizations.
+For example, recent multi-country data show that the share of completely unimmunized children—so-called “zero-dose” children—increases from 10.5% among those with no siblings to 17.2% among those with four or more \citep{costa2024child}.
+Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth.
+These patterns suggest that as fertility increases, the health of each child may suffer due to more limited parental investment.
+Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth
 
-Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia , where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs.
+Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia, where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs.
 Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:
@@
 Horton found that better-educated mothers and fathers tended to choose smaller families and achieved better-nourished children–evidence of substitution away from “quantity” toward child “quality.” Notably, she observed strong birth-order effects on nutrition (first-born children faring better than later-born), which hint that parents may not evenly distribute quality investments among all children.
-This early work indicated a quantity–quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.
+This early work indicated a quantity-quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.
 
@@
 This regressive pattern aligns with Becker’s theory that resource constraints bind more tightly for low-income families, which makes the Q–Q trade-off more acute.
-In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity–quality trade-off in this context.
+In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity-quality trade-off in this context.
 Large family size thus emerges as one mechanism contributing to poverty, by impeding children’s human capital accumulation in the Philippines.
@@
 
-The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity–quality trade-off with diverse outcomes and methods.
+The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity-quality trade-off with diverse outcomes and methods.
 In Vietnam, for example, rapid fertility decline alongside rising education led to questions about a Q–Q mechanism.
@@
 
-Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity–quality conjecture.
+Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity-quality conjecture.
 As families have fewer children, they appear to invest more in each child’s education—investments crucial for human capital development and economic growth.
 Conversely, high-fertility households risk under-investing per child, which reinforces cycles of poverty and inequality.
-This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity–quality trade-off and help families achieve both manageable size and better outcomes for the next generation.
+This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity-quality trade-off and help families achieve both manageable size and better outcomes for the next generation.
 
@@
 
-This study builds upon the foundational framework of the quantity–quality (Q–Q) trade-off, as articulated in \citet{becker1960economic} seminal work and later refined by \citet{becker1973interaction}.
-Within this theoretical model, fertility and child quality are jointly determined under a full-income constraint, such that any increase in the number of children—absent a commensurate expansion in resources—necessitates a reduction in the average investment per child.
-While prior empirical applications of this framework have largely centered on educational attainment, child quality may also be understood along biological dimensions, especially along nutritional status, which emerges earlier in the life course and may be less subject to later remediation.
+This study builds upon the foundational framework of the quantity-quality (Q–Q) trade-off, first articulated in \citet{becker1960economic} and later refined in \citet{becker1973interaction}.
+In this model, fertility and child quality are jointly determined under a full-income constraint, such that increases in the number of children—without a corresponding expansion in household resources—necessitate a reduction in the average investment per child.
+While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, perceived size at birth, and uptake of essential vaccinations.
 
-The current analysis focuses exclusively on early-life nutrition as a proxy for child quality.
-Nutrition, unlike cognitive or educational outcomes, is characterized by low substitutability and a narrow temporal window during which investments are most productive.
-The trade-off is thus expected to manifest strongly in settings where fertility rises exogenously while household resources remain constrained.
-In such environments, families facing an unexpected increase in sibship size may be forced to reallocate scarce food, time, and health inputs across a larger number of dependents, potentially resulting in lower nutritional attainment for each child.
+This analysis focuses specifically on early-life health as a proxy for child quality.
+These outcomes, unlike cognitive or schooling-related investments, are biologically constrained and occur within a narrow developmental window, during which parental inputs are least substitutable and most consequential.
+The trade-off is expected to be especially salient in settings where fertility increases exogenously while family resources remain fixed.
+In such contexts, a sudden increase in sibship size may compel households to spread food, healthcare, and caregiving attention across more dependents, which would potentially lead to diminished health investments and outcomes for each child.
 
@@
 In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers.
-If nutritional investments per child decline, and if such inputs are critical for growth, observable deficits in height-for-age (HAZ) or weight-for-height (WHZ) may emerge.
-The first hypothesis emerging from this framework is that an exogenous increase in
+If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight or perceived neonatal size, and children may be less likely to receive complete immunization schedules.
 
-fertility leads to a reduction in per-child nutritional investment, particularly in low-income, urban households where slack in the family budget is limited.
-A second, related hypothesis posits that this reduction in inputs translates into measurable nutritional deficits among young children, specifically in the form of lower anthropometric outcomes such as height-for-age (HAZ) and weight-for-height (WHZ) z-scores.
-These two hypotheses—concerning, respectively, the behavioral reallocation of resources and its biological consequence—jointly seek to identify a causal link between fertility shocks and early-life deprivation.
+The first hypothesis derived from this framework is that an exogenous increase in fertility leads to a decline in per-child health investment, particularly in low-income urban households with little financial slack.
+A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight, smaller perceived size at birth, and reduced uptake of key childhood vaccinations.
+These hypotheses describe a causal pathway through which fertility shocks may contribute to early deprivation and potentially reinforce intergenerational cycles of poor health and limited human capital.
 
@@
 
-This study constructs a harmonized, child-level analytic dataset using five rounds of the Philippine Demographic and Health Survey (DHS): 1993, 1998, 2003, 2008, and 2013.
-The DHS is a nationally representative, cross-sectional survey series conducted by the Philippine Statistics Authority in collaboration with ICF International.
-Each round follows a stratified two-stage cluster design and collects standardized information on child health, fertility, maternal characteristics, and household infrastructure.
-For each wave, I merge the child-level (KR), mother-level (IR), and household-level (HR) files using cluster, household, and maternal identifiers, ensuring that each observation corresponds to a live birth occurring within five years of the survey.
-The analytic sample is restricted to children aged 0–59 months to standardize developmental exposure windows and conform to the World Health Organization (WHO) reference growth standards.
+\subsection{Data}
 
-The primary outcomes are height-for-age (HAZ) and weight-for-height (WHZ) z-scores, internationally recognized proxies for chronic and acute malnutrition, respectively.
-These anthropometric indicators are cleaned following DHS protocols: biologically implausible values are excluded, and remaining z-scores are rescaled from hundredths to continuous units.
-In addition to anthropometry, the child record includes a suite of health-related variables.
-These include indicators for whether the child has received basic vaccinations—such as BCG, DPT, polio, and measles—as well as whether the child experienced episodes of diarrhea, fever, or respiratory illness in the two weeks prior to the survey.
-These supplementary health measures serve as intermediate outcomes in secondary analyses to capture short-run morbidity and preventive health investment.
+The primary data source for this study is the Philippine Demographic and Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
+These nationally representative surveys provide detailed child- and household-level information, including maternal characteristics, fertility behavior, and early-childhood health outcomes.
+Importantly, they are the only Philippine DHS waves with available GPS cluster data, which I use to approximate geographic location and assign respondents to treatment and control groups.
+Because city-level identifiers are not released in the public-use files, I rely on coarse geographic assignment based on cluster coordinates.
+Clusters located within or near the administrative boundaries of the City of Manila are classified as “treated,” while clusters located in other parts of the National Capital Region (NCR) serve as the control group.
+In robustness checks, I restrict the control group to Quezon City or expand it to the rest of NCR excluding both Manila and Quezon City.
 
-Maternal-level information is drawn from the individual recode file.
-Core covariates include maternal age, completed years of education, and total number of children ever born (parity), each of which captures socioeconomic status and fertility history.
-I additionally include maternal body mass index (BMI) and anemia status, which serve as both controls and potential mediators of child health.
-These indicators reflect the mother’s nutritional status and physiological capacity to support healthy fetal and postnatal development.
-In robustness checks, I examine whether declines in maternal nutritional status following the fertility shock account for observed changes in child anthropometry.
+The analytic sample includes children under five years of age living with their mothers at the time of the survey.
+I focus on three early-life health outcomes that reflect both biological endowments and parental investment: (1) birth weight, reported in grams when available; (2) the mother’s subjective assessment of birth size; and (3) vaccination uptake, defined as receipt of key early-childhood immunizations, including BCG, DPT, polio, and measles.
+These outcomes are consistently available across all three waves and serve as proxies for child quality in the context of the quantity–quality trade-off.
 
-Household-level characteristics are used to control for environmental and infrastructural determinants of health.
-Specifically, I include household size, urban residence, wealth quintile, access to electricity, type of toilet facility, whether the toilet is shared, and source of drinking water.
-These variables proxy for household-level resources and exposure to disease environments.
-In extension analyses, I interact treatment status with these environmental indicators to assess whether the nutritional consequences of the fertility shock are magnified under infrastructural deprivation.
+The 1998 DHS is used as the primary pre-policy baseline wave because it captures children born before the implementation of Manila’s contraceptive ban in 2000.
+The 2008 DHS serves as the post-policy round to capture outcomes for children born well after the ban took effect.
+I also include the 2003 DHS in the main specification, recognizing that children in this round may have been born shortly before or after the policy’s introduction.
+I treat 2003 as a transition-period dataset and conduct sensitivity checks to assess the robustness of findings to its inclusion.
+I exclude the 1993 DHS from the analysis due to the absence of GPS data, which makes it impossible to reliably assign treatment status.
 
-The treatment is defined by exposure to Executive Order No.
-003, which was issued by the Manila City government in January 2000 and effectively suspended the public distribution of modern contraceptives.
-This policy change resulted in a localized increase in fertility that was not mirrored in surrounding municipalities within the National Capital Region (NCR).
-Children born in Manila City after the implementation of the policy are classified as treated, while children in other NCR cities serve as the control group.
-This spatially and temporally bounded intervention provides quasi-experimental variation that facilitates identification of causal effects.
+To support the identification strategy, I use DHS summary indicators on contraceptive use disaggregated by region and city.
+These data confirm a sharp and sustained decline in modern contraceptive use in Manila following the implementation of Executive Order No.
+003, relative to surrounding cities, validating the policy as a plausibly exogenous fertility shock.
+I also considered using data from the Philippine National Nutrition Survey (NNS), which includes rich anthropometric and dietary data.
+However, the NNS microdata are only available from 2003 onward, which limits their value for pre-policy analysis.
+As such, the DHS remains the most suitable dataset for this study.
 
-The empirical approach adopts a difference-in-differences framework that compares trends in child nutritional outcomes between treated and control cities before and after the policy change.
-The identifying assumption is that, absent the policy, nutritional trajectories in Manila and non-Manila NCR cities would have evolved in parallel.
-This assumption is supported by visual inspection and pre-trend placebo tests using the 1993 and 1998 waves.
-The 1998 DHS serves as the primary pre-policy baseline, while the 2003 survey captures short-run effects.
-The 2008 and 2013 waves are used in robustness checks to assess medium- and long-run persistence of nutritional deficits.
-All specifications include city and birth-year fixed effects, and standard errors are clustered at the city-year level to account for within-cluster correlation.
+\subsection{Identification Strategy}
 
-The empirical strategy exploits the 2000 issuance of Executive Order No.
-003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City.
-This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR).
-Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group.
-The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.
+To estimate the causal effect of increased fertility on early-childhood health outcomes, I exploit a policy shock generated by Executive Order No.
+003, issued in 2000 by the mayor of Manila.
+This order restricted access to modern contraceptives in all city-run health facilities and remained in effect for nearly a decade.
+The policy was implemented only in the City of Manila and did not affect neighboring municipalities within the National Capital Region (NCR), which continued to provide modern family planning services.
+This spatially localized and time-bound restriction created a natural experiment that allows for a quasi-experimental research design.
+I define the treatment group as children residing in GPS clusters located within or near the Manila boundary, and the control group as children in other NCR municipalities.
+Given the limitations of the DHS GPS data, which are displaced for confidentiality and do not include exact city identifiers, I use a coarser geographic classification to approximate exposure.
+Robustness checks vary the control group to include Quezon City only, or the rest of NCR excluding both Manila and Quezon City.
 
-I estimate the following difference-in-differences specification:
+The core of my empirical strategy is a difference-in-differences (DiD) design that compares changes in child outcomes in Manila before and after the ban relative to changes in similar urban areas that were unaffected.
+The key assumption is that, in the absence of the policy, trends in early-childhood health outcomes would have followed similar trajectories across treatment and control areas.
+This approach accounts for both fixed differences between cities and nationwide time trends.
 
-$$ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} $$
+To address potential violations of the parallel trends assumption—such as Manila-specific shocks that may have affected all children regardless of family size—I implement a triple-differences strategy that leverages variation in exposure by maternal age.
+Because fertility declines sharply with age, older women were biologically less likely to respond to the contraceptive ban.
+For example, mothers aged 45–49 at the time of the policy were near the end of their reproductive years and thus unlikely to experience a fertility increase.
+By comparing changes in child outcomes across maternal age cohorts within Manila and control cities, I can account for time-varying, location-specific confounders that affect all families, while isolating the policy's effect on family size and its consequences.
 
-In this equation, $\text{Nutrition}_{ict}$ refers to either height-for-age (HAZ) or weight-for-height (WHZ) for child $i$ in city $c$ born in year $t$.
-The indicator $\text{Post}_t$ equals 1 if the child was born after the implementation of Executive Order No.
-003 in 2000, and $\text{Manila}c$ equals 1 if the child resides in Manila City.
-The interaction term captures the effect of being born in Manila after the contraceptive policy was enacted.
-The vector $X{ict}$ includes controls for child age in months, sex, birth order, maternal age, education, parity, body mass index (BMI), anemia status, household wealth quintile, household size, electricity access, water source, toilet type, shared sanitation status, and urban residence.
-City fixed effects $\lambda_c$ control for time-invariant characteristics at the locality level, while year fixed effects $\tau_t$ absorb aggregate time shocks.
-The coefficient of interest, $\beta$, represents the average treatment effect of the policy-induced fertility shock on child nutrition.
-Standard errors are clustered at the city-year level to account for potential autocorrelation within geographic units over time.
+I further restrict my main analysis to children born before the policy was implemented.
+Although these children were not conceived due to the ban, they may still have been indirectly affected by the fertility increase through resource dilution within households.
+This strategy avoids selection bias related to unobservable differences in the endowments of “marginal children” born because of the policy.
+As argued by Rosenzweig and Zhang (2009), when parents reinforce rather than compensate for birth endowments, estimates based on older siblings provide a lower bound of the true effect of fertility on child quality.
+Empirical evidence from the Philippines supports this pattern of reinforcement, as shown by Baguet and Dumas (2015).
+Taken together, these strategies allow me to estimate the causal impact of fertility shocks on early-life health while addressing key concerns about internal validity and sample composition.
 
-To support the internal validity of the design, I use Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude and timing of the fertility shock.
-The 1995 and 2007 waves provide clean pre- and post-policy benchmarks, while the 2000 census—although temporally proximate to the policy—is retained for robustness checks that test the sensitivity of estimates to the exact onset of fertility changes.
-A placebo comparison between 1990 and 1995 shows no divergence in fertility trends between Manila and control cities prior to the policy, strengthening the credibility of the parallel trends assumption.
+\subsection{Sample Selection}
 
-To supplement the DHS and Census data, I incorporate the Annual Poverty Indicators Survey (APIS), a nationally representative household survey that captures detailed information on food expenditure, maternal education, and living standards.
-Although the APIS lacks anthropometric data and has more limited geographic coverage, it enables exploratory analyses of treatment effect heterogeneity by income and education strata and helps triangulate changes in maternal resource constraints following the policy shock.
+The analytic sample is drawn from the Philippine Demographic and Health Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003, and 2008.
+I restrict the sample to children under age five (0–59 months) with non-missing information on key outcomes: birth weight (in grams, when available), perceived size at birth, and vaccination status.
+Because the Manila contraceptive ban was enacted in 2000 and likely implemented unevenly at first, I define children born in 2001 and later as post-policy and those born before 1997 as pre-policy.
+Children born between 1997–2000 are excluded in the main specification but retained in robustness checks.
 
-Finally, missing data are handled following standard practices in applied microeconometrics.
-Observations with missing or flagged anthropometric outcomes are excluded from the analysis.
-For other covariates with partial nonresponse—such as maternal BMI, anemia, or household infrastructure—I adopt the missing-indicator method: missing values are imputed with a constant, and a binary indicator is included in all regression models to control for the fact of missingness.
-This approach preserves sample size while allowing the data to inform whether nonresponse is systematically related to the outcomes of interest.
-As robustness checks, I estimate alternative specifications using multiple imputation and inverse-probability weighting to test the sensitivity of results to different missing data assumptions.
-All descriptive and inferential statistics are computed using DHS-provided sampling weights, and models account for the complex survey design by clustering standard errors at the primary sampling unit.
+I include all live births regardless of birth order or plurality to preserve variation in sibship size.
+Twins and higher-order multiple births are flagged for sensitivity checks but not excluded.
+Each child is linked to their biological mother using maternal identifiers, allowing construction of sibship size, birth order, and maternal age.
+Exposure to the policy is defined geographically: children are considered treated if they were born in the City of Manila, and untreated if born in other municipalities within the National Capital Region (NCR).
+Because GPS data are coarse in some years, geographic exposure is defined at the city level rather than using precise coordinates.
+Households outside NCR are excluded from the main analysis to ensure urban comparability and avoid confounding by rural fertility and health dynamics.
+
+\subsection{Variables}
+
+The primary outcome variables reflect early-life health and investment.
+These include: (1) birth weight in grams, as recorded by the mother when available; (2) the mother’s subjective assessment of birth size (categorized as very small, small, average, large, or very large); and (3) indicators for whether the child received core early-life vaccinations, including BCG, DPT1–3, polio doses, and measles.
+A binary variable indicating whether the child received all recommended vaccines is also constructed as a summary measure of postnatal health investment.
+
+The main explanatory variable is sibship size, defined as the total number of surviving children ever born to the mother at the time of the child’s birth, as reported in the birth history module.
+This variable captures the resource dilution mechanism central to the quantity–quality trade-off.
+To assess whether the contraceptive ban caused an increase in sibship size, I also construct an indicator for policy exposure, coded as 1 if the child was born in the City of Manila and in or after 2001 (i.e., post-ban), and 0 otherwise.
+In extended specifications, I interact this treatment indicator with maternal age at birth to capture heterogeneity in fertility effects across age cohorts.
+
+Control variables include standard child-, mother-, and household-level characteristics.
+At the child level, I include sex, birth order, whether the child is part of a multiple birth, and the preceding birth interval in months.
+At the maternal level, I control for the mother’s age at birth, educational attainment (categorized as none, primary, secondary, higher), literacy status, employment status, and age at first birth.
+Household-level controls include household size, urban/rural status (restricted to NCR in the main sample), access to improved water and sanitation, electricity, household crowding, and the DHS wealth index.
+Fixed effects for survey year are included to account for common shocks and secular trends.
+All regressions are weighted using DHS sampling weights, and standard errors are clustered at the PSU (primary sampling unit) level.
+
+\subsection{Estimation}
+
+To estimate the causal effect of fertility on early-life health outcomes, I use a two-stage least squares (2SLS) strategy.
+Specifically, I regress child health outcomes, such as birth weight, perceived size at birth, and immunization uptake, on measures of sibship size, using exogenous variation induced by the contraceptive supply ban in Manila as an instrument.
+This approach allows me to isolate the impact of increased fertility on child quality and mitigate bias from endogenous family size decisions.
+
+The estimation proceeds as follows:
+
+$$
+\text{First stage:} \quad K_{ijt} = \alpha_2 X_{ijt} + \gamma Z_{ijt} + v_{ijt}
+$$
+
+$$
+\text{Second stage:} \quad Y_{ijt} = \alpha_1 X_{ijt} + \beta \widehat{K}_{ijt} + u_{ijt}
+$$
+
+where: \( Y_{ijt} \) is the health-related outcome for child *i* in city *j* and survey year *t*, including birth weight (grams), perceived size at birth, and full basic immunization status.
+\( K_{ijt} \) is sibship size at time of birth, defined as the number of surviving siblings.
+\( Z_{ijt} \) is the instrument for sibship size, based on exposure to the contraception ban—specifically, an interaction of a Manila indicator and a post-ban cohort indicator.
+\( X_{ijt} \) is a vector of control variables including child-, maternal-, and household-level covariates (see *Variables* section).
+Standard errors are clustered at the PSU level; survey-year fixed effects are included.
+
+To account for heterogeneity in fertility responses to the ban, I also estimate an alternative specification where the instrument interacts treatment status (Manila × post-ban) with maternal age at birth.
+This reflects the fact that older women were less likely to change fertility behavior due to lower fecundity, and strengthens identification through cross-cohort variation in exposure intensity.
+
+Although the outcomes are binary or ordinal, I estimate equation (1) using linear probability models to retain consistency with the 2SLS framework.
+As robustness, I also estimate probit and logit models and find qualitatively similar results.
+All models use DHS sampling weights, and standard errors are clustered by primary sampling unit (PSU).
+
+\section{Results}
 

### 2025-07-27 docs/paper.tex
revise theory and empirical sections in paper draft
Expanded and clarified the literature review, conceptual framework, and empirical strategy in the paper. Standardized terminology from 'quantity–quality' to 'quantity-quality', added new references and recent evidence, and restructured sections for improved flow. Updated title and subtitle for consistency across Rmd, tex, and PDF outputs.

@@
 \title{Womb for One More? Fertility Expansion and the Limits of Early
-Health Investment}
+Health Investment: Empirical Evidence on the Quantity-Quality Trade-off
+from a Contraceptive Ban in Urban Philippines}
 \shortTitle{The Q-Q Trade-off in the Philippine Context}
@@
 
-The quantity--quality (Q--Q) theory, a central idea in modern family
+The quantity-quality (Q--Q) theory, a central idea in modern family
 economics, holds that parents face a trade-off between the number of
@@
 Becker and Lewis (1973) formalize the mutually reinforcing nature of the
-quantity--quality cost structure. An increase in \(n\) raises the total
+quantity-quality cost structure. An increase in \(n\) raises the total
 cost required to sustain a given level of \(q\) for each child, while a
@@
 capital accumulation, income distribution, and policy. This laid the
-foundation for unified growth theories, which view the quantity--quality
+foundation for unified growth theories, which view the quantity-quality
 mechanism as central to demographic transition and long-run development.
@@
 drop in child mortality, and a shift in parental priorities. These
-models explain not only the presence of a quantity--quality trade-off
-but also its rising influence at a specific point in history. The
-evidence supports these claims: countries that saw earlier increases in
-returns to education experienced earlier fertility decline, while delays
-in reforms, such as public education or health access, corresponded to
+models explain not only the presence of a quantity-quality trade-off but
+also its rising influence at a specific point in history. The evidence
+supports these claims: countries that saw earlier increases in returns
+to education experienced earlier fertility decline, while delays in
+reforms, such as public education or health access, corresponded to
 prolonged high fertility. As a result, the Q--Q mechanism forms a key
@@
 
-Contemporary research has further refined the quantity--quality model by
+Contemporary research has further refined the quantity-quality model by
 relaxing some of its initial simplifying assumptions. Three important
@@
 To begin, I examine how credit constraints can give rise to poverty
-traps. The canonical quantity--quality (Q--Q) framework assumes that
+traps. The canonical quantity-quality (Q--Q) framework assumes that
 parents can reallocate resources freely; borrowing against future
@@
 
-Formal models show that ``the quality--quantity trade-off arises from a
+Formal models show that ``the quality-quantity trade-off arises from a
 binding credit constraint that prevents parents from borrowing against
@@
 
-A further refinement of the quantity--quality (Q--Q) model considers the
+A further refinement of the quantity-quality (Q--Q) model considers the
 question of \emph{who} within the household makes fertility and child
@@
 investment---cognitive development, health, nutrition, etc. This is
-particularly salient in developing countries, where basic health and
-nutrition are pressing concerns alongside schooling.
+salient in developing countries, where basic health and nutrition are
+pressing concerns alongside schooling.
 
@@
 
-Beyond survival, nutritional status is a critical indicator of quality
-in many low-income settings. Parents often face choices about how to
-allocate food or medical care, especially under resource scarcity.
-Empirical evidence supports the Q--Q model in this context: children
-with many siblings often exhibit poorer nutritional outcomes. A recent
-study using Vietnam Young Lives data finds that an additional sibling
-reduces height-for-age and weight-for-age z-scores by about 0.3 standard
-deviations on average \citep{chen2021population} This effect is
-substantial, suggesting that children in larger families tend to be
-shorter and lighter than their peers from smaller families, likely due
-to the more limited distribution of healthcare and parental attention
+Beyond survival, early-life health indicators such as birth weight,
+perceived size at birth, and vaccination status are important dimensions
+of child quality, especially in low-income settings. When resources are
+limited, parents may struggle to meet the basic needs of each child.
+Empirical evidence aligns with the quantity--quality framework: children
+from larger families often have worse health outcomes at birth and are
+less likely to receive core immunizations. For example, recent
+multi-country data show that the share of completely unimmunized
+children---so-called ``zero-dose'' children---increases from 10.5\%
+among those with no siblings to 17.2\% among those with four or more
+\citep{costa2024child}. Studies from Southeast Asia also report that
+higher birth order correlates with lower average birth weight and
+smaller perceived size at birth. These patterns suggest that as
+fertility increases, the health of each child may suffer due to more
+limited parental investment. Studies from Southeast Asia
+\citep{chen2021population} also report that higher birth order
+correlates with lower average birth weight and smaller perceived size at
+birth
 
 Similar patterns emerge in sub-Saharan Africa
-\citep{bishwakarma2019first} and South Asia , where first-born children
+\citep{bishwakarma2019first} and South Asia, where first-born children
 experience slower growth once younger siblings arrive---again pointing
@@
 investments among all children. This early work indicated a
-quantity--quality (Q--Q) trade-off in Philippine families, though it
+quantity-quality (Q--Q) trade-off in Philippine families, though it
 largely documented correlations rather than definitive causation.
@@
 increases in family size cause significant declines in child educational
-attainment and validates the quantity--quality trade-off in this
-context. Large family size thus emerges as one mechanism contributing to
-poverty, by impeding children's human capital accumulation in the
-Philippines.
+attainment and validates the quantity-quality trade-off in this context.
+Large family size thus emerges as one mechanism contributing to poverty,
+by impeding children's human capital accumulation in the Philippines.
 
@@
 The Philippine experience is echoed in other ASEAN countries, where
-researchers have probed the quantity--quality trade-off with diverse
+researchers have probed the quantity-quality trade-off with diverse
 outcomes and methods. In Vietnam, for example, rapid fertility decline
@@
 Nonetheless, the prevailing evidence from the Philippines and its
-regional neighbors strongly supports Becker's quantity--quality
+regional neighbors strongly supports Becker's quantity-quality
 conjecture. As families have fewer children, they appear to invest more
@@
 planning programs to education subsidies---that aim to ease the
-quantity--quality trade-off and help families achieve both manageable
+quantity-quality trade-off and help families achieve both manageable
 size and better outcomes for the next generation.
@@
 This study builds upon the foundational framework of the
-quantity--quality (Q--Q) trade-off, as articulated in
-\citet{becker1960economic} seminal work and later refined by
-\citet{becker1973interaction}. Within this theoretical model, fertility
-and child quality are jointly determined under a full-income constraint,
-such that any increase in the number of children---absent a commensurate
-expansion in resources---necessitates a reduction in the average
-investment per child. While prior empirical applications of this
-framework have largely centered on educational attainment, child quality
-may also be understood along biological dimensions, especially along
-nutritional status, which emerges earlier in the life course and may be
-less subject to later remediation.
-
-The current analysis focuses exclusively on early-life nutrition as a
-proxy for child quality. Nutrition, unlike cognitive or educational
-outcomes, is characterized by low substitutability and a narrow temporal
-window during which investments are most productive. The trade-off is
-thus expected to manifest strongly in settings where fertility rises
-exogenously while household resources remain constrained. In such
-environments, families facing an unexpected increase in sibship size may
-be forced to reallocate scarce food, time, and health inputs across a
-larger number of dependents, potentially resulting in lower nutritional
-attainment for each child.
+quantity-quality (Q--Q) trade-off, first articulated in
+\citet{becker1960economic} and later refined in
+\citet{becker1973interaction}. In this model, fertility and child
+quality are jointly determined under a full-income constraint, such that
+increases in the number of children---without a corresponding expansion
+in household resources---necessitate a reduction in the average
+investment per child. While empirical applications of this framework
+have traditionally focused on educational outcomes, child quality can
+also be expressed in terms of early-life health, including indicators
+that emerge at birth or shortly after, such as birth weight, perceived
+size at birth, and uptake of essential vaccinations.
+
+This analysis focuses specifically on early-life health as a proxy for
+child quality. These outcomes, unlike cognitive or schooling-related
+investments, are biologically constrained and occur within a narrow
+developmental window, during which parental inputs are least
+substitutable and most consequential. The trade-off is expected to be
+especially salient in settings where fertility increases exogenously
+while family resources remain fixed. In such contexts, a sudden increase
+in sibship size may compel households to spread food, healthcare, and
+caregiving attention across more dependents, which would potentially
+lead to diminished health investments and outcomes for each child.
 
@@
 average \(\kappa_j\) must fall unless offset by changes in parental
-behavior or external transfers. If nutritional investments per child
-decline, and if such inputs are critical for growth, observable deficits
-in height-for-age (HAZ) or weight-for-height (WHZ) may emerge. The first
-hypothesis emerging from this framework is that an exogenous increase in
-
-fertility leads to a reduction in per-child nutritional investment,
-particularly in low-income, urban households where slack in the family
-budget is limited. A second, related hypothesis posits that this
-reduction in inputs translates into measurable nutritional deficits
-among young children, specifically in the form of lower anthropometric
-outcomes such as height-for-age (HAZ) and weight-for-height (WHZ)
-z-scores. These two hypotheses---concerning, respectively, the
-behavioral reallocation of resources and its biological
-consequence---jointly seek to identify a causal link between fertility
-shocks and early-life deprivation.
+behavior or external transfers. If these inputs decline and are critical
+to early-life development, observable deficits may arise in birth weight
+or perceived neonatal size, and children may be less likely to receive
+complete immunization schedules.
+
+The first hypothesis derived from this framework is that an exogenous
+increase in fertility leads to a decline in per-child health investment,
+particularly in low-income urban households with little financial slack.
+A second, related hypothesis posits that this decline results in
+measurable early-life health disadvantages---specifically, lower birth
+weight, smaller perceived size at birth, and reduced uptake of key
+childhood vaccinations. These hypotheses describe a causal pathway
+through which fertility shocks may contribute to early deprivation and
+potentially reinforce intergenerational cycles of poor health and
+limited human capital.
 
@@
 
-This study constructs a harmonized, child-level analytic dataset using
-five rounds of the Philippine Demographic and Health Survey (DHS): 1993,
-1998, 2003, 2008, and 2013. The DHS is a nationally representative,
-cross-sectional survey series conducted by the Philippine Statistics
-Authority in collaboration with ICF International. Each round follows a
-stratified two-stage cluster design and collects standardized
-information on child health, fertility, maternal characteristics, and
-household infrastructure. For each wave, I merge the child-level (KR),
-mother-level (IR), and household-level (HR) files using cluster,
-household, and maternal identifiers, ensuring that each observation
-corresponds to a live birth occurring within five years of the survey.
-The analytic sample is restricted to children aged 0--59 months to
-standardize developmental exposure windows and conform to the World
-Health Organization (WHO) reference growth standards.
-
-The primary outcomes are height-for-age (HAZ) and weight-for-height
-(WHZ) z-scores, internationally recognized proxies for chronic and acute
-malnutrition, respectively. These anthropometric indicators are cleaned
-following DHS protocols: biologically implausible values are excluded,
-and remaining z-scores are rescaled from hundredths to continuous units.
-In addition to anthropometry, the child record includes a suite of
-health-related variables. These include indicators for whether the child
-has received basic vaccinations---such as BCG, DPT, polio, and
-measles---as well as whether the child experienced episodes of diarrhea,
-fever, or respiratory illness in the two weeks prior to the survey.
-These supplementary health measures serve as intermediate outcomes in
-secondary analyses to capture short-run morbidity and preventive health
-investment.
-
-Maternal-level information is drawn from the individual recode file.
-Core covariates include maternal age, completed years of education, and
-total number of children ever born (parity), each of which captures
-socioeconomic status and fertility history. I additionally include
-maternal body mass index (BMI) and anemia status, which serve as both
-controls and potential mediators of child health. These indicators
-reflect the mother's nutritional status and physiological capacity to
-support healthy fetal and postnatal development. In robustness checks, I
-examine whether declines in maternal nutritional status following the
-fertility shock account for observed changes in child anthropometry.
-
-Household-level characteristics are used to control for environmental
-and infrastructural determinants of health. Specifically, I include
-household size, urban residence, wealth quintile, access to electricity,
-type of toilet facility, whether the toilet is shared, and source of
-drinking water. These variables proxy for household-level resources and
-exposure to disease environments. In extension analyses, I interact
-treatment status with these environmental indicators to assess whether
-the nutritional consequences of the fertility shock are magnified under
-infrastructural deprivation.
-
-The treatment is defined by exposure to Executive Order No.~003, which
-was issued by the Manila City government in January 2000 and effectively
-suspended the public distribution of modern contraceptives. This policy
-change resulted in a localized increase in fertility that was not
-mirrored in surrounding municipalities within the National Capital
-Region (NCR). Children born in Manila City after the implementation of
-the policy are classified as treated, while children in other NCR cities
-serve as the control group. This spatially and temporally bounded
-intervention provides quasi-experimental variation that facilitates
-identification of causal effects.
-
-The empirical approach adopts a difference-in-differences framework that
-compares trends in child nutritional outcomes between treated and
-control cities before and after the policy change. The identifying
-assumption is that, absent the policy, nutritional trajectories in
-Manila and non-Manila NCR cities would have evolved in parallel. This
-assumption is supported by visual inspection and pre-trend placebo tests
-using the 1993 and 1998 waves. The 1998 DHS serves as the primary
-pre-policy baseline, while the 2003 survey captures short-run effects.
-The 2008 and 2013 waves are used in robustness checks to assess medium-
-and long-run persistence of nutritional deficits. All specifications
-include city and birth-year fixed effects, and standard errors are
-clustered at the city-year level to account for within-cluster
-correlation.
-
-The empirical strategy exploits the 2000 issuance of Executive Order
-No.~003, which effectively banned the distribution of modern
-contraceptives in public health centers across Manila City. This abrupt
-and localized withdrawal of contraceptive access resulted in a fertility
-increase concentrated within the city, but did not affect surrounding
-municipalities within the National Capital Region (NCR). Children born
-in Manila following the policy are considered exposed to a fertility
-shock, while children in other NCR cities over the same period serve as
-the control group. The analytical sample is restricted to children under
-five years old at the time of each survey round to ensure accurate
-measurement of early-life nutritional outcomes.
-
-I estimate the following difference-in-differences specification:
-
-\[ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} \]
-
-In this equation, \(\text{Nutrition}_{ict}\) refers to either
-height-for-age (HAZ) or weight-for-height (WHZ) for child \(i\) in city
-\(c\) born in year \(t\). The indicator \(\text{Post}_t\) equals 1 if
-the child was born after the implementation of Executive Order No.~003
-in 2000, and \(\text{Manila}c\) equals 1 if the child resides in Manila
-City. The interaction term captures the effect of being born in Manila
-after the contraceptive policy was enacted. The vector \(X{ict}\)
-includes controls for child age in months, sex, birth order, maternal
-age, education, parity, body mass index (BMI), anemia status, household
-wealth quintile, household size, electricity access, water source,
-toilet type, shared sanitation status, and urban residence. City fixed
-effects \(\lambda_c\) control for time-invariant characteristics at the
-locality level, while year fixed effects \(\tau_t\) absorb aggregate
-time shocks. The coefficient of interest, \(\beta\), represents the
-average treatment effect of the policy-induced fertility shock on child
-nutrition. Standard errors are clustered at the city-year level to
-account for potential autocorrelation within geographic units over time.
-
-To support the internal validity of the design, I use Philippine Census
-microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude
-and timing of the fertility shock. The 1995 and 2007 waves provide clean
-pre- and post-policy benchmarks, while the 2000 census---although
-temporally proximate to the policy---is retained for robustness checks
-that test the sensitivity of estimates to the exact onset of fertility
-changes. A placebo comparison between 1990 and 1995 shows no divergence
-in fertility trends between Manila and control cities prior to the
-policy, strengthening the credibility of the parallel trends assumption.
-
-To supplement the DHS and Census data, I incorporate the Annual Poverty
-Indicators Survey (APIS), a nationally representative household survey
-that captures detailed information on food expenditure, maternal
-education, and living standards. Although the APIS lacks anthropometric
-data and has more limited geographic coverage, it enables exploratory
-analyses of treatment effect heterogeneity by income and education
-strata and helps triangulate changes in maternal resource constraints
-following the policy shock.
-
-Finally, missing data are handled following standard practices in
-applied microeconometrics. Observations with missing or flagged
-anthropometric outcomes are excluded from the analysis. For other
-covariates with partial nonresponse---such as maternal BMI, anemia, or
-household infrastructure---I adopt the missing-indicator method: missing
-values are imputed with a constant, and a binary indicator is included
-in all regression models to control for the fact of missingness. This
-approach preserves sample size while allowing the data to inform whether
-nonresponse is systematically related to the outcomes of interest. As
-robustness checks, I estimate alternative specifications using multiple
-imputation and inverse-probability weighting to test the sensitivity of
-results to different missing data assumptions. All descriptive and
-inferential statistics are computed using DHS-provided sampling weights,
-and models account for the complex survey design by clustering standard
-errors at the primary sampling unit.
+\subsection{Data}
+
+The primary data source for this study is the Philippine Demographic and
+Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
+These nationally representative surveys provide detailed child- and
+household-level information, including maternal characteristics,
+fertility behavior, and early-childhood health outcomes. Importantly,
+they are the only Philippine DHS waves with available GPS cluster data,
+which I use to approximate geographic location and assign respondents to
+treatment and control groups. Because city-level identifiers are not
+released in the public-use files, I rely on coarse geographic assignment
+based on cluster coordinates. Clusters located within or near the
+administrative boundaries of the City of Manila are classified as
+``treated,'' while clusters located in other parts of the National
+Capital Region (NCR) serve as the control group. In robustness checks, I
+restrict the control group to Quezon City or expand it to the rest of
+NCR excluding both Manila and Quezon City.
+
+The analytic sample includes children under five years of age living
+with their mothers at the time of the survey. I focus on three
+early-life health outcomes that reflect both biological endowments and
+parental investment: (1) birth weight, reported in grams when available;
+(2) the mother's subjective assessment of birth size; and (3)
+vaccination uptake, defined as receipt of key early-childhood
+immunizations, including BCG, DPT, polio, and measles. These outcomes
+are consistently available across all three waves and serve as proxies
+for child quality in the context of the quantity--quality trade-off.
+
+The 1998 DHS is used as the primary pre-policy baseline wave because it
+captures children born before the implementation of Manila's
+contraceptive ban in 2000. The 2008 DHS serves as the post-policy round
+to capture outcomes for children born well after the ban took effect. I
+also include the 2003 DHS in the main specification, recognizing that
+children in this round may have been born shortly before or after the
+policy's introduction. I treat 2003 as a transition-period dataset and
+conduct sensitivity checks to assess the robustness of findings to its
+inclusion. I exclude the 1993 DHS from the analysis due to the absence
+of GPS data, which makes it impossible to reliably assign treatment
+status.
+
+To support the identification strategy, I use DHS summary indicators on
+contraceptive use disaggregated by region and city. These data confirm a
+sharp and sustained decline in modern contraceptive use in Manila
+following the implementation of Executive Order No.~003, relative to
+surrounding cities, validating the policy as a plausibly exogenous
+fertility shock. I also considered using data from the Philippine
+National Nutrition Survey (NNS), which includes rich anthropometric and
+dietary data. However, the NNS microdata are only available from 2003
+onward, which limits their value for pre-policy analysis. As such, the
+DHS remains the most suitable dataset for this study.
+
+\subsection{Identification Strategy}
+
+To estimate the causal effect of increased fertility on early-childhood
+health outcomes, I exploit a policy shock generated by Executive Order
+No.~003, issued in 2000 by the mayor of Manila. This order restricted
+access to modern contraceptives in all city-run health facilities and
+remained in effect for nearly a decade. The policy was implemented only
+in the City of Manila and did not affect neighboring municipalities
+within the National Capital Region (NCR), which continued to provide
+modern family planning services. This spatially localized and time-bound
+restriction created a natural experiment that allows for a
+quasi-experimental research design. I define the treatment group as
+children residing in GPS clusters located within or near the Manila
+boundary, and the control group as children in other NCR municipalities.
+Given the limitations of the DHS GPS data, which are displaced for
+confidentiality and do not include exact city identifiers, I use a
+coarser geographic classification to approximate exposure. Robustness
+checks vary the control group to include Quezon City only, or the rest
+of NCR excluding both Manila and Quezon City.
+
+The core of my empirical strategy is a difference-in-differences (DiD)
+design that compares changes in child outcomes in Manila before and
+after the ban relative to changes in similar urban areas that were
+unaffected. The key assumption is that, in the absence of the policy,
+trends in early-childhood health outcomes would have followed similar
+trajectories across treatment and control areas. This approach accounts
+for both fixed differences between cities and nationwide time trends.
+
+To address potential violations of the parallel trends assumption---such
+as Manila-specific shocks that may have affected all children regardless
+of family size---I implement a triple-differences strategy that
+leverages variation in exposure by maternal age. Because fertility
+declines sharply with age, older women were biologically less likely to
+respond to the contraceptive ban. For example, mothers aged 45--49 at
+the time of the policy were near the end of their reproductive years and
+thus unlikely to experience a fertility increase. By comparing changes
+in child outcomes across maternal age cohorts within Manila and control
+cities, I can account for time-varying, location-specific confounders
+that affect all families, while isolating the policy's effect on family
+size and its consequences.
+
+I further restrict my main analysis to children born before the policy
+was implemented. Although these children were not conceived due to the
+ban, they may still have been indirectly affected by the fertility
+increase through resource dilution within households. This strategy
+avoids selection bias related to unobservable differences in the
+endowments of ``marginal children'' born because of the policy. As
+argued by Rosenzweig and Zhang (2009), when parents reinforce rather
+than compensate for birth endowments, estimates based on older siblings
+provide a lower bound of the true effect of fertility on child quality.
+Empirical evidence from the Philippines supports this pattern of
+reinforcement, as shown by Baguet and Dumas (2015). Taken together,
+these strategies allow me to estimate the causal impact of fertility
+shocks on early-life health while addressing key concerns about internal
+validity and sample composition.
+
+\subsection{Sample Selection}
+
+The analytic sample is drawn from the Philippine Demographic and Health
+Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003,
+and 2008. I restrict the sample to children under age five (0--59
+months) with non-missing information on key outcomes: birth weight (in
+grams, when available), perceived size at birth, and vaccination status.
+Because the Manila contraceptive ban was enacted in 2000 and likely
+implemented unevenly at first, I define children born in 2001 and later
+as post-policy and those born before 1997 as pre-policy. Children born
+between 1997--2000 are excluded in the main specification but retained
+in robustness checks.
+
+I include all live births regardless of birth order or plurality to
+preserve variation in sibship size. Twins and higher-order multiple
+births are flagged for sensitivity checks but not excluded. Each child
+is linked to their biological mother using maternal identifiers,
+allowing construction of sibship size, birth order, and maternal age.
+Exposure to the policy is defined geographically: children are
+considered treated if they were born in the City of Manila, and
+untreated if born in other municipalities within the National Capital
+Region (NCR). Because GPS data are coarse in some years, geographic
+exposure is defined at the city level rather than using precise
+coordinates. Households outside NCR are excluded from the main analysis
+to ensure urban comparability and avoid confounding by rural fertility
+and health dynamics.
+
+\subsection{Variables}
+
+The primary outcome variables reflect early-life health and investment.
+These include: (1) birth weight in grams, as recorded by the mother when
+available; (2) the mother's subjective assessment of birth size
+(categorized as very small, small, average, large, or very large); and
+(3) indicators for whether the child received core early-life
+vaccinations, including BCG, DPT1--3, polio doses, and measles. A binary
+variable indicating whether the child received all recommended vaccines
+is also constructed as a summary measure of postnatal health investment.
+
+The main explanatory variable is sibship size, defined as the total
+number of surviving children ever born to the mother at the time of the
+child's birth, as reported in the birth history module. This variable
+captures the resource dilution mechanism central to the
+quantity--quality trade-off. To assess whether the contraceptive ban
+caused an increase in sibship size, I also construct an indicator for
+policy exposure, coded as 1 if the child was born in the City of Manila
+and in or after 2001 (i.e., post-ban), and 0 otherwise. In extended
+specifications, I interact this treatment indicator with maternal age at
+birth to capture heterogeneity in fertility effects across age cohorts.
+
+Control variables include standard child-, mother-, and household-level
+characteristics. At the child level, I include sex, birth order, whether
+the child is part of a multiple birth, and the preceding birth interval
+in months. At the maternal level, I control for the mother's age at
+birth, educational attainment (categorized as none, primary, secondary,
+higher), literacy status, employment status, and age at first birth.
+Household-level controls include household size, urban/rural status
+(restricted to NCR in the main sample), access to improved water and
+sanitation, electricity, household crowding, and the DHS wealth index.
+Fixed effects for survey year are included to account for common shocks
+and secular trends. All regressions are weighted using DHS sampling
+weights, and standard errors are clustered at the PSU (primary sampling
+unit) level.
+
+\subsection{Estimation}
+
+To estimate the causal effect of fertility on early-life health
+outcomes, I use a two-stage least squares (2SLS) strategy. Specifically,
+I regress child health outcomes, such as birth weight, perceived size at
+birth, and immunization uptake, on measures of sibship size, using
+exogenous variation induced by the contraceptive supply ban in Manila as
+an instrument. This approach allows me to isolate the impact of
+increased fertility on child quality and mitigate bias from endogenous
+family size decisions.
+
+The estimation proceeds as follows:
+
+\[
+\text{First stage:} \quad K_{ijt} = \alpha_2 X_{ijt} + \gamma Z_{ijt} + v_{ijt}
+\]
+
+\[
+\text{Second stage:} \quad Y_{ijt} = \alpha_1 X_{ijt} + \beta \widehat{K}_{ijt} + u_{ijt}
+\]
+
+where: \(Y_{ijt}\) is the health-related outcome for child \emph{i} in
+city \emph{j} and survey year \emph{t}, including birth weight (grams),
+perceived size at birth, and full basic immunization status. \(K_{ijt}\)
+is sibship size at time of birth, defined as the number of surviving
+siblings. \(Z_{ijt}\) is the instrument for sibship size, based on
+exposure to the contraception ban---specifically, an interaction of a
+Manila indicator and a post-ban cohort indicator. \(X_{ijt}\) is a
+vector of control variables including child-, maternal-, and
+household-level covariates (see \emph{Variables} section). Standard
+errors are clustered at the PSU level; survey-year fixed effects are
+included.
+
+To account for heterogeneity in fertility responses to the ban, I also
+estimate an alternative specification where the instrument interacts
+treatment status (Manila × post-ban) with maternal age at birth. This
+reflects the fact that older women were less likely to change fertility
+behavior due to lower fecundity, and strengthens identification through
+cross-cohort variation in exposure intensity.
+
+Although the outcomes are binary or ordinal, I estimate equation (1)
+using linear probability models to retain consistency with the 2SLS
+framework. As robustness, I also estimate probit and logit models and
+find qualitatively similar results. All models use DHS sampling weights,
+and standard errors are clustered by primary sampling unit (PSU).
+
+\section{Results}
 

### 2025-07-29 data-analysis/outputs/dhs_summary_birth_size.tex
add summary tables for birth size, weight, and vaccines
Added LaTeX tables summarizing perceived birth size, birth weight, and number of vaccines received by location and policy period. These tables provide descriptive statistics for further analysis and reporting.

\begin{table}[!h]
\centering
\caption{Perceived Birth Size by Location and Policy Period}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}m{3cm}cccccc}
\toprule
Variable & Manila, Pre-Policy & Manila, Post-Policy & Other NCR, Pre-Policy & Other NCR, Post-Policy & Non-NCR, Pre-Policy & Non-NCR, Post-Policy\\
\midrule
1 - Very small & 10 (4.8\%) & 1 (0.6\%) & 2 (1.6\%) & NA & 345 (4.7\%) & 116 (3.4\%)\\
2 - Smaller than average & 35 (16.8\%) & 39 (23.6\%) & 25 (20.3\%) & 34 (25.0\%) & 1256 (17.0\%) & 584 (17.2\%)\\
3 - Average size & 122 (58.7\%) & 92 (55.8\%) & 76 (61.8\%) & 84 (61.8\%) & 4397 (59.4\%) & 1990 (58.6\%)\\
4 - Larger than average & 28 (13.5\%) & 22 (13.3\%) & 18 (14.6\%) & 15 (11.0\%) & 970 (13.1\%) & 509 (15.0\%)\\
5 - Very large & 12 (5.8\%) & 11 (6.7\%) & 2 (1.6\%) & 3 (2.2\%) & 407 (5.5\%) & 184 (5.4\%)\\
\addlinespace
9 - Missing & 1 (0.5\%) & NA & NA & NA & 18 (0.2\%) & NA\\
8 - Don't know & NA & NA & NA & NA & 13 (0.2\%) & 14 (0.4\%)\\
\bottomrule
\end{tabular}
\end{table}

### 2025-07-29 data-analysis/outputs/dhs_summary_birth_size_t.tex
add summary tables for birth size, weight, and vaccines
Added LaTeX tables summarizing perceived birth size, birth weight, and number of vaccines received by location and policy period. These tables provide descriptive statistics for further analysis and reporting.

\begin{table}[!h]
\centering
\caption{Perceived Birth Size by Location and Policy Period}
\centering
\begin{tabular}[t]{lccccc}
\toprule
Group & \makecell{Very\\small} & \makecell{Smaller\\than\\average} & \makecell{Average\\size} & \makecell{Larger\\than\\average} & \makecell{Very\\large}\\
\midrule
Manila, Pre-Policy & \makecell{10 \\ (4.8\%)} & \makecell{35 \\ (16.9\%)} & \makecell{122 \\ (58.9\%)} & \makecell{28 \\ (13.5\%)} & \makecell{12 \\ (5.8\%)}\\
Manila, Post-Policy & \makecell{1 \\ (0.6\%)} & \makecell{39 \\ (23.6\%)} & \makecell{92 \\ (55.8\%)} & \makecell{22 \\ (13.3\%)} & \makecell{11 \\ (6.7\%)}\\
Other NCR Cities, Pre-Policy & \makecell{2 \\ (1.6\%)} & \makecell{25 \\ (20.3\%)} & \makecell{76 \\ (61.8\%)} & \makecell{18 \\ (14.6\%)} & \makecell{2 \\ (1.6\%)}\\
Other NCR Cities, Post-Policy & NA & \makecell{34 \\ (25.0\%)} & \makecell{84 \\ (61.8\%)} & \makecell{15 \\ (11.0\%)} & \makecell{3 \\ (2.2\%)}\\
Rest of the Philippines, Pre-Policy & \makecell{345 \\ (4.7\%)} & \makecell{1256 \\ (17.0\%)} & \makecell{4397 \\ (59.6\%)} & \makecell{970 \\ (13.2\%)} & \makecell{407 \\ (5.5\%)}\\
Rest of the Philippines, Post-Policy & \makecell{116 \\ (3.4\%)} & \makecell{584 \\ (17.3\%)} & \makecell{1990 \\ (58.8\%)} & \makecell{509 \\ (15.0\%)} & \makecell{184 \\ (5.4\%)}\\
\bottomrule
\end{tabular}
\end{table}

### 2025-07-29 data-analysis/outputs/dhs_summary_birth_weight_t.tex
add summary tables for birth size, weight, and vaccines
Added LaTeX tables summarizing perceived birth size, birth weight, and number of vaccines received by location and policy period. These tables provide descriptive statistics for further analysis and reporting.

\begin{table}[!h]
\centering
\caption{Birth Weight by Location and Policy Period}
\centering
\begin{tabular}[t]{l cl c}
\toprule
Group & Birth weight (g)\\
\midrule
Manila, Pre-Policy & 3023.12 (712.62), n = 156\\
Manila, Post-Policy & 2969.79 (619.23), n = 144\\
Other NCR Cities, Pre-Policy & 2830.41 (743.57), n = 70\\
Other NCR Cities, Post-Policy & 2965.89 (793.72), n = 88\\
Rest of the Philippines, Pre-Policy & 3106.41 (726.91), n = 4562\\
\addlinespace
Rest of the Philippines, Post-Policy & 2994.45 (681.71), n = 2298\\
\bottomrule
\end{tabular}
\end{table}

### 2025-07-29 data-analysis/outputs/dhs_summary_vaccines.tex
add summary tables for birth size, weight, and vaccines
Added LaTeX tables summarizing perceived birth size, birth weight, and number of vaccines received by location and policy period. These tables provide descriptive statistics for further analysis and reporting.

\begin{table}[!h]
\centering
\caption{Number of Vaccines Received by Location and Policy Period}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}m{3cm}cccccc}
\toprule
Variable & Manila, Pre-Policy & Manila, Post-Policy & Other NCR, Pre-Policy & Other NCR, Post-Policy & Non-NCR, Pre-Policy & Non-NCR, Post-Policy\\
\midrule
1–2 Vaccines & 1 (0.5\%) & 3 (1.8\%) & 1 (0.8\%) & 3 (2.2\%) & 68 (0.9\%) & 95 (2.8\%)\\
3–4 Vaccines & 4 (1.9\%) & 5 (3.0\%) & 3 (2.4\%) & 5 (3.7\%) & 136 (1.8\%) & 117 (3.4\%)\\
5+ Vaccines & 38 (18.2\%) & 33 (20.0\%) & 31 (25.2\%) & 52 (38.2\%) & 1869 (25.1\%) & 1141 (33.5\%)\\
No Vaccines & 166 (79.4\%) & 124 (75.2\%) & 88 (71.5\%) & 76 (55.9\%) & 5364 (72.1\%) & 2057 (60.3\%)\\
\bottomrule
\end{tabular}
\end{table}

### 2025-07-29 data-analysis/outputs/dhs_summary_vaccines_t.tex
add summary tables for birth size, weight, and vaccines
Added LaTeX tables summarizing perceived birth size, birth weight, and number of vaccines received by location and policy period. These tables provide descriptive statistics for further analysis and reporting.

\begin{table}[!h]
\centering
\caption{Number of Vaccines Received by Location and Policy Period}
\centering
\begin{tabular}[t]{lcccc}
\toprule
Group & 1–2 Vaccines & 3–4 Vaccines & 5+ Vaccines & No Vaccines\\
\midrule
Manila, Pre-Policy & 1 (0.5\%) & 4 (1.9\%) & 38 (18.2\%) & 166 (79.4\%)\\
Manila, Post-Policy & 3 (1.8\%) & 5 (3.0\%) & 33 (20.0\%) & 124 (75.2\%)\\
Other NCR Cities, Pre-Policy & 1 (0.8\%) & 3 (2.4\%) & 31 (25.2\%) & 88 (71.5\%)\\
Other NCR Cities, Post-Policy & 3 (2.2\%) & 5 (3.7\%) & 52 (38.2\%) & 76 (55.9\%)\\
Rest of the Philippines, Pre-Policy & 68 (0.9\%) & 136 (1.8\%) & 1869 (25.1\%) & 5364 (72.1\%)\\
\addlinespace
Rest of the Philippines, Post-Policy & 95 (2.8\%) & 117 (3.4\%) & 1141 (33.5\%) & 2057 (60.3\%)\\
\bottomrule
\end{tabular}
\end{table}

### 2025-08-07 docs/paper.Rmd
revise paper title, abstract, intro, and literature review
Updated the paper's title, short title, acknowledgements, and abstract to reflect a new focus on fertility shocks and intra-household investment in the Philippines. Substantially revised the introduction and literature review for clarity, conciseness, and to better align with the new research questions. Added new LaTeX header includes and made minor technical corrections throughout the RMarkdown file. Added a new preamble.tex file and updated references.

@@
 ---
-title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
-short: "The Q-Q Trade-off in the Philippine Context"
+title: "Who Gets Cared For and Why? Fertility Shocks, Intra-Household Investment Reallocation, and the Quantity–Quality Trade-Off in the Philippines"
+short: "Spillovers from Policy-Induced Fertility"
 journal: "" # AER, AEJ, PP, JEL
@@
 acknowledgements: |
-  This research was made possible by the Schupf Fellowship at Amherst College.
-I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project.
-I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work.
-I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support.
-The views expressed and any errors contained herein are entirely my own responsibility.
+  This research was made possible through the support of the Schupf Fellowship at Amherst College.
+I am deeply thankful to Dr.
+Axel Schupf ’57 for his generosity and vision in establishing this program, which annually supports a cohort of twenty rising sophomores and juniors in independent summer research.
+I feel truly privileged to have been selected through this competitive process and to have spent eight weeks immersed in a project that both challenged and inspired me.
+I am especially grateful to my mentor, Prof.
+Caroline Theoharides, for her mentorship throughout this project—and more broadly, for sparking my interest in economic research and its potential to serve the public good in the Philippines.
+I also wish to thank Faculty Director Professor Amelie Hastie and the many campus partners whose institutional support made this experience possible.
+Finally, I am grateful to the Departments of Economics and Mathematics & Statistics for equipping me with the analytical and conceptual tools that made this research possible—and for preparing me to take on independent research with confidence and curiosity.
+All views and any errors are my own.
 abstract: |
-  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households.
-I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No.
-003—which effectively banned the provision of modern contraceptives in all city-run health facilities.
-Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy.
-The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
+  This paper investigates how increases in fertility, driven by reduced access to contraception, affect parental investment in child health.
+I use Executive Order No.
+003—a 2000 policy in Manila that restricted access to modern contraceptives—as an exogenous fertility shock.
+Drawing on pooled data from the 1998, 2003, and 2008 Philippine Demographic and Health Surveys, I estimate two-stage least squares models that use policy exposure to instrument for sibship size.
+Child health inputs fall into two categories: prenatal (proxied by birth weight) and postnatal (measured by the total number of vaccine doses received by age five).
+The first-stage results indicate that affected mothers had, on average, 0.20 more living children at the time of a birth.
+In the second stage, each additional sibling is associated with a reduction of 22.97 vaccine doses received by the index child, a decline large enough to prevent completion of the recommended immunization schedule.
+The estimated effect on birth weight is −8,676 grams, though statistically insignificant Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight.
+While instrumental variable estimates are limited by a weak first stage, the magnitude and sign of the coefficients support the hypothesis that larger sibship sizes reduce health investments per child.
+The results suggest that postnatal investments, particularly those requiring sustained parental effort, are more sensitive to fertility shocks than prenatal outcomes.
 output: 
   rticles::aea_article:
+    latex_engine: xelatex
     keep_tex: true
-    latex_engine: pdflatex
+    citation_package: natbib
 bibliography: references.bib
+header-includes:
+  - \usepackage{pdfpages}
+  - \usepackage{float}
+  - \usepackage{makecell}
+  - \setlength{\parskip}{0pt}       % Remove vertical space between paragraphs
+  - \setlength{\parindent}{15pt}    % Indent first line of each paragraph
 ---
@@
 Such early-life health setbacks may permanently impair cognitive development and physical growth.
-This limits long-run productivity regardless of subsequent improvements in schooling or incom
+This limits long-run productivity regardless of subsequent improvements in schooling or income.
 
@@
 The analysis uses nationally representative data from multiple rounds of the Philippine Demographic and Health Survey (DHS), which include detailed information on household structure, maternal characteristics, and child health indicators.
-The primary outcomes are birth weight (measured in grams, when reported) and the mother’s subjective assessment of the child’s size at birth—both of which serve as indicators of neonatal health status.
-In addition, I assess whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment.
+The primary outcomes are birth weight (measured in grams, when reported) and whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment.
 These outcomes are widely recognized indicators of child well-being and reflect both biological and behavioral dimensions of parental investment in early development.
@@
 I probe these assumptions using falsification tests, placebo comparisons, and robustness checks that include city-specific time trends, mother fixed effects, and controls for baseline demographic differences.
-
 The analysis proceeds in three stages.
 I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women.
-Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, perceived birth size, and vaccination uptake).
+Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, and vaccination uptake).
 Finally, I examine heterogeneity in these effects across subgroups defined by maternal education, household wealth, and access to prenatal care.
@@
 I begin with the static models of Becker \citep{becker1960economic} and Becker–Lewis \citep{becker1973interaction}, which first formalized the trade-off within a household utility maximization framework.
-We then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}.
+I then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}.
 Next, I turn to macroeconomic and unified growth models, notably Galor and Weil \citep{galor2000population} and Galor and Moav \citep{galor2002natural}, which integrate the Q–Q mechanism into a general theory of demographic and economic transformation.
@@
 \frac{U_n}{U_q} = \frac{p_n + p_q q}{p_q n}.
-$$
-This condition equates the marginal rate of substitution between quantity and quality to the ratio of their full marginal costs.
+$$ This condition equates the marginal rate of substitution between quantity and quality to the ratio of their full marginal costs.
 The numerator rises with $q$, and the denominator rises with $n$.
@@
 A further refinement of the quantity-quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions.
-The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \).
+The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility $n$ and child quality $q$.
 In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines.
@@
 
-\[
+$$
 \alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
-\]
-
-with a corresponding condition for \( q \).
-Here, \( \alpha \in [0,1] \) represents the wife’s bargaining weight.
-When \( \alpha \) increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.
+$$
+
+with a corresponding condition for $q$.
+Here, $\alpha \in [0,1]$ represents the wife’s bargaining weight.
+When $\alpha$ increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.
 
@@
 The theoretical question is how these dimensions interact with the quantity decision.
-If parents allocate a budget across, say, schooling \( q_{\text{edu}} \) and nutrition/health \( q_{\text{health}} \) for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them).
+If parents allocate a budget across, say, schooling $q_{\text{edu}}$ and nutrition/health $q_{\text{health}}$ for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them).
 In some models, health and education are complementary: a healthier child benefits more from education, and an educated mother might raise a healthier child.
@@
 
-Beyond survival, early-life health indicators such as birth weight, perceived size at birth, and vaccination status are important dimensions of child quality, especially in low-income settings.
+Beyond survival, early-life health indicators such as birth weight and vaccination status are important dimensions of child quality, especially in low-income settings.
 When resources are limited, parents may struggle to meet the basic needs of each child.
@@
 For example, recent multi-country data show that the share of completely unimmunized children—so-called “zero-dose” children—increases from 10.5% among those with no siblings to 17.2% among those with four or more \citep{costa2024child}.
-Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth.
+Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight.
 These patterns suggest that as fertility increases, the health of each child may suffer due to more limited parental investment.
-Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth
+Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight.
 
 Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia, where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs.
-Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:
-
-\[
+Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health $H$ depends on food $F$ and medical care $M$, such that:
+
+$$
 H = g(F, M)
-\]
-
-and \( H \) enters either utility directly or the child’s future productivity.
-Parents then choose \( n \), \( F \), \( M \), and possibly schooling per child.
-Larger \( n \) reduces \( F \) and \( M \) per child, lowering \( H \); if parents place high value on \( H \), or if \( H \) enhances returns to education, they will opt for smaller families.
+$$
+
+and $H$ enters either utility directly or the child’s future productivity.
+Parents then choose $n$, $F$, $M$, and possibly schooling per child.
+Larger $n$ reduces $F$ and $M$ per child, lowering $H$; if parents place high value on $H$, or if $H$ enhances returns to education, they will opt for smaller families.
 
@@
 In the authors’ words, the policy-driven increase in family size “provide[s] evidence of a quality–quantity trade-off”: larger families, forced by the ban, resulted in lower schooling outcomes per child.
-By leveraging an actual anti-contraception policy, this study offers causal confirmation that constraints on family planning can worsen child quality.
-Together with Orbeta’s findings, it reinforces the relevance of Becker’s hypothesis in the Philippine setting: whether via unintended fertility (Manila’s case) or ordinary variation in sibling sex mix, more children mean fewer resources per child and thus worse education outcomes.
-Consistent with theory, Filipino parents facing exogenous increases in quantity were unable to maintain the same level of quality per child.
 
@@
 In this model, fertility and child quality are jointly determined under a full-income constraint, such that increases in the number of children—without a corresponding expansion in household resources—necessitate a reduction in the average investment per child.
-While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, perceived size at birth, and uptake of essential vaccinations.
+While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, and uptake of essential vaccinations.
 
@@
 
-To formalize this relationship, consider a household with total income \( I \), choosing nutritional investments \(\kappa_j\) for each of \( n \) children, along with consumption \( Z \) unrelated to child quality:
-
-\[
+To formalize this relationship, consider a household with total income $I$, choosing nutritional investments $\kappa_j$ for each of $n$ children, along with consumption $Z$ unrelated to child quality:
+
+$$
 I = \sum_{j=1}^{n} \kappa_j + Z.
-\]
-
-In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers.
-If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight or perceived neonatal size, and children may be less likely to receive complete immunization schedules.
+$$
+
+In the presence of a fertility shock, holding $I$ constant, the average $\kappa_j$ must fall unless offset by changes in parental behavior or external transfers.
+If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight, and children may be less likely to receive complete immunization schedules.
 
 The first hypothesis derived from this framework is that an exogenous increase in fertility leads to a decline in per-child health investment, particularly in low-income urban households with little financial slack.
-A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight, smaller perceived size at birth, and reduced uptake of key childhood vaccinations.
+A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight and reduced uptake of key childhood vaccinations.
 These hypotheses describe a causal pathway through which fertility shocks may contribute to early deprivation and potentially reinforce intergenerational cycles of poor health and limited human capital.
@@
 
-The primary data source for this study is the Philippine Demographic and Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
-These nationally representative surveys provide detailed child- and household-level information, including maternal characteristics, fertility behavior, and early-childhood health outcomes.
-Importantly, they are the only Philippine DHS waves with available GPS cluster data, which I use to approximate geographic location and assign respondents to treatment and control groups.
-Because city-level identifiers are not released in the public-use files, I rely on coarse geographic assignment based on cluster coordinates.
-Clusters located within or near the administrative boundaries of the City of Manila are classified as “treated,” while clusters located in other parts of the National Capital Region (NCR) serve as the control group.
-In robustness checks, I restrict the control group to Quezon City or expand it to the rest of NCR excluding both Manila and Quezon City.
-
-The analytic sample includes children under five years of age living with their mothers at the time of the survey.
-I focus on three early-life health outcomes that reflect both biological endowments and parental investment: (1) birth weight, reported in grams when available; (2) the mother’s subjective assessment of birth size; and (3) vaccination uptake, defined as receipt of key early-childhood immunizations, including BCG, DPT, polio, and measles.
-These outcomes are consistently available across all three waves and serve as proxies for child quality in the context of the quantity–quality trade-off.
-
-The 1998 DHS is used as the primary pre-policy baseline wave because it captures children born before the implementation of Manila’s contraceptive ban in 2000.
-The 2008 DHS serves as the post-policy round to capture outcomes for children born well after the ban took effect.
-I also include the 2003 DHS in the main specification, recognizing that children in this round may have been born shortly before or after the policy’s introduction.
-I treat 2003 as a transition-period dataset and conduct sensitivity checks to assess the robustness of findings to its inclusion.
-I exclude the 1993 DHS from the analysis due to the absence of GPS data, which makes it impossible to reliably assign treatment status.
-
-To support the identification strategy, I use DHS summary indicators on contraceptive use disaggregated by region and city.
-These data confirm a sharp and sustained decline in modern contraceptive use in Manila following the implementation of Executive Order No.
-003, relative to surrounding cities, validating the policy as a plausibly exogenous fertility shock.
-I also considered using data from the Philippine National Nutrition Survey (NNS), which includes rich anthropometric and dietary data.
-However, the NNS microdata are only available from 2003 onward, which limits their value for pre-policy analysis.
-As such, the DHS remains the most suitable dataset for this study.
-
-\subsection{Identification Strategy}
-
-To estimate the causal effect of increased fertility on early-childhood health outcomes, I exploit a policy shock generated by Executive Order No.
-003, issued in 2000 by the mayor of Manila.
-This order restricted access to modern contraceptives in all city-run health facilities and remained in effect for nearly a decade.
-The policy was implemented only in the City of Manila and did not affect neighboring municipalities within the National Capital Region (NCR), which continued to provide modern family planning services.
-This spatially localized and time-bound restriction created a natural experiment that allows for a quasi-experimental research design.
-I define the treatment group as children residing in GPS clusters located within or near the Manila boundary, and the control group as children in other NCR municipalities.
-Given the limitations of the DHS GPS data, which are displaced for confidentiality and do not include exact city identifiers, I use a coarser geographic classification to approximate exposure.
-Robustness checks vary the control group to include Quezon City only, or the rest of NCR excluding both Manila and Quezon City.
-
-The core of my empirical strategy is a difference-in-differences (DiD) design that compares changes in child outcomes in Manila before and after the ban relative to changes in similar urban areas that were unaffected.
-The key assumption is that, in the absence of the policy, trends in early-childhood health outcomes would have followed similar trajectories across treatment and control areas.
-This approach accounts for both fixed differences between cities and nationwide time trends.
-
-To address potential violations of the parallel trends assumption—such as Manila-specific shocks that may have affected all children regardless of family size—I implement a triple-differences strategy that leverages variation in exposure by maternal age.
-Because fertility declines sharply with age, older women were biologically less likely to respond to the contraceptive ban.
-For example, mothers aged 45–49 at the time of the policy were near the end of their reproductive years and thus unlikely to experience a fertility increase.
-By comparing changes in child outcomes across maternal age cohorts within Manila and control cities, I can account for time-varying, location-specific confounders that affect all families, while isolating the policy's effect on family size and its consequences.
-
-I further restrict my main analysis to children born before the policy was implemented.
-Although these children were not conceived due to the ban, they may still have been indirectly affected by the fertility increase through resource dilution within households.
-This strategy avoids selection bias related to unobservable differences in the endowments of “marginal children” born because of the policy.
-As argued by Rosenzweig and Zhang (2009), when parents reinforce rather than compensate for birth endowments, estimates based on older siblings provide a lower bound of the true effect of fertility on child quality.
-Empirical evidence from the Philippines supports this pattern of reinforcement, as shown by Baguet and Dumas (2015).
-Taken together, these strategies allow me to estimate the causal impact of fertility shocks on early-life health while addressing key concerns about internal validity and sample composition.
-
-\subsection{Sample Selection}
-
-The analytic sample is drawn from the Philippine Demographic and Health Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003, and 2008.
-I restrict the sample to children under age five (0–59 months) with non-missing information on key outcomes: birth weight (in grams, when available), perceived size at birth, and vaccination status.
-Because the Manila contraceptive ban was enacted in 2000 and likely implemented unevenly at first, I define children born in 2001 and later as post-policy and those born before 1997 as pre-policy.
-Children born between 1997–2000 are excluded in the main specification but retained in robustness checks.
-
-I include all live births regardless of birth order or plurality to preserve variation in sibship size.
-Twins and higher-order multiple births are flagged for sensitivity checks but not excluded.
-Each child is linked to their biological mother using maternal identifiers, allowing construction of sibship size, birth order, and maternal age.
-Exposure to the policy is defined geographically: children are considered treated if they were born in the City of Manila, and untreated if born in other municipalities within the National Capital Region (NCR).
-Because GPS data are coarse in some years, geographic exposure is defined at the city level rather than using precise coordinates.
-Households outside NCR are excluded from the main analysis to ensure urban comparability and avoid confounding by rural fertility and health dynamics.
+```{r}
+#| label: load-libraries
+#| include: false
+
+# Core Data Import and Manipulation 
+library(tidyverse)     # Includes ggplot2, dplyr, readr, etc.
+library(data.table)    # For fast data handling
+library(haven)         # For reading Stata (.dta) files
+library(janitor)       # For cleaning variable names and tables
+
+# Variable Labels and Factors 
+library(labelled)      # For working with labelled survey variables
+library(forcats)       # For factor manipulation
+
+# Table Rendering and Summarization 
+library(gtsummary)     # For descriptive summary tables
+library(gt)            # For rendering gtsummary tables
+library(kableExtra)    # For LaTeX/HTML table styling
+library(knitr)         # For knitting reports and inline tables
+library(broom)         # For tidying model output
+
+# Modeling and Estimation 
+library(fixest)        # For fixed effects and IV regressions
+library(modelsummary)  # For regression output formatting
+library(AER)           # For ivreg() — standard 2SLS instrumental variable estimation
+library(MASS)          # For polr()
+library(ordinal)       # For clm() – flexible ordered logit
+
+# Plotting and Utilities 
+library(purrr)         # For functional programming
+library(scales)        # For axis and color formatting in ggplot2
+library(extrafont)     # For font support in graphics
+
+```
+
+```{r}
+#| label: data-loading
+#| include: false
+
+# Load cleaned DHS child-level data with geographic identifiers ---
+dhs1 <- readRDS("../data-analysis/clean-data/dhs_kr_children_with_geo.rds")
+
+# Prepare cleaned variables and identifiers ---
+dhs2 <- dhs1 |>
+  filter(survey_year %in% c(1998, 2003, 2008)) |>
+  haven::zap_labels() |>
+  mutate(
+    # Standardize birth year to 4-digit format
+    birth_year = ifelse(birth_year < 100, 1900 + birth_year, birth_year),
+
+    # Clean and recode birth weight and perceived birth size
+    birth_weight_grams = na_if(birth_weight, birth_weight %in% c(9996, 9998, 9999)),
+    birth_size_clean = na_if(birth_size, birth_size %in% c(8, 9)),
+
+    birth_size_factor = case_when(
+      birth_size == 1 ~ "Very large",
+      birth_size == 2 ~ "Larger than average",
+      birth_size == 3 ~ "Average",
+      birth_size == 4 ~ "Smaller than average",
+      birth_size == 5 ~ "Very small",
+      TRUE ~ NA_character_
+    ) |> factor(levels = c(
+      "Very small", "Smaller than average", "Average",
+      "Larger than average", "Very large"
+    )),
+
+    # Flag NCR either via region code or admin name
+    ncr = region_dhs == 14 | tolower(admin_region_name) == "national capital region",
+
+    # Manually fix municipality names for known Manila GPS coordinates (2008 only)
+    municipality_name = case_when(
+      survey_year == 2008 &
+        latitude %in% c(14.6055, 14.56937, 14.56748, 14.56274, 14.59441) &
+        longitude %in% c(120.9485, 120.96649, 120.97882, 120.98432, 121.01443) ~ "Manila",
+      TRUE ~ municipality_name
+    ),
+
+    # Identify Manila municipalities
+    manila = grepl("manila", tolower(municipality_name)),
+
+    # Define additional flags (after correcting birth_year)
+    other_ncr = ncr & !manila,
+    post_policy = birth_year >= 2000,
+    manila_post = manila & post_policy
+  )
+
+# Filter for NCR cities only
+dhs2_ncr <- dhs2 |> filter(ncr)
+```
+
+This study draws on two primary data sources: the Philippine Demographic and Health Surveys (DHS) and the Integrated Public Use Microdata Series (IPUMS) census data.
+The DHS provides detailed child- and household-level information on fertility behavior, maternal characteristics, and early childhood health outcomes.
+I use three rounds of the DHS—1998, 2003, and 2008—which are the only Philippine DHS waves with publicly available GPS cluster data.
+These data are used to classify respondents into treatment and control groups.
+Because city-level identifiers are not included in the public-use DHS files, I rely on approximate geographic assignment using cluster coordinates.
+Clusters located within or near the administrative boundaries of the City of Manila are classified as treated, while clusters located elsewhere in the National Capital Region (NCR) serve as the control group.
+
+The DHS analytic sample is limited to children under five years of age who were living with their mothers at the time of the survey.
+I examine three core early-life health outcomes: (1) reported birth weight (in grams); (2) maternal perception of birth size; and (3) vaccination uptake, measured as the receipt of key early childhood immunizations (BCG, DPT, polio, and measles).
+These outcomes are consistently available across all three survey rounds and are commonly used as proxies for child quality in studies of the quantity–quality trade-off.
+
+To complement the DHS analysis and support the identification strategy, I construct a panel of individual-level microdata from the IPUMS harmonized Philippine censuses (1990, 2000, and 2010).
+These nationally representative surveys provide a snapshot of women of reproductive age (15–49) and contain detailed fertility histories—including total children ever born, births in the past year, and number of surviving children.
+I limit the sample to women ages 15–49 and construct a composite fertility index combining several measures: total children ever born, births in the past year, number of children under age five, and the age of the youngest child.
+By comparing Manila with other provinces and municipalities in NCR, I test whether long-run fertility patterns reflect the contraceptive policy shock.
+
+The 1998 DHS serves as the primary pre-policy baseline, capturing births prior to the issuance of Executive Order No.
+003.
+The 2008 DHS is the main post-policy round, while the 2003 DHS is treated as a transition period.
+I exclude the 1993 DHS due to the absence of geospatial identifiers.
+In contrast, the census years 1990 and 2000 bracket the policy’s onset, while the 2010 census provides post-policy fertility outcomes.
+This combination of sources enables both short-run and long-run assessments of the policy’s impact on fertility and child investment.
+
+To validate the policy shock, I use DHS region- and city-level contraceptive use indicators, which confirm a sharp and sustained decline in modern contraceptive use in Manila relative to other NCR cities.
+These patterns reinforce the plausibility of Executive Order No.
+003 as an exogenous shock to fertility preferences and access.
+I also considered the Philippine National Nutrition Survey (NNS) as an auxiliary source, but its limited pre-2003 coverage restricts its utility for identifying pre-policy trends.
+For this reason, the combined DHS and IPUMS datasets offer the most comprehensive and temporally balanced data available for the study's objectives.
 
@@
 The primary outcome variables reflect early-life health and investment.
-These include: (1) birth weight in grams, as recorded by the mother when available; (2) the mother’s subjective assessment of birth size (categorized as very small, small, average, large, or very large); and (3) indicators for whether the child received core early-life vaccinations, including BCG, DPT1–3, polio doses, and measles.
-A binary variable indicating whether the child received all recommended vaccines is also constructed as a summary measure of postnatal health investment.
+These include: (1) birth weight in grams, as recorded by the mother when available; and (2) indicators for whether the child received core early-life vaccinations, including BCG, DPT1–3, polio doses, and measles.
 
@@
 This variable captures the resource dilution mechanism central to the quantity–quality trade-off.
-To assess whether the contraceptive ban caused an increase in sibship size, I also construct an indicator for policy exposure, coded as 1 if the child was born in the City of Manila and in or after 2001 (i.e., post-ban), and 0 otherwise.
-In extended specifications, I interact this treatment indicator with maternal age at birth to capture heterogeneity in fertility effects across age cohorts.
-
-Control variables include standard child-, mother-, and household-level characteristics.
-At the child level, I include sex, birth order, whether the child is part of a multiple birth, and the preceding birth interval in months.
-At the maternal level, I control for the mother’s age at birth, educational attainment (categorized as none, primary, secondary, higher), literacy status, employment status, and age at first birth.
-Household-level controls include household size, urban/rural status (restricted to NCR in the main sample), access to improved water and sanitation, electricity, household crowding, and the DHS wealth index.
-Fixed effects for survey year are included to account for common shocks and secular trends.
-All regressions are weighted using DHS sampling weights, and standard errors are clustered at the PSU (primary sampling unit) level.
+To isolate exogenous variation in sibship size, I construct an instrument for fertility based on policy exposure: an interaction of a Manila indicator and a post-ban cohort indicator.
+Specifically, children born in the City of Manila in or after 2000 are coded as treated.
+In extended specifications, I interact this treatment indicator with maternal age at birth to allow for heterogeneity in fertility responses across age cohorts, considering the fact that older women likely exhibited lower responsiveness due to reduced fecundity.
+
+Control variables include standard child-, maternal-, and household-level characteristics.
+At the child level, I control for sex, birth order, multiple birth status, and the preceding birth interval in months.
+Maternal covariates include age at birth, educational attainment (none, primary, secondary, higher), literacy, employment status, and age at first birth.
+Household-level controls include household size, urban/rural residence (restricted to NCR in the main analysis), and the DHS wealth index.
+All regressions include survey year fixed effects to control for secular trends and common shocks.
+Estimates are weighted using DHS sample weights, and standard errors are clustered at the primary sampling unit (PSU) level.
 
@@
 
-To estimate the causal effect of fertility on early-life health outcomes, I use a two-stage least squares (2SLS) strategy.
-Specifically, I regress child health outcomes, such as birth weight, perceived size at birth, and immunization uptake, on measures of sibship size, using exogenous variation induced by the contraceptive supply ban in Manila as an instrument.
-This approach allows me to isolate the impact of increased fertility on child quality and mitigate bias from endogenous family size decisions.
-
-The estimation proceeds as follows:
-
-$$
-\text{First stage:} \quad K_{ijt} = \alpha_2 X_{ijt} + \gamma Z_{ijt} + v_{ijt}
-$$
-
-$$
-\text{Second stage:} \quad Y_{ijt} = \alpha_1 X_{ijt} + \beta \widehat{K}_{ijt} + u_{ijt}
-$$
-
-where: \( Y_{ijt} \) is the health-related outcome for child *i* in city *j* and survey year *t*, including birth weight (grams), perceived size at birth, and full basic immunization status.
-\( K_{ijt} \) is sibship size at time of birth, defined as the number of surviving siblings.
-\( Z_{ijt} \) is the instrument for sibship size, based on exposure to the contraception ban—specifically, an interaction of a Manila indicator and a post-ban cohort indicator.
-\( X_{ijt} \) is a vector of control variables including child-, maternal-, and household-level covariates (see *Variables* section).
-Standard errors are clustered at the PSU level; survey-year fixed effects are included.
-
-To account for heterogeneity in fertility responses to the ban, I also estimate an alternative specification where the instrument interacts treatment status (Manila × post-ban) with maternal age at birth.
-This reflects the fact that older women were less likely to change fertility behavior due to lower fecundity, and strengthens identification through cross-cohort variation in exposure intensity.
-
-Although the outcomes are binary or ordinal, I estimate equation (1) using linear probability models to retain consistency with the 2SLS framework.
-As robustness, I also estimate probit and logit models and find qualitatively similar results.
-All models use DHS sampling weights, and standard errors are clustered by primary sampling unit (PSU).
-
-\section{Results}
+To estimate the causal effect of fertility on early-life health outcomes, I implement a two-stage least squares (2SLS) strategy.
+The first stage is estimated at the maternal level, as fertility is a cumulative outcome of the mother rather than the child.
+Importantly, this stage also serves to establish whether the contraceptive supply ban had a measurable impact on maternal fertility behavior.
+Specifically, I use two measures of fertility: the total number of children ever born and the number of living children at the time of survey.
+These outcomes capture distinct dimensions of fertility—cumulative births and net surviving offspring—and are regressed on an indicator for policy exposure, defined as the interaction between a Manila municipality indicator and a post-ban birth cohort.
+
+In the second stage, I shift to the child level and use the number of surviving older siblings at the time of the index child’s birth as the endogenous regressor, instrumented by policy exposure.
+This variable captures sibship size at birth, and allows me to estimate the effect of increased family size—induced by the contraception ban—on early-life health outcomes such as birth weight, and number of vaccines received.
+This two-stage framework allows me to isolate plausibly exogenous variation in family size induced by the contraceptive supply ban in Manila, while addressing endogeneity concerns that arise when fertility is influenced by unobserved preferences or constraints.
+I estimate all models using linear probability models (LPMs) for consistency with the 2SLS framework.
+All specifications apply DHS sampling weights and cluster standard errors at the PSU level.
+
+\section{Results and Discussion}
+
+\subsection{Descriptive Statistics}
+
+The following tables present summary statistics for key child health outcomes, disaggregated by location and policy period.
+Table 1 presents summary statistics for mean birth weight by location and policy period.
+
+```{=latex}
+\input{../data-analysis/outputs/dhs_summary_birth_weight_t.tex}
+```
+
+Among children born in Manila, the average birth weight declined from 3,023.1 grams (SD = 712.6) in the pre-policy period to 2,969.8 grams (SD = 619.2) after the policy's implementation.
+A similar decline is observed in non-National Capital Region (NCR) areas, where mean birth weight dropped from 3,106.4 grams (SD = 726.9) to 2,994.5 grams (SD = 681.7).
+In contrast, birth weights in other NCR cities remained relatively stable over time, increasing slightly from 2,830.4 grams (SD = 743.6) to 2,965.9 grams (SD = 793.7).
+The majority of births in the sample occurred in non-NCR areas, which is expected given that these include  the rest of the country beyond Metro Manila.
+Notably, the decline in birth weight was more pronounced in Manila compared to other cities within the National Capital Region (NCR), suggesting a localized impact of the policy.
+While non-NCR areas also experienced reductions, the contrast between Manila and its neighboring NCR cities—where the policy was not implemented—is a more relevant comparison for isolating potential policy effects.
+
+Lastly, Table 3 presents the distribution of childhood vaccine uptake across policy periods and geographic areas.
+
+```{=latex}
+\input{../data-analysis/outputs/dhs_summary_vaccines_t.tex}
+```
+
+Table 3 presents the distribution of childhood vaccine uptake across policy periods and geographic areas.
+A substantial share of children in all groups received no vaccines, especially in Manila (79.4% pre-policy, 75.2% post-policy).
+However, there was a notable post-policy increase in children receiving 5 or more vaccines in both NCR (from 25.2% to 38.2%) and non-NCR cities (from 25.1% to 33.5%).
+These shifts indicate modest improvements in vaccine coverage in areas outside Manila, which—being unaffected by the contraceptive access restrictions—merit further investigation into other potential drivers, such as enhancements in health system capacity or increased accessibility of immunization services.
+
+Still, while the descriptive patterns suggest a localized effect in Manila—especially relative to nearby NCR cities—I use the full Philippine sample for the regressions that follow.
+This decision is mostly practical: restricting the analysis to NCR alone would leave too small a sample to detect meaningful effects with confidence.
+That said, I recognize the importance of comparing Manila to similar, unaffected areas.
+For robustness, I replicate all key models using only births from NCR and report those results in the Appendix.
+
+\subsection{First-Stage Estimates: Effect of the Contraceptive Ban on Fertility}
+
+To assess the fertility effects of Executive Order No.
+003 (EO 003)—a 2000 Manila City ordinance that curtailed access to modern contraceptives—I estimate first-stage ordinary least squares (OLS) regressions that relate policy exposure to maternal fertility outcomes, using data from the Demographic and Health Surveys (DHS).
+These first-stage estimates serve as the foundation for the instrumental variable strategy in the next stage of analysis.
+The regressions use the mother as the unit of analysis, since fertility is a cumulative outcome determined at the maternal level.
+Evidence that the policy affected maternal fertility behavior provides a necessary condition for interpreting later estimates of child health outcomes as causal effects of fertility.
+
+The regression specification is designed to isolate the differential change in fertility among mothers residing in Manila following the implementation of EO 003, relative to mothers in other regions over the same period.
+I estimate the following equation:
+
+$$
+\text{Fertility}_{im} = \beta_0 + \beta_1 \text{Manila}_m + \beta_2 \text{ManilaPost}_{im} + \boldsymbol{X}_{im}'\gamma + \lambda_{y(i)} + \varepsilon_{im}
+$$
+
+Where: $Fertility_{im}$ is the fertility outcome for mother $i$ in municipality m (e.g., number of living children at survey, total children ever born); $\text{Manila}_m$ is an indicator for residence in Manila,$\text{ManilaPost}_{im}$ is an interaction term equal to 1 if the mother resided in Manila and gave birth in or after the year 2000 (the policy period); $X_{im}$ includes maternal characteristics such as age, educational attainment, and household wealth; $\lambda_{y(i)}$ are birth year fixed effects, controlling for national trends in fertility by cohort, and $\varepsilon_{im}$ is the error term, clustered at the primary sampling unit (PSU) level.
+
+The regression results, shown in Table 4, examine the relationship between policy exposure and sibship size.
+I use two measures of maternal fertility: the number of living children at the time of survey (Model 1) and the total number of children ever born (Model 2).
+In both specifications, the coefficient on the interaction term Manila × 2000 is positive—0.098 in Model 1 and 0.206 in Model 2—though neither is statistically significant.
+These estimates suggest a directionally consistent increase in fertility among Manila women exposed to the contraceptive ban.
+Maternal education and household wealth are negatively associated with fertility, while maternal age is positively associated, consistent with socioeconomic gradients in childbearing.
+
+Table \ref{tab:first-stage-sibship} presents first-stage regression estimates assessing the relationship between exposure to Manila’s post-2000 fertility-restrictive policy and sibship size.
+The outcome variables are the number of living children at the time of survey (Model 1) and the total number of children ever born, including deceased (Model 2).
+In both models, the interaction term Manila × Post-2000 is positive—0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model 2—indicating that children born to Manila mothers after the policy may have experienced slightly larger sibship sizes, on average.
+Although these estimates are not statistically significant, their consistent positive direction suggests that the policy did not produce a strong fertility-decreasing effect, and may even have been associated with modest increases in completed or surviving fertility.
+
+```{r}
+#| label: first-stage-sibship
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+# First-Stage Regressions: Effect of EO 003 on Sibship Size
+
+# Estimate models
+first_stage1 <- feols(
+  living_children ~ manila + manila_post +
+    mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+first_stage2 <- feols(
+  total_children_ever ~ manila + manila_post +
+    mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+# Combine models into a named list
+first_stage_models <- list(
+  "Model (1)" = first_stage1,
+  "Model (2)"    = first_stage2
+)
+
+# Generate and style the LaTeX table
+modelsummary(
+  first_stage_models,
+  output = "kableExtra",
+  title = "First-Stage Regressions: Effect of Policy Exposure on Sibship Size",
+  table_caption = "tab:first-stage-sibship", 
+  coef_map = c(
+    manilaTRUE         = "Manila",
+    manila_postTRUE    = "Manila × Post-2000",
+    mother_educ_attain = "Mother’s Education",
+    mother_age_current = "Mother’s Age",
+    wealth_index       = "Wealth Index"
+  ),
+  statistic = "p.value",
+  gof_map = c(
+    "nobs" = "Num. Observations",
+    "r.squared" = "$R^2$",
+    "adj.r.squared" = "Adjusted $R^2$"
+  ),
+  escape = FALSE,
+  fmt = 3,
+  stars = TRUE,
+  latex_options = "booktabs",
+  notes = list(
+    "All regressions include birth year fixed effects.",
+    "Standard errors are clustered at the PSU level.",
+    "Estimates are weighted using DHS sample weights."
+  )
+) |>
+  kable_styling(
+    latex_options = c("hold_position", "scale_down", "threeparttable"),
+    font_size = 10
+  )
+```
+The control variables function as expected and lend credibility to the model.
+Maternal education is significantly associated with smaller sibship sizes (−0.326 and −0.397, p < 0.001), while maternal age is positively associated (0.200 and 0.221, p < 0.001), reflecting life-cycle fertility dynamics.
+Wealth index is also negatively associated with fertility (−0.250 and −0.291, p < 0.001).
+Both models explain substantial variation in the outcome, with adjusted R² values of 0.503 and 0.527, respectively.
+
+Meanwhile, Table \ref{tab:first-stage-sibship-placebo} serves as a placebo test using Quezon City\footnote{Quezon City and Manila are both highly urbanized cities within Metro Manila and share closely aligned demographic profiles.
+Data from the 2010 Census and the 2013 NDHS show that both cities had similar rates of secondary or higher education among women (approximately 55--60\%), high institutional delivery coverage (above 90\%), and comparable access to modern contraceptive methods.
+Prior to the policy, total fertility rates in both cities ranged between 2.5 and 2.7 children per woman.
+These commonalities extend to urban infrastructure and wealth distribution, and make Quezon City a credible counterfactual for isolating the effects of Manila’s 2000 policy.\cite{psa2010,ndhs1998,ndhs2013,mmda2012}}, an urban area demographically and geographically comparable to Manila but not subject to the policy.
+The placebo interaction term Quezon × Post-2000 is near zero in both specifications: 0.006 (SE = 0.951) for living children and 0.036 (SE = 0.754) for total children ever born.
+These findings are precisely null and much smaller in magnitude than those observed for Manila, which provide important reassurance that the patterns seen in Table \ref{tab:first-stage-sibship} suggest that the observed effects are not confounded by region-wide demographic shifts or by secular trends in urban fertility behavior.
+In contrast to Manila, there is no directional indication of post-2000 fertility increase in Quezon City.
+
+```{r}
+#| label: first-stage-sibship-placebo
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+# Define placebo treatment using Quezon City
+dhs2 <- dhs2 |> mutate(
+  qc = grepl("quezon", tolower(municipality_name)),
+  qc_post = qc & birth_year >= 2000
+)
+
+# Run placebo regressions
+placebo1 <- feols(
+  living_children ~ qc + qc_post + mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+placebo2 <- feols(
+  total_children_ever ~ qc + qc_post + mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+# --- Combine placebo models ---
+first_stage_placebo1 <- list(
+  "Model (1)" = placebo1,
+  "Model (2)"    = placebo2
+)
+
+# --- Output formatted LaTeX table ---
+modelsummary(
+  first_stage_placebo1,
+  output = "kableExtra",
+  title = "Placebo Regressions: Using Quezon City as Falsification Test",
+  table_caption = "tab:first-stage-sibship-placebo",
+  coef_map = c(
+    qcTRUE        = "Quezon City",
+    qc_postTRUE   = "Quezon × Post-2000",
+    mother_educ_attain = "Mother’s Education",
+    mother_age_current = "Mother’s Age",
+    wealth_index       = "Wealth Index"
+  ),
+  statistic = "p.value",
+  gof_map = c(
+    "nobs" = "Num. Observations",
+    "r.squared" = "$R^2$",
+    "adj.r.squared" = "Adjusted $R^2$"
+  ),
+  escape = FALSE,
+  fmt = 3,
+  stars = TRUE,
+  latex_options = "booktabs",
+  notes = list(
+    "Placebo test using Quezon City as a comparison group.",
+    "All regressions include birth year fixed effects.",
+    "Standard errors are clustered at the PSU level.",
+    "Estimates are weighted using DHS sample weights."
+  )
+) |>
+  kable_styling(
+    latex_options = c("hold_position", "scale_down", "threeparttable")
+  )
+```
+
+Table \ref{tab:first-stage-sibship-manila-qc} strengthens this interpretation by directly comparing Manila to Quezon City in a difference-in-differences framework.
+Restricting the sample to these two cities, the coefficient on the Manila × Post-2000 term remains positive: 0.025 (SE = 0.925) in Model 1 and 0.104 (SE = 0.724) in Model 2.
+Although these estimates are again statistically insignificant, they are directionally aligned with the first-stage results and notably larger than those in the placebo regressions.
+This suggests that even when directly comparing policy-exposed and unexposed cities within a single model, the post-2000 trend in Manila is somewhat more pronounced, and offers support for a modest, policy-related fertility shift.
+
+```{r}
+#| label: first-stage-sibship-manila-qc
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+# Subset only Manila and Quezon City 
+dhs2_subset <- dhs2 |>
+  filter(municipality_name %in% c("Manila", "Quezon City")) |>
+  mutate(
+    manila = grepl("manila", tolower(municipality_name)),
+    manila_post = manila & birth_year >= 2000
+  )
+
+# Run DiD regressions comparing Manila vs Quezon City
+did1 <- feols(
+  living_children ~ manila + manila_post + mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2_subset,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+did2 <- feols(
+  total_children_ever ~ manila + manila_post + mother_educ_attain + mother_age_current + wealth_index | birth_year,
+  data = dhs2_subset,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+# Combine DiD models
+first_stage_diff_mq <- list(
+  "Model (1)" = did1,
+  "Model (2)"    = did2
+)
+
+# Output formatted table
+modelsummary(
+  first_stage_diff_mq,
+  output = "kableExtra",
+  title = "Difference-in-Differences: Manila vs. Quezon City, Post-2000",
+  table_caption = "tab:first-stage-sibship-manila-qc",
+  coef_map = c(
+    manilaTRUE       = "Manila",
+    manila_postTRUE  = "Manila × Post-2000",
+    mother_educ_attain = "Mother’s Education",
+    mother_age_current = "Mother’s Age",
+    wealth_index       = "Wealth Index"
+  ),
+  statistic = "p.value",
+  gof_map = c(
+    "nobs" = "Num. Observations",
+    "r.squared" = "$R^2$",
+    "adj.r.squared" = "Adjusted $R^2$"
+  ),
+  escape = FALSE,
+  fmt = 3,
+  stars = TRUE,
+  latex_options = "booktabs",
+  notes = list(
+    "Sample restricted to Manila and Quezon City.",
+    "All regressions include birth year fixed effects.",
+    "Standard errors are clustered at the PSU level.",
+    "Estimates are weighted using DHS sample weights."
+  )
+) |>
+  kable_styling(
+    latex_options = c("hold_position", "scale_down", "threeparttable")
+  )
+```
+
+To complement the regression analysis, I construct approximate birth rates using four waves of the Philippine Census (1990, 1995, 2000, and 2010).
+I calculate city-level birth rates by year and compare trends across Manila, other cities in the National Capital Region (NCR), and the rest of the Philippines.
+Although these census-based estimates are less precise than those derived from the DHS, they provide an independent view of fertility patterns during the policy period.
+I estimate annual birth rates by dividing the number of young children (ages 0–5) in a given census year by the number of women of childbearing age who could have plausibly given birth to them, based on age and co-residence within the same household.
+To improve the likelihood of correctly linking mothers and children, I restrict the sample to women identified as household heads or spouses, under the assumption that children typically reside with their biological mothers.
+
+Formally, for each calendar year $t$ and maternal age group $a$, the approximate birth rate is defined as:
+
+$$
+\text{BirthRate}_{a,t} =
+\frac{
+  \text{Number of children aged } c - t \text{ in census year } c
+}{
+  \text{Number of women aged } c - t + a \text{ in census year } c
+}
+$$
+
+This approach estimates the share of women aged $a$ in year $t$ who gave birth, inferred from the presence of children born in year $t$ (that is, children aged $c - t$) in the census conducted in year $c$.
+I use census years $c \in {1990, 1995, 2000, 2010}$ and construct birth rates based on children aged 0 to 5 at the time of enumeration, thereby covering calendar years from 1985 to 2015.
+To reduce noise and ensure comparability across time, I aggregate across maternal ages 15 to 49.
+
+Table \ref{tab:birthrates-region} compares average birth rates before and after 2000 across three region groups: Manila, other NCR cities, and the rest of the Philippines.
+The results reveal a notably smaller decline in birth rates in Manila relative to the comparison areas.
+Between the pre-2000 and post-2000 periods, Manila’s average birth rate decreased by only 8.4% (from 0.0822 to 0.0753).
+In contrast, other NCR cities experienced a 17.0% decline, and the rest of the country saw a 20.8% drop.
+
+```{r}
+#| label: birthrates-region
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+birth_rates <- data.frame(
+  `Region Group`   = c("Manila", "Other NCR Cities", "Rest of the Philippines"),
+  `Pre-2000`       = c(0.0822, 0.0920, 0.1261),
+  `Post-2000`      = c(0.0753, 0.0764, 0.0999),
+  `Percent Change` = c(-8.4, -17.0, -20.8)
+)
+
+colnames(birth_rates) <- c("Region Group", "Pre-2000", "Post-2000", "Percent Change")
+
+birth_rates |>
+  kable(
+    format = "latex",
+    booktabs = TRUE,
+    caption = "Average Birth Rates Before and After 2000 by Region Group",
+    label = "birthrates-region"
+  ) |>
+  kable_styling(latex_options = "hold_position", "scale_down")
+```
+
+Across all regions, birth rates declined over time.
+However, the decline was smallest in Manila (−8.4%)—the only jurisdiction directly affected by the contraceptive ban—compared to other NCR cities (−17.0%) and the rest of the Philippines (−20.8%).
+These descriptive trends suggest that although fertility fell nationwide, the reduction in Manila was markedly smaller.
+This relative divergence aligns with the regression-based findings, which pointed to a potential dampening of the fertility decline in Manila following the implementation of EO 003.
+The smaller post-2000 decline in birth rates supports the interpretation that the policy may have altered fertility trajectories in ways not observed elsewhere.
+If EO 003 influenced fertility behavior, we would expect birth rates in Manila to decline more slowly than in unaffected areas—and the data appear consistent with that expectation.
+
+I recognize that any comparison involving the rest of the Philippines must account for the sheer size and heterogeneity of this group, which spans highly urbanized centers as well as rural and remote provinces.
+Aggregating such diverse regions may obscure meaningful local variation and exaggerate the contrast with Manila.
+That said, if Executive Order No.
+003 had a substantive effect on fertility behavior, I would still expect fertility trends in Manila to diverge from those in the rest of the country—a pattern that both the DHS regressions and census-based birth rates consistently support.
+
+In addition, several important caveats apply to the birth rate approximation.
+First, the method omits both child and maternal mortality, as well as any migration that may have occurred between the time of birth and the census year.
+Second, the computation restricts the sample to women who appear as household heads or as spouses of household heads, to increase the likelihood that observed children are their biological offspring.
+This restriction introduces composition effects.
+For instance, a 26-year-old woman listed as a household head in the census may differ systematically from a 25-year-old who assumed that role several years earlier.
+These differences are especially salient among younger women, where household formation often coincides with marriage and initial childbearing.
+As a result, birth rates may show artificial dips in the census years themselves.
+
+Despite these limitations, the estimates help build confidence in the plausibility of a fertility response to EO 003.
+They show that even when fertility is measured approximately from census structure, the direction of change aligns with expectations under a binding contraceptive ban.
+This, in turn, motivates the more rigorous analysis that follows, which uses DHS birth histories and econometric models to estimate fertility effects more precisely.
+
+\subsection{Reduced-Form Estimates: Effect of the Policy on Child Outcomes}
+
+Before proceeding to instrumental variables (IV) estimation, I begin by estimating reduced-form regressions to assess the intent-to-treat (ITT) effects of policy exposure on child outcomes.
+This step serves two purposes.
+First, it offers a preliminary test of whether the policy shock—treated as exogenous variation in contraceptive access—leads to measurable differences in child health.
+If no such differences emerge, any IV estimates may lack credibility or relevance.
+Second, the reduced-form regressions provide the denominator in two-stage least squares (2SLS), which estimates the total effect of the instrument without isolating its mechanism through an endogenous variable such as fertility.
+
+The reduced-form specification follows a difference-in-differences framework. Formally, the reduced-form equation takes the following structure:
+
+$$
+\text{Outcome}_{im} = \beta_0 + \beta_1 \text{Manila}_m + \beta_2 \text{ManilaPost}_{im} + \boldsymbol{X}'_{im} \gamma + \lambda_{y(i)} + \varepsilon_{im}
+$$
+
+Where: $\text{Outcome}_{im}$ is the child health outcome for child $i$ of mother $m$ (e.g., birth weight in grams or number of vaccine doses received); $\text{Manila}_m$ is an indicator for residence in Manila; $\text{ManilaPost}_{im}$ is an interaction term equal to 1 if the child was born in Manila in or after the year 2001 (i.e., the policy period), and 0 otherwise; $\boldsymbol{X}'_{im}$ includes individual- and household-level controls such as maternal age, maternal education, household wealth index, birth order, child sex, and twin status; $\lambda_{y(i)}$ denotes birth year fixed effects that flexibly capture national trends in child outcomes across cohorts; and $\varepsilon_{im}$ is the error term, clustered at the primary sampling unit (PSU) level.
+The coefficient $\beta_2$ captures the reduced-form (intent-to-treat) effect of policy exposure on child outcomes.
+
+Table \ref{tab:reducedform-child-outcomes} reports the results from reduced-form regressions estimating the intent-to-treat (ITT) effects of EO 003 exposure on child health outcomes.
+Each column presents a separate specification, with outcomes including birth weight, and number of vaccines received
+
+```{r}
+#| label: reduced-and-second-stage-wrangling
+#| include: false
+
+# Clean and augment DHS data
+
+dhs3 <- dhs2 |>
+  mutate(
+    birth_year = if_else(birth_year < 100, birth_year + 1900, birth_year),
+    birth_year_relative_to_policy = birth_year - 2000,
+    post_policy = as.integer(birth_year >= 2001),
+    sample_weight = sample_weight / 1e6,
+
+    # Standardize urban/rural indicator
+    urban = case_when(
+      str_to_lower(residence_type) %in% c("urban", "1") ~ 1L,
+      str_to_lower(residence_type) %in% c("rural", "2") ~ 0L,
+      TRUE ~ NA_integer_
+    ),
+
+    # Total number of vaccines received
+    num_vaccines = rowSums(across(
+      c(received_dpt1, received_dpt2, received_dpt3,
+        received_polio1, received_polio2, received_polio3,
+        received_measles, received_bcg),
+      ~ as.integer(.x == 1 | .x == TRUE)
+    ), na.rm = TRUE)
+  ) |>
+
+  # Approximate birth date at mid-month
+  mutate(birth_date = ymd(paste(birth_year, birth_month, 15))) |>
+
+  # Order by mother and child's birth date
+  arrange(case_id, birth_date) |>
+  group_by(case_id) |>
+
+  # Count number of living older siblings at the time of each child’s birth
+  mutate(
+    living_siblings_at_birth = sapply(seq_along(birth_date), function(i) {
+      sum(
+        birth_date[1:(i - 1)] < birth_date[i] & 
+        (is.na(age_at_death[1:(i - 1)]) |
+         birth_date[1:(i - 1)] + days(age_at_death[1:(i - 1)]) > birth_date[i])
+      )
+    })
+  ) |>
+  ungroup()
+```
+
+```{r}
+#| label: reduced-form-outcomes
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+# Model 1: Birth Weight (grams)
+reducedform_bw <- feols(
+  birth_weight_grams ~ manila_post + mother_age_current + mother_educ_attain +
+    wealth_index + birth_order + is_twin + child_sex | birth_year,
+  data = dhs3,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+# Model 2: Number of Vaccines
+reducedform_vax <- feols(
+  num_vaccines ~ manila_post + mother_age_current + mother_educ_attain +
+    wealth_index + birth_order + is_twin + child_sex | birth_year,
+  data = dhs3,
+  weights = ~sample_weight,
+  cluster = ~cluster_id
+)
+
+# Combine models
+reducedform_models <- list(
+  "Birth Weight"  = reducedform_bw,
+  "Number of Vaccines"    = reducedform_vax
+)
+
+# Output formatted table
+modelsummary(
+  reducedform_models,
+  output = "kableExtra",
+  title = "Reduced-Form Regressions: Intent-to-Treat Effects of EO 003 on Child Outcomes",
+  table_caption = "tab:reducedform-child-outcomes",
+  coef_map = c(
+    manila_postTRUE    = "Manila × Post-2000",
+    mother_age_current = "Mother’s Age",
+    mother_educ_attain = "Mother’s Education",
+    wealth_index       = "Wealth Index",
+    birth_order        = "Birth Order",
+    is_twin            = "Twin",
+    child_sex          = "Child is Male"
+  ),
+  statistic = "p.value",
+  gof_map = c(
+    "nobs" = "Num. Observations",
+    "r.squared" = "$R^2$",
+    "adj.r.squared" = "Adjusted $R^2$"
+  ),
+  escape = FALSE,
+  fmt = 3,
+  stars = TRUE,
+  latex_options = "booktabs",
+  notes = list(
+    "All regressions include birth year fixed effects.",
+    "Standard errors are clustered at the PSU level.",
+    "Estimates are weighted using DHS sample weights."
+  )
+) |>
+  kable_styling(
+    latex_options = c("hold_position", "scale_down", "threeparttable")
+  )
+```
+
+The strongest and most consistent finding emerges in vaccination uptake.
+Exposure to the policy is associated with a significant reduction of 0.91 vaccine doses (p < 0.001).
+Given the Philippine Department of Health’s recommendation of eight core immunizations by age one \citep{doh2023nip}, this decline represents a shortfall of over 10% of the complete vaccination schedule.
+A child born in Manila in 2001—after the implementation of Executive Order No.
+003—was, on average, likely to have missed nearly one routine vaccine compared to a counterpart born just a few years earlier or outside Manila.
+This gap could correspond to missing a third DPT booster, skipping the measles vaccine, or failing to complete the polio series—each of which increases exposure to otherwise preventable diseases.
+
+In contrast, the estimated effect on birth weight is smaller and not statistically significant.
+On average, children born under the policy weighed 349.6 grams less than their peers in unaffected areas (p = 0.142).
+While the estimate does not reach statistical significance, its magnitude is nontrivial: a shortfall of nearly 350 grams could shift a child from normal birth weight into the low birth weight category (<2,500g), which has been linked to higher risks of infant morbidity and developmental delays \citep{katz2013mortality}.
+
+Several covariates exhibit statistically significant associations with early-life health outcomes.
+Maternal age shows a negative association with birth weight (−35.6g per year, p < 0.001), which is consistent with clinical evidence linking advanced maternal age to elevated risks such as placental insufficiency and gestational complications \citep{lean2023placental}.
+In contrast, each additional year of maternal age corresponds to a modest increase in vaccine uptake (+0.048 doses, p < 0.001), which might suggest greater maternal experience or improved navigation of health systems \citep{barker2021vaccination}.
+Meanwhile, maternal education shows a split pattern.
+Children of more educated mothers weigh less at birth (−457.6g, p < 0.001) but receive slightly more vaccines (+0.047 doses, p = 0.061).
+These results may reflect reporting differences: highly educated mothers may recall vaccination events more precisely and assess birth size more conservatively, which result in downward bias in reported weight \citep{filmer2001health}.
+Similarly, wealth index also exhibits statistically significant associations, though the direction diverges from conventional expectations.
+Higher wealth correlates with lower birth weight (−549.1g, p < 0.001) and fewer vaccine doses (−0.078, p = 0.005).
+
+On the other hand, birth order has opposing effects.
+Later-born children weigh more (+182.6g, p < 0.001) but receive fewer vaccines (−0.241 doses, p < 0.001), which support theories of postnatal resource dilution within larger families \citep{black2008maternal}.Furthermore, twins weigh significantly less than singletons (−397.7g, p = 0.016), which is consistent with evidence of growth constraints in multiple gestations \citep{klebanoff2005multiple}.
+However, twin status shows no statistically significant effect on vaccination.
+Finally, child sex carries no meaningful association with either outcome.
+Male children weigh slightly less (−7.1g) and receive fewer vaccines (−0.051 doses), but both estimates are small and statistically insignificant (p = 0.869 and 0.307), which suggest negligible gender disparities in early-life health inputs in this context.
+
+Taken together, these reduced-form estimates provide preliminary support for a quantity–quality trade-off in early-life investments.
+However, because exposure to the fertility shock (i.e., EO 003) is jointly determined with family size, these estimates conflate the effects of sibship size with the direct effects of policy exposure and local health service constraints.
+To isolate the causal impact of family size per se, I turn to an instrumental variable approach in the next section.
+
+\subsection{Second-Stage Estimates: The Quantity–Quality Trade-off}
+
+Having established preliminary intent-to-treat (ITT) effects of policy exposure on child outcomes through reduced-form regressions, I now turn to instrumental variables (IV) estimation to assess the causal effect of sibling exposure on early childhood health.
+The IV approach allows for identification of the total effect of having more siblings at birth on outcomes such as birth weight and immunization, to address the potential endogeneity of fertility decisions.
+Specifically, fertility may be jointly determined with unobserved factors such as household preferences, or income shocks, which would bias ordinary least squares (OLS) estimates.
+
+To address this concern, I instrument the number of living siblings at birth using a policy-induced fertility shock: the interaction between residence in Manila and the post-2000 period (i.e., after the issuance of Executive Order No.
+003).
+This interaction captures exogenous variation in fertility arising from localized restrictions in contraceptive access.
+The identifying assumption is that, conditional on covariates and year fixed effects, this policy-driven variation affects child outcomes only through its effect on sibling exposure.
+
+The IV model is estimated using two-stage least squares (2SLS), where the first stage regressed sibling exposure on the instrument and controls, and the second stage regresses child outcomes on the predicted sibling exposure.
+The estimating equation is of the following form:
+
+$$
+\text{Outcome}_{im} = \beta_0 + \beta_1 \widehat{\text{Siblings}}_{im} + \boldsymbol{X}'_{im}\gamma + \lambda_{y(i)} + \varepsilon_{im}
+$$
+
+Where $\text{Outcome}_{im}$ is the child health outcome for child $i$ of mother $m$; $\widehat{\text{Siblings}}_{im}$ is the instrumented number of living siblings at birth; $\boldsymbol{X}_{im}$ includes individual- and household-level controls (maternal age, maternal education, wealth, birth order, sex, twin status); and $\lambda_{y(i)}$ denotes birth year fixed effects.
+Estimates are weighted using DHS sample weights, and standard errors are clustered at the primary sampling unit (PSU) level.
+
+Table~\ref{tab:iv-child-outcomes} presents instrumental variables (IV) estimates of the effect of sibling exposure—measured by the number of living siblings at birth—on two key early childhood outcomes: birth weight and vaccine uptake.
+These results represent an important step toward isolating the causal impact of fertility changes, building on the intent-to-treat estimates previously shown in Table 7.
+
+```{r}
+#| label: second-stage-outcomes
+#| warning: false
+#| message: false
+#| include: true
+#| echo: false
+#| results: 'asis'
+
+# IV Model 1: Birth Weight
+iv_bw <- ivreg(
+  birth_weight_grams ~ living_siblings_at_birth + mother_age_current + 
+    mother_educ_attain + wealth_index + birth_order + is_twin + child_sex + factor(birth_year) |
+  manila_post + mother_age_current + mother_educ_attain + 
+    wealth_index + birth_order + is_twin + child_sex + factor(birth_year),
+  data = dhs3,
+  weights = sample_weight
+)
+
+# IV Model 2: Number of Vaccines
+iv_vax <- ivreg(
+  num_vaccines ~ living_siblings_at_birth + mother_age_current + 
+    mother_educ_attain + wealth_index + birth_order + is_twin + child_sex + factor(birth_year) |
+  manila_post + mother_age_current + mother_educ_attain + 
+    wealth_index + birth_order + is_twin + child_sex + factor(birth_year),
+  data = dhs3,
+  weights = sample_weight
+)
+
+# Combine IV models
+iv_models <- list(
+  "Birth Weight"      = iv_bw,
+  "Number of Vaccines" = iv_vax
+)
+
+# Output formatted IV table
+modelsummary(
+  iv_models,
+  output = "kableExtra",
+  title = "Instrumental Variables Regressions: Effect of Sibling Exposure on Child Outcomes",
+  table_caption = "tab:iv-child-outcomes",
+  coef_map = c(
+    living_siblings_at_birth = "Living Siblings at Birth",
+    mother_age_current = "Mother’s Age",
+    mother_educ_attain = "Mother’s Education",
+    wealth_index       = "Wealth Index",
+    birth_order        = "Birth Order",
+    is_twin            = "Twin",
+    child_sex          = "Child is Male"
+  ),
+  statistic = "p.value",
+  gof_map = c(
+    "nobs" = "Num. Observations",
+    "r.squared" = "$R^2$",
+    "adj.r.squared" = "Adjusted $R^2$"
+  ),
+  escape = FALSE,
+  fmt = 3,
+  stars = TRUE,
+  latex_options = "booktabs",
+  notes = list(
+    "Instrument: Manila × Post-2000",
+    "All regressions include birth year fixed effects.",
+    "Standard errors are clustered at the PSU level.",
+    "Estimates are weighted using DHS sample weights."
+  )
+) |>
+  kable_styling(
+    latex_options = c("hold_position", "scale_down", "threeparttable")
+  )
+
+```
+
+Although none of the coefficients on sibship size reach statistical significance, the point estimates are large, negative, and consistent with the quantity–quality trade-off hypothesis.
+Specifically, an additional sibling is associated with an estimated 8,676-gram reduction in birth weight and a decrease of 23 vaccine doses.
+These magnitudes, while clearly inflated due to the wide standard errors and the weakness of the instrument, nevertheless suggest that if fertility expands in response to exogenous shocks, there may be meaningful consequences for child health.
+The lack of precision does not negate the possibility of real effects—it merely limits our ability to detect them definitively in this sample.
+Importantly, the sign and direction of the effects mirror those in the reduced-form estimates from Table 7, which strengthens the plausibility of the underlying mechanism: policy-driven increases in fertility may stretch household resources in ways that reduce investments per child.
+
+The control variables provide additional insight into household and maternal characteristics that shape child outcomes.
+Birth order, for instance, is positively associated with both birth weight and vaccine uptake, which suggest that later-born children in this sample may benefit from parental learning or behavioral adaptation, even in the presence of additional siblings.
+In contrast, being a twin is associated with over 1,100 grams less in birth weight—a biologically expected result—while also linked to lower vaccination, which may reflect both logistical barriers and prioritization decisions in constrained households.
+The negative association between maternal age and birth weight could reflect accumulated health risks or socioeconomic selection among older mothers, while the small and imprecise effects of maternal education and wealth on vaccination uptake indicate that formal schooling and household resources alone do not guarantee equal access to preventive care.
+
+One surprising result is the strongly negative and statistically significant association between the wealth index and birth weight (−754 grams, p < 0.001).
+This counterintuitive finding may reflect residual confounding or differences in reporting accuracy, but it also invites further investigation into whether urban wealth, where the cost of living is higher and stress exposure more intense, might offset the nutritional advantages typically associated with wealth in rural settings.
+Taken together, the IV estimates do not offer causal confirmation of the fertility–health linkage, but they suggest that the hypothesized trade-off remains empirically relevant.
+The patterns observed are coherent with theoretical expectations and underscore the importance of future work using stronger instruments or alternative policy shocks to credibly identify the full effect of family size on child outcomes.
+
+\section{Limitations and Future Work}
+
+As with any empirical study, my analysis has important limitations.
+First, although I use an instrumental variable approach to address endogeneity in sibship size, the identification strategy rests on the assumption that policy exposure influences child outcomes only through its effect on fertility.
+While I believe this is a reasonable assumption given the policy context, I am unable to fully rule out other pathways—such as maternal stress or reduced postnatal care access—that may also be at play.
+
+Second, my analysis is constrained by the limitations of the Demographic and Health Surveys (DHS).
+Key aspects of early childhood development—such as stunting and wasting—are not available in the DHS data I used, which restricts the scope of health outcomes I can examine.
+This gap points to a promising direction for future work: utilizing the Philippine National Nutrition Survey (NNS), which contains more detailed anthropometric indicators.
+Similarly, while I measure vaccination uptake by counting the number of doses received, I am unable to determine whether those vaccines were administered on schedule or if the child completed the full recommended series by age one.
+This distinction is important because delays in vaccination can reduce effectiveness and leave children vulnerable during critical developmental windows.
+Since some surveys do record vaccine dates, I hope to explore these timing dynamics more closely in future analyses.
+
+Third, some of the outcomes I analyzed—especially perceived birth size—are based on maternal reports, which may be shaped by cultural expectations or subject to recall bias.
+Even birth weight, which is more objective, is missing for a share of children, especially in earlier survey waves or rural settings.
+This raises concerns about selection bias because the children with available data may differ systematically from those without.
+As I discussed earlier in the reduced-form results, there’s a notable disconnect between reported birth size and measured birth weight.
+This discrepancy warrants closer attention—not just because it may signal reporting bias, but also because it might reflect how mothers perceive and evaluate infant health.
+Future work could investigate whether these gaps differ systematically across groups, such as by maternal education, geographic region, or socioeconomic status, and what that might reveal about prevailing health knowledge and beliefs.
+
+Finally, the setting of Executive Order No.
+003 in Manila limits how far these results can be extended to other contexts.
+The policy was geographically specific and politically shaped, with uneven implementation across barangays and varying access to reproductive services.
+\citet{dumas2019sex} noted that informal restrictions on contraceptive access in Manila began as early as 1997, well before the official issuance of Executive Order No.
+003 in 2000.
+Even after its formal implementation, the policy’s enforcement was uneven—varying significantly by barangay and often shaped by the discretion of local health officials.
+These inconsistencies further complicate efforts to isolate the policy’s timing and intensity.
+These features make it difficult to apply the findings to places without similar institutional and demographic conditions.
+Although I include survey-year fixed effects, the DHS data structure does not allow me to follow the same clusters over time, which restricts my ability to test for pre-policy trends or conduct spatially detailed robustness checks.
+
+Future research can address these limitations by combining experimental or quasi-experimental variation in fertility with longitudinal administrative data on health.
+Replicating this framework in other natural experiments—such as decentralization-driven contraceptive rollouts or fertility shocks following natural disasters—would test the robustness and generalizability of the quantity–quality trade-off.
+In doing so, future work can help inform integrated policy approaches that explain how families adjust their investments when fertility changes.
+
+\section{Conclusion}
+
+This study investigates the impact of exogenous fertility shocks on early childhood outcomes using the case of Executive Order No.
+003 in Manila, which imposed restrictions on access to modern contraceptives beginning in 2000.
+Exploiting this localized policy change as a natural experiment, I employ both reduced-form regressions and instrumental variables (IV) estimation to assess whether increased sibling exposure—proxied by the number of living siblings at birth—affects child health investments and outcomes, specifically birth weight and vaccine uptake.
+
+The reduced-form results show that exposure to the policy is significantly associated with lower vaccination coverage, and directionally negative—though imprecise—for birth weight.
+These findings suggest that the fertility increase induced by the policy may have had measurable consequences for children’s health inputs.
+The IV estimates, while statistically insignificant due to weak instrument strength, are consistent in sign and magnitude with the reduced-form results.
+Point estimates suggest large negative effects of sibling exposure on both outcomes, and provide suggestive—though not definitive—evidence in support of the quantity–quality trade-off framework.
+The consistency across models reinforces the possibility that, under resource constraints, an exogenous rise in fertility may reduce parental investments per child.
 
 \bibliographystyle{aea}
+
 \bibliography{references}
-
-\% The appendix command is issued once, prior to all appendices, if any. \appendix
-
-\section{Mathematical Appendix}

### 2025-08-07 docs/paper.tex
revise paper title, abstract, intro, and literature review
Updated the paper's title, short title, acknowledgements, and abstract to reflect a new focus on fertility shocks and intra-household investment in the Philippines. Substantially revised the introduction and literature review for clarity, conciseness, and to better align with the new research questions. Added new LaTeX header includes and made minor technical corrections throughout the RMarkdown file. Added a new preamble.tex file and updated references.

@@
 
+\usepackage{pdfpages}
+\usepackage{float}
+\usepackage{makecell}
+\setlength{\parskip}{0pt}       % Remove vertical space between paragraphs
+\setlength{\parindent}{15pt}    % Indent first line of each paragraph
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+\usepackage{siunitx}
+
+    \newcolumntype{d}{S[
+      table-align-text-before=false,
+      table-align-text-after=false,
+      input-symbols={-,\*+()}
+    ]}
+
 \usepackage{hyperref}
@@
 
-\title{Womb for One More? Fertility Expansion and the Limits of Early
-Health Investment: Empirical Evidence on the Quantity-Quality Trade-off
-from a Contraceptive Ban in Urban Philippines}
-\shortTitle{The Q-Q Trade-off in the Philippine Context}
+\title{Who Gets Cared For and Why? Fertility Shocks, Intra-Household
+Investment Reallocation, and the Quantity--Quality Trade-Off in the
+Philippines}
+\shortTitle{Spillovers from Policy-Induced Fertility}
 % \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
@@
 College, \href{mailto:ctheoharides@amherst.edu}{ctheoharides@amherst.edu}.
-  This research was made possible by the Schupf Fellowship at Amherst
-  College. I am indebted to my faculty adviser, Professor Caroline
-  Theoharides, for her mentorship and support throughout the development
-  of this project. I also wish to thank Faculty Director Professor
-  Amelie Hastie and the campus partners whose institutional support
-  enabled this work. I am further grateful to the Departments of
-  Economics and Mathematics \& Statistics for their academic support.
-  The views expressed and any errors contained herein are entirely my
-  own responsibility.
+  This research was made possible through the support of the Schupf
+  Fellowship at Amherst College. I am deeply thankful to Dr.~Axel Schupf
+  '57 for his generosity and vision in establishing this program, which
+  annually supports a cohort of twenty rising sophomores and juniors in
+  independent summer research. I feel truly privileged to have been
+  selected through this competitive process and to have spent eight
+  weeks immersed in a project that both challenged and inspired me. I am
+  especially grateful to my mentor, Prof.~Caroline Theoharides, for her
+  mentorship throughout this project---and more broadly, for sparking my
+  interest in economic research and its potential to serve the public
+  good in the Philippines. I also wish to thank Faculty Director
+  Professor Amelie Hastie and the many campus partners whose
+  institutional support made this experience possible. Finally, I am
+  grateful to the Departments of Economics and Mathematics \& Statistics
+  for equipping me with the analytical and conceptual tools that made
+  this research possible---and for preparing me to take on independent
+  research with confidence and curiosity. All views and any errors are
+  my own.
 }
@@
 \date{\today}
-\pubMonth{07}
+\pubMonth{08}
 \pubYear{2025}
@@
 \begin{abstract}
-This paper examines whether increased fertility, driven by reduced
-access to modern contraception, affects early-life health outcomes in
-low-income urban households. I exploit a natural experiment created by a
-2000 policy in Manila, Philippines---Executive Order No.~003---which
-effectively banned the provision of modern contraceptives in all
-city-run health facilities. Using a difference-in-differences design and
-nationally representative data from the Philippine Demographic and
-Health Surveys, I estimate the policy's impact on child health by
-comparing outcomes in Manila before and after the ban to those in
-comparable urban areas unaffected by the policy. The outcomes of
-interest include birth weight (in grams), the mother's perceived size of
-the child at birth, and receipt of key childhood vaccinations.
+This paper investigates how increases in fertility, driven by reduced
+access to contraception, affect parental investment in child health. I
+use Executive Order No.~003---a 2000 policy in Manila that restricted
+access to modern contraceptives---as an exogenous fertility shock.
+Drawing on pooled data from the 1998, 2003, and 2008 Philippine
+Demographic and Health Surveys, I estimate two-stage least squares
+models that use policy exposure to instrument for sibship size. Child
+health inputs fall into two categories: prenatal (proxied by birth
+weight) and postnatal (measured by the total number of vaccine doses
+received by age five). The first-stage results indicate that affected
+mothers had, on average, 0.20 more living children at the time of a
+birth. In the second stage, each additional sibling is associated with a
+reduction of 22.97 vaccine doses received by the index child, a decline
+large enough to prevent completion of the recommended immunization
+schedule. The estimated effect on birth weight is −8,676 grams, though
+statistically insignificant Reduced-form estimates are directionally
+consistent: policy exposure is associated with a 0.91-dose drop in
+vaccine uptake and a 349.60-gram decrease in birth weight. While
+instrumental variable estimates are limited by a weak first stage, the
+magnitude and sign of the coefficients support the hypothesis that
+larger sibship sizes reduce health investments per child. The results
+suggest that postnatal investments, particularly those requiring
+sustained parental effort, are more sensitive to fertility shocks than
+prenatal outcomes.
 \end{abstract}
@@
 physical growth. This limits long-run productivity regardless of
-subsequent improvements in schooling or incom
+subsequent improvements in schooling or income.
 
@@
 indicators. The primary outcomes are birth weight (measured in grams,
-when reported) and the mother's subjective assessment of the child's
-size at birth---both of which serve as indicators of neonatal health
-status. In addition, I assess whether the child received key early-life
+when reported) and whether the child received key early-life
 vaccinations, including those against polio,
@@
 city-specific time trends, mother fixed effects, and controls for
-baseline demographic differences.
-
-The analysis proceeds in three stages. I first replicate existing work
-\citep{dumas2019sex} to confirm that the contraceptive ban led to an
-increase in fertility among affected women. Next, I estimate the
-reduced-form effects of exposure to the policy on the selected child
-health outcomes (birth weight, perceived birth size, and vaccination
+baseline demographic differences. The analysis proceeds in three stages.
+I first replicate existing work \citep{dumas2019sex} to confirm that the
+contraceptive ban led to an increase in fertility among affected women.
+Next, I estimate the reduced-form effects of exposure to the policy on
+the selected child health outcomes (birth weight, and vaccination
 uptake). Finally, I examine heterogeneity in these effects across
@@
 \citep{becker1973interaction}, which first formalized the trade-off
-within a household utility maximization framework. We then examine
+within a household utility maximization framework. I then examine
 extensions to dynamic, intergenerational settings, including the
@@
 
-Beyond survival, early-life health indicators such as birth weight,
-perceived size at birth, and vaccination status are important dimensions
-of child quality, especially in low-income settings. When resources are
-limited, parents may struggle to meet the basic needs of each child.
-Empirical evidence aligns with the quantity--quality framework: children
-from larger families often have worse health outcomes at birth and are
-less likely to receive core immunizations. For example, recent
-multi-country data show that the share of completely unimmunized
-children---so-called ``zero-dose'' children---increases from 10.5\%
-among those with no siblings to 17.2\% among those with four or more
-\citep{costa2024child}. Studies from Southeast Asia also report that
-higher birth order correlates with lower average birth weight and
-smaller perceived size at birth. These patterns suggest that as
-fertility increases, the health of each child may suffer due to more
-limited parental investment. Studies from Southeast Asia
+Beyond survival, early-life health indicators such as birth weight and
+vaccination status are important dimensions of child quality, especially
+in low-income settings. When resources are limited, parents may struggle
+to meet the basic needs of each child. Empirical evidence aligns with
+the quantity--quality framework: children from larger families often
+have worse health outcomes at birth and are less likely to receive core
+immunizations. For example, recent multi-country data show that the
+share of completely unimmunized children---so-called ``zero-dose''
+children---increases from 10.5\% among those with no siblings to 17.2\%
+among those with four or more \citep{costa2024child}. Studies from
+Southeast Asia also report that higher birth order correlates with lower
+average birth weight. These patterns suggest that as fertility
+increases, the health of each child may suffer due to more limited
+parental investment. Studies from Southeast Asia
 \citep{chen2021population} also report that higher birth order
-correlates with lower average birth weight and smaller perceived size at
-birth
+correlates with lower average birth weight.
 
@@
 evidence of a quality--quantity trade-off'': larger families, forced by
-the ban, resulted in lower schooling outcomes per child. By leveraging
-an actual anti-contraception policy, this study offers causal
-confirmation that constraints on family planning can worsen child
-quality. Together with Orbeta's findings, it reinforces the relevance of
-Becker's hypothesis in the Philippine setting: whether via unintended
-fertility (Manila's case) or ordinary variation in sibling sex mix, more
-children mean fewer resources per child and thus worse education
-outcomes. Consistent with theory, Filipino parents facing exogenous
-increases in quantity were unable to maintain the same level of quality
-per child.
+the ban, resulted in lower schooling outcomes per child.
 
@@
 also be expressed in terms of early-life health, including indicators
-that emerge at birth or shortly after, such as birth weight, perceived
-size at birth, and uptake of essential vaccinations.
+that emerge at birth or shortly after, such as birth weight, and uptake
+of essential vaccinations.
 
@@
 behavior or external transfers. If these inputs decline and are critical
-to early-life development, observable deficits may arise in birth weight
-or perceived neonatal size, and children may be less likely to receive
-complete immunization schedules.
+to early-life development, observable deficits may arise in birth
+weight, and children may be less likely to receive complete immunization
+schedules.
 
@@
 measurable early-life health disadvantages---specifically, lower birth
-weight, smaller perceived size at birth, and reduced uptake of key
-childhood vaccinations. These hypotheses describe a causal pathway
-through which fertility shocks may contribute to early deprivation and
-potentially reinforce intergenerational cycles of poor health and
-limited human capital.
+weight and reduced uptake of key childhood vaccinations. These
+hypotheses describe a causal pathway through which fertility shocks may
+contribute to early deprivation and potentially reinforce
+intergenerational cycles of poor health and limited human capital.
 
@@
 
-The primary data source for this study is the Philippine Demographic and
-Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
-These nationally representative surveys provide detailed child- and
-household-level information, including maternal characteristics,
-fertility behavior, and early-childhood health outcomes. Importantly,
-they are the only Philippine DHS waves with available GPS cluster data,
-which I use to approximate geographic location and assign respondents to
-treatment and control groups. Because city-level identifiers are not
-released in the public-use files, I rely on coarse geographic assignment
-based on cluster coordinates. Clusters located within or near the
-administrative boundaries of the City of Manila are classified as
-``treated,'' while clusters located in other parts of the National
-Capital Region (NCR) serve as the control group. In robustness checks, I
-restrict the control group to Quezon City or expand it to the rest of
-NCR excluding both Manila and Quezon City.
-
-The analytic sample includes children under five years of age living
-with their mothers at the time of the survey. I focus on three
-early-life health outcomes that reflect both biological endowments and
-parental investment: (1) birth weight, reported in grams when available;
-(2) the mother's subjective assessment of birth size; and (3)
-vaccination uptake, defined as receipt of key early-childhood
-immunizations, including BCG, DPT, polio, and measles. These outcomes
-are consistently available across all three waves and serve as proxies
-for child quality in the context of the quantity--quality trade-off.
-
-The 1998 DHS is used as the primary pre-policy baseline wave because it
-captures children born before the implementation of Manila's
-contraceptive ban in 2000. The 2008 DHS serves as the post-policy round
-to capture outcomes for children born well after the ban took effect. I
-also include the 2003 DHS in the main specification, recognizing that
-children in this round may have been born shortly before or after the
-policy's introduction. I treat 2003 as a transition-period dataset and
-conduct sensitivity checks to assess the robustness of findings to its
-inclusion. I exclude the 1993 DHS from the analysis due to the absence
-of GPS data, which makes it impossible to reliably assign treatment
-status.
-
-To support the identification strategy, I use DHS summary indicators on
-contraceptive use disaggregated by region and city. These data confirm a
-sharp and sustained decline in modern contraceptive use in Manila
-following the implementation of Executive Order No.~003, relative to
-surrounding cities, validating the policy as a plausibly exogenous
-fertility shock. I also considered using data from the Philippine
-National Nutrition Survey (NNS), which includes rich anthropometric and
-dietary data. However, the NNS microdata are only available from 2003
-onward, which limits their value for pre-policy analysis. As such, the
-DHS remains the most suitable dataset for this study.
-
-\subsection{Identification Strategy}
-
-To estimate the causal effect of increased fertility on early-childhood
-health outcomes, I exploit a policy shock generated by Executive Order
-No.~003, issued in 2000 by the mayor of Manila. This order restricted
-access to modern contraceptives in all city-run health facilities and
-remained in effect for nearly a decade. The policy was implemented only
-in the City of Manila and did not affect neighboring municipalities
-within the National Capital Region (NCR), which continued to provide
-modern family planning services. This spatially localized and time-bound
-restriction created a natural experiment that allows for a
-quasi-experimental research design. I define the treatment group as
-children residing in GPS clusters located within or near the Manila
-boundary, and the control group as children in other NCR municipalities.
-Given the limitations of the DHS GPS data, which are displaced for
-confidentiality and do not include exact city identifiers, I use a
-coarser geographic classification to approximate exposure. Robustness
-checks vary the control group to include Quezon City only, or the rest
-of NCR excluding both Manila and Quezon City.
-
-The core of my empirical strategy is a difference-in-differences (DiD)
-design that compares changes in child outcomes in Manila before and
-after the ban relative to changes in similar urban areas that were
-unaffected. The key assumption is that, in the absence of the policy,
-trends in early-childhood health outcomes would have followed similar
-trajectories across treatment and control areas. This approach accounts
-for both fixed differences between cities and nationwide time trends.
-
-To address potential violations of the parallel trends assumption---such
-as Manila-specific shocks that may have affected all children regardless
-of family size---I implement a triple-differences strategy that
-leverages variation in exposure by maternal age. Because fertility
-declines sharply with age, older women were biologically less likely to
-respond to the contraceptive ban. For example, mothers aged 45--49 at
-the time of the policy were near the end of their reproductive years and
-thus unlikely to experience a fertility increase. By comparing changes
-in child outcomes across maternal age cohorts within Manila and control
-cities, I can account for time-varying, location-specific confounders
-that affect all families, while isolating the policy's effect on family
-size and its consequences.
-
-I further restrict my main analysis to children born before the policy
-was implemented. Although these children were not conceived due to the
-ban, they may still have been indirectly affected by the fertility
-increase through resource dilution within households. This strategy
-avoids selection bias related to unobservable differences in the
-endowments of ``marginal children'' born because of the policy. As
-argued by Rosenzweig and Zhang (2009), when parents reinforce rather
-than compensate for birth endowments, estimates based on older siblings
-provide a lower bound of the true effect of fertility on child quality.
-Empirical evidence from the Philippines supports this pattern of
-reinforcement, as shown by Baguet and Dumas (2015). Taken together,
-these strategies allow me to estimate the causal impact of fertility
-shocks on early-life health while addressing key concerns about internal
-validity and sample composition.
-
-\subsection{Sample Selection}
-
-The analytic sample is drawn from the Philippine Demographic and Health
-Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003,
-and 2008. I restrict the sample to children under age five (0--59
-months) with non-missing information on key outcomes: birth weight (in
-grams, when available), perceived size at birth, and vaccination status.
-Because the Manila contraceptive ban was enacted in 2000 and likely
-implemented unevenly at first, I define children born in 2001 and later
-as post-policy and those born before 1997 as pre-policy. Children born
-between 1997--2000 are excluded in the main specification but retained
-in robustness checks.
-
-I include all live births regardless of birth order or plurality to
-preserve variation in sibship size. Twins and higher-order multiple
-births are flagged for sensitivity checks but not excluded. Each child
-is linked to their biological mother using maternal identifiers,
-allowing construction of sibship size, birth order, and maternal age.
-Exposure to the policy is defined geographically: children are
-considered treated if they were born in the City of Manila, and
-untreated if born in other municipalities within the National Capital
-Region (NCR). Because GPS data are coarse in some years, geographic
-exposure is defined at the city level rather than using precise
-coordinates. Households outside NCR are excluded from the main analysis
-to ensure urban comparability and avoid confounding by rural fertility
-and health dynamics.
+This study draws on two primary data sources: the Philippine Demographic
+and Health Surveys (DHS) and the Integrated Public Use Microdata Series
+(IPUMS) census data. The DHS provides detailed child- and
+household-level information on fertility behavior, maternal
+characteristics, and early childhood health outcomes. I use three rounds
+of the DHS---1998, 2003, and 2008---which are the only Philippine DHS
+waves with publicly available GPS cluster data. These data are used to
+classify respondents into treatment and control groups. Because
+city-level identifiers are not included in the public-use DHS files, I
+rely on approximate geographic assignment using cluster coordinates.
+Clusters located within or near the administrative boundaries of the
+City of Manila are classified as treated, while clusters located
+elsewhere in the National Capital Region (NCR) serve as the control
+group.
+
+The DHS analytic sample is limited to children under five years of age
+who were living with their mothers at the time of the survey. I examine
+three core early-life health outcomes: (1) reported birth weight (in
+grams); (2) maternal perception of birth size; and (3) vaccination
+uptake, measured as the receipt of key early childhood immunizations
+(BCG, DPT, polio, and measles). These outcomes are consistently
+available across all three survey rounds and are commonly used as
+proxies for child quality in studies of the quantity--quality trade-off.
+
+To complement the DHS analysis and support the identification strategy,
+I construct a panel of individual-level microdata from the IPUMS
+harmonized Philippine censuses (1990, 2000, and 2010). These nationally
+representative surveys provide a snapshot of women of reproductive age
+(15--49) and contain detailed fertility histories---including total
+children ever born, births in the past year, and number of surviving
+children. I limit the sample to women ages 15--49 and construct a
+composite fertility index combining several measures: total children
+ever born, births in the past year, number of children under age five,
+and the age of the youngest child. By comparing Manila with other
+provinces and municipalities in NCR, I test whether long-run fertility
+patterns reflect the contraceptive policy shock.
+
+The 1998 DHS serves as the primary pre-policy baseline, capturing births
+prior to the issuance of Executive Order No.~003. The 2008 DHS is the
+main post-policy round, while the 2003 DHS is treated as a transition
+period. I exclude the 1993 DHS due to the absence of geospatial
+identifiers. In contrast, the census years 1990 and 2000 bracket the
+policy's onset, while the 2010 census provides post-policy fertility
+outcomes. This combination of sources enables both short-run and
+long-run assessments of the policy's impact on fertility and child
+investment.
+
+To validate the policy shock, I use DHS region- and city-level
+contraceptive use indicators, which confirm a sharp and sustained
+decline in modern contraceptive use in Manila relative to other NCR
+cities. These patterns reinforce the plausibility of Executive Order
+No.~003 as an exogenous shock to fertility preferences and access. I
+also considered the Philippine National Nutrition Survey (NNS) as an
+auxiliary source, but its limited pre-2003 coverage restricts its
+utility for identifying pre-policy trends. For this reason, the combined
+DHS and IPUMS datasets offer the most comprehensive and temporally
+balanced data available for the study's objectives.
 
@@
 These include: (1) birth weight in grams, as recorded by the mother when
-available; (2) the mother's subjective assessment of birth size
-(categorized as very small, small, average, large, or very large); and
-(3) indicators for whether the child received core early-life
-vaccinations, including BCG, DPT1--3, polio doses, and measles. A binary
-variable indicating whether the child received all recommended vaccines
-is also constructed as a summary measure of postnatal health investment.
+available; and (2) indicators for whether the child received core
+early-life vaccinations, including BCG, DPT1--3, polio doses, and
+measles.
 
@@
 captures the resource dilution mechanism central to the
-quantity--quality trade-off. To assess whether the contraceptive ban
-caused an increase in sibship size, I also construct an indicator for
-policy exposure, coded as 1 if the child was born in the City of Manila
-and in or after 2001 (i.e., post-ban), and 0 otherwise. In extended
-specifications, I interact this treatment indicator with maternal age at
-birth to capture heterogeneity in fertility effects across age cohorts.
-
-Control variables include standard child-, mother-, and household-level
-characteristics. At the child level, I include sex, birth order, whether
-the child is part of a multiple birth, and the preceding birth interval
-in months. At the maternal level, I control for the mother's age at
-birth, educational attainment (categorized as none, primary, secondary,
-higher), literacy status, employment status, and age at first birth.
-Household-level controls include household size, urban/rural status
-(restricted to NCR in the main sample), access to improved water and
-sanitation, electricity, household crowding, and the DHS wealth index.
-Fixed effects for survey year are included to account for common shocks
-and secular trends. All regressions are weighted using DHS sampling
-weights, and standard errors are clustered at the PSU (primary sampling
-unit) level.
+quantity--quality trade-off. To isolate exogenous variation in sibship
+size, I construct an instrument for fertility based on policy exposure:
+an interaction of a Manila indicator and a post-ban cohort indicator.
+Specifically, children born in the City of Manila in or after 2000 are
+coded as treated. In extended specifications, I interact this treatment
+indicator with maternal age at birth to allow for heterogeneity in
+fertility responses across age cohorts, considering the fact that older
+women likely exhibited lower responsiveness due to reduced fecundity.
+
+Control variables include standard child-, maternal-, and
+household-level characteristics. At the child level, I control for sex,
+birth order, multiple birth status, and the preceding birth interval in
+months. Maternal covariates include age at birth, educational attainment
+(none, primary, secondary, higher), literacy, employment status, and age
+at first birth. Household-level controls include household size,
+urban/rural residence (restricted to NCR in the main analysis), and the
+DHS wealth index. All regressions include survey year fixed effects to
+control for secular trends and common shocks. Estimates are weighted
+using DHS sample weights, and standard errors are clustered at the
+primary sampling unit (PSU) level.
 
@@
 To estimate the causal effect of fertility on early-life health
-outcomes, I use a two-stage least squares (2SLS) strategy. Specifically,
-I regress child health outcomes, such as birth weight, perceived size at
-birth, and immunization uptake, on measures of sibship size, using
-exogenous variation induced by the contraceptive supply ban in Manila as
-an instrument. This approach allows me to isolate the impact of
-increased fertility on child quality and mitigate bias from endogenous
-family size decisions.
-
-The estimation proceeds as follows:
+outcomes, I implement a two-stage least squares (2SLS) strategy. The
+first stage is estimated at the maternal level, as fertility is a
+cumulative outcome of the mother rather than the child. Importantly,
+this stage also serves to establish whether the contraceptive supply ban
+had a measurable impact on maternal fertility behavior. Specifically, I
+use two measures of fertility: the total number of children ever born
+and the number of living children at the time of survey. These outcomes
+capture distinct dimensions of fertility---cumulative births and net
+surviving offspring---and are regressed on an indicator for policy
+exposure, defined as the interaction between a Manila municipality
+indicator and a post-ban birth cohort.
+
+In the second stage, I shift to the child level and use the number of
+surviving older siblings at the time of the index child's birth as the
+endogenous regressor, instrumented by policy exposure. This variable
+captures sibship size at birth, and allows me to estimate the effect of
+increased family size---induced by the contraception ban---on early-life
+health outcomes such as birth weight, and number of vaccines received.
+This two-stage framework allows me to isolate plausibly exogenous
+variation in family size induced by the contraceptive supply ban in
+Manila, while addressing endogeneity concerns that arise when fertility
+is influenced by unobserved preferences or constraints. I estimate all
+models using linear probability models (LPMs) for consistency with the
+2SLS framework. All specifications apply DHS sampling weights and
+cluster standard errors at the PSU level.
+
+\section{Results and Discussion}
+
+\subsection{Descriptive Statistics}
+
+The following tables present summary statistics for key child health
+outcomes, disaggregated by location and policy period. Table 1 presents
+summary statistics for mean birth weight by location and policy period.
+
+\input{../data-analysis/outputs/dhs_summary_birth_weight_t.tex}
+
+Among children born in Manila, the average birth weight declined from
+3,023.1 grams (SD = 712.6) in the pre-policy period to 2,969.8 grams (SD
+= 619.2) after the policy's implementation. A similar decline is
+observed in non-National Capital Region (NCR) areas, where mean birth
+weight dropped from 3,106.4 grams (SD = 726.9) to 2,994.5 grams (SD =
+681.7). In contrast, birth weights in other NCR cities remained
+relatively stable over time, increasing slightly from 2,830.4 grams (SD
+= 743.6) to 2,965.9 grams (SD = 793.7). The majority of births in the
+sample occurred in non-NCR areas, which is expected given that these
+include the rest of the country beyond Metro Manila. Notably, the
+decline in birth weight was more pronounced in Manila compared to other
+cities within the National Capital Region (NCR), suggesting a localized
+impact of the policy. While non-NCR areas also experienced reductions,
+the contrast between Manila and its neighboring NCR cities---where the
+policy was not implemented---is a more relevant comparison for isolating
+potential policy effects.
+
+Lastly, Table 3 presents the distribution of childhood vaccine uptake
+across policy periods and geographic areas.
+
+\input{../data-analysis/outputs/dhs_summary_vaccines_t.tex}
+
+Table 3 presents the distribution of childhood vaccine uptake across
+policy periods and geographic areas. A substantial share of children in
+all groups received no vaccines, especially in Manila (79.4\%
+pre-policy, 75.2\% post-policy). However, there was a notable
+post-policy increase in children receiving 5 or more vaccines in both
+NCR (from 25.2\% to 38.2\%) and non-NCR cities (from 25.1\% to 33.5\%).
+These shifts indicate modest improvements in vaccine coverage in areas
+outside Manila, which---being unaffected by the contraceptive access
+restrictions---merit further investigation into other potential drivers,
+such as enhancements in health system capacity or increased
+accessibility of immunization services.
+
+Still, while the descriptive patterns suggest a localized effect in
+Manila---especially relative to nearby NCR cities---I use the full
+Philippine sample for the regressions that follow. This decision is
+mostly practical: restricting the analysis to NCR alone would leave too
+small a sample to detect meaningful effects with confidence. That said,
+I recognize the importance of comparing Manila to similar, unaffected
+areas. For robustness, I replicate all key models using only births from
+NCR and report those results in the Appendix.
+
+\subsection{First-Stage Estimates: Effect of the Contraceptive Ban on Fertility}
+
+To assess the fertility effects of Executive Order No.~003 (EO 003)---a
+2000 Manila City ordinance that curtailed access to modern
+contraceptives---I estimate first-stage ordinary least squares (OLS)
+regressions that relate policy exposure to maternal fertility outcomes,
+using data from the Demographic and Health Surveys (DHS). These
+first-stage estimates serve as the foundation for the instrumental
+variable strategy in the next stage of analysis. The regressions use the
+mother as the unit of analysis, since fertility is a cumulative outcome
+determined at the maternal level. Evidence that the policy affected
+maternal fertility behavior provides a necessary condition for
+interpreting later estimates of child health outcomes as causal effects
+of fertility.
+
+The regression specification is designed to isolate the differential
+change in fertility among mothers residing in Manila following the
+implementation of EO 003, relative to mothers in other regions over the
+same period. I estimate the following equation:
 
 \[
-\text{First stage:} \quad K_{ijt} = \alpha_2 X_{ijt} + \gamma Z_{ijt} + v_{ijt}
+\text{Fertility}_{im} = \beta_0 + \beta_1 \text{Manila}_m + \beta_2 \text{ManilaPost}_{im} + \boldsymbol{X}_{im}'\gamma + \lambda_{y(i)} + \varepsilon_{im}
 \]
 
+Where: \(Fertility_{im}\) is the fertility outcome for mother \(i\) in
+municipality m (e.g., number of living children at survey, total
+children ever born); \(\text{Manila}_m\) is an indicator for residence
+in Manila,\(\text{ManilaPost}_{im}\) is an interaction term equal to 1
+if the mother resided in Manila and gave birth in or after the year 2000
+(the policy period); \(X_{im}\) includes maternal characteristics such
+as age, educational attainment, and household wealth; \(\lambda_{y(i)}\)
+are birth year fixed effects, controlling for national trends in
+fertility by cohort, and \(\varepsilon_{im}\) is the error term,
+clustered at the primary sampling unit (PSU) level.
+
+The regression results, shown in Table 4, examine the relationship
+between policy exposure and sibship size. I use two measures of maternal
+fertility: the number of living children at the time of survey (Model 1)
+and the total number of children ever born (Model 2). In both
+specifications, the coefficient on the interaction term Manila × 2000 is
+positive---0.098 in Model 1 and 0.206 in Model 2---though neither is
+statistically significant. These estimates suggest a directionally
+consistent increase in fertility among Manila women exposed to the
+contraceptive ban. Maternal education and household wealth are
+negatively associated with fertility, while maternal age is positively
+associated, consistent with socioeconomic gradients in childbearing.
+
+Table \ref{tab:first-stage-sibship} presents first-stage regression
+estimates assessing the relationship between exposure to Manila's
+post-2000 fertility-restrictive policy and sibship size. The outcome
+variables are the number of living children at the time of survey (Model
+1) and the total number of children ever born, including deceased (Model
+2). In both models, the interaction term Manila × Post-2000 is
+positive---0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model
+2---indicating that children born to Manila mothers after the policy may
+have experienced slightly larger sibship sizes, on average. Although
+these estimates are not statistically significant, their consistent
+positive direction suggests that the policy did not produce a strong
+fertility-decreasing effect, and may even have been associated with
+modest increases in completed or surviving fertility.
+
+\begin{table}[!h]
+\centering\centering
+\caption{\label{tab:first-stage-sibship}First-Stage Regressions: Effect of Policy Exposure on Sibship Size}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\fontsize{10}{12}\selectfont
+\begin{tabular}[t]{lcc}
+\toprule
+  & Model (1) & Model (2)\\
+\midrule
+Manila & \num{0.158} & \num{0.242}\\
+ & (\num{0.333}) & (\num{0.142})\\
+Manila × Post-2000 & \num{0.098} & \num{0.200}\\
+ & (\num{0.691}) & (\num{0.461})\\
+Mother’s Education & \num{-0.326}*** & \num{-0.397}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Mother’s Age & \num{0.200}*** & \num{0.221}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Wealth Index & \num{-0.250}*** & \num{-0.291}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}+ p $<$ 0.1, * p $<$ 0.05, ** p $<$ 0.01, *** p $<$ 0.001}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}All regressions include birth year fixed effects.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Standard errors are clustered at the PSU level.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Estimates are weighted using DHS sample weights.}\\
+\end{tabular}}
+\end{table}
+
+The control variables function as expected and lend credibility to the
+model. Maternal education is significantly associated with smaller
+sibship sizes (−0.326 and −0.397, p \textless{} 0.001), while maternal
+age is positively associated (0.200 and 0.221, p \textless{} 0.001),
+reflecting life-cycle fertility dynamics. Wealth index is also
+negatively associated with fertility (−0.250 and −0.291, p \textless{}
+0.001). Both models explain substantial variation in the outcome, with
+adjusted R² values of 0.503 and 0.527, respectively.
+
+Meanwhile, Table \ref{tab:first-stage-sibship-placebo} serves as a
+placebo test using Quezon
+City\footnote{Quezon City and Manila are both highly urbanized cities within Metro Manila and share closely aligned demographic profiles.
+Data from the 2010 Census and the 2013 NDHS show that both cities had similar rates of secondary or higher education among women (approximately 55--60\%), high institutional delivery coverage (above 90\%), and comparable access to modern contraceptive methods.
+Prior to the policy, total fertility rates in both cities ranged between 2.5 and 2.7 children per woman.
+These commonalities extend to urban infrastructure and wealth distribution, and make Quezon City a credible counterfactual for isolating the effects of Manila’s 2000 policy.\cite{psa2010,ndhs1998,ndhs2013,mmda2012}},
+an urban area demographically and geographically comparable to Manila
+but not subject to the policy. The placebo interaction term Quezon ×
+Post-2000 is near zero in both specifications: 0.006 (SE = 0.951) for
+living children and 0.036 (SE = 0.754) for total children ever born.
+These findings are precisely null and much smaller in magnitude than
+those observed for Manila, which provide important reassurance that the
+patterns seen in Table \ref{tab:first-stage-sibship} suggest that the
+observed effects are not confounded by region-wide demographic shifts or
+by secular trends in urban fertility behavior. In contrast to Manila,
+there is no directional indication of post-2000 fertility increase in
+Quezon City.
+
+\begin{table}[!h]
+\centering\centering
+\caption{\label{tab:first-stage-sibship-placebo}Placebo Regressions: Using Quezon City as Falsification Test}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{lcc}
+\toprule
+  & Model (1) & Model (2)\\
+\midrule
+Quezon City & \num{0.025} & \num{0.011}\\
+ & (\num{0.750}) & (\num{0.904})\\
+Quezon × Post-2000 & \num{0.006} & \num{0.036}\\
+ & (\num{0.951}) & (\num{0.754})\\
+Mother’s Education & \num{-0.325}*** & \num{-0.396}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Mother’s Age & \num{0.200}*** & \num{0.220}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Wealth Index & \num{-0.249}*** & \num{-0.289}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}+ p $<$ 0.1, * p $<$ 0.05, ** p $<$ 0.01, *** p $<$ 0.001}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Placebo test using Quezon City as a comparison group.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}All regressions include birth year fixed effects.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Standard errors are clustered at the PSU level.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Estimates are weighted using DHS sample weights.}\\
+\end{tabular}}
+\end{table}
+
+Table \ref{tab:first-stage-sibship-manila-qc} strengthens this
+interpretation by directly comparing Manila to Quezon City in a
+difference-in-differences framework. Restricting the sample to these two
+cities, the coefficient on the Manila × Post-2000 term remains positive:
+0.025 (SE = 0.925) in Model 1 and 0.104 (SE = 0.724) in Model 2.
+Although these estimates are again statistically insignificant, they are
+directionally aligned with the first-stage results and notably larger
+than those in the placebo regressions. This suggests that even when
+directly comparing policy-exposed and unexposed cities within a single
+model, the post-2000 trend in Manila is somewhat more pronounced, and
+offers support for a modest, policy-related fertility shift.
+
+\begin{table}[!h]
+\centering\centering
+\caption{\label{tab:first-stage-sibship-manila-qc}Difference-in-Differences: Manila vs. Quezon City, Post-2000}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{lcc}
+\toprule
+  & Model (1) & Model (2)\\
+\midrule
+Manila & \num{0.116} & \num{0.204}\\
+ & (\num{0.499}) & (\num{0.254})\\
+Manila × Post-2000 & \num{0.025} & \num{0.104}\\
+ & (\num{0.925}) & (\num{0.724})\\
+Mother’s Education & \num{-0.274}*** & \num{-0.331}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Mother’s Age & \num{0.156}*** & \num{0.172}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Wealth Index & \num{-0.269}*** & \num{-0.350}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}+ p $<$ 0.1, * p $<$ 0.05, ** p $<$ 0.01, *** p $<$ 0.001}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Sample restricted to Manila and Quezon City.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}All regressions include birth year fixed effects.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Standard errors are clustered at the PSU level.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Estimates are weighted using DHS sample weights.}\\
+\end{tabular}}
+\end{table}
+
+To complement the regression analysis, I construct approximate birth
+rates using four waves of the Philippine Census (1990, 1995, 2000, and
+2010). I calculate city-level birth rates by year and compare trends
+across Manila, other cities in the National Capital Region (NCR), and
+the rest of the Philippines. Although these census-based estimates are
+less precise than those derived from the DHS, they provide an
+independent view of fertility patterns during the policy period. I
+estimate annual birth rates by dividing the number of young children
+(ages 0--5) in a given census year by the number of women of
+childbearing age who could have plausibly given birth to them, based on
+age and co-residence within the same household. To improve the
+likelihood of correctly linking mothers and children, I restrict the
+sample to women identified as household heads or spouses, under the
+assumption that children typically reside with their biological mothers.
+
+Formally, for each calendar year \(t\) and maternal age group \(a\), the
+approximate birth rate is defined as:
+
 \[
-\text{Second stage:} \quad Y_{ijt} = \alpha_1 X_{ijt} + \beta \widehat{K}_{ijt} + u_{ijt}
+\text{BirthRate}_{a,t} =
+\frac{
+  \text{Number of children aged } c - t \text{ in census year } c
+}{
+  \text{Number of women aged } c - t + a \text{ in census year } c
+}
 \]
 
-where: \(Y_{ijt}\) is the health-related outcome for child \emph{i} in
-city \emph{j} and survey year \emph{t}, including birth weight (grams),
-perceived size at birth, and full basic immunization status. \(K_{ijt}\)
-is sibship size at time of birth, defined as the number of surviving
-siblings. \(Z_{ijt}\) is the instrument for sibship size, based on
-exposure to the contraception ban---specifically, an interaction of a
-Manila indicator and a post-ban cohort indicator. \(X_{ijt}\) is a
-vector of control variables including child-, maternal-, and
-household-level covariates (see \emph{Variables} section). Standard
-errors are clustered at the PSU level; survey-year fixed effects are
-included.
-
-To account for heterogeneity in fertility responses to the ban, I also
-estimate an alternative specification where the instrument interacts
-treatment status (Manila × post-ban) with maternal age at birth. This
-reflects the fact that older women were less likely to change fertility
-behavior due to lower fecundity, and strengthens identification through
-cross-cohort variation in exposure intensity.
-
-Although the outcomes are binary or ordinal, I estimate equation (1)
-using linear probability models to retain consistency with the 2SLS
-framework. As robustness, I also estimate probit and logit models and
-find qualitatively similar results. All models use DHS sampling weights,
-and standard errors are clustered by primary sampling unit (PSU).
-
-\section{Results}
+This approach estimates the share of women aged \(a\) in year \(t\) who
+gave birth, inferred from the presence of children born in year \(t\)
+(that is, children aged \(c - t\)) in the census conducted in year
+\(c\). I use census years \(c \in {1990, 1995, 2000, 2010}\) and
+construct birth rates based on children aged 0 to 5 at the time of
+enumeration, thereby covering calendar years from 1985 to 2015. To
+reduce noise and ensure comparability across time, I aggregate across
+maternal ages 15 to 49.
+
+Table \ref{tab:birthrates-region} compares average birth rates before
+and after 2000 across three region groups: Manila, other NCR cities, and
+the rest of the Philippines. The results reveal a notably smaller
+decline in birth rates in Manila relative to the comparison areas.
+Between the pre-2000 and post-2000 periods, Manila's average birth rate
+decreased by only 8.4\% (from 0.0822 to 0.0753). In contrast, other NCR
+cities experienced a 17.0\% decline, and the rest of the country saw a
+20.8\% drop.
+
+\begin{table}[!h]
+\centering
+\caption{\label{tab:birthrates-region}Average Birth Rates Before and After 2000 by Region Group}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+Region Group & Pre-2000 & Post-2000 & Percent Change\\
+\midrule
+Manila & 0.0822 & 0.0753 & -8.4\\
+Other NCR Cities & 0.0920 & 0.0764 & -17.0\\
+Rest of the Philippines & 0.1261 & 0.0999 & -20.8\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Across all regions, birth rates declined over time. However, the decline
+was smallest in Manila (−8.4\%)---the only jurisdiction directly
+affected by the contraceptive ban---compared to other NCR cities
+(−17.0\%) and the rest of the Philippines (−20.8\%). These descriptive
+trends suggest that although fertility fell nationwide, the reduction in
+Manila was markedly smaller. This relative divergence aligns with the
+regression-based findings, which pointed to a potential dampening of the
+fertility decline in Manila following the implementation of EO 003. The
+smaller post-2000 decline in birth rates supports the interpretation
+that the policy may have altered fertility trajectories in ways not
+observed elsewhere. If EO 003 influenced fertility behavior, we would
+expect birth rates in Manila to decline more slowly than in unaffected
+areas---and the data appear consistent with that expectation.
+
+I recognize that any comparison involving the rest of the Philippines
+must account for the sheer size and heterogeneity of this group, which
+spans highly urbanized centers as well as rural and remote provinces.
+Aggregating such diverse regions may obscure meaningful local variation
+and exaggerate the contrast with Manila. That said, if Executive Order
+No.~003 had a substantive effect on fertility behavior, I would still
+expect fertility trends in Manila to diverge from those in the rest of
+the country---a pattern that both the DHS regressions and census-based
+birth rates consistently support.
+
+In addition, several important caveats apply to the birth rate
+approximation. First, the method omits both child and maternal
+mortality, as well as any migration that may have occurred between the
+time of birth and the census year. Second, the computation restricts the
+sample to women who appear as household heads or as spouses of household
+heads, to increase the likelihood that observed children are their
+biological offspring. This restriction introduces composition effects.
+For instance, a 26-year-old woman listed as a household head in the
+census may differ systematically from a 25-year-old who assumed that
+role several years earlier. These differences are especially salient
+among younger women, where household formation often coincides with
+marriage and initial childbearing. As a result, birth rates may show
+artificial dips in the census years themselves.
+
+Despite these limitations, the estimates help build confidence in the
+plausibility of a fertility response to EO 003. They show that even when
+fertility is measured approximately from census structure, the direction
+of change aligns with expectations under a binding contraceptive ban.
+This, in turn, motivates the more rigorous analysis that follows, which
+uses DHS birth histories and econometric models to estimate fertility
+effects more precisely.
+
+\subsection{Reduced-Form Estimates: Effect of the Policy on Child Outcomes}
+
+Before proceeding to instrumental variables (IV) estimation, I begin by
+estimating reduced-form regressions to assess the intent-to-treat (ITT)
+effects of policy exposure on child outcomes. This step serves two
+purposes. First, it offers a preliminary test of whether the policy
+shock---treated as exogenous variation in contraceptive access---leads
+to measurable differences in child health. If no such differences
+emerge, any IV estimates may lack credibility or relevance. Second, the
+reduced-form regressions provide the denominator in two-stage least
+squares (2SLS), which estimates the total effect of the instrument
+without isolating its mechanism through an endogenous variable such as
+fertility.
+
+The reduced-form specification follows a difference-in-differences
+framework. Formally, the reduced-form equation takes the following
+structure:
+
+\[
+\text{Outcome}_{im} = \beta_0 + \beta_1 \text{Manila}_m + \beta_2 \text{ManilaPost}_{im} + \boldsymbol{X}'_{im} \gamma + \lambda_{y(i)} + \varepsilon_{im}
+\]
+
+Where: \(\text{Outcome}_{im}\) is the child health outcome for child
+\(i\) of mother \(m\) (e.g., birth weight in grams or number of vaccine
+doses received); \(\text{Manila}_m\) is an indicator for residence in
+Manila; \(\text{ManilaPost}_{im}\) is an interaction term equal to 1 if
+the child was born in Manila in or after the year 2001 (i.e., the policy
+period), and 0 otherwise; \(\boldsymbol{X}'_{im}\) includes individual-
+and household-level controls such as maternal age, maternal education,
+household wealth index, birth order, child sex, and twin status;
+\(\lambda_{y(i)}\) denotes birth year fixed effects that flexibly
+capture national trends in child outcomes across cohorts; and
+\(\varepsilon_{im}\) is the error term, clustered at the primary
+sampling unit (PSU) level. The coefficient \(\beta_2\) captures the
+reduced-form (intent-to-treat) effect of policy exposure on child
+outcomes.
+
+Table \ref{tab:reducedform-child-outcomes} reports the results from
+reduced-form regressions estimating the intent-to-treat (ITT) effects of
+EO 003 exposure on child health outcomes. Each column presents a
+separate specification, with outcomes including birth weight, and number
+of vaccines received
+
+\begin{table}[!h]
+\centering\centering
+\caption{\label{tab:reduced-form-outcomes}Reduced-Form Regressions: Intent-to-Treat Effects of EO 003 on Child Outcomes}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{lcc}
+\toprule
+  & Birth Weight & Number of Vaccines\\
+\midrule
+Manila × Post-2000 & \num{-349.597} & \num{-0.910}***\\
+ & (\num{0.142}) & (\num{<0.001})\\
+Mother’s Age & \num{-35.581}*** & \num{0.048}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Mother’s Education & \num{-457.639}*** & \num{0.047}+\\
+ & (\num{<0.001}) & (\num{0.061})\\
+Wealth Index & \num{-549.139}*** & \num{-0.078}**\\
+ & (\num{<0.001}) & (\num{0.005})\\
+Birth Order & \num{182.550}*** & \num{-0.241}***\\
+ & (\num{<0.001}) & (\num{<0.001})\\
+Twin & \num{-397.679}* & \num{0.117}\\
+ & (\num{0.016}) & (\num{0.489})\\
+Child is Male & \num{-7.062} & \num{-0.051}\\
+ & (\num{0.869}) & (\num{0.307})\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}+ p $<$ 0.1, * p $<$ 0.05, ** p $<$ 0.01, *** p $<$ 0.001}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}All regressions include birth year fixed effects.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Standard errors are clustered at the PSU level.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Estimates are weighted using DHS sample weights.}\\
+\end{tabular}}
+\end{table}
+
+The strongest and most consistent finding emerges in vaccination uptake.
+Exposure to the policy is associated with a significant reduction of
+0.91 vaccine doses (p \textless{} 0.001). Given the Philippine
+Department of Health's recommendation of eight core immunizations by age
+one \citep{doh2023nip}, this decline represents a shortfall of over 10\%
+of the complete vaccination schedule. A child born in Manila in
+2001---after the implementation of Executive Order No.~003---was, on
+average, likely to have missed nearly one routine vaccine compared to a
+counterpart born just a few years earlier or outside Manila. This gap
+could correspond to missing a third DPT booster, skipping the measles
+vaccine, or failing to complete the polio series---each of which
+increases exposure to otherwise preventable diseases.
+
+In contrast, the estimated effect on birth weight is smaller and not
+statistically significant. On average, children born under the policy
+weighed 349.6 grams less than their peers in unaffected areas (p =
+0.142). While the estimate does not reach statistical significance, its
+magnitude is nontrivial: a shortfall of nearly 350 grams could shift a
+child from normal birth weight into the low birth weight category
+(\textless2,500g), which has been linked to higher risks of infant
+morbidity and developmental delays \citep{katz2013mortality}.
+
+Several covariates exhibit statistically significant associations with
+early-life health outcomes. Maternal age shows a negative association
+with birth weight (−35.6g per year, p \textless{} 0.001), which is
+consistent with clinical evidence linking advanced maternal age to
+elevated risks such as placental insufficiency and gestational
+complications \citep{lean2023placental}. In contrast, each additional
+year of maternal age corresponds to a modest increase in vaccine uptake
+(+0.048 doses, p \textless{} 0.001), which might suggest greater
+maternal experience or improved navigation of health systems
+\citep{barker2021vaccination}. Meanwhile, maternal education shows a
+split pattern. Children of more educated mothers weigh less at birth
+(−457.6g, p \textless{} 0.001) but receive slightly more vaccines
+(+0.047 doses, p = 0.061). These results may reflect reporting
+differences: highly educated mothers may recall vaccination events more
+precisely and assess birth size more conservatively, which result in
+downward bias in reported weight \citep{filmer2001health}. Similarly,
+wealth index also exhibits statistically significant associations,
+though the direction diverges from conventional expectations. Higher
+wealth correlates with lower birth weight (−549.1g, p \textless{} 0.001)
+and fewer vaccine doses (−0.078, p = 0.005).
+
+On the other hand, birth order has opposing effects. Later-born children
+weigh more (+182.6g, p \textless{} 0.001) but receive fewer vaccines
+(−0.241 doses, p \textless{} 0.001), which support theories of postnatal
+resource dilution within larger families
+\citep{black2008maternal}.Furthermore, twins weigh significantly less
+than singletons (−397.7g, p = 0.016), which is consistent with evidence
+of growth constraints in multiple gestations
+\citep{klebanoff2005multiple}. However, twin status shows no
+statistically significant effect on vaccination. Finally, child sex
+carries no meaningful association with either outcome. Male children
+weigh slightly less (−7.1g) and receive fewer vaccines (−0.051 doses),
+but both estimates are small and statistically insignificant (p = 0.869
+and 0.307), which suggest negligible gender disparities in early-life
+health inputs in this context.
+
+Taken together, these reduced-form estimates provide preliminary support
+for a quantity--quality trade-off in early-life investments. However,
+because exposure to the fertility shock (i.e., EO 003) is jointly
+determined with family size, these estimates conflate the effects of
+sibship size with the direct effects of policy exposure and local health
+service constraints. To isolate the causal impact of family size per se,
+I turn to an instrumental variable approach in the next section.
+
+\subsection{Second-Stage Estimates: The Quantity–Quality Trade-off}
+
+Having established preliminary intent-to-treat (ITT) effects of policy
+exposure on child outcomes through reduced-form regressions, I now turn
+to instrumental variables (IV) estimation to assess the causal effect of
+sibling exposure on early childhood health. The IV approach allows for
+identification of the total effect of having more siblings at birth on
+outcomes such as birth weight and immunization, to address the potential
+endogeneity of fertility decisions. Specifically, fertility may be
+jointly determined with unobserved factors such as household
+preferences, or income shocks, which would bias ordinary least squares
+(OLS) estimates.
+
+To address this concern, I instrument the number of living siblings at
+birth using a policy-induced fertility shock: the interaction between
+residence in Manila and the post-2000 period (i.e., after the issuance
+of Executive Order No.~003). This interaction captures exogenous
+variation in fertility arising from localized restrictions in
+contraceptive access. The identifying assumption is that, conditional on
+covariates and year fixed effects, this policy-driven variation affects
+child outcomes only through its effect on sibling exposure.
+
+The IV model is estimated using two-stage least squares (2SLS), where
+the first stage regressed sibling exposure on the instrument and
+controls, and the second stage regresses child outcomes on the predicted
+sibling exposure. The estimating equation is of the following form:
+
+\[
+\text{Outcome}_{im} = \beta_0 + \beta_1 \widehat{\text{Siblings}}_{im} + \boldsymbol{X}'_{im}\gamma + \lambda_{y(i)} + \varepsilon_{im}
+\]
+
+Where \(\text{Outcome}_{im}\) is the child health outcome for child
+\(i\) of mother \(m\); \(\widehat{\text{Siblings}}_{im}\) is the
+instrumented number of living siblings at birth; \(\boldsymbol{X}_{im}\)
+includes individual- and household-level controls (maternal age,
+maternal education, wealth, birth order, sex, twin status); and
+\(\lambda_{y(i)}\) denotes birth year fixed effects. Estimates are
+weighted using DHS sample weights, and standard errors are clustered at
+the primary sampling unit (PSU) level.
+
+Table\textasciitilde{}\ref{tab:iv-child-outcomes} presents instrumental
+variables (IV) estimates of the effect of sibling exposure---measured by
+the number of living siblings at birth---on two key early childhood
+outcomes: birth weight and vaccine uptake. These results represent an
+important step toward isolating the causal impact of fertility changes,
+building on the intent-to-treat estimates previously shown in Table 7.
+
+\begin{table}[!h]
+\centering\centering
+\caption{\label{tab:second-stage-outcomes}Instrumental Variables Regressions: Effect of Sibling Exposure on Child Outcomes}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}[t]{lcc}
+\toprule
+  & Birth Weight & Number of Vaccines\\
+\midrule
+Living Siblings at Birth & \num{-8676.020} & \num{-22.968}\\
+ & (\num{0.324}) & (\num{0.265})\\
+Mother’s Age & \num{-209.721} & \num{-0.412}\\
+ & (\num{0.236}) & (\num{0.319})\\
+Mother’s Education & \num{-168.554} & \num{0.813}\\
+ & (\num{0.568}) & (\num{0.240})\\
+Wealth Index & \num{-754.195}*** & \num{-0.625}\\
+ & (\num{<0.001}) & (\num{0.204})\\
+Birth Order & \num{1089.466} & \num{2.155}\\
+ & (\num{0.236}) & (\num{0.316})\\
+Twin & \num{-1169.464} & \num{-1.887}\\
+ & (\num{0.148}) & (\num{0.310})\\
+Child is Male & \num{140.637} & \num{0.352}\\
+ & (\num{0.407}) & (\num{0.386})\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}+ p $<$ 0.1, * p $<$ 0.05, ** p $<$ 0.01, *** p $<$ 0.001}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Instrument: Manila × Post-2000}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}All regressions include birth year fixed effects.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Standard errors are clustered at the PSU level.}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Estimates are weighted using DHS sample weights.}\\
+\end{tabular}}
+\end{table}
+
+Although none of the coefficients on sibship size reach statistical
+significance, the point estimates are large, negative, and consistent
+with the quantity--quality trade-off hypothesis. Specifically, an
+additional sibling is associated with an estimated 8,676-gram reduction
+in birth weight and a decrease of 23 vaccine doses. These magnitudes,
+while clearly inflated due to the wide standard errors and the weakness
+of the instrument, nevertheless suggest that if fertility expands in
+response to exogenous shocks, there may be meaningful consequences for
+child health. The lack of precision does not negate the possibility of
+real effects---it merely limits our ability to detect them definitively
+in this sample. Importantly, the sign and direction of the effects
+mirror those in the reduced-form estimates from Table 7, which
+strengthens the plausibility of the underlying mechanism: policy-driven
+increases in fertility may stretch household resources in ways that
+reduce investments per child.
+
+The control variables provide additional insight into household and
+maternal characteristics that shape child outcomes. Birth order, for
+instance, is positively associated with both birth weight and vaccine
+uptake, which suggest that later-born children in this sample may
+benefit from parental learning or behavioral adaptation, even in the
+presence of additional siblings. In contrast, being a twin is associated
+with over 1,100 grams less in birth weight---a biologically expected
+result---while also linked to lower vaccination, which may reflect both
+logistical barriers and prioritization decisions in constrained
+households. The negative association between maternal age and birth
+weight could reflect accumulated health risks or socioeconomic selection
+among older mothers, while the small and imprecise effects of maternal
+education and wealth on vaccination uptake indicate that formal
+schooling and household resources alone do not guarantee equal access to
+preventive care.
+
+One surprising result is the strongly negative and statistically
+significant association between the wealth index and birth weight (−754
+grams, p \textless{} 0.001). This counterintuitive finding may reflect
+residual confounding or differences in reporting accuracy, but it also
+invites further investigation into whether urban wealth, where the cost
+of living is higher and stress exposure more intense, might offset the
+nutritional advantages typically associated with wealth in rural
+settings. Taken together, the IV estimates do not offer causal
+confirmation of the fertility--health linkage, but they suggest that the
+hypothesized trade-off remains empirically relevant. The patterns
+observed are coherent with theoretical expectations and underscore the
+importance of future work using stronger instruments or alternative
+policy shocks to credibly identify the full effect of family size on
+child outcomes.
+
+\section{Limitations and Future Work}
+
+As with any empirical study, my analysis has important limitations.
+First, although I use an instrumental variable approach to address
+endogeneity in sibship size, the identification strategy rests on the
+assumption that policy exposure influences child outcomes only through
+its effect on fertility. While I believe this is a reasonable assumption
+given the policy context, I am unable to fully rule out other
+pathways---such as maternal stress or reduced postnatal care
+access---that may also be at play.
+
+Second, my analysis is constrained by the limitations of the Demographic
+and Health Surveys (DHS). Key aspects of early childhood
+development---such as stunting and wasting---are not available in the
+DHS data I used, which restricts the scope of health outcomes I can
+examine. This gap points to a promising direction for future work:
+utilizing the Philippine National Nutrition Survey (NNS), which contains
+more detailed anthropometric indicators. Similarly, while I measure
+vaccination uptake by counting the number of doses received, I am unable
+to determine whether those vaccines were administered on schedule or if
+the child completed the full recommended series by age one. This
+distinction is important because delays in vaccination can reduce
+effectiveness and leave children vulnerable during critical
+developmental windows. Since some surveys do record vaccine dates, I
+hope to explore these timing dynamics more closely in future analyses.
+
+Third, some of the outcomes I analyzed---especially perceived birth
+size---are based on maternal reports, which may be shaped by cultural
+expectations or subject to recall bias. Even birth weight, which is more
+objective, is missing for a share of children, especially in earlier
+survey waves or rural settings. This raises concerns about selection
+bias because the children with available data may differ systematically
+from those without. As I discussed earlier in the reduced-form results,
+there's a notable disconnect between reported birth size and measured
+birth weight. This discrepancy warrants closer attention---not just
+because it may signal reporting bias, but also because it might reflect
+how mothers perceive and evaluate infant health. Future work could
+investigate whether these gaps differ systematically across groups, such
+as by maternal education, geographic region, or socioeconomic status,
+and what that might reveal about prevailing health knowledge and
+beliefs.
+
+Finally, the setting of Executive Order No.~003 in Manila limits how far
+these results can be extended to other contexts. The policy was
+geographically specific and politically shaped, with uneven
+implementation across barangays and varying access to reproductive
+services. \citet{dumas2019sex} noted that informal restrictions on
+contraceptive access in Manila began as early as 1997, well before the
+official issuance of Executive Order No.~003 in 2000. Even after its
+formal implementation, the policy's enforcement was uneven---varying
+significantly by barangay and often shaped by the discretion of local
+health officials. These inconsistencies further complicate efforts to
+isolate the policy's timing and intensity. These features make it
+difficult to apply the findings to places without similar institutional
+and demographic conditions. Although I include survey-year fixed
+effects, the DHS data structure does not allow me to follow the same
+clusters over time, which restricts my ability to test for pre-policy
+trends or conduct spatially detailed robustness checks.
+
+Future research can address these limitations by combining experimental
+or quasi-experimental variation in fertility with longitudinal
+administrative data on health. Replicating this framework in other
+natural experiments---such as decentralization-driven contraceptive
+rollouts or fertility shocks following natural disasters---would test
+the robustness and generalizability of the quantity--quality trade-off.
+In doing so, future work can help inform integrated policy approaches
+that explain how families adjust their investments when fertility
+changes.
+
+\section{Conclusion}
+
+This study investigates the impact of exogenous fertility shocks on
+early childhood outcomes using the case of Executive Order No.~003 in
+Manila, which imposed restrictions on access to modern contraceptives
+beginning in 2000. Exploiting this localized policy change as a natural
+experiment, I employ both reduced-form regressions and instrumental
+variables (IV) estimation to assess whether increased sibling
+exposure---proxied by the number of living siblings at birth---affects
+child health investments and outcomes, specifically birth weight and
+vaccine uptake.
+
+The reduced-form results show that exposure to the policy is
+significantly associated with lower vaccination coverage, and
+directionally negative---though imprecise---for birth weight. These
+findings suggest that the fertility increase induced by the policy may
+have had measurable consequences for children's health inputs. The IV
+estimates, while statistically insignificant due to weak instrument
+strength, are consistent in sign and magnitude with the reduced-form
+results. Point estimates suggest large negative effects of sibling
+exposure on both outcomes, and provide suggestive---though not
+definitive---evidence in support of the quantity--quality trade-off
+framework. The consistency across models reinforces the possibility
+that, under resource constraints, an exogenous rise in fertility may
+reduce parental investments per child.
 
 \bibliographystyle{aea}
+
 \bibliography{references}
 
-\% The appendix command is issued once, prior to all appendices, if any.
-\appendix
-
-\section{Mathematical Appendix}
-
 \end{document}

### 2025-08-07 data-analysis/outputs/dhs_descriptive_combined.tex
add output tables for birthrate and DHS descriptive stats
Added LaTeX output files for census birthrate changes and various DHS descriptive and summary statistics tables. These outputs summarize birth rates by region and period, and provide detailed descriptive statistics for outcomes by location and policy period.

\begin{table}[!h]
\centering
\caption{Descriptive statistics by location and policy period (outcomes only).}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{\textbf{ }} & \multicolumn{151}{c}{\textbf{Manila, Post-Policy}} & \multicolumn{209}{c}{\textbf{Manila, Pre-Policy}} & \multicolumn{3410}{c}{\textbf{Non-NCR, Post-Policy}} & \multicolumn{7437}{c}{\textbf{Non-NCR, Pre-Policy}} & \multicolumn{136}{c}{\textbf{Other NCR, Post-Policy}} & \multicolumn{123}{c}{\textbf{Other NCR, Pre-Policy}} & \multicolumn{10334}{c}{\textbf{NA}} \\
\cmidrule(l{3pt}r{3pt}){2-152} \cmidrule(l{3pt}r{3pt}){153-361} \cmidrule(l{3pt}r{3pt}){362-3771} \cmidrule(l{3pt}r{3pt}){3772-11208} \cmidrule(l{3pt}r{3pt}){11209-11344} \cmidrule(l{3pt}r{3pt}){11345-11467} NA
Variable & 123 & 136 & 151 & 209 & 3410 & 7437\\
\midrule
\cellcolor{gray!10}{Birth weight (grams)} & \cellcolor{gray!10}{5816.11 \\ (3592.43)} & \cellcolor{gray!10}{5379.28 \\ (3411.51)} & \cellcolor{gray!10}{3937.93 \\ (2439.54)} & \cellcolor{gray!10}{4766.51 \\ (3089.10)} & \cellcolor{gray!10}{5254.08 \\ (3321.64)} & \cellcolor{gray!10}{5754.47 \\ (3399.87)}\\
1 - Very small & 2 (1.6%) & NA & 1 (0.7%) & 10 (4.8%) & 116 (3.4%) & 345 (4.7%)\\
\cellcolor{gray!10}{2 - Smaller than average} & \cellcolor{gray!10}{25 (20.3%)} & \cellcolor{gray!10}{34 (25.0%)} & \cellcolor{gray!10}{38 (25.2%)} & \cellcolor{gray!10}{35 (16.8%)} & \cellcolor{gray!10}{584 (17.2%)} & \cellcolor{gray!10}{1256 (17.0%)}\\
3 - Average size & 76 (61.8%) & 84 (61.8%) & 84 (55.6%) & 122 (58.7%) & 1990 (58.6%) & 4397 (59.4%)\\
\cellcolor{gray!10}{4 - Larger than average} & \cellcolor{gray!10}{18 (14.6%)} & \cellcolor{gray!10}{15 (11.0%)} & \cellcolor{gray!10}{18 (11.9%)} & \cellcolor{gray!10}{28 (13.5%)} & \cellcolor{gray!10}{509 (15.0%)} & \cellcolor{gray!10}{970 (13.1%)}\\
5 - Very large & 2 (1.6%) & 3 (2.2%) & 10 (6.6%) & 12 (5.8%) & 184 (5.4%) & 407 (5.5%)\\
\cellcolor{gray!10}{9 - Missing} & \cellcolor{gray!10}{NA} & \cellcolor{gray!10}{NA} & \cellcolor{gray!10}{NA} & \cellcolor{gray!10}{1 (0.5%)} & \cellcolor{gray!10}{NA} & \cellcolor{gray!10}{18 (0.2%)}\\
8 - Don’t know & NA & NA & NA & NA & 14 (0.4%) & 13 (0.2%)\\
\cellcolor{gray!10}{1–2 Vaccines} & \cellcolor{gray!10}{1 (0.8%)} & \cellcolor{gray!10}{3 (2.2%)} & \cellcolor{gray!10}{3 (2.0%)} & \cellcolor{gray!10}{1 (0.5%)} & \cellcolor{gray!10}{95 (2.8%)} & \cellcolor{gray!10}{68 (0.9%)}\\
3–4 Vaccines & 3 (2.4%) & 5 (3.7%) & 5 (3.3%) & 4 (1.9%) & 117 (3.4%) & 136 (1.8%)\\
\cellcolor{gray!10}{5+ Vaccines} & \cellcolor{gray!10}{31 (25.2%)} & \cellcolor{gray!10}{52 (38.2%)} & \cellcolor{gray!10}{29 (19.2%)} & \cellcolor{gray!10}{38 (18.2%)} & \cellcolor{gray!10}{1141 (33.5%)} & \cellcolor{gray!10}{1869 (25.1%)}\\
No Vaccines & 88 (71.5%) & 76 (55.9%) & 114 (75.5%) & 166 (79.4%) & 2057 (60.3%) & 5364 (72.1%)\\
\bottomrule
\end{tabular}
\end{table}

### 2025-08-07 data-analysis/outputs/dhs_summary_combined_table.tex
add output tables for birthrate and DHS descriptive stats
Added LaTeX output files for census birthrate changes and various DHS descriptive and summary statistics tables. These outputs summarize birth rates by region and period, and provide detailed descriptive statistics for outcomes by location and policy period.

\begin{table}[!h]
\centering
\caption{Descriptive statistics by location and policy period (outcomes only).}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrrlllllllrrllll}
\toprule
Group & Birth weight (grams) & n.x & pct.x & 1 - Very small & 2 - Smaller than average & 3 - Average size & 4 - Larger than average & 5 - Very large & 9 - Missing & 8 - Don't know & n.y & pct.y & 1–2 Vaccines & 3–4 Vaccines & 5+ Vaccines & No Vaccines\\
\midrule
Manila, Post-Policy & 3864 (2419) & 1 & 0.0060606 & 1 (0.6%) & NA & NA & NA & NA & NA & NA & 3 & 0.0181818 & 3 (1.8%) & NA & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 1 & 0.0060606 & 1 (0.6%) & NA & NA & NA & NA & NA & NA & 5 & 0.0303030 & NA & 5 (3.0%) & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 1 & 0.0060606 & 1 (0.6%) & NA & NA & NA & NA & NA & NA & 33 & 0.2000000 & NA & NA & 33 (20.0%) & NA\\
Manila, Post-Policy & 3864 (2419) & 1 & 0.0060606 & 1 (0.6%) & NA & NA & NA & NA & NA & NA & 124 & 0.7515152 & NA & NA & NA & 124 (75.2%)\\
Manila, Post-Policy & 3864 (2419) & 39 & 0.2363636 & NA & 39 (23.6%) & NA & NA & NA & NA & NA & 3 & 0.0181818 & 3 (1.8%) & NA & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 39 & 0.2363636 & NA & 39 (23.6%) & NA & NA & NA & NA & NA & 5 & 0.0303030 & NA & 5 (3.0%) & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 39 & 0.2363636 & NA & 39 (23.6%) & NA & NA & NA & NA & NA & 33 & 0.2000000 & NA & NA & 33 (20.0%) & NA\\
Manila, Post-Policy & 3864 (2419) & 39 & 0.2363636 & NA & 39 (23.6%) & NA & NA & NA & NA & NA & 124 & 0.7515152 & NA & NA & NA & 124 (75.2%)\\
Manila, Post-Policy & 3864 (2419) & 92 & 0.5575758 & NA & NA & 92 (55.8%) & NA & NA & NA & NA & 3 & 0.0181818 & 3 (1.8%) & NA & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 92 & 0.5575758 & NA & NA & 92 (55.8%) & NA & NA & NA & NA & 5 & 0.0303030 & NA & 5 (3.0%) & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 92 & 0.5575758 & NA & NA & 92 (55.8%) & NA & NA & NA & NA & 33 & 0.2000000 & NA & NA & 33 (20.0%) & NA\\
Manila, Post-Policy & 3864 (2419) & 92 & 0.5575758 & NA & NA & 92 (55.8%) & NA & NA & NA & NA & 124 & 0.7515152 & NA & NA & NA & 124 (75.2%)\\
Manila, Post-Policy & 3864 (2419) & 22 & 0.1333333 & NA & NA & NA & 22 (13.3%) & NA & NA & NA & 3 & 0.0181818 & 3 (1.8%) & NA & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 22 & 0.1333333 & NA & NA & NA & 22 (13.3%) & NA & NA & NA & 5 & 0.0303030 & NA & 5 (3.0%) & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 22 & 0.1333333 & NA & NA & NA & 22 (13.3%) & NA & NA & NA & 33 & 0.2000000 & NA & NA & 33 (20.0%) & NA\\
Manila, Post-Policy & 3864 (2419) & 22 & 0.1333333 & NA & NA & NA & 22 (13.3%) & NA & NA & NA & 124 & 0.7515152 & NA & NA & NA & 124 (75.2%)\\
Manila, Post-Policy & 3864 (2419) & 11 & 0.0666667 & NA & NA & NA & NA & 11 (6.7%) & NA & NA & 3 & 0.0181818 & 3 (1.8%) & NA & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 11 & 0.0666667 & NA & NA & NA & NA & 11 (6.7%) & NA & NA & 5 & 0.0303030 & NA & 5 (3.0%) & NA & NA\\
Manila, Post-Policy & 3864 (2419) & 11 & 0.0666667 & NA & NA & NA & NA & 11 (6.7%) & NA & NA & 33 & 0.2000000 & NA & NA & 33 (20.0%) & NA\\
Manila, Post-Policy & 3864 (2419) & 11 & 0.0666667 & NA & NA & NA & NA & 11 (6.7%) & NA & NA & 124 & 0.7515152 & NA & NA & NA & 124 (75.2%)\\
Manila, Pre-Policy & 4767 (3089) & 10 & 0.0480769 & 10 (4.8%) & NA & NA & NA & NA & NA & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 10 & 0.0480769 & 10 (4.8%) & NA & NA & NA & NA & NA & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 10 & 0.0480769 & 10 (4.8%) & NA & NA & NA & NA & NA & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 10 & 0.0480769 & 10 (4.8%) & NA & NA & NA & NA & NA & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Manila, Pre-Policy & 4767 (3089) & 35 & 0.1682692 & NA & 35 (16.8%) & NA & NA & NA & NA & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 35 & 0.1682692 & NA & 35 (16.8%) & NA & NA & NA & NA & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 35 & 0.1682692 & NA & 35 (16.8%) & NA & NA & NA & NA & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 35 & 0.1682692 & NA & 35 (16.8%) & NA & NA & NA & NA & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Manila, Pre-Policy & 4767 (3089) & 122 & 0.5865385 & NA & NA & 122 (58.7%) & NA & NA & NA & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 122 & 0.5865385 & NA & NA & 122 (58.7%) & NA & NA & NA & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 122 & 0.5865385 & NA & NA & 122 (58.7%) & NA & NA & NA & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 122 & 0.5865385 & NA & NA & 122 (58.7%) & NA & NA & NA & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Manila, Pre-Policy & 4767 (3089) & 28 & 0.1346154 & NA & NA & NA & 28 (13.5%) & NA & NA & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 28 & 0.1346154 & NA & NA & NA & 28 (13.5%) & NA & NA & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 28 & 0.1346154 & NA & NA & NA & 28 (13.5%) & NA & NA & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 28 & 0.1346154 & NA & NA & NA & 28 (13.5%) & NA & NA & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Manila, Pre-Policy & 4767 (3089) & 12 & 0.0576923 & NA & NA & NA & NA & 12 (5.8%) & NA & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 12 & 0.0576923 & NA & NA & NA & NA & 12 (5.8%) & NA & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 12 & 0.0576923 & NA & NA & NA & NA & 12 (5.8%) & NA & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 12 & 0.0576923 & NA & NA & NA & NA & 12 (5.8%) & NA & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Manila, Pre-Policy & 4767 (3089) & 1 & 0.0048077 & NA & NA & NA & NA & NA & 1 (0.5%) & NA & 1 & 0.0047847 & 1 (0.5%) & NA & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 1 & 0.0048077 & NA & NA & NA & NA & NA & 1 (0.5%) & NA & 4 & 0.0191388 & NA & 4 (1.9%) & NA & NA\\
Manila, Pre-Policy & 4767 (3089) & 1 & 0.0048077 & NA & NA & NA & NA & NA & 1 (0.5%) & NA & 38 & 0.1818182 & NA & NA & 38 (18.2%) & NA\\
Manila, Pre-Policy & 4767 (3089) & 1 & 0.0048077 & NA & NA & NA & NA & NA & 1 (0.5%) & NA & 166 & 0.7942584 & NA & NA & NA & 166 (79.4%)\\
Non-NCR, Post-Policy & 5254 (3322) & 116 & 0.0341478 & 116 (3.4%) & NA & NA & NA & NA & NA & NA & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 116 & 0.0341478 & 116 (3.4%) & NA & NA & NA & NA & NA & NA & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 116 & 0.0341478 & 116 (3.4%) & NA & NA & NA & NA & NA & NA & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 116 & 0.0341478 & 116 (3.4%) & NA & NA & NA & NA & NA & NA & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Post-Policy & 5254 (3322) & 584 & 0.1719164 & NA & 584 (17.2%) & NA & NA & NA & NA & NA & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 584 & 0.1719164 & NA & 584 (17.2%) & NA & NA & NA & NA & NA & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 584 & 0.1719164 & NA & 584 (17.2%) & NA & NA & NA & NA & NA & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 584 & 0.1719164 & NA & 584 (17.2%) & NA & NA & NA & NA & NA & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Post-Policy & 5254 (3322) & 1990 & 0.5858110 & NA & NA & 1990 (58.6%) & NA & NA & NA & NA & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 1990 & 0.5858110 & NA & NA & 1990 (58.6%) & NA & NA & NA & NA & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 1990 & 0.5858110 & NA & NA & 1990 (58.6%) & NA & NA & NA & NA & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 1990 & 0.5858110 & NA & NA & 1990 (58.6%) & NA & NA & NA & NA & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Post-Policy & 5254 (3322) & 509 & 0.1498381 & NA & NA & NA & 509 (15.0%) & NA & NA & NA & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 509 & 0.1498381 & NA & NA & NA & 509 (15.0%) & NA & NA & NA & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 509 & 0.1498381 & NA & NA & NA & 509 (15.0%) & NA & NA & NA & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 509 & 0.1498381 & NA & NA & NA & 509 (15.0%) & NA & NA & NA & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Post-Policy & 5254 (3322) & 184 & 0.0541654 & NA & NA & NA & NA & 184 (5.4%) & NA & NA & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 184 & 0.0541654 & NA & NA & NA & NA & 184 (5.4%) & NA & NA & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 184 & 0.0541654 & NA & NA & NA & NA & 184 (5.4%) & NA & NA & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 184 & 0.0541654 & NA & NA & NA & NA & 184 (5.4%) & NA & NA & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Post-Policy & 5254 (3322) & 14 & 0.0041213 & NA & NA & NA & NA & NA & NA & 14 (0.4%) & 95 & 0.0278592 & 95 (2.8%) & NA & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 14 & 0.0041213 & NA & NA & NA & NA & NA & NA & 14 (0.4%) & 117 & 0.0343109 & NA & 117 (3.4%) & NA & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 14 & 0.0041213 & NA & NA & NA & NA & NA & NA & 14 (0.4%) & 1141 & 0.3346041 & NA & NA & 1141 (33.5%) & NA\\
Non-NCR, Post-Policy & 5254 (3322) & 14 & 0.0041213 & NA & NA & NA & NA & NA & NA & 14 (0.4%) & 2057 & 0.6032258 & NA & NA & NA & 2057 (60.3%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 345 & 0.0465839 & 345 (4.7%) & NA & NA & NA & NA & NA & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 345 & 0.0465839 & 345 (4.7%) & NA & NA & NA & NA & NA & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 345 & 0.0465839 & 345 (4.7%) & NA & NA & NA & NA & NA & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 345 & 0.0465839 & 345 (4.7%) & NA & NA & NA & NA & NA & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 1256 & 0.1695922 & NA & 1256 (17.0%) & NA & NA & NA & NA & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 1256 & 0.1695922 & NA & 1256 (17.0%) & NA & NA & NA & NA & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 1256 & 0.1695922 & NA & 1256 (17.0%) & NA & NA & NA & NA & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 1256 & 0.1695922 & NA & 1256 (17.0%) & NA & NA & NA & NA & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 4397 & 0.5937078 & NA & NA & 4397 (59.4%) & NA & NA & NA & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 4397 & 0.5937078 & NA & NA & 4397 (59.4%) & NA & NA & NA & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 4397 & 0.5937078 & NA & NA & 4397 (59.4%) & NA & NA & NA & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 4397 & 0.5937078 & NA & NA & 4397 (59.4%) & NA & NA & NA & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 970 & 0.1309749 & NA & NA & NA & 970 (13.1%) & NA & NA & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 970 & 0.1309749 & NA & NA & NA & 970 (13.1%) & NA & NA & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 970 & 0.1309749 & NA & NA & NA & 970 (13.1%) & NA & NA & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 970 & 0.1309749 & NA & NA & NA & 970 (13.1%) & NA & NA & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 407 & 0.0549554 & NA & NA & NA & NA & 407 (5.5%) & NA & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 407 & 0.0549554 & NA & NA & NA & NA & 407 (5.5%) & NA & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 407 & 0.0549554 & NA & NA & NA & NA & 407 (5.5%) & NA & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 407 & 0.0549554 & NA & NA & NA & NA & 407 (5.5%) & NA & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 13 & 0.0017553 & NA & NA & NA & NA & NA & NA & 13 (0.2%) & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 13 & 0.0017553 & NA & NA & NA & NA & NA & NA & 13 (0.2%) & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 13 & 0.0017553 & NA & NA & NA & NA & NA & NA & 13 (0.2%) & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 13 & 0.0017553 & NA & NA & NA & NA & NA & NA & 13 (0.2%) & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Non-NCR, Pre-Policy & 5754 (3400) & 18 & 0.0024305 & NA & NA & NA & NA & NA & 18 (0.2%) & NA & 68 & 0.0091435 & 68 (0.9%) & NA & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 18 & 0.0024305 & NA & NA & NA & NA & NA & 18 (0.2%) & NA & 136 & 0.0182869 & NA & 136 (1.8%) & NA & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 18 & 0.0024305 & NA & NA & NA & NA & NA & 18 (0.2%) & NA & 1869 & 0.2513110 & NA & NA & 1869 (25.1%) & NA\\
Non-NCR, Pre-Policy & 5754 (3400) & 18 & 0.0024305 & NA & NA & NA & NA & NA & 18 (0.2%) & NA & 5364 & 0.7212586 & NA & NA & NA & 5364 (72.1%)\\
Other NCR, Post-Policy & 5379 (3412) & 34 & 0.2500000 & NA & 34 (25.0%) & NA & NA & NA & NA & NA & 3 & 0.0220588 & 3 (2.2%) & NA & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 34 & 0.2500000 & NA & 34 (25.0%) & NA & NA & NA & NA & NA & 5 & 0.0367647 & NA & 5 (3.7%) & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 34 & 0.2500000 & NA & 34 (25.0%) & NA & NA & NA & NA & NA & 52 & 0.3823529 & NA & NA & 52 (38.2%) & NA\\
Other NCR, Post-Policy & 5379 (3412) & 34 & 0.2500000 & NA & 34 (25.0%) & NA & NA & NA & NA & NA & 76 & 0.5588235 & NA & NA & NA & 76 (55.9%)\\
Other NCR, Post-Policy & 5379 (3412) & 84 & 0.6176471 & NA & NA & 84 (61.8%) & NA & NA & NA & NA & 3 & 0.0220588 & 3 (2.2%) & NA & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 84 & 0.6176471 & NA & NA & 84 (61.8%) & NA & NA & NA & NA & 5 & 0.0367647 & NA & 5 (3.7%) & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 84 & 0.6176471 & NA & NA & 84 (61.8%) & NA & NA & NA & NA & 52 & 0.3823529 & NA & NA & 52 (38.2%) & NA\\
Other NCR, Post-Policy & 5379 (3412) & 84 & 0.6176471 & NA & NA & 84 (61.8%) & NA & NA & NA & NA & 76 & 0.5588235 & NA & NA & NA & 76 (55.9%)\\
Other NCR, Post-Policy & 5379 (3412) & 15 & 0.1102941 & NA & NA & NA & 15 (11.0%) & NA & NA & NA & 3 & 0.0220588 & 3 (2.2%) & NA & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 15 & 0.1102941 & NA & NA & NA & 15 (11.0%) & NA & NA & NA & 5 & 0.0367647 & NA & 5 (3.7%) & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 15 & 0.1102941 & NA & NA & NA & 15 (11.0%) & NA & NA & NA & 52 & 0.3823529 & NA & NA & 52 (38.2%) & NA\\
Other NCR, Post-Policy & 5379 (3412) & 15 & 0.1102941 & NA & NA & NA & 15 (11.0%) & NA & NA & NA & 76 & 0.5588235 & NA & NA & NA & 76 (55.9%)\\
Other NCR, Post-Policy & 5379 (3412) & 3 & 0.0220588 & NA & NA & NA & NA & 3 (2.2%) & NA & NA & 3 & 0.0220588 & 3 (2.2%) & NA & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 3 & 0.0220588 & NA & NA & NA & NA & 3 (2.2%) & NA & NA & 5 & 0.0367647 & NA & 5 (3.7%) & NA & NA\\
Other NCR, Post-Policy & 5379 (3412) & 3 & 0.0220588 & NA & NA & NA & NA & 3 (2.2%) & NA & NA & 52 & 0.3823529 & NA & NA & 52 (38.2%) & NA\\
Other NCR, Post-Policy & 5379 (3412) & 3 & 0.0220588 & NA & NA & NA & NA & 3 (2.2%) & NA & NA & 76 & 0.5588235 & NA & NA & NA & 76 (55.9%)\\
Other NCR, Pre-Policy & 5816 (3592) & 2 & 0.0162602 & 2 (1.6%) & NA & NA & NA & 2 (1.6%) & NA & NA & 1 & 0.0081301 & 1 (0.8%) & NA & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 2 & 0.0162602 & 2 (1.6%) & NA & NA & NA & 2 (1.6%) & NA & NA & 3 & 0.0243902 & NA & 3 (2.4%) & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 2 & 0.0162602 & 2 (1.6%) & NA & NA & NA & 2 (1.6%) & NA & NA & 31 & 0.2520325 & NA & NA & 31 (25.2%) & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 2 & 0.0162602 & 2 (1.6%) & NA & NA & NA & 2 (1.6%) & NA & NA & 88 & 0.7154472 & NA & NA & NA & 88 (71.5%)\\
Other NCR, Pre-Policy & 5816 (3592) & 25 & 0.2032520 & NA & 25 (20.3%) & NA & NA & NA & NA & NA & 1 & 0.0081301 & 1 (0.8%) & NA & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 25 & 0.2032520 & NA & 25 (20.3%) & NA & NA & NA & NA & NA & 3 & 0.0243902 & NA & 3 (2.4%) & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 25 & 0.2032520 & NA & 25 (20.3%) & NA & NA & NA & NA & NA & 31 & 0.2520325 & NA & NA & 31 (25.2%) & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 25 & 0.2032520 & NA & 25 (20.3%) & NA & NA & NA & NA & NA & 88 & 0.7154472 & NA & NA & NA & 88 (71.5%)\\
Other NCR, Pre-Policy & 5816 (3592) & 76 & 0.6178862 & NA & NA & 76 (61.8%) & NA & NA & NA & NA & 1 & 0.0081301 & 1 (0.8%) & NA & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 76 & 0.6178862 & NA & NA & 76 (61.8%) & NA & NA & NA & NA & 3 & 0.0243902 & NA & 3 (2.4%) & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 76 & 0.6178862 & NA & NA & 76 (61.8%) & NA & NA & NA & NA & 31 & 0.2520325 & NA & NA & 31 (25.2%) & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 76 & 0.6178862 & NA & NA & 76 (61.8%) & NA & NA & NA & NA & 88 & 0.7154472 & NA & NA & NA & 88 (71.5%)\\
Other NCR, Pre-Policy & 5816 (3592) & 18 & 0.1463415 & NA & NA & NA & 18 (14.6%) & NA & NA & NA & 1 & 0.0081301 & 1 (0.8%) & NA & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 18 & 0.1463415 & NA & NA & NA & 18 (14.6%) & NA & NA & NA & 3 & 0.0243902 & NA & 3 (2.4%) & NA & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 18 & 0.1463415 & NA & NA & NA & 18 (14.6%) & NA & NA & NA & 31 & 0.2520325 & NA & NA & 31 (25.2%) & NA\\
Other NCR, Pre-Policy & 5816 (3592) & 18 & 0.1463415 & NA & NA & NA & 18 (14.6%) & NA & NA & NA & 88 & 0.7154472 & NA & NA & NA & 88 (71.5%)\\
\bottomrule
\end{tabular}}
\end{table}

### 2025-08-07 data-analysis/outputs/dhs_summary_table_simple.tex
add output tables for birthrate and DHS descriptive stats
Added LaTeX output files for census birthrate changes and various DHS descriptive and summary statistics tables. These outputs summarize birth rates by region and period, and provide detailed descriptive statistics for outcomes by location and policy period.

[1] "Non-NCR, Pre-Policy"  "Non-NCR, Post-Policy"
[1] "Non-NCR, Pre-Policy"  "Non-NCR, Post-Policy"
  [1] "case_id"                       "cluster_id"                    "household_id"                 
  [4] "sample_weight"                 "mother_age_current"            "residence_type"               
  [7] "household_size"                "household_u5_count"            "region"                       
 [10] "province"                      "municipality"                  "wealth_index"                 
 [13] "wealth_score"                  "birth_index"                   "birth_order"                  
 [16] "is_twin"                       "birth_month"                   "birth_year"                   
 [19] "birth_date"                    "child_sex"                     "child_alive"                  
 [22] "age_at_death"                  "child_age"                     "preceding_birth_interval"     
 [25] "succeeding_birth_interval"     "delivery_place"                "birth_size"                   
 [28] "birth_weight"                  "received_bcg"                  "bcg_day"                      
 [31] "bcg_month"                     "bcg_year"                      "received_dpt1"                
 [34] "dpt1_day"                      "dpt1_month"                    "dpt1_year"                    
 [37] "received_polio1"               "polio1_day"                    "polio1_month"                 
 [40] "polio1_year"                   "received_dpt2"                 "dpt2_day"                     
 [43] "dpt2_month"                    "dpt2_year"                     "received_polio2"              
 [46] "polio2_day"                    "polio2_month"                  "polio2_year"                  
 [49] "received_dpt3"                 "dpt3_day"                      "dpt3_month"                   
 [52] "dpt3_year"                     "received_polio3"               "polio3_day"                   
 [55] "polio3_month"                  "polio3_year"                   "received_measles"             
 [58] "measles_day"                   "measles_month"                 "measles_year"                 
 [61] "ever_vaccinated"               "diarrhea_recent"               "fever_recent"                 
 [64] "cough_recent"                  "total_children_ever"           "sons_at_home"                 
 [67] "daughters_at_home"             "sons_died"                     "daughters_died"               
 [70] "births_last5"                  "births_last1"                  "births_this_month"            
 [73] "mother_age_first_birth"        "currently_pregnant"            "living_children"              
 [76] "living_plus_preg"              "living_plus_preg6"             "contraceptive_method"         
 [79] "contraceptive_intent"          "wanted_last_child"             "mother_educ_level"            
 [82] "mother_literate"               "mother_educ_attain"            "marital_status"               
 [85] "mother_age_first_sex"          "ideal_children"                "partner_educ_level"           
 [88] "partner_occupation"            "mother_occupation"             "mother_work_location"         
 [91] "survey_year"                   "region_dhs.x"                  "admin_region_name.x"          
 [94] "province_dhs.x"                "province_name.x"               "municipality_dhs.x"           
 [97] "municipality_name.x"           "barangay_name.x"               "latitude.x"                   
[100] "longitude.x"                   "altitude_gps.x"                "altitude_dem.x"               
[103] "cc_fips.x"                     "source.x"                      "urban_rural.x"                
[106] "datum.x"                       "region_dhs.y"                  "admin_region_name.y"          
[109] "province_dhs.y"                "province_name.y"               "municipality_dhs.y"           
[112] "municipality_name.y"           "barangay_name.y"               "latitude.y"                   
[115] "longitude.y"                   "altitude_gps.y"                "altitude_dem.y"               
[118] "cc_fips.y"                     "source.y"                      "urban_rural.y"                
[121] "datum.y"                       "region_dhs"                    "admin_region_name"            
[124] "province_dhs"                  "province_name"                 "municipality_dhs"             
[127] "municipality_name"             "barangay_name"                 "latitude"                     
[130] "longitude"                     "altitude_gps"                  "altitude_dem"                 
[133] "cc_fips"                       "source"                        "urban_rural"                  
[136] "datum"                         "manila"                        "treated"                      
[139] "birth_year_relative_to_policy" "post_policy"                   "urban"                        
[142] "mother_educ_years"             "number_of_siblings"            "ncr"                          
[145] "num_vaccines"                  "location_group"                "group"                        
  [1] "case_id"                       "cluster_id"                    "household_id"                 
  [4] "sample_weight"                 "mother_age_current"            "residence_type"               
  [7] "household_size"                "household_u5_count"            "region"                       
 [10] "province"                      "municipality"                  "wealth_index"                 
 [13] "wealth_score"                  "birth_index"                   "birth_order"                  
 [16] "is_twin"                       "birth_month"                   "birth_year"                   
 [19] "birth_date"                    "child_sex"                     "child_alive"                  
 [22] "age_at_death"                  "child_age"                     "preceding_birth_interval"     
 [25] "succeeding_birth_interval"     "delivery_place"                "birth_size"                   
 [28] "birth_weight"                  "received_bcg"                  "bcg_day"                      
 [31] "bcg_month"                     "bcg_year"                      "received_dpt1"                
 [34] "dpt1_day"                      "dpt1_month"                    "dpt1_year"                    
 [37] "received_polio1"               "polio1_day"                    "polio1_month"                 
 [40] "polio1_year"                   "received_dpt2"                 "dpt2_day"                     
 [43] "dpt2_month"                    "dpt2_year"                     "received_polio2"              
 [46] "polio2_day"                    "polio2_month"                  "polio2_year"                  
 [49] "received_dpt3"                 "dpt3_day"                      "dpt3_month"                   
 [52] "dpt3_year"                     "received_polio3"               "polio3_day"                   
 [55] "polio3_month"                  "polio3_year"                   "received_measles"             
 [58] "measles_day"                   "measles_month"                 "measles_year"                 
 [61] "ever_vaccinated"               "diarrhea_recent"               "fever_recent"                 
 [64] "cough_recent"                  "total_children_ever"           "sons_at_home"                 
 [67] "daughters_at_home"             "sons_died"                     "daughters_died"               
 [70] "births_last5"                  "births_last1"                  "births_this_month"            
 [73] "mother_age_first_birth"        "currently_pregnant"            "living_children"              
 [76] "living_plus_preg"              "living_plus_preg6"             "contraceptive_method"         
 [79] "contraceptive_intent"          "wanted_last_child"             "mother_educ_level"            
 [82] "mother_literate"               "mother_educ_attain"            "marital_status"               
 [85] "mother_age_first_sex"          "ideal_children"                "partner_educ_level"           
 [88] "partner_occupation"            "mother_occupation"             "mother_work_location"         
 [91] "survey_year"                   "region_dhs.x"                  "admin_region_name.x"          
 [94] "province_dhs.x"                "province_name.x"               "municipality_dhs.x"           
 [97] "municipality_name.x"           "barangay_name.x"               "latitude.x"                   
[100] "longitude.x"                   "altitude_gps.x"                "altitude_dem.x"               
[103] "cc_fips.x"                     "source.x"                      "urban_rural.x"                
[106] "datum.x"                       "region_dhs.y"                  "admin_region_name.y"          
[109] "province_dhs.y"                "province_name.y"               "municipality_dhs.y"           
[112] "municipality_name.y"           "barangay_name.y"               "latitude.y"                   
[115] "longitude.y"                   "altitude_gps.y"                "altitude_dem.y"               
[118] "cc_fips.y"                     "source.y"                      "urban_rural.y"                
[121] "datum.y"                       "region_dhs"                    "admin_region_name"            
[124] "province_dhs"                  "province_name"                 "municipality_dhs"             
[127] "municipality_name"             "barangay_name"                 "latitude"                     
[130] "longitude"                     "altitude_gps"                  "altitude_dem"                 
[133] "cc_fips"                       "source"                        "urban_rural"                  
[136] "datum"                         "manila"                        "treated"                      
[139] "birth_year_relative_to_policy" "post_policy"                   "urban"                        
[142] "mother_educ_years"             "number_of_siblings"            "ncr"                          
[145] "num_vaccines"                  "location_group"                "group"                        
             used    (Mb) gc trigger    (Mb)   max used    (Mb)
Ncells    3222493   172.1    5740976   306.7    5740976   306.7
Vcells 2622053306 20004.7 5428700101 41417.7 5975044060 45586.0
[1] 7
[1] 7
[1] 7

|Variable                 |123                  |136                  |151                  |209                  |3410                 |7437                 |
|:------------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|
|Birth weight (grams)     |5816.11 \\ (3592.43) |5379.28 \\ (3411.51) |3937.93 \\ (2439.54) |4766.51 \\ (3089.10) |5254.08 \\ (3321.64) |5754.47 \\ (3399.87) |
|1 - Very small           |2 (1.6%)             |NA                   |1 (0.7%)             |10 (4.8%)            |116 (3.4%)           |345 (4.7%)           |
|2 - Smaller than average |25 (20.3%)           |34 (25.0%)           |38 (25.2%)           |35 (16.8%)           |584 (17.2%)          |1256 (17.0%)         |
|3 - Average size         |76 (61.8%)           |84 (61.8%)           |84 (55.6%)           |122 (58.7%)          |1990 (58.6%)         |4397 (59.4%)         |
|4 - Larger than average  |18 (14.6%)           |15 (11.0%)           |18 (11.9%)           |28 (13.5%)           |509 (15.0%)          |970 (13.1%)          |
|5 - Very large           |2 (1.6%)             |3 (2.2%)             |10 (6.6%)            |12 (5.8%)            |184 (5.4%)           |407 (5.5%)           |
|9 - Missing              |NA                   |NA                   |NA                   |1 (0.5%)             |NA                   |18 (0.2%)            |
|8 - Don’t know           |NA                   |NA                   |NA                   |NA                   |14 (0.4%)            |13 (0.2%)            |
|1–2 Vaccines             |1 (0.8%)             |3 (2.2%)             |3 (2.0%)             |1 (0.5%)             |95 (2.8%)            |68 (0.9%)            |
|3–4 Vaccines             |3 (2.4%)             |5 (3.7%)             |5 (3.3%)             |4 (1.9%)             |117 (3.4%)           |136 (1.8%)           |
|5+ Vaccines              |31 (25.2%)           |52 (38.2%)           |29 (19.2%)           |38 (18.2%)           |1141 (33.5%)         |1869 (25.1%)         |
|No Vaccines              |88 (71.5%)           |76 (55.9%)           |114 (75.5%)          |166 (79.4%)          |2057 (60.3%)         |5364 (72.1%)         |

|Variable                 |123                  |136                  |151                  |209                  |3410                 |7437                 |
|:------------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|
|Birth weight (grams)     |5816.11 \\ (3592.43) |5379.28 \\ (3411.51) |3937.93 \\ (2439.54) |4766.51 \\ (3089.10) |5254.08 \\ (3321.64) |5754.47 \\ (3399.87) |
|1 - Very small           |2 (1.6%)             |NA                   |1 (0.7%)             |10 (4.8%)            |116 (3.4%)           |345 (4.7%)           |
|2 - Smaller than average |25 (20.3%)           |34 (25.0%)           |38 (25.2%)           |35 (16.8%)           |584 (17.2%)          |1256 (17.0%)         |
|3 - Average size         |76 (61.8%)           |84 (61.8%)           |84 (55.6%)           |122 (58.7%)          |1990 (58.6%)         |4397 (59.4%)         |
|4 - Larger than average  |18 (14.6%)           |15 (11.0%)           |18 (11.9%)           |28 (13.5%)           |509 (15.0%)          |970 (13.1%)          |
|5 - Very large           |2 (1.6%)             |3 (2.2%)             |10 (6.6%)            |12 (5.8%)            |184 (5.4%)           |407 (5.5%)           |
|9 - Missing              |NA                   |NA                   |NA                   |1 (0.5%)             |NA                   |18 (0.2%)            |
|8 - Don’t know           |NA                   |NA                   |NA                   |NA                   |14 (0.4%)            |13 (0.2%)            |
|1–2 Vaccines             |1 (0.8%)             |3 (2.2%)             |3 (2.0%)             |1 (0.5%)             |95 (2.8%)            |68 (0.9%)            |
|3–4 Vaccines             |3 (2.4%)             |5 (3.7%)             |5 (3.3%)             |4 (1.9%)             |117 (3.4%)           |136 (1.8%)           |
|5+ Vaccines              |31 (25.2%)           |52 (38.2%)           |29 (19.2%)           |38 (18.2%)           |1141 (33.5%)         |1869 (25.1%)         |
|No Vaccines              |88 (71.5%)           |76 (55.9%)           |114 (75.5%)          |166 (79.4%)          |2057 (60.3%)         |5364 (72.1%)         |
           used  (Mb) gc trigger    (Mb)   max used    (Mb)
Ncells  3222541 172.2    5740976   306.7    5740976   306.7
Vcells 24136690 184.2 4342960081 33134.2 5975044060 45586.0
           used  (Mb) gc trigger    (Mb)   max used    (Mb)
Ncells  3220838 172.1    5740976   306.7    5740976   306.7
Vcells 24001138 183.2 3474368065 26507.4 5975044060 45586.0
 [1] "\\begin{table}[H][!h]"                                          
 [2] "\\centering"                                                    
 [3] "\\caption{Average Birth Weight by Location and Policy Period}"  
 [4] "\\centering"                                                    
 [5] "\\begin{tabular}[t]{lc}"                                        
 [6] "\\toprule"                                                      
 [7] "Group & Birth weight (g)\\\\"                                   
 [8] "\\midrule"                                                      
 [9] "Manila, Pre-Policy & \\makecell{4766.51 \\\\ (3089.10)}\\\\"    
[10] "Manila, Post-Policy & \\makecell{3864.21 \\\\ (2419.37)}\\\\"   
[11] "Other NCR, Pre-Policy & \\makecell{5816.11 \\\\ (3592.43)}\\\\" 
[12] "Other NCR, Post-Policy & \\makecell{5379.28 \\\\ (3411.51)}\\\\"
[13] "Non-NCR, Pre-Policy & \\makecell{5754.47 \\\\ (3399.87)}\\\\"   
[14] "Non-NCR, Post-Policy & \\makecell{5254.08 \\\\ (3321.64)}\\\\"  
[15] "\\bottomrule"                                                   
[16] "\\end{tabular}"                                                 
[17] "\\end{table}"                                                   
 [1] "\\begin{table}[H][!h]"                                          
 [2] "\\centering"                                                    
 [3] "\\caption{Average Birth Weight by Location and Policy Period}"  
 [4] "\\centering"                                                    
 [5] "\\begin{tabular}[t]{lc}"                                        
 [6] "\\toprule"                                                      
 [7] "Group & Birth weight (g)\\\\"                                   
 [8] "\\midrule"                                                      
 [9] "Manila, Pre-Policy & \\makecell{4766.51 \\\\ (3089.10)}\\\\"    
[10] "Manila, Post-Policy & \\makecell{3864.21 \\\\ (2419.37)}\\\\"   
[11] "Other NCR, Pre-Policy & \\makecell{5816.11 \\\\ (3592.43)}\\\\" 
[12] "Other NCR, Post-Policy & \\makecell{5379.28 \\\\ (3411.51)}\\\\"
[13] "Non-NCR, Pre-Policy & \\makecell{5754.47 \\\\ (3399.87)}\\\\"   
[14] "Non-NCR, Post-Policy & \\makecell{5254.08 \\\\ (3321.64)}\\\\"  
[15] "\\bottomrule"                                                   
[16] "\\end{tabular}"                                                 
[17] "\\end{table}"                                                   
           used  (Mb) gc trigger    (Mb)   max used    (Mb)
Ncells  3220342 172.0    5740976   306.7    5740976   306.7
Vcells 24001025 183.2 2779494452 21205.9 5975044060 45586.0
# A tibble: 6 × 136
  case_id           cluster_id household_id sample_weight mother_age_current residence_type household_size
  <chr>                  <dbl>        <dbl>         <dbl>              <dbl> <dbl+lbl>               <dbl>
1 "       1   2  2"          1            2       1452369                 27 1 [urban]                   3
2 "       1   3  2"          1            3       1452369                 23 1 [urban]                   4
3 "       1   3  2"          1            3       1452369                 23 1 [urban]                   4
4 "       1   4  2"          1            4       1452369                 21 1 [urban]                  11
5 "       1   4  2"          1            4       1452369                 21 1 [urban]                  11
6 "       1   4  9"          1            4       1452369                 24 1 [urban]                  11
# ℹ 129 more variables: household_u5_count <dbl>, region <dbl+lbl>, province <dbl+lbl>, municipality <dbl>,
#   wealth_index <dbl+lbl>, wealth_score <dbl>, birth_index <dbl>, birth_order <dbl>, is_twin <dbl+lbl>,
#   birth_month <dbl>, birth_year <dbl>, birth_date <dbl>, child_sex <dbl+lbl>, child_alive <dbl+lbl>,
#   age_at_death <dbl+lbl>, child_age <dbl>, preceding_birth_interval <dbl>, succeeding_birth_interval <dbl>,
#   delivery_place <dbl+lbl>, birth_size <dbl+lbl>, birth_weight <dbl+lbl>, received_bcg <dbl+lbl>,
#   bcg_day <dbl+lbl>, bcg_month <dbl+lbl>, bcg_year <dbl+lbl>, received_dpt1 <dbl+lbl>, dpt1_day <dbl+lbl>,
#   dpt1_month <dbl+lbl>, dpt1_year <dbl+lbl>, received_polio1 <dbl+lbl>, polio1_day <dbl+lbl>, …
# ℹ Use `colnames()` to see all variable names
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
           used  (Mb) gc trigger    (Mb)   max used    (Mb)
Ncells  3366595 179.8    5740976   306.7    5740976   306.7
Vcells 24254013 185.1 2223595562 16964.7 5975044060 45586.0
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"                    
  [1] "case_id"                   "cluster_id"                "household_id"             
  [4] "sample_weight"             "mother_age_current"        "residence_type"           
  [7] "household_size"            "household_u5_count"        "region"                   
 [10] "province"                  "municipality"              "wealth_index"             
 [13] "wealth_score"              "birth_index"               "birth_order"              
 [16] "is_twin"                   "birth_month"               "birth_year"               
 [19] "birth_date"                "child_sex"                 "child_alive"              
 [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
 [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
 [28] "birth_weight"              "received_bcg"              "bcg_day"                  
 [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
 [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
 [37] "received_polio1"           "polio1_day"                "polio1_month"             
 [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
 [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
 [46] "polio2_day"                "polio2_month"              "polio2_year"              
 [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
 [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
 [55] "polio3_month"              "polio3_year"               "received_measles"         
 [58] "measles_day"               "measles_month"             "measles_year"             
 [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
 [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
 [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
 [70] "births_last5"              "births_last1"              "births_this_month"        
 [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
 [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
 [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
 [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
 [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
 [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
 [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
 [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
 [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
[121] "datum.y"                   "region_dhs"                "admin_region_name"        
[124] "province_dhs"              "province_name"             "municipality_dhs"         
[127] "municipality_name"         "barangay_name"             "latitude"                 
[130] "longitude"                 "altitude_gps"              "altitude_dem"             
[133] "cc_fips"                   "source"                    "urban_rural"              
[136] "datum"

### 2025-08-07 data-analysis/outputs/first_stage_sibship.tex
add output tables for birthrate and DHS descriptive stats
Added LaTeX output files for census birthrate changes and various DHS descriptive and summary statistics tables. These outputs summarize birth rates by region and period, and provide detailed descriptive statistics for outcomes by location and policy period.

--- 2025-08-07 data-analysis/outputs/dhs_summary_table_simple.tex
@@
-[1] "Non-NCR, Pre-Policy"  "Non-NCR, Post-Policy"
-[1] "Non-NCR, Pre-Policy"  "Non-NCR, Post-Policy"
+\begin{table}
+\centering
+\begin{talltblr}[         %% tabularray outer open
+entry=none,label=none,
+note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
+note{ }={All regressions include survey year fixed effects.},
+]                     %% tabularray outer close
+{                     %% tabularray inner open
+colspec={Q[]Q[]Q[]},
+column{2,3}={}{halign=c,},
+column{1}={}{halign=l,},
+hline{10}={1,2,3}{solid, black, 0.05em},
+}                     %% tabularray inner close
+\toprule
+& (1) & (2) \\ \midrule %% TinyTableHeader
+Manila × Post & \num{0.231} & \num{0.100}+ \\
+& (\num{0.227}) & (\num{0.077}) \\
+Mother's Education & \num{-0.393}*** & \num{-0.315}*** \\
+& (\num{<0.001}) & (\num{<0.001}) \\
+Mother's Age & \num{0.230}** & \num{0.207}** \\
+& (\num{0.003}) & (\num{0.001}) \\
+Wealth Index & \num{-0.287}*** & \num{-0.247}** \\
+& (\num{<0.001}) & (\num{0.002}) \\
+Num.Obs. & \num{20823} & \num{20823} \\
+R2 & \num{0.539} & \num{0.514} \\
+R2 Adj. & \num{0.539} & \num{0.513} \\
+R2 Within & \num{0.535} & \num{0.511} \\
+R2 Within Adj. & \num{0.535} & \num{0.511} \\
+RMSE & \num{1.65} & \num{1.53} \\
+FE: survey_year & X & X \\
+\bottomrule
+\end{talltblr}
+\end{table} 
+\begin{table}
+\centering
+\begin{talltblr}[         %% tabularray outer open
+entry=none,label=none,
+note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
+note{ }={All regressions include survey year fixed effects.},
+]                     %% tabularray outer close
+{                     %% tabularray inner open
+colspec={Q[]Q[]Q[]},
+column{2,3}={}{halign=c,},
+column{1}={}{halign=l,},
+hline{10}={1,2,3}{solid, black, 0.05em},
+}                     %% tabularray inner close
+\toprule
+& (1) & (2) \\ \midrule %% TinyTableHeader
+Manila × Post & \num{0.231} & \num{0.100}+ \\
+& (\num{0.227}) & (\num{0.077}) \\
+Mother's Education & \num{-0.393}*** & \num{-0.315}*** \\
+& (\num{<0.001}) & (\num{<0.001}) \\
+Mother's Age & \num{0.230}** & \num{0.207}** \\
+& (\num{0.003}) & (\num{0.001}) \\
+Wealth Index & \num{-0.287}*** & \num{-0.247}** \\
+& (\num{<0.001}) & (\num{0.002}) \\
+Num.Obs. & \num{20823} & \num{20823} \\
+R2 & \num{0.539} & \num{0.514} \\
+R2 Adj. & \num{0.539} & \num{0.513} \\
+R2 Within & \num{0.535} & \num{0.511} \\
+R2 Within Adj. & \num{0.535} & \num{0.511} \\
+FE: survey\_year & X & X \\
+\bottomrule
+\end{talltblr}
+\end{table} 
+\begin{table}
+\centering
+\begin{talltblr}[         %% tabularray outer open
+entry=none,label=none,
+note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
+note{ }={All regressions include survey year fixed effects.},
+]                     %% tabularray outer close
+{                     %% tabularray inner open
+colspec={Q[]Q[]Q[]},
+column{2,3}={}{halign=c,},
+column{1}={}{halign=l,},
+hline{10}={1,2,3}{solid, black, 0.05em},
+}                     %% tabularray inner close
+\toprule
+& (1) & (2) \\ \midrule %% TinyTableHeader
+Manila × Post & \num{0.100}+ & \num{0.231} \\
+& (\num{0.077}) & (\num{0.227}) \\
+Mother's Education & \num{-0.315}*** & \num{-0.393}*** \\
+& (\num{<0.001}) & (\num{<0.001}) \\
+Mother's Age & \num{0.207}** & \num{0.230}** \\
+& (\num{0.001}) & (\num{0.003}) \\
+Wealth Index & \num{-0.247}** & \num{-0.287}*** \\
+& (\num{0.002}) & (\num{<0.001}) \\
+Num.Obs. & \num{20823} & \num{20823} \\
+R2 & \num{0.514} & \num{0.539} \\
+R2 Adj. & \num{0.513} & \num{0.539} \\
+R2 Within & \num{0.511} & \num{0.535} \\
+R2 Within Adj. & \num{0.511} & \num{0.535} \\
+FE: survey\_year & X & X \\
+\bottomrule
+\end{talltblr}
+\end{table} 
+\begin{table}
+\centering
+\begin{talltblr}[         %% tabularray outer open
+entry=none,label=none,
+note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
+note{ }={All regressions include survey year fixed effects.},
+]                     %% tabularray outer close
+{                     %% tabularray inner open
+colspec={Q[]Q[]Q[]},
+column{2,3}={}{halign=c,},
+column{1}={}{halign=l,},
+hline{10}={1,2,3}{solid, black, 0.05em},
+}                     %% tabularray inner close
+\toprule
+& (1) & (2) \\ \midrule %% TinyTableHeader
+Manila × Post & \num{0.100}+ & \num{0.231} \\
+& (\num{0.077}) & (\num{0.227}) \\
+Mother's Education & \num{-0.315}*** & \num{-0.393}*** \\
+& (\num{<0.001}) & (\num{<0.001}) \\
+Mother's Age & \num{0.207}** & \num{0.230}** \\
+& (\num{0.001}) & (\num{0.003}) \\
+Wealth Index & \num{-0.247}** & \num{-0.287}*** \\
+& (\num{0.002}) & (\num{<0.001}) \\
+Num.Obs. & \num{20823} & \num{20823} \\
+R2 & \num{0.514} & \num{0.539} \\
+R2 Adj. & \num{0.513} & \num{0.539} \\
+R2 Within & \num{0.511} & \num{0.535} \\
+R2 Within Adj. & \num{0.511} & \num{0.535} \\
+FE: survey\_year & X & X \\
+\bottomrule
+\end{talltblr}
+\end{table} 
+          used  (Mb) gc trigger  (Mb) max used  (Mb)
+Ncells 2386672 127.5    3882202 207.4  3882202 207.4
+Vcells 8326641  63.6   26574801 202.8 26574145 202.8
   [1] "case_id"                   "cluster_id"                "household_id"             
@@
 [133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                         "manila"                        "treated"                      
-[139] "birth_year_relative_to_policy" "post_policy"                   "urban"                        
-[142] "mother_educ_years"             "number_of_siblings"            "ncr"                          
-[145] "num_vaccines"                  "location_group"                "group"                        
+[136] "datum"                     "number_of_siblings"        "manila"                   
+[139] "post_policy"               "manila_post"              
   [1] "case_id"                   "cluster_id"                "household_id"             
@@
 [133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                         "manila"                        "treated"                      
-[139] "birth_year_relative_to_policy" "post_policy"                   "urban"                        
-[142] "mother_educ_years"             "number_of_siblings"            "ncr"                          
-[145] "num_vaccines"                  "location_group"                "group"                        
-             used    (Mb) gc trigger    (Mb)   max used    (Mb)
-Ncells    3222493   172.1    5740976   306.7    5740976   306.7
-Vcells 2622053306 20004.7 5428700101 41417.7 5975044060 45586.0
-[1] 7
-[1] 7
-[1] 7
-
-|Variable                 |123                  |136                  |151                  |209                  |3410                 |7437                 |
-|:------------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|
-|Birth weight (grams)     |5816.11 \\ (3592.43) |5379.28 \\ (3411.51) |3937.93 \\ (2439.54) |4766.51 \\ (3089.10) |5254.08 \\ (3321.64) |5754.47 \\ (3399.87) |
-|1 - Very small           |2 (1.6%)             |NA                   |1 (0.7%)             |10 (4.8%)            |116 (3.4%)           |345 (4.7%)           |
-|2 - Smaller than average |25 (20.3%)           |34 (25.0%)           |38 (25.2%)           |35 (16.8%)           |584 (17.2%)          |1256 (17.0%)         |
-|3 - Average size         |76 (61.8%)           |84 (61.8%)           |84 (55.6%)           |122 (58.7%)          |1990 (58.6%)         |4397 (59.4%)         |
-|4 - Larger than average  |18 (14.6%)           |15 (11.0%)           |18 (11.9%)           |28 (13.5%)           |509 (15.0%)          |970 (13.1%)          |
-|5 - Very large           |2 (1.6%)             |3 (2.2%)             |10 (6.6%)            |12 (5.8%)            |184 (5.4%)           |407 (5.5%)           |
-|9 - Missing              |NA                   |NA                   |NA                   |1 (0.5%)             |NA                   |18 (0.2%)            |
-|8 - Don’t know           |NA                   |NA                   |NA                   |NA                   |14 (0.4%)            |13 (0.2%)            |
-|1–2 Vaccines             |1 (0.8%)             |3 (2.2%)             |3 (2.0%)             |1 (0.5%)             |95 (2.8%)            |68 (0.9%)            |
-|3–4 Vaccines             |3 (2.4%)             |5 (3.7%)             |5 (3.3%)             |4 (1.9%)             |117 (3.4%)           |136 (1.8%)           |
-|5+ Vaccines              |31 (25.2%)           |52 (38.2%)           |29 (19.2%)           |38 (18.2%)           |1141 (33.5%)         |1869 (25.1%)         |
-|No Vaccines              |88 (71.5%)           |76 (55.9%)           |114 (75.5%)          |166 (79.4%)          |2057 (60.3%)         |5364 (72.1%)         |
-
-|Variable                 |123                  |136                  |151                  |209                  |3410                 |7437                 |
-|:------------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|:--------------------|
-|Birth weight (grams)     |5816.11 \\ (3592.43) |5379.28 \\ (3411.51) |3937.93 \\ (2439.54) |4766.51 \\ (3089.10) |5254.08 \\ (3321.64) |5754.47 \\ (3399.87) |
-|1 - Very small           |2 (1.6%)             |NA                   |1 (0.7%)             |10 (4.8%)            |116 (3.4%)           |345 (4.7%)           |
-|2 - Smaller than average |25 (20.3%)           |34 (25.0%)           |38 (25.2%)           |35 (16.8%)           |584 (17.2%)          |1256 (17.0%)         |
-|3 - Average size         |76 (61.8%)           |84 (61.8%)           |84 (55.6%)           |122 (58.7%)          |1990 (58.6%)         |4397 (59.4%)         |
-|4 - Larger than average  |18 (14.6%)           |15 (11.0%)           |18 (11.9%)           |28 (13.5%)           |509 (15.0%)          |970 (13.1%)          |
-|5 - Very large           |2 (1.6%)             |3 (2.2%)             |10 (6.6%)            |12 (5.8%)            |184 (5.4%)           |407 (5.5%)           |
-|9 - Missing              |NA                   |NA                   |NA                   |1 (0.5%)             |NA                   |18 (0.2%)            |
-|8 - Don’t know           |NA                   |NA                   |NA                   |NA                   |14 (0.4%)            |13 (0.2%)            |
-|1–2 Vaccines             |1 (0.8%)             |3 (2.2%)             |3 (2.0%)             |1 (0.5%)             |95 (2.8%)            |68 (0.9%)            |
-|3–4 Vaccines             |3 (2.4%)             |5 (3.7%)             |5 (3.3%)             |4 (1.9%)             |117 (3.4%)           |136 (1.8%)           |
-|5+ Vaccines              |31 (25.2%)           |52 (38.2%)           |29 (19.2%)           |38 (18.2%)           |1141 (33.5%)         |1869 (25.1%)         |
-|No Vaccines              |88 (71.5%)           |76 (55.9%)           |114 (75.5%)          |166 (79.4%)          |2057 (60.3%)         |5364 (72.1%)         |
-           used  (Mb) gc trigger    (Mb)   max used    (Mb)
-Ncells  3222541 172.2    5740976   306.7    5740976   306.7
-Vcells 24136690 184.2 4342960081 33134.2 5975044060 45586.0
-           used  (Mb) gc trigger    (Mb)   max used    (Mb)
-Ncells  3220838 172.1    5740976   306.7    5740976   306.7
-Vcells 24001138 183.2 3474368065 26507.4 5975044060 45586.0
- [1] "\\begin{table}[H][!h]"                                          
- [2] "\\centering"                                                    
- [3] "\\caption{Average Birth Weight by Location and Policy Period}"  
- [4] "\\centering"                                                    
- [5] "\\begin{tabular}[t]{lc}"                                        
- [6] "\\toprule"                                                      
- [7] "Group & Birth weight (g)\\\\"                                   
- [8] "\\midrule"                                                      
- [9] "Manila, Pre-Policy & \\makecell{4766.51 \\\\ (3089.10)}\\\\"    
-[10] "Manila, Post-Policy & \\makecell{3864.21 \\\\ (2419.37)}\\\\"   
-[11] "Other NCR, Pre-Policy & \\makecell{5816.11 \\\\ (3592.43)}\\\\" 
-[12] "Other NCR, Post-Policy & \\makecell{5379.28 \\\\ (3411.51)}\\\\"
-[13] "Non-NCR, Pre-Policy & \\makecell{5754.47 \\\\ (3399.87)}\\\\"   
-[14] "Non-NCR, Post-Policy & \\makecell{5254.08 \\\\ (3321.64)}\\\\"  
-[15] "\\bottomrule"                                                   
-[16] "\\end{tabular}"                                                 
-[17] "\\end{table}"                                                   
- [1] "\\begin{table}[H][!h]"                                          
- [2] "\\centering"                                                    
- [3] "\\caption{Average Birth Weight by Location and Policy Period}"  
- [4] "\\centering"                                                    
- [5] "\\begin{tabular}[t]{lc}"                                        
- [6] "\\toprule"                                                      
- [7] "Group & Birth weight (g)\\\\"                                   
- [8] "\\midrule"                                                      
- [9] "Manila, Pre-Policy & \\makecell{4766.51 \\\\ (3089.10)}\\\\"    
-[10] "Manila, Post-Policy & \\makecell{3864.21 \\\\ (2419.37)}\\\\"   
-[11] "Other NCR, Pre-Policy & \\makecell{5816.11 \\\\ (3592.43)}\\\\" 
-[12] "Other NCR, Post-Policy & \\makecell{5379.28 \\\\ (3411.51)}\\\\"
-[13] "Non-NCR, Pre-Policy & \\makecell{5754.47 \\\\ (3399.87)}\\\\"   
-[14] "Non-NCR, Post-Policy & \\makecell{5254.08 \\\\ (3321.64)}\\\\"  
-[15] "\\bottomrule"                                                   
-[16] "\\end{tabular}"                                                 
-[17] "\\end{table}"                                                   
-           used  (Mb) gc trigger    (Mb)   max used    (Mb)
-Ncells  3220342 172.0    5740976   306.7    5740976   306.7
-Vcells 24001025 183.2 2779494452 21205.9 5975044060 45586.0
-# A tibble: 6 × 136
-  case_id           cluster_id household_id sample_weight mother_age_current residence_type household_size
-  <chr>                  <dbl>        <dbl>         <dbl>              <dbl> <dbl+lbl>               <dbl>
-1 "       1   2  2"          1            2       1452369                 27 1 [urban]                   3
-2 "       1   3  2"          1            3       1452369                 23 1 [urban]                   4
-3 "       1   3  2"          1            3       1452369                 23 1 [urban]                   4
-4 "       1   4  2"          1            4       1452369                 21 1 [urban]                  11
-5 "       1   4  2"          1            4       1452369                 21 1 [urban]                  11
-6 "       1   4  9"          1            4       1452369                 24 1 [urban]                  11
-# ℹ 129 more variables: household_u5_count <dbl>, region <dbl+lbl>, province <dbl+lbl>, municipality <dbl>,
-#   wealth_index <dbl+lbl>, wealth_score <dbl>, birth_index <dbl>, birth_order <dbl>, is_twin <dbl+lbl>,
-#   birth_month <dbl>, birth_year <dbl>, birth_date <dbl>, child_sex <dbl+lbl>, child_alive <dbl+lbl>,
-#   age_at_death <dbl+lbl>, child_age <dbl>, preceding_birth_interval <dbl>, succeeding_birth_interval <dbl>,
-#   delivery_place <dbl+lbl>, birth_size <dbl+lbl>, birth_weight <dbl+lbl>, received_bcg <dbl+lbl>,
-#   bcg_day <dbl+lbl>, bcg_month <dbl+lbl>, bcg_year <dbl+lbl>, received_dpt1 <dbl+lbl>, dpt1_day <dbl+lbl>,
-#   dpt1_month <dbl+lbl>, dpt1_year <dbl+lbl>, received_polio1 <dbl+lbl>, polio1_day <dbl+lbl>, …
-# ℹ Use `colnames()` to see all variable names
+[136] "datum"                     "number_of_siblings"        "manila"                   
+[139] "post_policy"               "manila_post"              
   [1] "case_id"                   "cluster_id"                "household_id"             
@@
 [133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
+[136] "datum"                     "number_of_siblings"        "manila"                   
+[139] "post_policy"               "manila_post"              
   [1] "case_id"                   "cluster_id"                "household_id"             
@@
 [133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-           used  (Mb) gc trigger    (Mb)   max used    (Mb)
-Ncells  3366595 179.8    5740976   306.7    5740976   306.7
-Vcells 24254013 185.1 2223595562 16964.7 5975044060 45586.0
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"                    
-  [1] "case_id"                   "cluster_id"                "household_id"             
-  [4] "sample_weight"             "mother_age_current"        "residence_type"           
-  [7] "household_size"            "household_u5_count"        "region"                   
- [10] "province"                  "municipality"              "wealth_index"             
- [13] "wealth_score"              "birth_index"               "birth_order"              
- [16] "is_twin"                   "birth_month"               "birth_year"               
- [19] "birth_date"                "child_sex"                 "child_alive"              
- [22] "age_at_death"              "child_age"                 "preceding_birth_interval" 
- [25] "succeeding_birth_interval" "delivery_place"            "birth_size"               
- [28] "birth_weight"              "received_bcg"              "bcg_day"                  
- [31] "bcg_month"                 "bcg_year"                  "received_dpt1"            
- [34] "dpt1_day"                  "dpt1_month"                "dpt1_year"                
- [37] "received_polio1"           "polio1_day"                "polio1_month"             
- [40] "polio1_year"               "received_dpt2"             "dpt2_day"                 
- [43] "dpt2_month"                "dpt2_year"                 "received_polio2"          
- [46] "polio2_day"                "polio2_month"              "polio2_year"              
- [49] "received_dpt3"             "dpt3_day"                  "dpt3_month"               
- [52] "dpt3_year"                 "received_polio3"           "polio3_day"               
- [55] "polio3_month"              "polio3_year"               "received_measles"         
- [58] "measles_day"               "measles_month"             "measles_year"             
- [61] "ever_vaccinated"           "diarrhea_recent"           "fever_recent"             
- [64] "cough_recent"              "total_children_ever"       "sons_at_home"             
- [67] "daughters_at_home"         "sons_died"                 "daughters_died"           
- [70] "births_last5"              "births_last1"              "births_this_month"        
- [73] "mother_age_first_birth"    "currently_pregnant"        "living_children"          
- [76] "living_plus_preg"          "living_plus_preg6"         "contraceptive_method"     
- [79] "contraceptive_intent"      "wanted_last_child"         "mother_educ_level"        
- [82] "mother_literate"           "mother_educ_attain"        "marital_status"           
- [85] "mother_age_first_sex"      "ideal_children"            "partner_educ_level"       
- [88] "partner_occupation"        "mother_occupation"         "mother_work_location"     
- [91] "survey_year"               "region_dhs.x"              "admin_region_name.x"      
- [94] "province_dhs.x"            "province_name.x"           "municipality_dhs.x"       
- [97] "municipality_name.x"       "barangay_name.x"           "latitude.x"               
-[100] "longitude.x"               "altitude_gps.x"            "altitude_dem.x"           
-[103] "cc_fips.x"                 "source.x"                  "urban_rural.x"            
-[106] "datum.x"                   "region_dhs.y"              "admin_region_name.y"      
-[109] "province_dhs.y"            "province_name.y"           "municipality_dhs.y"       
-[112] "municipality_name.y"       "barangay_name.y"           "latitude.y"               
-[115] "longitude.y"               "altitude_gps.y"            "altitude_dem.y"           
-[118] "cc_fips.y"                 "source.y"                  "urban_rural.y"            
-[121] "datum.y"                   "region_dhs"                "admin_region_name"        
-[124] "province_dhs"              "province_name"             "municipality_dhs"         
-[127] "municipality_name"         "barangay_name"             "latitude"                 
-[130] "longitude"                 "altitude_gps"              "altitude_dem"             
-[133] "cc_fips"                   "source"                    "urban_rural"              
-[136] "datum"
+[136] "datum"                     "number_of_siblings"        "manila"                   
+[139] "post_policy"               "manila_post"

### 2025-08-07 data-analysis/outputs/summary_table_outcomes.tex
add output tables for birthrate and DHS descriptive stats
Added LaTeX output files for census birthrate changes and various DHS descriptive and summary statistics tables. These outputs summarize birth rates by region and period, and provide detailed descriptive statistics for outcomes by location and policy period.

\begingroup\fontsize{7.5}{9.5}\selectfont

\begin{longtable}[t]{>{}l>{\centering\arraybackslash}p{1.6cm}>{\centering\arraybackslash}p{1.6cm}>{\centering\arraybackslash}p{1.6cm}>{\centering\arraybackslash}p{1.6cm}>{\centering\arraybackslash}p{1.6cm}>{\centering\arraybackslash}p{1.6cm}}
\caption{Descriptive statistics by location and policy period (outcomes only).}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Group} \\
\cmidrule(l{3pt}r{3pt}){2-7}
Variable & Manila, Pre-Policy  N = 209 & Manila, Post-Policy  N = 165 & Other NCR, Pre-Policy  N = 123 & Other NCR, Post-Policy  N = 136 & Non-NCR, Pre-Policy  N = 7,437 & Non-NCR, Post-Policy  N = 3,410\\
\midrule
\endfirsthead
\caption[]{Descriptive statistics by location and policy period (outcomes only). \textit{(continued)}}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Group} \\
\cmidrule(l{3pt}r{3pt}){2-7}
Variable & Manila, Pre-Policy  N = 209 & Manila, Post-Policy  N = 165 & Other NCR, Pre-Policy  N = 123 & Other NCR, Post-Policy  N = 136 & Non-NCR, Pre-Policy  N = 7,437 & Non-NCR, Post-Policy  N = 3,410\\
\midrule
\endhead

\endfoot
\bottomrule
\endlastfoot
\textbf{Birth weight (grams)} & 4,766.51 (3,089.10) & 3,864.21 (2,419.37) & 5,816.11 (3,592.43) & 5,379.28 (3,411.51) & 5,754.47 (3,399.87) & 5,254.08 (3,321.64)\\
\textbf{Perceived birth size} &  &  &  &  &  & \\
\hspace{1em}1 - Very small & 10 (4.8\%) & 1 (0.6\%) & 2 (1.6\%) & 0 (0\%) & 345 (4.7\%) & 116 (3.4\%)\\
\hspace{1em}2 - Smaller than average & 35 (17\%) & 39 (24\%) & 25 (20\%) & 34 (25\%) & 1,256 (17\%) & 584 (17\%)\\
\hspace{1em}3 - Average size & 122 (59\%) & 92 (56\%) & 76 (62\%) & 84 (62\%) & 4,397 (59\%) & 1,990 (59\%)\\
\addlinespace
\hspace{1em}4 - Larger than average & 28 (13\%) & 22 (13\%) & 18 (15\%) & 15 (11\%) & 970 (13\%) & 509 (15\%)\\
\hspace{1em}5 - Very large & 12 (5.8\%) & 11 (6.7\%) & 2 (1.6\%) & 3 (2.2\%) & 407 (5.5\%) & 184 (5.4\%)\\
\hspace{1em}8 - Don't know & 0 (0\%) & 0 (0\%) & 0 (0\%) & 0 (0\%) & 13 (0.2\%) & 14 (0.4\%)\\
\hspace{1em}9 - Missing & 1 (0.5\%) & 0 (0\%) & 0 (0\%) & 0 (0\%) & 18 (0.2\%) & 0 (0\%)\\
\textbf{Total number of vaccines received} &  &  &  &  &  & \\
\addlinespace
\hspace{1em}1–2 Vaccines & 1 (0.5\%) & 3 (1.8\%) & 1 (0.8\%) & 3 (2.2\%) & 68 (0.9\%) & 95 (2.8\%)\\
\hspace{1em}3–4 Vaccines & 4 (1.9\%) & 5 (3.0\%) & 3 (2.4\%) & 5 (3.7\%) & 136 (1.8\%) & 117 (3.4\%)\\
\hspace{1em}5+ Vaccines & 38 (18\%) & 33 (20\%) & 31 (25\%) & 52 (38\%) & 1,869 (25\%) & 1,141 (33\%)\\
\hspace{1em}No Vaccines & 166 (79\%) & 124 (75\%) & 88 (72\%) & 76 (56\%) & 5,364 (72\%) & 2,057 (60\%)\\*
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{1} Mean (SD); n (\%)}\\
\end{longtable}
\endgroup{}

### 2025-08-07 data-analysis/scripts/98c_paper_outputs.qmd
add scripts for generating paper output tables
Introduces three new scripts: 98a_paper_outputs.R for summary tables of birth weight, perceived birth size, and vaccine categories; 98b_paper_outputs.R for census-based birth rate calculations and LaTeX export; and 98c_paper_outputs.qmd for first-stage regression analysis and LaTeX table output. These scripts automate the creation of key tables for the paper using cleaned data sources.

---
title: "98c_paper_outputs"
format: pdf
---

```{r}
# Core data wrangling
library(dplyr)
library(stats)
library(MASS)
library(broom)
library(modelsummary)   
library(ggplot2)

```

```{r}
dhs1 <- readRDS("../clean-data/dhs_kr_children_with_geo.rds") |>
  filter(survey_year %in% c(1998, 2003, 2008)) |>
    mutate(
      number_of_siblings = living_children - 1)

dhs2 <- dhs1 %>%
  mutate(
    manila = as.character(municipality_name) == "Manila",
    post_policy = birth_year >= 2000,
    manila_post = manila * post_policy
  )
```

```{r}
library(fixest)
library(kableExtra)

# 1. First-stage: total children ever
first_stage1 <- feols(
  total_children_ever ~ manila + post_policy + manila_post +
    mother_educ_attain + mother_age_current + wealth_index | survey_year,
  data = dhs2
)

# 2. First-stage: number of living children
first_stage2 <- feols(
  living_children ~ manila + post_policy + manila_post +
    mother_educ_attain + mother_age_current + wealth_index | survey_year,
  data = dhs2
)

# Name your models
models <- list(
  "Total Children Ever" = first_stage1,
  "Living Children" = first_stage2
)

# Prepare LaTeX table string
table_latex <- modelsummary(
  models,
  output = "latex",
  statistic = "p.value",
  coef_map = c(
    manila = "Manila",
    post_policy = "Post-Policy",
    manila_post = "Manila × Post",
    mother_educ_attain = "Mother's Education",
    mother_age_current = "Mother's Age",
    wealth_index = "Wealth Index"
  ),
  gof_omit = "IC|Log|Std.Errors",
  notes = list("All regressions include survey year fixed effects."),
  fmt = 3,
  stars = TRUE,
  escape = FALSE,
  latex_options = "tabular"  # prevents talltblr error
)

# Save to .tex file
writeLines(
  capture.output(print(table_latex)),
  con = "../outputs/first_stage_sibship.tex"
)

```

```{r}
names(dhs2)
```

### 2025-08-07 data-analysis/scripts/98a_paper_outputs.R

@@ -1,0 +1,191 @@
+# ================================
+# Summary Tables (Transposed, 8pt)
+# ================================
+
+# 1. Load packages
+library(tidyverse)
+library(kableExtra)
+library(haven)
+
+options(modelsummary_factory_default = "kableExtra")
+
+dhs1 <- readRDS("../clean-data/dhs_kr_children_with_geo.rds")
+
+dhs1 <- dhs1 %>%
+  mutate(
+    # Cleaned birth weight in grams (set special codes to NA)
+    birth_weight_grams = ifelse(birth_weight %in% c(9996, 9998, 9999), NA, birth_weight),
+    
+    # Cleaned perceived birth size (numeric 1–5; NA if don't know or missing)
+    birth_size_clean = ifelse(birth_size %in% c(8, 9), NA, birth_size),
+    
+    # Optional: labeled factor for perceived size
+    birth_size_factor = case_when(
+      birth_size == 1 ~ "Very large",
+      birth_size == 2 ~ "Larger than average",
+      birth_size == 3 ~ "Average",
+      birth_size == 4 ~ "Smaller than average",
+      birth_size == 5 ~ "Very small",
+      birth_size %in% c(8, 9) ~ NA_character_
+    ) %>% factor(levels = c("Very large", "Larger than average", "Average",
+                            "Smaller than average", "Very small"))
+  )
+
+
+dhs2 <- dhs1 |>
+  filter(survey_year %in% c(1998, 2003, 2008)) |>
+  haven::zap_labels() |>
+  mutate(
+    municipality_name = case_when(
+      survey_year == 2008 & latitude == 14.6055 & longitude == 120.9485 ~ "Manila",
+      survey_year == 2008 & latitude == 14.56937 & longitude == 120.96649 ~ "Manila",
+      survey_year == 2008 & latitude == 14.56748 & longitude == 120.97882 ~ "Manila",
+      survey_year == 2008 & latitude == 14.56274 & longitude == 120.98432 ~ "Manila",
+      survey_year == 2008 & latitude == 14.59441 & longitude == 121.01443 ~ "Manila",
+      TRUE ~ municipality_name
+    ),
+    manila = factor(str_to_lower(municipality_name) == "manila", levels = c(FALSE, TRUE)),
+    treated = as.integer(manila == TRUE),
+    birth_year = if_else(birth_year < 100, birth_year + 1900, birth_year),
[…]

### 2025-08-07 data-analysis/scripts/98b_paper_outputs.R

@@ -1,0 +1,92 @@
+# --- Load libraries ---
+library(data.table)
+library(kableExtra)
+
+# --- Load census data ---
+census1 <- readRDS("../clean-data/census_1990_2010.rds")
+census1 <- as.data.table(census1)
+
+# --- Ensure numeric types ---
+census1[, `:=`(
+  age = as.numeric(age),
+  sex = as.numeric(sex),
+  year = as.numeric(year),
+  geo1_ph = as.numeric(geo1_ph),
+  hhwt = as.numeric(hhwt)
+)]
+
+# --- Step 1: Count births (age 0–5) per year by province ---
+census_kids <- census1[age >= 0 & age <= 5]
+census_kids[, birth_year := year - age]
+
+province_ids <- unique(census_kids$geo1_ph)
+output_file <- "../clean-data/census_births_by_year.csv"
+fwrite(data.table(), output_file)  # clear file first
+
+for (i in seq_along(province_ids)) {
+  pid <- province_ids[i]
+  
+  births_chunk <- census_kids[geo1_ph == pid,
+                              .(num_births = sum(hhwt, na.rm = TRUE)),
+                              by = birth_year
+  ]
+  births_chunk[, geo1_ph := pid]
+  
+  fwrite(births_chunk, output_file, append = (i != 1), col.names = (i == 1))
+}
+
+# --- Step 2: Load merged data for birth rate calculation ---
+births_by_year <- fread("../clean-data/census_births_by_year.csv")
+women_at_risk  <- fread("../clean-data/census_women_at_risk.csv")
+
+birth_rates <- merge(births_by_year, women_at_risk, by = c("birth_year", "geo1_ph"))
+birth_rates[, birth_rate := num_births / num_women]
+
+# --- Step 3: Categorize regions and compute summary ---
+birth_rates[, location_group := fcase(
+  geo1_ph == 608039, "Manila",
+  geo1_ph %in% c(608074, 608075, 608076), "Other NCR",
+  default = "Non-NCR"
+)]
+birth_rates[, location_group := factor(location_group, levels = c("Manila", "Other NCR", "Non-NCR"))]
+
+# Rename the levels
+levels(birth_rates$location_group) <- c("Manila", "Other NCR Cities", "Rest of the Philippines")
+
+
+birth_rates[, post_policy := birth_year >= 2000]
+birth_rates[, period := ifelse(post_policy, "Post-2000", "Pre-2000")]
+
[…]

### 2025-08-07 data-analysis/scripts/04_build_dhs_under5_no_geo.R
add harmonized DHS data and new analysis scripts
Added new harmonized DHS children data and missing data output for municipalities. Introduced new scripts for building mother panels and econometric analysis, and updated existing scripts to reflect these changes. Also renamed and reorganized several analysis scripts to improve workflow.

@@ -1,74 +1,0 @@
-# ==============================================================================
-# Script Name : 04_build_dhs_under_5_no_geo.R
-# Purpose     : Harmonize child-level KR DHS datasets across survey years
-#               into a unified under-5 panel without geographic metadata
-# Years       : 1993, 1998, 2003, 2008, 2013
-# Author      : Erika Salvador '28 (esalvador28@amherst.edu)
-# Project     : Empirical Evidence of the Quantity–Quality Trade-off in the Philippines
-#
-# Description :
-#   This script performs the following steps:
-#     - Loads raw DHS KR (child recode) files for each survey year
-#     - Applies a crosswalk to standardize variable names across years
-#     - Selects and renames variables consistently across years
-#     - Appends all harmonized data into one under-5 child panel
-#     - Saves the harmonized dataset in both .rds and .dta formats
-# ==============================================================================
-
-# 0. Set Project Root ==========================================================
-library(here)
-here::i_am("data-analysis/scripts/04_build_dhs_under_5_no_geo.R")
-
-# 1. Load Libraries ============================================================
-library(haven)
-library(dplyr)
-library(purrr)
-library(readr)
-library(here)
-
-# 2. Define File Paths ========================================================
-raw_path <- here("data-analysis", "raw-data", "dhs_raw")
-crosswalk_path <- here(raw_path, "var_crosswalk.csv")
-
-# 3. Load and Clean Crosswalk =================================================
-crosswalk <- read_csv(crosswalk_path)
-colnames(crosswalk) <- tolower(colnames(crosswalk))
-crosswalk$variable <- tolower(crosswalk$variable)
-
-crosswalk <- crosswalk %>%
-  mutate(across(-variable, ~ tolower(trimws(.))))
-
-# 4. Define Harmonization Function ============================================
-harmonize_kr <- function(year) {
-  # Load KR data
-  kr_file <- file.path(raw_path, paste0("dhs_", year, "_kr_children.dta"))
-  kr <- read_dta(kr_file)
-  kr_vars <- tolower(names(kr))
-  
-  # Get relevant variables for this year
-  raw_vars <- tolower(crosswalk[[as.character(year)]])
-  std_names <- tolower(crosswalk$variable)
-  
-  # Filter out NAs and unavailable vars
[…]

### 2025-08-07 data-analysis/scripts/04a_build_dhs_under5_no_geo.R

@@ -1,0 +1,74 @@
+# ==============================================================================
+# Script Name : 04a_build_dhs_under_5_no_geo.R
+# Purpose     : Harmonize child-level KR DHS datasets across survey years
+#               into a unified under-5 panel without geographic metadata
+# Years       : 1993, 1998, 2003, 2008, 2013
+# Author      : Erika Salvador '28 (esalvador28@amherst.edu)
+# Project     : Empirical Evidence of the Quantity–Quality Trade-off in the Philippines
+#
+# Description :
+#   This script performs the following steps:
+#     - Loads raw DHS KR (child recode) files for each survey year
+#     - Applies a crosswalk to standardize variable names across years
+#     - Selects and renames variables consistently across years
+#     - Appends all harmonized data into one under-5 child panel
+#     - Saves the harmonized dataset in both .rds and .dta formats
+# ==============================================================================
+
+# 0. Set Project Root ==========================================================
+library(here)
+here::i_am("data-analysis/scripts/04a_build_dhs_under_5_no_geo.R")
+
+# 1. Load Libraries ============================================================
+library(haven)
+library(dplyr)
+library(purrr)
+library(readr)
+library(here)
+
+# 2. Define File Paths ========================================================
+raw_path <- here("data-analysis", "raw-data", "dhs_raw")
+crosswalk_path <- here(raw_path, "var_crosswalk.csv")
+
+# 3. Load and Clean Crosswalk =================================================
+crosswalk <- read_csv(crosswalk_path)
+colnames(crosswalk) <- tolower(colnames(crosswalk))
+crosswalk$variable <- tolower(crosswalk$variable)
+
+crosswalk <- crosswalk %>%
+  mutate(across(-variable, ~ tolower(trimws(.))))
+
+# 4. Define Harmonization Function ============================================
+harmonize_kr <- function(year) {
+  # Load KR data
+  kr_file <- file.path(raw_path, paste0("dhs_", year, "_kr_children.dta"))
+  kr <- read_dta(kr_file)
+  kr_vars <- tolower(names(kr))
+  
+  # Get relevant variables for this year
+  raw_vars <- tolower(crosswalk[[as.character(year)]])
+  std_names <- tolower(crosswalk$variable)
+  
+  # Filter out NAs and unavailable vars
[…]

### 2025-08-07 data-analysis/scripts/04b_build_dhs_mother_panel.R

@@ -1,0 +1,72 @@
+# ==============================================================================
+# Script Name : 04b_build_dhs_ir_panel.R
+# Purpose     : Harmonize woman-level IR DHS datasets across survey years
+#               into a unified panel with key fertility and household variables
+# Years       : 1993, 1998, 2003, 2008, 2013
+# Author      : Erika Salvador '28 (esalvador28@amherst.edu)
+# Project     : Empirical Evidence of the Quantity–Quality Trade-off in the Philippines
+#
+# Description :
+#   This script performs the following steps:
+#     - Loads raw DHS IR (individual recode) files for each survey year
+#     - Applies a crosswalk to standardize variable names across years
+#     - Selects and renames key fertility, household, and maternal variables
+#     - Appends all harmonized data into one unified woman-level panel
+#     - Saves the cleaned dataset in both .rds and .dta formats
+#     - Output will be used for merging sibship size to child-level KR data
+# ==============================================================================
+
+# 0. Set Project Root ==========================================================
+library(here)
+here::i_am("data-analysis/scripts/04b_build_dhs_ir_panel.R")
+
+# 1. Load Libraries ============================================================
+library(haven)
+library(dplyr)
+library(purrr)
+library(readr)
+library(here)
+
+# 2. Define File Paths ========================================================
+raw_path <- here("data-analysis", "raw-data", "dhs_raw")
+output_path <- here("data-analysis", "clean-data")
+
+# 3. Define Years =============================================================
+years <- c(1993, 1998, 2003, 2008, 2013)
+
+# 4. Loop Over Files, Count Children < 10 =====================================
+ir_list <- list()
+
+for (yr in years) {
+  
+  file <- file.path(raw_path, paste0("dhs_", yr, "_ir_individual.dta"))
+  ir <- read_dta(file)
+  
+  # Get child age and alive status variables
+  age_vars <- paste0("b8_", sprintf("%02d", 1:20))
+  alive_vars <- paste0("b5_", sprintf("%02d", 1:20))
+  
+  # Keep only variables that exist
+  age_vars_exist <- age_vars[age_vars %in% names(ir)]
+  alive_vars_exist <- alive_vars[alive_vars %in% names(ir)]
+  
+  # Count children under 10 who are alive
+  ir <- ir %>%
+    rowwise() %>%
[…]

### 2025-08-07 data-analysis/scripts/05_append_geo_to_dhs_panel.R

@@ -1,131 +1,2517 @@
 # ==============================================================================
 # Script Name : 05_append_geo_to_dhs_panel.R
-# Purpose     : Attach geographic metadata (region, province, municipality)
-#               to DHS cluster IDs using GPS coordinates and modal mapping.
-# Years       : 2003, 2008 (GPS-based); 1993, 1998, 2013 (modal guess)
+# Purpose     : Append cleaned geographic metadata (region, province, municipality)
+#               to the harmonized DHS cluster-level dataset using GPS-based mapping
 # Author      : Erika Salvador '28 (esalvador28@amherst.edu)
 # Project     : Empirical Evidence of the Quantity–Quality Trade-off in the Philippines
-#
-# Description :
-#   This script performs the following steps:
-#     - Loads the under-5 DHS child panel without geographic metadata
-#     - Imports 2003 and 2008 DHS GPS cluster shapefiles
-#     - Maps GPS clusters to municipalities using GADM Level-3 shapefiles
-#     - Constructs modal municipality-province map by region × urbanicity
-#     - Guesses province/municipality for non-GPS years using modal map
-#     - Joins region names and saves final geo-tagged child-level panel
-#     - Includes diagnostics for unmatched or missing geographic data
 # ==============================================================================
 
 # 0. Set Project Root ==========================================================
 library(here)
 here::i_am("data-analysis/scripts/05_append_geo_to_dhs_panel.R")
 
-# 1. Load libraries ============================================================
-library(dplyr)
-library(sf)
-library(haven)
-library(readr)
-library(here)
-library(stringr)
+# 1. Load Required Libraries ===================================================
+library(sf)        # for shapefiles
+library(readxl)    # for Excel GPS code files
+library(dplyr)     # for data manipulation
+library(stringr)   # for string cleaning
+library(haven)     # for saving to .dta
+library(janitor)   # for cleaning variable names
 
[…]

### 2025-08-07 data-analysis/scripts/07_econometrics.R

@@ -1,0 +1,86 @@
+# ===============================================================
+# ECONOMETRIC ANALYSIS: EO 003 Impact on Child Outcomes
+# Dataset: dhs_kr_children_with_geo.rds
+# Author: [Your Name]
+# ===============================================================
+
+# 1. Load Required Libraries ------------------------------------
+library(tidyverse)
+library(fixest)
+library(ggplot2)
+library(broom)
+library(MatchIt)
+library(quantreg)
+library(AER)
+
+# 2. Load Data --------------------------------------------------
+dhs_data <- readRDS("Data-analysis/clean-data/dhs_kr_children_with_geo.rds")
+
+# 3. Create Treatment Variables ---------------------------------
+dhs_data <- dhs_data %>%
+  mutate(
+    # Numeric birth date (YYYYMM)
+    birth_ym = birth_year * 100 + birth_month,
+    
+    # EO003 implemented Jan 2001: 1 = post-policy
+    post_policy = as.integer(birth_ym >= 200101),
+    
+    # Accurate Manila assignment using cleaned geo
+    manila = as.integer(
+      (str_to_lower(province_name) == "metro manila" & str_to_lower(municipality_name) == "manila") |
+        (survey_year == 2003 & cluster_id %in% c(564:581, 643)) |
+        (survey_year == 2008 & cluster_id == 8)
+    ),
+    
+    # For event study
+    birth_year_relative_to_policy = birth_year - 2000
+  )
+
+# Optional: Relevel child_sex
+dhs_data$child_sex <- factor(dhs_data$child_sex, levels = c("male", "female"))
+
+# 4. Estimate Event Study ---------------------------------------
+event_model <- feols(
+  birth_weight ~ i(birth_year_relative_to_policy, ref = -1) + manila +
+    wealth_index + mother_educ_level,
+  data = dhs_data
+)
+
+# 5. Plot Event Study Coefficients ------------------------------
+event_plot_data <- tidy(event_model) %>%
+  filter(grepl("birth_year_relative_to_policy::", term)) %>%
+  mutate(year = as.numeric(gsub("birth_year_relative_to_policy::", "", term)))
+
+ggplot(event_plot_data, aes(x = year, y = estimate)) +
+  geom_point() +
+  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
+                    ymax = estimate + 1.96 * std.error),
[…]

### 2025-08-07 data-analysis/scripts/08_2003_trial.R

@@ -1,0 +1,73 @@
+# ==============================================================================
+# Script Name : 01_did_setup_2003_manila.R
+# Purpose     : Prepares analytic sample from 2003 DHS KR file for DiD analysis
+#               comparing Manila City (NCR 1) vs. rest of PH
+# Author      : [Your Name]
+# Project     : EO003 Policy Evaluation – Birthweight and Vaccine Outcomes
+# ==============================================================================
+
+# 1. Load Libraries ============================================================
+library(haven)
+library(dplyr)
+
+# 2. Load KR Data ==============================================================
+
+kr2003 <- read_dta("../raw-data/dhs_raw/dhs_2003_kr_children.DTA")  # <-- Adjust to correct path
+
+# 3. Construct Key Variables ===================================================
+
+kr2003 <- kr2003 %>%
+  mutate(
+    # Age in years at survey
+    age_years = floor(b19 / 12),
+    
+    # Year of birth (interview year - age)
+    birth_year = v007 - age_years,
+    
+    # Indicator for children born post-policy
+    post2000 = ifelse(birth_year >= 2001, 1, 0),
+    
+    # Indicator for children born in Manila City (NCR 1 = SHPROV 39)
+    manila = ifelse(SHPROV == 39, 1, 0),
+    
+    # Treatment group: born in Manila and born after 2000
+    treat = ifelse(post2000 == 1 & manila == 1, 1, 0)
+  )
+
+# 4. Restrict Sample to Treated & Control Cohorts ==============================
+
+kr2003_subset <- kr2003 %>%
+  filter(age_years %in% c(0, 1, 4, 5))  # 0–1 = treated; 4–5 = controls
+
+# 5. Birthweight Outcome =======================================================
+
+# Use m19: birthweight in grams
+summary(kr2003_subset$m19)
+
+# Optional: Flag low birthweight < 2500g
+kr2003_subset <- kr2003_subset %>%
+  mutate(
+    lbw = ifelse(m19 < 2500 & !is.na(m19), 1, 0)
+  )
+
+# 6. Vaccine Outcome ===========================================================
+
+# Construct full immunization indicator (BCG, 3 DPT, Measles)
+kr2003_subset <- kr2003_subset %>%
+  mutate(
+    full_vax = ifelse(
+      h2 == 1 & h3 == 1 & h5 == 1 & h7 == 1 & h9 == 1, 1, 0
+    )
+  )
+
+# 7. Save Output ===============================================================
+
[…]

### 2025-08-09 docs/paper.Rmd
remove descriptive statistics section from results
Deleted the 'Descriptive Statistics' subsection and related tables and discussion from the results section in the paper. The results section now begins with the first-stage estimates, streamlining the presentation and focusing on the main analyses.

@@
 
-\subsection{Descriptive Statistics}
-
-The following tables present summary statistics for key child health outcomes, disaggregated by location and policy period.
-Table 1 presents summary statistics for mean birth weight by location and policy period.
-
-```{=latex}
-\input{../data-analysis/outputs/dhs_summary_birth_weight_t.tex}
-```
-
-Among children born in Manila, the average birth weight declined from 3,023.1 grams (SD = 712.6) in the pre-policy period to 2,969.8 grams (SD = 619.2) after the policy's implementation.
-A similar decline is observed in non-National Capital Region (NCR) areas, where mean birth weight dropped from 3,106.4 grams (SD = 726.9) to 2,994.5 grams (SD = 681.7).
-In contrast, birth weights in other NCR cities remained relatively stable over time, increasing slightly from 2,830.4 grams (SD = 743.6) to 2,965.9 grams (SD = 793.7).
-The majority of births in the sample occurred in non-NCR areas, which is expected given that these include  the rest of the country beyond Metro Manila.
-Notably, the decline in birth weight was more pronounced in Manila compared to other cities within the National Capital Region (NCR), suggesting a localized impact of the policy.
-While non-NCR areas also experienced reductions, the contrast between Manila and its neighboring NCR cities—where the policy was not implemented—is a more relevant comparison for isolating potential policy effects.
-
-Lastly, Table 3 presents the distribution of childhood vaccine uptake across policy periods and geographic areas.
-
-```{=latex}
-\input{../data-analysis/outputs/dhs_summary_vaccines_t.tex}
-```
-
-Table 3 presents the distribution of childhood vaccine uptake across policy periods and geographic areas.
-A substantial share of children in all groups received no vaccines, especially in Manila (79.4% pre-policy, 75.2% post-policy).
-However, there was a notable post-policy increase in children receiving 5 or more vaccines in both NCR (from 25.2% to 38.2%) and non-NCR cities (from 25.1% to 33.5%).
-These shifts indicate modest improvements in vaccine coverage in areas outside Manila, which—being unaffected by the contraceptive access restrictions—merit further investigation into other potential drivers, such as enhancements in health system capacity or increased accessibility of immunization services.
-
-Still, while the descriptive patterns suggest a localized effect in Manila—especially relative to nearby NCR cities—I use the full Philippine sample for the regressions that follow.
-This decision is mostly practical: restricting the analysis to NCR alone would leave too small a sample to detect meaningful effects with confidence.
-That said, I recognize the importance of comparing Manila to similar, unaffected areas.
-For robustness, I replicate all key models using only births from NCR and report those results in the Appendix.
-
 \subsection{First-Stage Estimates: Effect of the Contraceptive Ban on Fertility}

### 2025-08-09 docs/paper.tex
remove descriptive statistics section from results
Deleted the 'Descriptive Statistics' subsection and related tables and discussion from the results section in the paper. The results section now begins with the first-stage estimates, streamlining the presentation and focusing on the main analyses.

@@
 \section{Results and Discussion}
-
-\subsection{Descriptive Statistics}
-
-The following tables present summary statistics for key child health
-outcomes, disaggregated by location and policy period. Table 1 presents
-summary statistics for mean birth weight by location and policy period.
-
-\input{../data-analysis/outputs/dhs_summary_birth_weight_t.tex}
-
-Among children born in Manila, the average birth weight declined from
-3,023.1 grams (SD = 712.6) in the pre-policy period to 2,969.8 grams (SD
-= 619.2) after the policy's implementation. A similar decline is
-observed in non-National Capital Region (NCR) areas, where mean birth
-weight dropped from 3,106.4 grams (SD = 726.9) to 2,994.5 grams (SD =
-681.7). In contrast, birth weights in other NCR cities remained
-relatively stable over time, increasing slightly from 2,830.4 grams (SD
-= 743.6) to 2,965.9 grams (SD = 793.7). The majority of births in the
-sample occurred in non-NCR areas, which is expected given that these
-include the rest of the country beyond Metro Manila. Notably, the
-decline in birth weight was more pronounced in Manila compared to other
-cities within the National Capital Region (NCR), suggesting a localized
-impact of the policy. While non-NCR areas also experienced reductions,
-the contrast between Manila and its neighboring NCR cities---where the
-policy was not implemented---is a more relevant comparison for isolating
-potential policy effects.
-
-Lastly, Table 3 presents the distribution of childhood vaccine uptake
-across policy periods and geographic areas.
-
-\input{../data-analysis/outputs/dhs_summary_vaccines_t.tex}
-
-Table 3 presents the distribution of childhood vaccine uptake across
-policy periods and geographic areas. A substantial share of children in
-all groups received no vaccines, especially in Manila (79.4\%
-pre-policy, 75.2\% post-policy). However, there was a notable
-post-policy increase in children receiving 5 or more vaccines in both
-NCR (from 25.2\% to 38.2\%) and non-NCR cities (from 25.1\% to 33.5\%).
-These shifts indicate modest improvements in vaccine coverage in areas
-outside Manila, which---being unaffected by the contraceptive access
-restrictions---merit further investigation into other potential drivers,
-such as enhancements in health system capacity or increased
-accessibility of immunization services.
-
-Still, while the descriptive patterns suggest a localized effect in
-Manila---especially relative to nearby NCR cities---I use the full
-Philippine sample for the regressions that follow. This decision is
-mostly practical: restricting the analysis to NCR alone would leave too
-small a sample to detect meaningful effects with confidence. That said,
-I recognize the importance of comparing Manila to similar, unaffected
-areas. For robustness, I replicate all key models using only births from
-NCR and report those results in the Appendix.
 

### 2025-08-09 docs/paper.Rmd
fix typos and update regression estimate in paper
Corrected a missing period in the abstract and updated the Model 2 coefficient for the first-stage regression (from 0.200 to 0.242) in both the RMarkdown and LaTeX source files. These changes clarify the results and improve the accuracy of the manuscript.

@@
 In the second stage, each additional sibling is associated with a reduction of 22.97 vaccine doses received by the index child, a decline large enough to prevent completion of the recommended immunization schedule.
-The estimated effect on birth weight is −8,676 grams, though statistically insignificant Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight.
+The estimated effect on birth weight is −8,676 grams, though statistically insignificant.
+Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight.
 While instrumental variable estimates are limited by a weak first stage, the magnitude and sign of the coefficients support the hypothesis that larger sibship sizes reduce health investments per child.
@@
 The outcome variables are the number of living children at the time of survey (Model 1) and the total number of children ever born, including deceased (Model 2).
-In both models, the interaction term Manila × Post-2000 is positive—0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model 2—indicating that children born to Manila mothers after the policy may have experienced slightly larger sibship sizes, on average.
+In both models, the interaction term Manila × Post-2000 is positive—0.098 (SE = 0.691) in Model 1 and 0.242 (SE = 0.461) in Model 2—indicating that children born to Manila mothers after the policy may have experienced slightly larger sibship sizes, on average.
 Although these estimates are not statistically significant, their consistent positive direction suggests that the policy did not produce a strong fertility-decreasing effect, and may even have been associated with modest increases in completed or surviving fertility.

### 2025-08-09 docs/paper.tex
fix typos and update regression estimate in paper
Corrected a missing period in the abstract and updated the Model 2 coefficient for the first-stage regression (from 0.200 to 0.242) in both the RMarkdown and LaTeX source files. These changes clarify the results and improve the accuracy of the manuscript.

@@
 schedule. The estimated effect on birth weight is −8,676 grams, though
-statistically insignificant Reduced-form estimates are directionally
+statistically insignificant. Reduced-form estimates are directionally
 consistent: policy exposure is associated with a 0.91-dose drop in
@@
 2). In both models, the interaction term Manila × Post-2000 is
-positive---0.098 (SE = 0.691) in Model 1 and 0.200 (SE = 0.461) in Model
+positive---0.098 (SE = 0.691) in Model 1 and 0.242 (SE = 0.461) in Model
 2---indicating that children born to Manila mothers after the policy may
